### 2020-01-29 README.Rmd
add README

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Casting multiple shadows: interactive data visualisation with tours and embeddings

`liminal` aims to increase the effective use of embeddings by combining
them with interactive and dynamic graphics (via the vegawidget and shiny
packages) and with a technique from multivariate stastics called the tour (via
the tourr package).  Briefly, a tour is a sequence of interploated projections
of multivariate data onto lower dimensional space. The sequence is displayed as
a dynamic visualisation, and enables us to see the shadows the high dimensional
data makes in a lower dimensional view. By combining the tour with embedding
algorithms, we can see the following:

1. whether distances in the embedding view are meaningful 
2. the local and global structure of the data 
3. identify 'interesting' shapes or points in the data

## Outline

- review of embedding algorithms, interactive graphics, tours
- technology
- design
- examples/benchmarks
  - Kevin moon
  - pdfsense
  - single cell
  - quick draw

### 2020-01-29 Rmd/paper.Rmd
paper outline

---
title: " Casting multiple shadows: interactive data visualisation with tours and embeddings"
description: |
  A new article created using the Distill format.
author:
  - name: Stuart Lee
    affiliation: Monash University
  - name: Di Cook
    affiliation: Monash University
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### 2020-01-30 Rmd/paper.Rmd
outline in place

@@
 ```
+
+<!--
+- review of embedding algorithms, interactive graphics, tours
+- technology
+- design
+- examples/benchmarks
+  - Kevin moon
+  - pdfsense
+  - single cell
+  - quick draw
+-->
+
+# Introduction
+
+## Goals of high dimensional vis
+
+## How do analysts actually use these methods
+
+# Related Work
+
+## Embedding methods
+
+## Tours
+
+## Interactive graphics
+
+# Methods
+
+## Tours as a streaming data problem
+
+## Multiple views, multiple contexts
+
+## Interactions and visual analytics for embeddings
+
+# Results
+
+## Case Study 1: Fake trees
+
+## Case Study 2: non-spherical clusters, pdfsense
+
+## Case Study 3: well-clustered data, single cell
+
+## Case Study 4: quickdraw

### 2020-01-30 paper.Rmd
move paper to top level

--- 2020-01-30 Rmd/paper.Rmd
@@
 title: " Casting multiple shadows: interactive data visualisation with tours and embeddings"
-description: |
-  A new article created using the Distill format.
 author:
@@
 output: distill::distill_article
+bibliography: liminal.bib
 ---
@@
 ## Case Study 4: quickdraw
+
+# Discussion
+
+# References

### 2020-01-30 scripts/liminal-pbmc3k.R
extract R code from Rmd

@@ -1,0 +1,96 @@
+## ---- include = FALSE------------------------------
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
+
+
+## ----setup-----------------------------------------
+library(TENxPBMCData)
+library(scater)
+library(scran)
+
+pbmc3k <- TENxPBMCData("pbmc3k")
+pbmc3k
+
+
+## ----qc--------------------------------------------
+is.mito <- grep("MT", rowData(pbmc3k)$Symbol_TENx)
+stats <- perCellQCMetrics(pbmc3k, subsets=list(Mito=is.mito))
+high.mito <- isOutlier(stats$subsets_Mito_percent, nmads=3, type="higher")
+pbmc3k <- pbmc3k[,!high.mito]
+
+
+## --------------------------------------------------
+pbmc3k <- logNormCounts(pbmc3k)
+
+
+## --------------------------------------------------
+dec3k <- modelGeneVar(pbmc3k)
+chosen.hvgs <- which(dec3k$bio > 0)
+
+
+## --------------------------------------------------
+set.seed(2099)
+pbmc3k <- runPCA(pbmc3k, 
+                 ncomponents = 15,
+                 subset_row = chosen.hvgs,
+                 BSPARAM = BiocSingular::RandomParam(),
+                 name = "pca")
+
+set.seed(1999)
+pbmc3k <- runTSNE(pbmc3k,
+                  subset_row = chosen.hvgs,
+                  perplexity = 30,
+                  name = "tsne")
+
+
+## --------------------------------------------------
+g <- buildSNNGraph(pbmc3k, k=15, use.dimred = 'pca')
+clust <- igraph::cluster_louvain(g)
+pbmc3k$cluster <- factor(igraph::membership(clust))
+
+
+
+## ----flatten---------------------------------------
+tsne_tbl <- data.frame(
+  tsne_x = reducedDim(pbmc3k, "tsne")[,1],
+  tsne_y = reducedDim(pbmc3k, "tsne")[,2],
+  cluster = pbmc3k$cluster
+)
+
+express <- t(logcounts(pbmc3k)[chosen.hvgs,])
+colnames(express) <- rowData(pbmc3k)[chosen.hvgs, "Symbol_TENx"]
+express <- as.data.frame(express)
+tsne_tbl <- dplyr::bind_cols(
+  tsne_tbl,
+  express
+)
+
+
+## --------------------------------------------------
[…]

### 2020-02-05 paper.Rmd
draft paragraph

@@
 
+Broadly characterised by @Brehmer2014-hk, characterised two main workflow
+steps that an analyst performs when using embedding methods: dimension reduction
+and cluster orientation. The first relates to 
+dimension reduction achieved by using an embedding method, here an analyst
+wants to characterise and map meaning onto the embedded form, for example
+identifying batch effects from a high throughput sequencing experiment, or
+identifying a gradient along the embedded form. The second relates to using
+embeddings as part of a clustering workflow. Here analysts are interested
+in identifying  and naming clusters and verifying them by either applying 
+known labels or colouring by variables that are a-priori known to distinguish 
+clusters.
+
 # Related Work
@@
 
-## Case Study 2: non-spherical clusters, pdfsense
+## Case Study 2: well-clustered smallish single cell
 
-## Case Study 3: well-clustered data, single cell
+## Case Study 3: largeish single cell data
 
-## Case Study 4: quickdraw
+## Case Study 4: non-spherical clusters, pdfsense
 

### 2020-02-05 paper.Rmd
add in another sentence
refinement, adding in JCGS references

@@
 
-Broadly characterised by @Brehmer2014-hk, characterised two main workflow
+@Brehmer2014-hk characterised two main workflow
 steps that an analyst performs when using embedding methods: dimension reduction
@@
 known labels or colouring by variables that are a-priori known to distinguish 
-clusters.
+clusters. Both of these workflow steps rely on the embedding being 'faithful' to
+the underlying high dimensional data, and become much more difficult when
+there is no underlying ground truth.
+A further empirical study by @Sedlmair2013-pn, showed that when using embedding teqhinques, simple 2d scatter plots were an apporiate visualisation to help analysts perform these tasks but they did note that
+depending on the underlying goal multiple dimensionality reduction techniques
+are necessary.
 
@@
 ## Embedding methods
+
+- PCA
+- tSNE
+- UMAP
+- PHATE
 

### 2020-02-05 paper.Rmd
more words

--- 2020-02-05 paper.Rmd
@@
 
-## Goals of high dimensional vis
+High dimensional data is increasingly prevalent in
+the natural sciences and beyond but presents a challenge to the analyst in terms of both data
+cleaning / preprocessing and visualisation. Methods
+to embed data from a high-dimesional space into a low-dimensional one now form a core step of the data analysis workflow where they are used to ascertain hidden structure and denoise data for downstream analysis (thereby nullifying the 'curse of dimensionality').
 
-## How do analysts actually use these methods
-
-Broadly characterised by @Brehmer2014-hk, characterised two main workflow
-steps that an analyst performs when using embedding methods: dimension reduction
+Choosing an appropriate embedding presents a 
+challenge to the analyst. How does an analyst know
+whether the embedding has captured the underlying
+topology and geometry of the high dimensional space?
+The answer depends on the analyst's workflow. 
+@Brehmer2014-hk characterised two main workflow
+steps that an analyst performs when using embedding techniques: dimension reduction
 and cluster orientation. The first relates to 
@@
 identifying batch effects from a high throughput sequencing experiment, or
-identifying a gradient along the embedded form. The second relates to using
+identifying a gradient or trajectory along the embedded form. The second relates to using
 embeddings as part of a clustering workflow. Here analysts are interested
@@
 known labels or colouring by variables that are a-priori known to distinguish 
-clusters.
+clusters. Both of these workflow steps rely on the embedding being 'faithful' or the original high dimensional dataset, and become much more difficult when there is no underlying ground truth.
 
@@
 ## Embedding methods
+
+We discuss the use of
+
+- PCA
+- tSNE
+- UMAP
+- PHATE
 

### 2020-02-06 paper.Rmd
more words in intro

@@
 
-High dimensional data is increasingly prevalent in
-the natural sciences and beyond but presents a challenge to the analyst in terms of both data
-cleaning / preprocessing and visualisation. Methods
-to embed data from a high-dimesional space into a low-dimensional one now form a core step of the data analysis workflow where they are used to ascertain hidden structure and denoise data for downstream analysis (thereby nullifying the 'curse of dimensionality').
+High dimensional data is increasingly prevalent in the natural sciences and
+beyond but presents a challenge to the analyst in terms of both data cleaning /
+preprocessing and visualisation. Methods to embed data from a high-dimesional
+space into a low-dimensional one now form a core step of the data analysis
+workflow where they are used to ascertain hidden structure and denoise data for
+downstream analysis (thereby nullifying the 'curse of dimensionality').
 
-Choosing an appropriate embedding presents a 
-challenge to the analyst. How does an analyst know
-whether the embedding has captured the underlying
-topology and geometry of the high dimensional space?
-The answer depends on the analyst's workflow. 
-@Brehmer2014-hk characterised two main workflow
-steps that an analyst performs when using embedding techniques: dimension reduction
-and cluster orientation. The first relates to 
-dimension reduction achieved by using an embedding method, here an analyst
-wants to characterise and map meaning onto the embedded form, for example
-identifying batch effects from a high throughput sequencing experiment, or
-identifying a gradient or trajectory along the embedded form. The second relates to using
-embeddings as part of a clustering workflow. Here analysts are interested
-in identifying  and naming clusters and verifying them by either applying 
-known labels or colouring by variables that are a-priori known to distinguish 
-clusters. Both of these workflow steps rely on the embedding being 'faithful' or the original high dimensional dataset, and become much more difficult when there is no underlying ground truth.
+Choosing an appropriate embedding presents a challenge to the analyst. How does
+an analyst know whether the embedding has captured the underlying topology and
+geometry of the high dimensional space? The answer depends on the analyst's
+workflow.  @Brehmer2014-hk characterised two main workflow steps that an
+analyst performs when using embedding techniques: dimension reduction and
+cluster orientation. The first relates to dimension reduction achieved by using
+an embedding method, here an analyst wants to characterise and map meaning onto
+the embedded form, for example identifying batch effects from a high throughput
+sequencing experiment, or identifying a gradient or trajectory along the
+embedded form. The second relates to using embeddings as part of a clustering
+workflow. Here analysts are interested in identifying and naming clusters and
+verifying them by either applying known labels or colouring by variables that
+are a-priori known to distinguish clusters. Both of these workflow steps rely
+on the embedding being 'faithful' or the original high dimensional dataset, and
+become much more difficult when there is no underlying ground truth.
 
-# Related Work
+As part of a visualization workflow, it's important to consider the perception 
+and interpretation of embedding methods as well. @Sedlmair2013-pn showed that 2D
+scatter plots were mostly suffucient for detecting class seperation, however
+they noted that often multiple embeddings were required. For the task of cluster
+identification, @Lewis2012-ai showed experimentally that novice
+users of non-linear embedding techniques were more likely to consider clusters
+of points on a 2d scatter plot to be the result of a spurious embedding
+compared to advanced users who were aware of the inner workings of the embedding
+algorithm. @Wattenberg2016-ji noted that in the case of t-distributed stochastic
+neighbourhood embeddings (t-SNE) it was important for users to modify algorithm
+parameters in order to preserve global structure in the low dimensional
+embedding.
 
-## Embedding methods
+Embedding methods can be classified into two broad groups: linear
+and non-linear methods. Linear methods perform a linear transformation
+of the data; one example is principal components analysis (PCA)
+which performs an eigendecomposition of the estimated sample covariance matrix.
+The eigenvalues are sorted in decreasing order and represent the variance
+explained by each component (eigenvector). A common approach to deciding on
+the number of principal components to retain is to plot the proportion of variance explained by each component and choose a cut-off. Non-linear methods
+generally perform preprocessing on the high-dimensional data such as generating
+a neighborhood graph and perform transformations on the preprocessed form. We
+restrict our attention to three methods that are commonly used in
+high-throughput biology (though this is by no means exhaustive): t-SNE, 
+uniform manifold alignment and projection (UMAP), and PHATE.
 
-We discuss the use of
-
-- PCA
-- tSNE
-- UMAP
-- PHATE
+Clearly, finding an appropriate embedding is a difficult and somewhat
+poorly defined problem. We propose a more pragmatic workflow inspired by (wickham blindfold paper) by incorporating
+interactive graphics and tours with embeddings that allows users to see a
+global overview of their high dimensional data and assists them with cluster
+orientation tasks. This workflow is incorporated into an R package
+called `liminal` (Available: https://github.com/sa-lee/liminal).
 

### 2020-02-06 paper.Rmd
write up t-SNE section

@@
 compared to advanced users who were aware of the inner workings of the embedding
-algorithm. @Wattenberg2016-ji noted that in the case of t-distributed stochastic
-neighbourhood embeddings (t-SNE) it was important for users to modify algorithm
-parameters in order to preserve global structure in the low dimensional
-embedding.
+algorithm.
 
@@
 explained by each component (eigenvector). A common approach to deciding on
-the number of principal components to retain is to plot the proportion of variance explained by each component and choose a cut-off. Non-linear methods
+the number of principal components to retain is to plot the proportion of variance explained by each component and choose a cut-off.
+
+Non-linear methods
 generally perform preprocessing on the high-dimensional data such as generating
@@
 restrict our attention to three methods that are commonly used in
-high-throughput biology (though this is by no means exhaustive): t-SNE, 
+high-throughput biology: t-distributed stochastic neighbor embedding (t-SNE), 
 uniform manifold alignment and projection (UMAP), and PHATE.
+The t-SNE algorithm estimates the similarity of (Euclidean) distances of points in a high dimensional space using a Gaussian distribution and then
+estimates a configuration in the low dimensional embedding space by modelling
+similarities using a t-distribution with 1 degree of freedom. The resulting
+configuration is the one that minimizes the Kulback-Leibler divergence
+between the two distributions. A recent theoretical contribution by @Linderman2019-dq proved that t-SNE can recover spherical and well separated 
+cluster shapes, and proposed new approaches for tuning the optimisation parameters. It is a known problem that t-SNE can have trouble recovering
+global structure and that configurations can be highly dependent on
+how the algorithm is initialised and parameterised 
+[@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
+related to LargeVis, and like t-SNE acts on the k-nearest neighbor graph. Its
+main differences are that it uses a different cost function (cross entropy) which is optimized using stochastic gradient descent and defines a different
+kernel for similiarites in the low dimensional space. Finally, PHATE is based
 
-Clearly, finding an appropriate embedding is a difficult and somewhat
-poorly defined problem. We propose a more pragmatic workflow inspired by (wickham blindfold paper) by incorporating
+There is no one-size fits all: finding an appropriate embedding for a given
+dataset is a difficult and somewhat poorly defined problem.
+For non-linear methods, there are a lot of parameters to explore that can have an effeft on the resulting visualisation and interpretation.
+We propose a more pragmatic workflow inspired by (wickham blindfold paper) by incorporating
 interactive graphics and tours with embeddings that allows users to see a
@@
 called `liminal` (Available: https://github.com/sa-lee/liminal).
+
+https://jlmelville.github.io/smallvis/init.html#other_initialization_options
 

### 2020-02-06 paper.Rmd
rejigging and write up about methods used.

@@
 
-As part of a visualization workflow, it's important to consider the perception 
-and interpretation of embedding methods as well. @Sedlmair2013-pn showed that 2D
-scatter plots were mostly suffucient for detecting class seperation, however
-they noted that often multiple embeddings were required. For the task of cluster
-identification, @Lewis2012-ai showed experimentally that novice
-users of non-linear embedding techniques were more likely to consider clusters
-of points on a 2d scatter plot to be the result of a spurious embedding
-compared to advanced users who were aware of the inner workings of the embedding
-algorithm.
-
 Embedding methods can be classified into two broad groups: linear
@@
 high-throughput biology: t-distributed stochastic neighbor embedding (t-SNE), 
-uniform manifold alignment and projection (UMAP), and PHATE.
+uniform manifold alignment and projection (UMAP), and potential of heat-diffusion for affinity-based transition embedding (PHATE).
 The t-SNE algorithm estimates the similarity of (Euclidean) distances of points in a high dimensional space using a Gaussian distribution and then
@@
 main differences are that it uses a different cost function (cross entropy) which is optimized using stochastic gradient descent and defines a different
-kernel for similiarites in the low dimensional space. Finally, PHATE is based
+kernel for similiarites in the low dimensional space. Due to it's computational
+speed it's possible to generate UMAP embeddings in more than three dimensions.
+Finally, PHATE, inspired by diffusion maps, is based on estimating an affinity matrix via a distance matrix and k-nearest neighbors graph.
+The algorithm
+denoises estimated distances in high dimensional space via transforming the
+affinity matrix into a Markov transition probaility matrix and diffusing this 
+matrix over a fixed number of time steps. Then the diffused probabilities
+are transformed once more to construct a distance matrix, and multidimensional
+scaling is performed to construct a 2d embedding for visualization.
+
+As part of a visualization workflow, it's important to consider the perception 
+and interpretation of embedding methods as well. @Sedlmair2013-pn showed that 2D
+scatter plots were mostly suffucient for detecting class seperation, however
+they noted that often multiple embeddings were required. For the task of cluster
+identification, @Lewis2012-ai showed experimentally that novice
+users of non-linear embedding techniques were more likely to consider clusters
+of points on a 2d scatter plot to be the result of a spurious embedding
+compared to advanced users who were aware of the inner workings of the embedding
+algorithm.
 

### 2020-02-06 paper.Rmd
reordering stuff
paragraph on tours

@@
 
+A complimentary approach for visualizing structure in high dimensional data
+is the tour. A tour is a sequence of projections of a high dimensional dataset
+onto a low-dimensional orthonormal basis matrix, that is represented as a dynamic visualization.
+The sequence of generated bases are interpolated to form the tour path, allowing a user to explore the subspace of projections.
+A
+grand tour corresponds to choosing new bases at random, and can give an
+overview of the structure in the data. Instead of picking projections at random, a guided tour can be used to generate a sequence 'interesting' projections
+as quantified by an index function. The tour has been used previously 
+by @Wickham2015-cx as tool for exploring statistical model fits.
+
+## Interactive graphics
+
 As part of a visualization workflow, it's important to consider the perception 
@@
 
-## Tours
+# Methods
 
-## Interactive graphics
-
-# Methods
+my contribution + design of interface
 

### 2020-02-06 paper.Rmd
remove whitespace
finished first draft of intro!

@@
 
-A complimentary approach for visualizing structure in high dimensional data
-is the tour. A tour is a sequence of projections of a high dimensional dataset
-onto a low-dimensional orthonormal basis matrix, that is represented as a dynamic visualization.
-The sequence of generated bases are interpolated to form the tour path, allowing a user to explore the subspace of projections.
-A
-grand tour corresponds to choosing new bases at random, and can give an
-overview of the structure in the data. Instead of picking projections at random, a guided tour can be used to generate a sequence 'interesting' projections
-as quantified by an index function. The tour has been used previously 
-by @Wickham2015-cx as tool for exploring statistical model fits.
-
-## Interactive graphics
-
 As part of a visualization workflow, it's important to consider the perception 
@@
 
+A complimentary approach for visualizing structure in high dimensional data
+is the tour. A tour is a sequence of projections of a high dimensional dataset
+onto a low-dimensional orthonormal basis matrix, that is represented as a dynamic visualization.
+The sequence of generated bases are interpolated to form the tour path, allowing a user to explore the subspace of projections.
+A
+grand tour corresponds to choosing new bases at random, and can give an
+overview of the structure in the data. Instead of picking projections at random, a guided tour can be used to generate a sequence 'interesting' projections
+as quantified by an index function.
+Given the dynamic nature of the tour, user interaction is important for controlling and exploring the visualisation: the tour has been used previously by @Wickham2015-cx as tool for exploring statistical model fits and by @Buja1996-fk for exploring factorial experimental
+designs.
+
+While there has been much work on the algorithmic details of the aformentioned
+embedding methods, there has been relatively few tools designed to assist
+users to interact with these techniques and assist them in making comparsions
+between embeddings and performing the aforementioned cluster orientation tasks. 
+Several interactive interfaces have been proposed for evaluating or using
+embedding techniques: the Sleepwalk interface provides a click and highlight visualisation for colouring points in an embedding according to their distance
+in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by 
+@Pezzotti2017-cz provides a user guided and modified form of the t-SNE algorithm, that allows users to modified optimisation parameters in real-time.
+Similarly, the embedding projector is a web interface to running UMAP, t-SNE
+or PCA live in the browser and provides interactions to color points, and
+highlights nearest neighbors [@Smilkov2016-hp].
+
 There is no one-size fits all: finding an appropriate embedding for a given
@@
 For non-linear methods, there are a lot of parameters to explore that can have an effeft on the resulting visualisation and interpretation.
-We propose a more pragmatic workflow inspired by (wickham blindfold paper) by incorporating
+Interfaces for evaluating embeddings
+require interaction but should also be able to be incorporated into an analysts workflow. We propose a more pragmatic workflow inspired by incorporating
 interactive graphics and tours with embeddings that allows users to see a

### 2020-02-07 paper.Rmd
restructure as html doc; reclassify structure

@@
   - name: Stuart Lee
-    affiliation: Monash University
   - name: Di Cook
-    affiliation: Monash University
 date: "`r Sys.Date()`"
-output: distill::distill_article
+output: bookdown::html_document2
 bibliography: liminal.bib
@@
   - quick draw
+
+https://jlmelville.github.io/smallvis/init.html#other_initialization_options
+
 -->
@@
 
-https://jlmelville.github.io/smallvis/init.html#other_initialization_options
-
-# Methods
+# Design
 
 my contribution + design of interface
-
-## Tours as a streaming data problem
 
@@
 
-# Results
+# Implementation
+
+The `liminal` software is implemented as an open source R package @baseR. All 
+visualisations produced by the package were constructed using the `vegawidget`
+package, which is  an R interface to the Vega-Lite grammar of interactive graphics. Ensemble graphics for arranging tour and embedding views 
+are generated using the `shiny` packages. We have made extensive use of the 
+`tourr` package for constructing tour paths.
+
+## Tours as a streaming data problem
+
+## Interactivity for embedding exploration
+
+# Case Studies
 

### 2020-02-07 paper.Rmd
add citations

@@
 sequencing experiment, or identifying a gradient or trajectory along the
-embedded form. The second relates to using embeddings as part of a clustering
+embedded form @Nguyen2019-yh. The second relates to using embeddings as part of a clustering
 workflow. Here analysts are interested in identifying and naming clusters and
@@
 
-my contribution + design of interface
-
 ## Multiple views, multiple contexts

### 2020-02-07 paper.Rmd
adding citations
re wrap
add section on design

@@
 sequencing experiment, or identifying a gradient or trajectory along the
-embedded form @Nguyen2019-yh. The second relates to using embeddings as part of a clustering
-workflow. Here analysts are interested in identifying and naming clusters and
-verifying them by either applying known labels or colouring by variables that
-are a-priori known to distinguish clusters. Both of these workflow steps rely
-on the embedding being 'faithful' or the original high dimensional dataset, and
-become much more difficult when there is no underlying ground truth.
+embedded form @Nguyen2019-yh. The second relates to using embeddings as part of
+a clustering workflow. Here analysts are interested in identifying and naming
+clusters and verifying them by either applying known labels or colouring by
+variables that are a-priori known to distinguish clusters. Both of these
+workflow steps rely on the embedding being 'faithful' or the original high
+dimensional dataset, and become much more difficult when there is no underlying
+ground truth.
 
-Embedding methods can be classified into two broad groups: linear
-and non-linear methods. Linear methods perform a linear transformation
-of the data; one example is principal components analysis (PCA)
-which performs an eigendecomposition of the estimated sample covariance matrix.
-The eigenvalues are sorted in decreasing order and represent the variance
-explained by each component (eigenvector). A common approach to deciding on
-the number of principal components to retain is to plot the proportion of variance explained by each component and choose a cut-off.
+Embedding methods can be classified into two broad groups: linear and
+non-linear methods. Linear methods perform a linear transformation of the data;
+one example is principal components analysis (PCA) which performs an
+eigendecomposition of the estimated sample covariance matrix. The eigenvalues
+are sorted in decreasing order and represent the variance explained by each
+component (eigenvector). A common approach to deciding on the number of
+principal components to retain is to plot the proportion of variance explained
+by each component and choose a cut-off.
 
-Non-linear methods
-generally perform preprocessing on the high-dimensional data such as generating
-a neighborhood graph and perform transformations on the preprocessed form. We
-restrict our attention to three methods that are commonly used in
-high-throughput biology: t-distributed stochastic neighbor embedding (t-SNE), 
-uniform manifold alignment and projection (UMAP), and potential of heat-diffusion for affinity-based transition embedding (PHATE).
-The t-SNE algorithm estimates the similarity of (Euclidean) distances of points in a high dimensional space using a Gaussian distribution and then
-estimates a configuration in the low dimensional embedding space by modelling
-similarities using a t-distribution with 1 degree of freedom. The resulting
-configuration is the one that minimizes the Kulback-Leibler divergence
-between the two distributions. A recent theoretical contribution by @Linderman2019-dq proved that t-SNE can recover spherical and well separated 
-cluster shapes, and proposed new approaches for tuning the optimisation parameters. It is a known problem that t-SNE can have trouble recovering
-global structure and that configurations can be highly dependent on
-how the algorithm is initialised and parameterised 
-[@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
-related to LargeVis, and like t-SNE acts on the k-nearest neighbor graph. Its
-main differences are that it uses a different cost function (cross entropy) which is optimized using stochastic gradient descent and defines a different
-kernel for similiarites in the low dimensional space. Due to it's computational
-speed it's possible to generate UMAP embeddings in more than three dimensions.
-Finally, PHATE, inspired by diffusion maps, is based on estimating an affinity matrix via a distance matrix and k-nearest neighbors graph.
-The algorithm
-denoises estimated distances in high dimensional space via transforming the
-affinity matrix into a Markov transition probaility matrix and diffusing this 
-matrix over a fixed number of time steps. Then the diffused probabilities
-are transformed once more to construct a distance matrix, and multidimensional
-scaling is performed to construct a 2d embedding for visualization.
+Non-linear methods generally perform preprocessing on the high-dimensional data
+such as generating a neighborhood graph and perform transformations on the
+preprocessed form. We restrict our attention to three methods that are commonly
+used in high-throughput biology: t-distributed stochastic neighbor embedding
+(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
+heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
+algorithm estimates the similarity of (Euclidean) distances of points in a high
+dimensional space using a Gaussian distribution and then estimates a
+configuration in the low dimensional embedding space by modelling similarities
+using a t-distribution with 1 degree of freedom. The resulting configuration is
+the one that minimizes the Kulback-Leibler divergence between the two
+distributions. A recent theoretical contribution by @Linderman2019-dq proved
+that t-SNE can recover spherical and well separated cluster shapes, and
+proposed new approaches for tuning the optimisation parameters. It is a known
+problem that t-SNE can have trouble recovering global structure and that
+configurations can be highly dependent on how the algorithm is initialised and
+parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
+related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
+neighbor graph. Its main differences are that it uses a different cost function
+(cross entropy) which is optimized using stochastic gradient descent and
+defines a different kernel for similiarites in the low dimensional space. Due
+to it's computational speed it's possible to generate UMAP embeddings in more
+than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
+estimating an affinity matrix via a distance matrix and k-nearest neighbors
+graph. The algorithm denoises estimated distances in high dimensional space via
+transforming the affinity matrix into a Markov transition probaility matrix and
+diffusing this matrix over a fixed number of time steps. Then the diffused
+probabilities are transformed once more to construct a distance matrix, and
+multidimensional scaling is performed to construct a 2d embedding for
+visualization.
 
 As part of a visualization workflow, it's important to consider the perception
-and interpretation of embedding methods as well. @Sedlmair2013-pn showed that 2D
-scatter plots were mostly suffucient for detecting class seperation, however
-they noted that often multiple embeddings were required. For the task of cluster
-identification, @Lewis2012-ai showed experimentally that novice
-users of non-linear embedding techniques were more likely to consider clusters
-of points on a 2d scatter plot to be the result of a spurious embedding
-compared to advanced users who were aware of the inner workings of the embedding
+and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
+2D scatter plots were mostly suffucient for detecting class seperation, however
+they noted that often multiple embeddings were required. For the task of
+cluster identification, @Lewis2012-ai showed experimentally that novice users
+of non-linear embedding techniques were more likely to consider clusters of
+points on a 2d scatter plot to be the result of a spurious embedding compared
+to advanced users who were aware of the inner workings of the embedding
 algorithm.
 
-A complimentary approach for visualizing structure in high dimensional data
-is the tour. A tour is a sequence of projections of a high dimensional dataset
-onto a low-dimensional orthonormal basis matrix, that is represented as a dynamic visualization.
-The sequence of generated bases are interpolated to form the tour path, allowing a user to explore the subspace of projections.
-A
-grand tour corresponds to choosing new bases at random, and can give an
-overview of the structure in the data. Instead of picking projections at random, a guided tour can be used to generate a sequence 'interesting' projections
-as quantified by an index function.
-Given the dynamic nature of the tour, user interaction is important for controlling and exploring the visualisation: the tour has been used previously by @Wickham2015-cx as tool for exploring statistical model fits and by @Buja1996-fk for exploring factorial experimental
-designs.
+A complimentary approach for visualizing structure in high dimensional data is
+the tour. A tour is a sequence of projections of a high dimensional dataset
+onto a low-dimensional orthonormal basis matrix, that is represented as a
+dynamic visualization. The sequence of generated bases are interpolated to form
+the tour path, allowing a user to explore the subspace of projections. A grand
+tour corresponds to choosing new bases at random, and can give an overview of
+the structure in the data. Instead of picking projections at random, a guided
+tour can be used to generate a sequence 'interesting' projections as quantified
+by an index function. Given the dynamic nature of the tour, user interaction is
+important for controlling and exploring the visualisation: the tour has been
+used previously by @Wickham2015-cx as tool for exploring statistical model fits
+and by @Buja1996-fk for exploring factorial experimental designs.
 
 While there has been much work on the algorithmic details of the aformentioned
-embedding methods, there has been relatively few tools designed to assist
-users to interact with these techniques and assist them in making comparsions
-between embeddings and performing the aforementioned cluster orientation tasks. 
-Several interactive interfaces have been proposed for evaluating or using
-embedding techniques: the Sleepwalk interface provides a click and highlight visualisation for colouring points in an embedding according to their distance
+embedding methods, there has been relatively few tools designed to assist users
+to interact with these techniques and assist them in making comparsions between
+embeddings and performing the aforementioned cluster orientation tasks. Several
+interactive interfaces have been proposed for evaluating or using embedding
+techniques: the Sleepwalk interface provides a click and highlight
+visualisation for colouring points in an embedding according to their distance
 in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
-@Pezzotti2017-cz provides a user guided and modified form of the t-SNE algorithm, that allows users to modified optimisation parameters in real-time.
-Similarly, the embedding projector is a web interface to running UMAP, t-SNE
-or PCA live in the browser and provides interactions to color points, and
+@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
+algorithm, that allows users to modified optimisation parameters in real-time.
+Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
+PCA live in the browser and provides interactions to color points, and
 highlights nearest neighbors [@Smilkov2016-hp].
@@
 There is no one-size fits all: finding an appropriate embedding for a given
-dataset is a difficult and somewhat poorly defined problem.
-For non-linear methods, there are a lot of parameters to explore that can have an effeft on the resulting visualisation and interpretation.
-Interfaces for evaluating embeddings
-require interaction but should also be able to be incorporated into an analysts workflow. We propose a more pragmatic workflow inspired by incorporating
-interactive graphics and tours with embeddings that allows users to see a
-global overview of their high dimensional data and assists them with cluster
-orientation tasks. This workflow is incorporated into an R package
+dataset is a difficult and somewhat poorly defined problem. For non-linear
+methods, there are a lot of parameters to explore that can have an effeft on
+the resulting visualisation and interpretation. Interfaces for evaluating
+embeddings require interaction but should also be able to be incorporated into
+an analysts workflow. We propose a more pragmatic workflow inspired by
+incorporating interactive graphics and tours with embeddings that allows users
+to see a global overview of their high dimensional data and assists them with
+cluster orientation tasks. This workflow is incorporated into an R package
 called `liminal` (Available: https://github.com/sa-lee/liminal).
@@
 
-## Multiple views, multiple contexts
+We propose using tours as part of an analyst's workflow in in performing
+dimensionality reduction tasks. We have made extensive use of ensemble
+graphics, that is aligning related plots alongside each other to provide
+context. As we will see in the case studies, this allows analysts to quickly
+compare views from embedding methods and allows them to see how the embedding
+method alters the global structure of their data. Using ensembles allows the
+use of interaction techniques, that allow analysts to perform cluster
+orientation tasks via linking multiple views. This approach allows our
+interface, to achieve the three principles for interactive high-dimensional
+data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
+and making comparisons.
 
-## Interactions and visual analytics for embeddings
+## Finding Gestalt: focus and context
+
+To investigate latent structure and the shape of a high dimensional dataset, a
+tour can be run without the use of an external embedding. It is often useful to
+first run principal components on the input as an intial dimension reduction
+step, and then tour a subset of those components instead, i.e. by selecting
+them from a scree plot. The default tour layout is a scatter plot with an axis
+layout displaying the magnitude and direction of each basis vector. Since the
+tour is dynamic, it is often useful to be able to pause and highlight a
+particular view. In our interface, brushing will pause the tour path, allowing
+users to identify 'interesting' projections. The domain of the axis scales from
+running a tour is called the half range, and is computed by rescaling the input
+data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
+event, allowing a user to pan and zoom on the tour view dynamically. This is
+useful for peeling back dense clumps of points to reveal structure.
+
+## Posing Queries: multiple views, multiple contexts
+
+## Making comparsions
 
@@
 visualisations produced by the package were constructed using the `vegawidget`
-package, which is  an R interface to the Vega-Lite grammar of interactive graphics. Ensemble graphics for arranging tour and embedding views 
-are generated using the `shiny` packages. We have made extensive use of the 
-`tourr` package for constructing tour paths.
+package, which is an R interface to the Vega-Lite grammar of interactive
+graphics. Ensemble graphics for arranging tour and embedding views are
+generated using the `shiny` packages. We have made extensive use of the `tourr`
+package for constructing tour paths.
 

### 2020-02-07 paper.Rmd
add posing queries

@@
 
+We combine the tour view in a side by side layout with a scatter
+plot view from an embedding algorithm. The views are linked; analysts can brush a group of points in the embedding view, and see its shape in the tour view.
+This can be used for either manually identifying clusers, or verifying the location of clusters relative to other groupings. Consequently, simple linked
+brushing can alleviate a downfall of non-linear methods such as UMAP or t-SNE
+that break hierachial structure of clusterings. Aside from simple
+linked brushing, we can also pose more complex queries via interrogating
+the neighbourhood graph. Using a linked neighbourhood brush, we can 
+visually investigate the nearest neighbour relationships in the 
+high-dimensional space via brushing in the embedding view. The user can select
+the number of nearest neighbours directly, and modify the distance metric 
+used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
+can be used to pose queries in either the tour view or the embedding view;
+interface controls allow these brushes to combine using logical operators
+such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
+single point highlights or colorings, allowing analysts to get a more holistic
+view of the effect of an embedding algorithm.
+
 ## Making comparsions

### 2020-03-24 paper.Rmd
wording
rejigging sections
update paper

--- 2020-02-07 paper.Rmd
@@
 ---
-title: " Casting multiple shadows: interactive data visualisation with tours and embeddings"
+title: " Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings"
 author:
@@
 date: "`r Sys.Date()`"
-output: bookdown::html_document2
+output: 
+  bookdown::html_document2:
+    toc: false
+    number_sections: no
 bibliography: liminal.bib
@@
 
-<!--
-- review of embedding algorithms, interactive graphics, tours
-- technology
-- design
-- examples/benchmarks
-  - Kevin moon
-  - pdfsense
-  - single cell
-  - quick draw
+There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows.
+The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of
+local and global structure in the resulting visualisation.
+We present a visual analytics framework for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
+A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
+The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view.
+By linking the tour to a view
+obtained from an NLDR method, we can preserve global structure and through user interactions like spatial linked brushing observe where
+the NLDR view may be misleading. We display several case 
+studies from single cell genomics, that shows our approach is useful for
+cluster orientation tasks. 
+The implementation of our framework is available as an R package called `liminal` available at https://github.com/sa-lee/liminal.
 
-https://jlmelville.github.io/smallvis/init.html#other_initialization_options
-
--->
-
-# Introduction
+## Introduction
 
@@
 beyond but presents a challenge to the analyst in terms of both data cleaning /
-preprocessing and visualisation. Methods to embed data from a high-dimesional
+pre-processing and visualisation. Methods to embed data from a high-dimensional
 space into a low-dimensional one now form a core step of the data analysis
-workflow where they are used to ascertain hidden structure and denoise data for
+workflow where they are used to ascertain hidden structure and de-noise data for
 downstream analysis (thereby nullifying the 'curse of dimensionality').
@@
 
-Non-linear methods generally perform preprocessing on the high-dimensional data
+Non-linear methods generally perform pre-processing on the high-dimensional data
 such as generating a neighborhood graph and perform transformations on the
-preprocessed form. We restrict our attention to three methods that are commonly
-used in high-throughput biology: t-distributed stochastic neighbor embedding
-(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
-heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
-algorithm estimates the similarity of (Euclidean) distances of points in a high
+pre-processed form. We restrict our attention to two methods that are commonly
+used in the literature: t-distributed stochastic neighbor embedding
+(t-SNE) and uniform manifold alignment and projection (UMAP). The t-SNE
+algorithm estimates the pairwise similarity of (Euclidean) distances of points in a high
 dimensional space using a Gaussian distribution and then estimates a
@@
 using a t-distribution with 1 degree of freedom. The resulting configuration is
-the one that minimizes the Kulback-Leibler divergence between the two
+the one that minimizes the Kullback-Leibler divergence between the two
 distributions. A recent theoretical contribution by @Linderman2019-dq proved
@@
 configurations can be highly dependent on how the algorithm is initialised and
-parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
+parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
 related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
@@
 (cross entropy) which is optimized using stochastic gradient descent and
-defines a different kernel for similiarites in the low dimensional space. Due
+defines a different kernel for similarities in the low dimensional space. Due
 to it's computational speed it's possible to generate UMAP embeddings in more
-than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
-estimating an affinity matrix via a distance matrix and k-nearest neighbors
-graph. The algorithm denoises estimated distances in high dimensional space via
-transforming the affinity matrix into a Markov transition probaility matrix and
-diffusing this matrix over a fixed number of time steps. Then the diffused
-probabilities are transformed once more to construct a distance matrix, and
-multidimensional scaling is performed to construct a 2d embedding for
-visualization.
+than three dimensions.
 
@@
 and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
-2D scatter plots were mostly suffucient for detecting class seperation, however
+2D scatter plots were mostly sufficient for detecting class separation, however
 they noted that often multiple embeddings were required. For the task of
@@
 
-A complimentary approach for visualizing structure in high dimensional data is
+A complementary approach for visualizing structure in high dimensional data is
 the tour. A tour is a sequence of projections of a high dimensional dataset
@@
 
-While there has been much work on the algorithmic details of the aformentioned
+While there has been much work on the algorithmic details of the aforementioned
 embedding methods, there has been relatively few tools designed to assist users
-to interact with these techniques and assist them in making comparsions between
+to interact with these techniques and assist them in making comparisons between
 embeddings and performing the aforementioned cluster orientation tasks. Several
@@
 dataset is a difficult and somewhat poorly defined problem. For non-linear
-methods, there are a lot of parameters to explore that can have an effeft on
+methods, there are a lot of parameters to explore that can have an effect on
 the resulting visualisation and interpretation. Interfaces for evaluating
@@
 
-# Design
+## Dimensionality Reduction
+
+## Visual Design
 
@@
 
-## Finding Gestalt: focus and context
+### Finding Gestalt: focus and context
 
@@
 tour can be run without the use of an external embedding. It is often useful to
-first run principal components on the input as an intial dimension reduction
+first run principal components on the input as an initial dimension reduction
 step, and then tour a subset of those components instead, i.e. by selecting
@@
 running a tour is called the half range, and is computed by rescaling the input
-data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
+data onto hyper-dimensional unit cube. We bind the half range to a mouse wheel
 event, allowing a user to pan and zoom on the tour view dynamically. This is
@@
 
-## Posing Queries: multiple views, multiple contexts
+### Posing Queries: multiple views, many contexts
 
-## Making comparsions
+We have combined the tour view in a side by side layout with 
+a scatter plot view that represents the output of an embedding algorithm. 
+These views are linked; analysts can brush regions or
+highlight collections of points in either view. Linked highlighting can
+be performed when points have been previously labeled according to some
+discrete structure, i.e. cluster labels are available. This is achieved
+via the analyst clicking on groups in the legend, which causes unselected groupings to have their points become less opaque. Consequently, 
+simple linked highlighting can alleviate a known downfall of methods such as 
+UMAP or t-SNE: that is distances between clusters are misleading. By
+highlighting corresponding clusters in the tour view, the analyst can
+see the relationship between clusters, and therefore obtain a more accurate 
+representation of the topology of their data.
 
-# Implementation
+Simple linked brushing is achieved via mouse-click and drag movements. By
+default, when brushing occurs in the tour view, the current projection is 
+paused and corresponding points in the embedding view are highlighted. 
+Likewise, when brushing occurs in the embedding view, corresponding points
+in the tour view are highlighted. In this case, an analyst can use brushing
+for manually identifying clusters and verifying cluster locations and shapes:
+brushing in the embedding view gives analysts a sense of the shape and 
+proximity of cluster in high-dimensional space.
 
-The `liminal` software is implemented as an open source R package @baseR. All
-visualisations produced by the package were constructed using the `vegawidget`
-package, which is an R interface to the Vega-Lite grammar of interactive
-graphics. Ensemble graphics for arranging tour and embedding views are
-generated using the `shiny` packages. We have made extensive use of the `tourr`
-package for constructing tour paths.
+### Making comparisons: revising embeddings
 
-## Tours as a streaming data problem
+As mentioned previously, when using any DR method, we are assuming
+the embedding is representative of the high-dimensional dataset it was
+computed from. Defining what it means for embedding to be 'representative` or
+'faithful' to high-dimensional data is ill-posed and depends on the underlying
+task an analyst is trying to achieve. At the very minimum, we are interested
+in distortions and diffusions of the high-dimensional data. Distortions
+occur when points that are near each other in the embedding view are 
+far from each other in the original dataset. This implies that the embedding 
+is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
+is reliant on the distance metric used; distortions and diffusions can
+be thought of as the preservation of distances or the nearest neighbors
+graphs between the high-dimensional space and the embedding space. As distances
+can be noisy in high-dimensions, ranks can be used instead as has been 
+proposed by ... Identifying distortions and diffusions allows an analyst
+to investigate the quality of their embedding and revise them iteratively. We
+propose this can be done visually using our side-by-side tour and embedding
+views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
+brushes and brush composition.
 
-## Interactivity for embedding exploration
+ Look up nearest neighbours graph from points that lie in a brushing region.
 
-# Case Studies
+Highlight the corresponding neighbours using colour or transparency in the linked view.
 
-## Case Study 1: Fake trees
+- The $k$ nearest neighbours graph can be pre-computed quickly, for either $X$, $Y$ or both.
+- Instead of using the neighbour indices, we could use the neighbour distances instead.
+- Composition of multiple brushes could be used to show where there are matches/mismatches between nearest neighbour graphs.
 
-## Case Study 2: well-clustered smallish single cell
+Using a linked neighbourhood brush, we can 
+visually investigate the nearest neighbour relationships in the 
+high-dimensional space via brushing in the embedding view. The user can select
+the number of nearest neighbours directly, and modify the distance metric 
+used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
+can be used to pose queries in either the tour view or the embedding view;
+interface controls allow these brushes to combine using logical operators
+such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
+simple color highlights, allowing analysts to get a more holistic
+view of the effect of an embedding algorithm.
 
-## Case Study 3: largeish single cell data
+## Software Infrastructure
 
-## Case Study 4: non-spherical clusters, pdfsense
+We have implemented the above design as an open source R package 
+called `liminal` [@baseR]. The package allows analysts to construct
+concatenated visualisations, via the
+Vega-Lite grammar of interactive graphics (using `vegawidget` package)
+and provides an interface for constructing brushes and manipulating tour paths
+using the `shiny` and `tourr` packages. `liminal` also provides a stand-alone
+interface to the tour, in addition to the linked scatterplot interfaces 
+discussed above. Below we display the `liminal` API, and our approach to
+generating tour paths and user interactions.
 
-# Discussion
+### Tours as a streaming data problem
 
-# References
+- don't need to realise the entire sequence
+- instead generate new bases according to a fixed frame rate
+- allows user interactions to play/pause on the current view
+
+### Linking views via brushes
+
+## Case Studies
+
+### Case Study 1: Exploring tree structured data with tours and t-SNE
+
+(ref:fake-trees) Example from PHATE
+
+```{r fake-trees, echo = FALSE, out.width = "50%", fig.cap="(ref:fake-trees)"}
+knitr::include_graphics(here::here("img", "tree.png"))
+```
+
+#### Using t-SNE
+
+- there's a lot of parameters to tweak
+- emphasises locality, distance between clusters of points can be misleading
+- size of clusters can be misleading 
+- may require a few different runs to capture topology of data
+- if there are clusters it will find them
+
+### Case Study 2: Clustering PBMC 10x single cell RNA-seq data
+
+### Case Study 3: Single cell mouse retina data
+
+## Discussion
+
+## Acknowledgements
+
+## References

### 2020-03-24 scripts/liminal-trees.R

@@ -1,0 +1,57 @@
+## ----init-----------------------------------------------------------
+library(liminal)
+dim(fake_trees)
+
+
+## ----pc-view--------------------------------------------------------
+pcs  <- prcomp(fake_trees[, -ncol(fake_trees)])
+# var explained
+summary(pcs)$importance[3,1:12]
+
+## ----pca-xy---------------------------------------------------------
+fake_trees <- dplyr::bind_cols(fake_trees, as.data.frame(pcs$x))
+limn_xy(fake_trees, x = PC1, y = PC2, color = branches)
+
+
+## ----tsne-xy--------------------------------------------------------
+set.seed(2099)
+tsne <- Rtsne::Rtsne(dplyr::select(fake_trees, dplyr::starts_with("dim")))
+
+tsne_df <- data.frame(tsneX = tsne$Y[,1],
+                      tsneY = tsne$Y[,2],
+                      branches = fake_trees$branches)
+
+limn_xy(tsne_df, x = tsneX, y = tsneY, color = branches)
+
+# now together
+limn_tour_xylink(
+  dplyr::select(fake_trees, PC1:PC12, branches),
+  tsne_df,
+  x_color = branches,
+  y_color = branches
+)
+
+
+
+## ----pca-init--------------------------------------------------------
+tsne_v2 <- Rtsne::Rtsne(
+  dplyr::select(fake_trees, dplyr::starts_with("dim")),
+  Y_init = clamp_sd(as.matrix(dplyr::select(fake_trees, PC1,PC2)), sd = 1e-4),
+  perplexity = nrow(fake_trees) / 100,
+  eta = nrow(fake_trees) / 48,
+  max_iter = 3000
+)
+
+tsne_v2_df <- data.frame(
+  tsneX = tsne_v2$Y[,1], 
+  tsneY = tsne_v2$Y[,2], 
+  branches = fake_trees$branches
+)
+
+limn_tour_xylink(
+  dplyr::select(fake_trees, PC1:PC12, branches),
+  tsne_v2_df,
+  x_color = branches,
+  y_color = branches
+)
+

### 2020-04-17 paper.Rmd
update bibliography; wrap lines
section on DR methods in progress

@@
 
-There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows.
-The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of
-local and global structure in the resulting visualisation.
-We present a visual analytics framework for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
-A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
-The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view.
-By linking the tour to a view
-obtained from an NLDR method, we can preserve global structure and through user interactions like spatial linked brushing observe where
-the NLDR view may be misleading. We display several case 
-studies from single cell genomics, that shows our approach is useful for
-cluster orientation tasks. 
-The implementation of our framework is available as an R package called `liminal` available at https://github.com/sa-lee/liminal.
+There has been a rapid uptake in the use of non-linear dimensionality reduction
+(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
+the natural sciences as part of cluster orientation and dimension reduction
+workflows. The appropriate use of these methods is made difficult by their
+complex parameterisations and the multitude of decisions required to balance
+the preservation of local and global structure in the resulting visualisation.
+We present a visual analytics framework for the pragmatic usage of NLDR methods
+by combining them with a technique called the tour. A tour is a sequence of
+interpolated linear projections of multivariate data onto a lower dimensional
+space. The sequence is displayed as a dynamic visualisation, allowing a user to
+see the shadows the high-dimensional data casts in a lower dimensional view. By
+linking the tour to a view obtained from an NLDR method, we can preserve global
+structure and through user interactions like spatial linked brushing observe
+where the NLDR view may be misleading. We display several case studies from
+single cell genomics, that shows our approach is useful for cluster orientation
+tasks.  The implementation of our framework is available as an R package called
+`liminal` available at https://github.com/sa-lee/liminal.
 
@@
 space into a low-dimensional one now form a core step of the data analysis
-workflow where they are used to ascertain hidden structure and de-noise data for
-downstream analysis (thereby nullifying the 'curse of dimensionality').
+workflow where they are used to ascertain hidden structure and de-noise data
+for downstream analysis (thereby nullifying the 'curse of dimensionality').
 
@@
 ground truth.
-
-Embedding methods can be classified into two broad groups: linear and
-non-linear methods. Linear methods perform a linear transformation of the data;
-one example is principal components analysis (PCA) which performs an
-eigendecomposition of the estimated sample covariance matrix. The eigenvalues
-are sorted in decreasing order and represent the variance explained by each
-component (eigenvector). A common approach to deciding on the number of
-principal components to retain is to plot the proportion of variance explained
-by each component and choose a cut-off.
-
-Non-linear methods generally perform pre-processing on the high-dimensional data
-such as generating a neighborhood graph and perform transformations on the
-pre-processed form. We restrict our attention to two methods that are commonly
-used in the literature: t-distributed stochastic neighbor embedding
-(t-SNE) and uniform manifold alignment and projection (UMAP). The t-SNE
-algorithm estimates the pairwise similarity of (Euclidean) distances of points in a high
-dimensional space using a Gaussian distribution and then estimates a
-configuration in the low dimensional embedding space by modelling similarities
-using a t-distribution with 1 degree of freedom. The resulting configuration is
-the one that minimizes the Kullback-Leibler divergence between the two
-distributions. A recent theoretical contribution by @Linderman2019-dq proved
-that t-SNE can recover spherical and well separated cluster shapes, and
-proposed new approaches for tuning the optimisation parameters. It is a known
-problem that t-SNE can have trouble recovering global structure and that
-configurations can be highly dependent on how the algorithm is initialised and
-parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
-related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
-neighbor graph. Its main differences are that it uses a different cost function
-(cross entropy) which is optimized using stochastic gradient descent and
-defines a different kernel for similarities in the low dimensional space. Due
-to it's computational speed it's possible to generate UMAP embeddings in more
-than three dimensions.
 
@@
 
-## Dimensionality Reduction
+## Dimensionality Reduction Overview
+
+In the following section, we provide a brief overview of the goals
+of DR methods with respect to the data analyst, in addition to their high level the 
+mathematical details. To begin we suppose the data is in the form of an $n \times p$ 
+rectangular matrix of real numbers, $X$, where $n$ is the number of  observations
+in $p$ dimensions. The purpose of any DR algorithm is to find low-dimensonal
+representation of the data, $Y$, such that $Y$ is an $n \times d$ matrix 
+where $d \ll p$. The aspiration is that the DR procedure will remove noise
+in the data while revealing hidden patterns and structure.
+
+DR methods can be classified into two broad groups: linear and
+non-linear methods. Linear methods perform a linear transformation of the data,
+that is, $Y$ is a linear transformation of $X$;
+one example is principal components analysis (PCA) which performs an
+eigendecomposition of the estimated sample covariance matrix. The eigenvalues
+are sorted in decreasing order and represent the variance explained by each
+component (eigenvector). A common approach to deciding on the number of
+principal components to retain is to plot the proportion of variance explained
+by each component and choose a cut-off.
+
+For non-linear methods $Y$ is generated via a pre-processed form of the input 
+$X$ such as the $k$-nearest neighbours graph or via a kernel transformation. 
+In this paper we restrict our attention to two methods that are commonly
+used in the literature: t-distributed stochastic neighbor embedding (t-SNE) and
+uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
+the pairwise similarity of (Euclidean) distances of points in a high
+dimensional space using a Gaussian distribution and then estimates a
+configuration in the low dimensional embedding space by modelling similarities
+using a t-distribution with 1 degree of freedom. The resulting configuration is
+the one that minimizes the Kullback-Leibler divergence between the two
+distributions. A recent theoretical contribution by @Linderman2019-dq proved
+that t-SNE can recover spherical and well separated cluster shapes, and
+proposed new approaches for tuning the optimisation parameters. It is a known
+problem that t-SNE can have trouble recovering global structure and that
+configurations can be highly dependent on how the algorithm is initialised and
+parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
+related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
+neighbor graph. Its main differences are that it uses a different cost function
+(cross entropy) which is optimized using stochastic gradient descent and
+defines a different kernel for similarities in the low dimensional space. Due
+to it's computational speed it's possible to generate UMAP embeddings in more
+than three dimensions.
+
+- there's a lot of parameters to tweak
+- emphasises locality, distance between clusters of points can be misleading
+- size of clusters can be misleading
+- may require a few different runs to capture topology of data
+- if there are clusters it will find them
 
@@
 
-We have combined the tour view in a side by side layout with 
-a scatter plot view that represents the output of an embedding algorithm. 
-These views are linked; analysts can brush regions or
-highlight collections of points in either view. Linked highlighting can
-be performed when points have been previously labeled according to some
-discrete structure, i.e. cluster labels are available. This is achieved
-via the analyst clicking on groups in the legend, which causes unselected groupings to have their points become less opaque. Consequently, 
-simple linked highlighting can alleviate a known downfall of methods such as 
-UMAP or t-SNE: that is distances between clusters are misleading. By
-highlighting corresponding clusters in the tour view, the analyst can
-see the relationship between clusters, and therefore obtain a more accurate 
-representation of the topology of their data.
+We have combined the tour view in a side by side layout with a scatter plot
+view that represents the output of an embedding algorithm.  These views are
+linked; analysts can brush regions or highlight collections of points in either
+view. Linked highlighting can be performed when points have been previously
+labeled according to some discrete structure, i.e. cluster labels are
+available. This is achieved via the analyst clicking on groups in the legend,
+which causes unselected groupings to have their points become less opaque.
+Consequently, simple linked highlighting can alleviate a known downfall of
+methods such as UMAP or t-SNE: that is distances between clusters are
+misleading. By highlighting corresponding clusters in the tour view, the
+analyst can see the relationship between clusters, and therefore obtain a more
+accurate representation of the topology of their data.
 
@@
 paused and corresponding points in the embedding view are highlighted.
-Likewise, when brushing occurs in the embedding view, corresponding points
-in the tour view are highlighted. In this case, an analyst can use brushing
-for manually identifying clusters and verifying cluster locations and shapes:
+Likewise, when brushing occurs in the embedding view, corresponding points in
+the tour view are highlighted. In this case, an analyst can use brushing for
+manually identifying clusters and verifying cluster locations and shapes:
 brushing in the embedding view gives analysts a sense of the shape and
@@
 
-As mentioned previously, when using any DR method, we are assuming
-the embedding is representative of the high-dimensional dataset it was
-computed from. Defining what it means for embedding to be 'representative` or
-'faithful' to high-dimensional data is ill-posed and depends on the underlying
-task an analyst is trying to achieve. At the very minimum, we are interested
-in distortions and diffusions of the high-dimensional data. Distortions
-occur when points that are near each other in the embedding view are 
-far from each other in the original dataset. This implies that the embedding 
-is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
-is reliant on the distance metric used; distortions and diffusions can
-be thought of as the preservation of distances or the nearest neighbors
-graphs between the high-dimensional space and the embedding space. As distances
-can be noisy in high-dimensions, ranks can be used instead as has been 
-proposed by ... Identifying distortions and diffusions allows an analyst
-to investigate the quality of their embedding and revise them iteratively. We
-propose this can be done visually using our side-by-side tour and embedding
-views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
-brushes and brush composition.
+As mentioned previously, when using any DR method, we are assuming the
+embedding is representative of the high-dimensional dataset it was computed
+from. Defining what it means for embedding to be 'representative` or 'faithful'
+to high-dimensional data is ill-posed and depends on the underlying task an
+analyst is trying to achieve. At the very minimum, we are interested in
+distortions and diffusions of the high-dimensional data. Distortions occur when
+points that are near each other in the embedding view are far from each other
+in the original dataset. This implies that the embedding is not continuous.
+Diffusions occur when points are far from each other in the embedding view are
+near in the original data.  Whether, points are near or far is reliant on the
+distance metric used; distortions and diffusions can be thought of as the
+preservation of distances or the nearest neighbors graphs between the
+high-dimensional space and the embedding space. As distances can be noisy in
+high-dimensions, ranks can be used instead as has been proposed by ...
+Identifying distortions and diffusions allows an analyst to investigate the
+quality of their embedding and revise them iteratively. We propose this can be
+done visually using our side-by-side tour and embedding views. Some quality
+checks, such as finding distortions can be identified using simple linked
+brushing, however we can also interrogate them via spatial brushes and brush
+composition.
 
@@
 
-Highlight the corresponding neighbours using colour or transparency in the linked view.
+Highlight the corresponding neighbours using colour or transparency in the
+linked view.
 
@@
 
-Using a linked neighbourhood brush, we can 
-visually investigate the nearest neighbour relationships in the 
-high-dimensional space via brushing in the embedding view. The user can select
-the number of nearest neighbours directly, and modify the distance metric 
-used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
-can be used to pose queries in either the tour view or the embedding view;
-interface controls allow these brushes to combine using logical operators
-such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
-simple color highlights, allowing analysts to get a more holistic
-view of the effect of an embedding algorithm.
+Using a linked neighbourhood brush, we can visually investigate the nearest
+neighbour relationships in the high-dimensional space via brushing in the
+embedding view. The user can select the number of nearest neighbours directly,
+and modify the distance metric used for determining the neighbours. This allows
+users to interrogate the stability of clusters generated in the embedding view.
+Multiple brushes can be used to pose queries in either the tour view or the
+embedding view; interface controls allow these brushes to combine using logical
+operators such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
+simple color highlights, allowing analysts to get a more holistic view of the
+effect of an embedding algorithm.
 
@@
 
-We have implemented the above design as an open source R package 
-called `liminal` [@baseR]. The package allows analysts to construct
-concatenated visualisations, via the
-Vega-Lite grammar of interactive graphics (using `vegawidget` package)
-and provides an interface for constructing brushes and manipulating tour paths
-using the `shiny` and `tourr` packages. `liminal` also provides a stand-alone
-interface to the tour, in addition to the linked scatterplot interfaces 
-discussed above. Below we display the `liminal` API, and our approach to
-generating tour paths and user interactions.
+We have implemented the above design as an open source R package called
+`liminal` [@baseR]. The package allows analysts to construct concatenated
+visualisations, via the Vega-Lite grammar of interactive graphics (using
+`vegawidget` package) and provides an interface for constructing brushes and
+manipulating tour paths using the `shiny` and `tourr` packages. `liminal` also
+provides a stand-alone interface to the tour, in addition to the linked
+scatterplot interfaces discussed above. Below we display the `liminal` API, and
+our approach to generating tour paths and user interactions.
 
@@
 
-#### Using t-SNE
-
-- there's a lot of parameters to tweak
-- emphasises locality, distance between clusters of points can be misleading
-- size of clusters can be misleading 
-- may require a few different runs to capture topology of data
-- if there are clusters it will find them
-
 ### Case Study 2: Clustering PBMC 10x single cell RNA-seq data
 
-### Case Study 3: Single cell mouse retina data
+### Case Study 3: QuickDraw images
 

### 2020-04-17 paper.Rmd
first draft of DR section

@@
 of DR methods with respect to the data analyst, in addition to their high level the 
-mathematical details. To begin we suppose the data is in the form of an $n \times p$ 
-rectangular matrix of real numbers, $X$, where $n$ is the number of  observations
-in $p$ dimensions. The purpose of any DR algorithm is to find low-dimensonal
-representation of the data, $Y$, such that $Y$ is an $n \times d$ matrix 
-where $d \ll p$. The aspiration is that the DR procedure will remove noise
-in the data while revealing hidden patterns and structure.
+mathematical details. To begin we suppose the data is in the form 
+rectangular matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, 
+where $n$ is the number of observations in $p$ dimensions. 
+The purpose of any DR algorithm is to find low-dimensonal
+representation of the data, $Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that
+$Y$ is an $n \times d$ matrix where $d \ll p$. The hope of the analyst is that
+the DR procedure to produces $Y$ will remove noise in the original dataset while
+retaining latent structure.
 
@@
 $X$ such as the $k$-nearest neighbours graph or via a kernel transformation. 
-In this paper we restrict our attention to two methods that are commonly
+Here we restrict our attention to two recent methods that are commonly
 used in the literature: t-distributed stochastic neighbor embedding (t-SNE) and
@@
 configuration in the low dimensional embedding space by modelling similarities
-using a t-distribution with 1 degree of freedom. The resulting configuration is
-the one that minimizes the Kullback-Leibler divergence between the two
-distributions. A recent theoretical contribution by @Linderman2019-dq proved
+using a t-distribution with 1 degree of freedom. There are several subtleties
+to the to use of the algorithm that are revealed by stepping through its
+machinery.
+
+To begin, t-SNE transforms pairwise distances between  
+$\mathbf{x_i}$ and $\mathbf{x_j}$  to similiarties using a Gaussian kernel:
+
+$$
+p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 / 2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 / 2\sigma_i^2)}
+$$
+
+The conditional probabilities are then normalised and symmetrised to form a 
+joint probability  distrubition via averaging:
+
+$$
+p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n}
+$$
+
+The variance parameter of the 
+Gaussian kernel is controlled by the analyst using a fixed value of perplexity
+for all observations:
+
+$$
+\text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i}))
+$$
+As the perplexity increases, $sigma_^2_{i}$ increases, until its bounded
+above by the number of observations , $n-1$, in the data, corresponding to  
+$\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a 
+nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all 
+observations that are not in the $\mathcal{O}(\text{perplexity}_i)$
+neighbourhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].
+
+Next, the target low-dimensional space, $Y$, pairwise distances between 
+$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability 
+distrubtion using a Cauchy kernel:
+
+$$
+q_{ij} = \frac{w_{ij}}{Z} \\
+\text{where } w_{ij} = \frac{1}{ 1 + \lVert \mathbf{y_i - y_j} \rVert ^ 2} 
+\text{ and } Z = \sum_{k \ne l} w_{kl}.
+$$
+
+The resulting configuration $Y$ is
+the one that minimizes the Kullback-Leibler divergence between 
+the probability distrubtions formed via similiarties of observations in $X$, $\mathcal{P}$ 
+and similiarities of observations in $Y$, $\mathcal{Q}$:
+
+$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log \frac{p_{ij}}{q_{ij}}$$
+
+Re-writing the loss function in terms of attractive (right) and repulsive (left) 
+forces we obtain:
+
+$$
+\mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} + \log\sum_{i \ne j} w_{ij}
+$$
+
+So when the the loss function is minimised this corresponds to large
+attractive forces, that is, the pairwise distances in $Y$ are small when
+there are non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The
+repulsive force should also be small when the loss function is minimised, 
+that is, when pairwise distances in $Y$ are large irregardless of the magnitude
+of the corresponding distances in $X$.
+
+Taken together, these details reveal the sheer number of decisions that an 
+anlayst must make. How does one choose the perplexity? How should the parameters
+that control the optimisation of the loss function (done with stochastic
+gradient descent), like the number of iterations, or early exaggeration (
+a multiplier of the attractive force at the beginning of the optimisation),
+or the learning rate be selected?  It is a known
+problem that t-SNE can have trouble recovering topology and that
+configurations can be highly dependent on how the algorithm is initialised and
+parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. If the goal is cluster
+orientation a recent theoretical contribution by @Linderman2019-dq proved
 that t-SNE can recover spherical and well separated cluster shapes, and
-proposed new approaches for tuning the optimisation parameters. It is a known
-problem that t-SNE can have trouble recovering global structure and that
-configurations can be highly dependent on how the algorithm is initialised and
-parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
+proposed new approaches for tuning the optimisation parameters. However,
+the cluster sizes and their relative orientation from a $t-SNE$ view can be 
+misleading perceptually, due to the algorithms emphasis on locality.
+
+Another recent method UMAP has seen a large rise in popularity (at least in
+single cell transcriptomics) [@McInnes2018-co]. is a method that is
 related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
@@
 to it's computational speed it's possible to generate UMAP embeddings in more
-than three dimensions.
-
-- there's a lot of parameters to tweak
-- emphasises locality, distance between clusters of points can be misleading
-- size of clusters can be misleading
-- may require a few different runs to capture topology of data
-- if there are clusters it will find them
+than three dimensions. It appears to suffer from the same perceptual issues
+as t-SNE, however it supposedly preservers global structure better
+than t-SNE [@Coenen2019-to].
 
@@
 
-We propose using tours as part of an analyst's workflow in in performing
-dimensionality reduction tasks. We have made extensive use of ensemble
+We propose using tours as part of an analyst's workflow when using embedding
+methods like t-SNE. We have made extensive use of ensemble
 graphics, that is aligning related plots alongside each other to provide
@@
 view. Linked highlighting can be performed when points have been previously
-labeled according to some discrete structure, i.e. cluster labels are
+labelled according to some discrete structure, i.e. cluster labels are
 available. This is achieved via the analyst clicking on groups in the legend,
@@
 distance metric used; distortions and diffusions can be thought of as the
-preservation of distances or the nearest neighbors graphs between the
+preservation of distances or the nearest neighbours graphs between the
 high-dimensional space and the embedding space. As distances can be noisy in
@@
 operators such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
-simple color highlights, allowing analysts to get a more holistic view of the
+simple colour highlights, allowing analysts to get a more holistic view of the
 effect of an embedding algorithm.

### 2020-04-17 paper.Rmd
add citations

@@
 configurations can be highly dependent on how the algorithm is initialised and
-parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. If the goal is cluster
-orientation a recent theoretical contribution by @Linderman2019-dq proved
+parameterized [@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is
+cluster orientation a recent theoretical contribution by @Linderman2019-dq proved
 that t-SNE can recover spherical and well separated cluster shapes, and

### 2020-04-17 paper.Rmd
update section

@@
 
-To begin, t-SNE transforms pairwise distances between  
-$\mathbf{x_i}$ and $\mathbf{x_j}$  to similiarties using a Gaussian kernel:
+To begin, t-SNE transforms pairwise distances between   $\mathbf{x_i}$ and 
+$\mathbf{x_j}$  to similiarties using a Gaussian kernel:
 
@@
 $$
-As the perplexity increases, $sigma_^2_{i}$ increases, until its bounded
+As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded
 above by the number of observations , $n-1$, in the data, corresponding to  

### 2020-04-20 paper.Rmd
update manuscript

@@
 where the NLDR view may be misleading. We display several case studies from
-single cell genomics, that shows our approach is useful for cluster orientation
-tasks.  The implementation of our framework is available as an R package called
+single cell transciptomics and image processing, that shows our approach is 
+useful for cluster orientation tasks.  The implementation of our framework 
+is available as an R package called
 `liminal` available at https://github.com/sa-lee/liminal.
@@
 
+The next section steps through three use cases of our approach 
+using simulated and real data. In the first case study, we step through
+data that lies on tree structure (similar to trajectories in single
+cell transcriptomics) to that highlights an iterative approach to modifying
+t-SNE views based on what we learn in the high-dimensional space. We show
+how through simple linked brushing, we can see when the assumptions
+of continuity of the embedding map are broken. In the second case study,
+we explore a well-clustered single cell transcriptomics data of mice retina...
+Finally, in the last case study we explore a subset of the QuickDraw dataset,
+we use `liminal` to identify instances of images from different labels that
+appear similiar.
+
 ### Case Study 1: Exploring tree structured data with tours and t-SNE
 
-(ref:fake-trees) Example from PHATE
+(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
+The data lies 2-d tree like structure consisting of ten branches.
+This data is available in the `phateR` package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
 
-```{r fake-trees, echo = FALSE, out.width = "50%", fig.cap="(ref:fake-trees)"}
+```{r fake-trees, echo = FALSE, out.width = "50%", fig.align = "center", fig.cap="(ref:fake-trees)"}
 knitr::include_graphics(here::here("img", "tree.png"))
@@
 
-### Case Study 2: Clustering PBMC 10x single cell RNA-seq data
+### Case Study 2: Clustering single cell RNA-seq data
 

### 2020-04-22 paper.Rmd
update case study section

@@
 
+?? include details of tour here
+
 ## Visual Design
@@
 
+In this case study we explore some simulated noisy tree structured data (figure \@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break topology of the data, and then seeing if we can resolve this by tweaking the default parameters
+with reference to the global view of the data set.
+
 (ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
@@
 ```
+
+First, we apply principal components and restrict the results down to the first twelve principal components (which makes up approximately 70% of the variance explained in the data) to use with the grand tour.
+
+<!-- can we show R code? -->
+
+Moreover,
+we run t-SNE using the default arguments on the complete data (this keeps the first 50 PCs, sets the perpelxity to equal 30 and performs random initalisation) using the `Rtsne` package [@Rtsne].
+We then create a linked tour with t-SNE layout with `liminal` as shown in the following video :
+
+<embededded video 1>
+
+From the video, we see that the t-SNE view has been unable to recapitulate the topology of the tree - the backbone (blue) branch has been split into two fragments.
+We can see this immediately via the linked highlighting over both plots.
+If we click on the legend for the zeroth branch, the blue colored points on each view are higlighted and the remaining points are made transparent.
+From, here it becomes apparent from the tour view that the blue branch forms the backbone of the tree and is connected to all other branches.
 

### 2020-04-22 paper.Rmd
update case studies section

@@
 
-From the video, we see that the t-SNE view has been unable to recapitulate the topology of the tree - the backbone (blue) branch has been split into two fragments.
+From the video, we see that the t-SNE view has been unable to recapitulate the topology of the tree - the backbone (blue) branch has been split into three fragments.
 We can see this immediately via the linked highlighting over both plots.
@@
 From, here it becomes apparent from the tour view that the blue branch forms the backbone of the tree and is connected to all other branches.
+We can also again see that cluster sizes formed via t-SNE can be misleading; from the tour view there is a lot of noise along the branches, while this does not appear to
+be the case for the t-SNE result.
+
+From the first view, we modify the inputs to the t-SNE view, to try and produce a better trade-off between local structure and retain the topology of the data.
+We keep every parameter the same except that we initialise $Y$ with the first two PCs (scaled to have standard deviation 1e-4) instead of the default random initialisation and increase the perplexity from 30 to 100.
+We then combine these results
+with our tour view as displayed in the video:
+
+<embedded video 2>
+
+The video shows that this selection of parameters results
+in the the tips of the branch
 

### 2020-04-22 scripts/liminal-trees.R

@@ -37,9 +37,7 @@ limn_tour_xylink(
 tsne_v2 <- Rtsne::Rtsne(
   dplyr::select(fake_trees, dplyr::starts_with("dim")),
   Y_init = clamp_sd(as.matrix(dplyr::select(fake_trees, PC1,PC2)), sd = 1e-4),
-  perplexity = nrow(fake_trees) / 100,
-  eta = nrow(fake_trees) / 48,
-  max_iter = 3000
+  perplexity = 100
 )
 
 tsne_v2_df <- data.frame(

### 2020-04-23 paper.Rmd
update case study one

@@
 single cell transciptomics and image processing, that shows our approach is
-useful for cluster orientation tasks.  The implementation of our framework 
-is available as an R package called
-`liminal` available at https://github.com/sa-lee/liminal.
+useful for cluster orientation tasks.  The implementation of our framework is
+available as an R package called `liminal` available at
+https://github.com/sa-lee/liminal.
 
@@
 Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
-PCA live in the browser and provides interactions to color points, and
-highlights nearest neighbors [@Smilkov2016-hp].
+PCA live in the browser and provides interactions to colour points, and
+highlights nearest neighbours [@Smilkov2016-hp].
 
@@
 
-In the following section, we provide a brief overview of the goals
-of DR methods with respect to the data analyst, in addition to their high level the 
-mathematical details. To begin we suppose the data is in the form 
-rectangular matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, 
-where $n$ is the number of observations in $p$ dimensions. 
-The purpose of any DR algorithm is to find low-dimensonal
-representation of the data, $Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that
-$Y$ is an $n \times d$ matrix where $d \ll p$. The hope of the analyst is that
-the DR procedure to produces $Y$ will remove noise in the original dataset while
-retaining latent structure.
+In the following section, we provide a brief overview of the goals of DR
+methods with respect to the data analyst, in addition to their high level the
+mathematical details. To begin we suppose the data is in the form rectangular
+matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
+the number of observations in $p$ dimensions.  The purpose of any DR algorithm
+is to find low-dimensional representation of the data, $Y = [\mathbf{y_1},
+\dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$ matrix where $d \ll p$.
+The hope of the analyst is that the DR procedure to produces $Y$ will remove
+noise in the original dataset while retaining latent structure.
 
-DR methods can be classified into two broad groups: linear and
-non-linear methods. Linear methods perform a linear transformation of the data,
-that is, $Y$ is a linear transformation of $X$;
-one example is principal components analysis (PCA) which performs an
-eigendecomposition of the estimated sample covariance matrix. The eigenvalues
-are sorted in decreasing order and represent the variance explained by each
-component (eigenvector). A common approach to deciding on the number of
-principal components to retain is to plot the proportion of variance explained
-by each component and choose a cut-off.
+DR methods can be classified into two broad groups: linear and non-linear
+methods. Linear methods perform a linear transformation of the data, that is,
+$Y$ is a linear transformation of $X$; one example is principal components
+analysis (PCA) which performs an eigendecomposition of the estimated sample
+covariance matrix. The eigenvalues are sorted in decreasing order and represent
+the variance explained by each component (eigenvector). A common approach to
+deciding on the number of principal components to retain is to plot the
+proportion of variance explained by each component and choose a cut-off.
 
@@
 $X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
-Here we restrict our attention to two recent methods that are commonly
-used in the literature: t-distributed stochastic neighbor embedding (t-SNE) and
+Here we restrict our attention to two recent methods that are commonly used in
+the literature: t-distributed stochastic neighbour embedding (t-SNE) and
 uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
-the pairwise similarity of (Euclidean) distances of points in a high
+the pair-wise similarity of (Euclidean) distances of points in a high
 dimensional space using a Gaussian distribution and then estimates a
@@
 
-To begin, t-SNE transforms pairwise distances between   $\mathbf{x_i}$ and 
-$\mathbf{x_j}$  to similiarties using a Gaussian kernel:
+To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
+$\mathbf{x_j}$ to similarities using a Gaussian kernel:
 
-$$
-p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 / 2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 / 2\sigma_i^2)}
-$$
+$$ p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 /
+2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 /
+2\sigma_i^2)} $$
 
 The conditional probabilities are then normalised and symmetrised to form a
-joint probability  distrubition via averaging:
+joint probability distribution via averaging:
 
-$$
-p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n}
-$$
+$$ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n} $$
 
-The variance parameter of the 
-Gaussian kernel is controlled by the analyst using a fixed value of perplexity
-for all observations:
+The variance parameter of the Gaussian kernel is controlled by the analyst
+using a fixed value of perplexity for all observations:
 
-$$
-\text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i}))
-$$
-As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded
-above by the number of observations , $n-1$, in the data, corresponding to  
+$$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
+As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded above
+by the number of observations , $n-1$, in the data, corresponding to
 $\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a
@@
 
-Next, the target low-dimensional space, $Y$, pairwise distances between 
+Next, the target low-dimensional space, $Y$, pair-wise distances between
 $\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
-distrubtion using a Cauchy kernel:
+distribution using a Cauchy kernel:
 
-$$
-q_{ij} = \frac{w_{ij}}{Z} \\
-\text{where } w_{ij} = \frac{1}{ 1 + \lVert \mathbf{y_i - y_j} \rVert ^ 2} 
-\text{ and } Z = \sum_{k \ne l} w_{kl}.
-$$
+$$ q_{ij} = \frac{w_{ij}}{Z} \\ \text{where } w_{ij} = \frac{1}{ 1 + \lVert
+\mathbf{y_i - y_j} \rVert ^ 2} \text{ and } Z = \sum_{k \ne l} w_{kl}. $$
 
-The resulting configuration $Y$ is
-the one that minimizes the Kullback-Leibler divergence between 
-the probability distrubtions formed via similiarties of observations in $X$, $\mathcal{P}$ 
-and similiarities of observations in $Y$, $\mathcal{Q}$:
+The resulting configuration $Y$ is the one that minimizes the Kullback-Leibler
+divergence between the probability distributions formed via similarities of
+observations in $X$, $\mathcal{P}$ and similarities of observations in $Y$,
+$\mathcal{Q}$:
 
-$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log \frac{p_{ij}}{q_{ij}}$$
+$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log
+\frac{p_{ij}}{q_{ij}}$$
 
-Re-writing the loss function in terms of attractive (right) and repulsive (left) 
-forces we obtain:
+Re-writing the loss function in terms of attractive (right) and repulsive
+(left) forces we obtain:
 
-$$
-\mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} + \log\sum_{i \ne j} w_{ij}
-$$
+$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} +
+\log\sum_{i \ne j} w_{ij} $$
 
-So when the the loss function is minimised this corresponds to large
-attractive forces, that is, the pairwise distances in $Y$ are small when
-there are non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The
-repulsive force should also be small when the loss function is minimised, 
-that is, when pairwise distances in $Y$ are large irregardless of the magnitude
-of the corresponding distances in $X$.
+So when the the loss function is minimised this corresponds to large attractive
+forces, that is, the pair-wise distances in $Y$ are small when there are
+non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
+should also be small when the loss function is minimised, that is, when
+pair-wise distances in $Y$ are large regardless of the magnitude of the
+corresponding distances in $X$.
 
 Taken together, these details reveal the sheer number of decisions that an
-anlayst must make. How does one choose the perplexity? How should the parameters
-that control the optimisation of the loss function (done with stochastic
-gradient descent), like the number of iterations, or early exaggeration (
-a multiplier of the attractive force at the beginning of the optimisation),
-or the learning rate be selected?  It is a known
-problem that t-SNE can have trouble recovering topology and that
-configurations can be highly dependent on how the algorithm is initialised and
-parameterized [@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is
-cluster orientation a recent theoretical contribution by @Linderman2019-dq proved
-that t-SNE can recover spherical and well separated cluster shapes, and
-proposed new approaches for tuning the optimisation parameters. However,
-the cluster sizes and their relative orientation from a $t-SNE$ view can be 
-misleading perceptually, due to the algorithms emphasis on locality.
+analyst must make. How does one choose the perplexity? How should the
+parameters that control the optimisation of the loss function (done with
+stochastic gradient descent), like the number of iterations, or early
+exaggeration ( a multiplier of the attractive force at the beginning of the
+optimisation), or the learning rate be selected?  It is a known problem that
+t-SNE can have trouble recovering topology and that configurations can be
+highly dependent on how the algorithm is initialised and parameterized
+[@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is cluster
+orientation a recent theoretical contribution by @Linderman2019-dq proved that
+t-SNE can recover spherical and well separated cluster shapes, and proposed new
+approaches for tuning the optimisation parameters. However, the cluster sizes
+and their relative orientation from a $t-SNE$ view can be misleading
+perceptually, due to the algorithms emphasis on locality.
 
 Another recent method UMAP has seen a large rise in popularity (at least in
-single cell transcriptomics) [@McInnes2018-co]. is a method that is
-related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
-neighbor graph. Its main differences are that it uses a different cost function
-(cross entropy) which is optimized using stochastic gradient descent and
-defines a different kernel for similarities in the low dimensional space. Due
-to it's computational speed it's possible to generate UMAP embeddings in more
-than three dimensions. It appears to suffer from the same perceptual issues
-as t-SNE, however it supposedly preservers global structure better
-than t-SNE [@Coenen2019-to].
+single cell transcriptomics) [@McInnes2018-co]. is a method that is related to
+LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
+Its main differences are that it uses a different cost function (cross entropy)
+which is optimized using stochastic gradient descent and defines a different
+kernel for similarities in the low dimensional space. Due to it's computational
+speed it's possible to generate UMAP embeddings in more than three dimensions.
+It appears to suffer from the same perceptual issues as t-SNE, however it
+supposedly preservers global structure better than t-SNE [@Coenen2019-to].
 
@@
 We propose using tours as part of an analyst's workflow when using embedding
-methods like t-SNE. We have made extensive use of ensemble
-graphics, that is aligning related plots alongside each other to provide
-context. As we will see in the case studies, this allows analysts to quickly
-compare views from embedding methods and allows them to see how the embedding
-method alters the global structure of their data. Using ensembles allows the
-use of interaction techniques, that allow analysts to perform cluster
-orientation tasks via linking multiple views. This approach allows our
-interface, to achieve the three principles for interactive high-dimensional
-data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
-and making comparisons.
+methods like t-SNE. We have made extensive use of ensemble graphics, that is
+aligning related plots alongside each other to provide context. As we will see
+in the case studies, this allows analysts to quickly compare views from
+embedding methods and allows them to see how the embedding method alters the
+global structure of their data. Using ensembles allows the use of interaction
+techniques, that allow analysts to perform cluster orientation tasks via
+linking multiple views. This approach allows our interface, to achieve the
+three principles for interactive high-dimensional data visualisation outlined
+by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.
 
@@
 
-The next section steps through three use cases of our approach 
-using simulated and real data. In the first case study, we step through
-data that lies on tree structure (similar to trajectories in single
-cell transcriptomics) to that highlights an iterative approach to modifying
-t-SNE views based on what we learn in the high-dimensional space. We show
-how through simple linked brushing, we can see when the assumptions
-of continuity of the embedding map are broken. In the second case study,
-we explore a well-clustered single cell transcriptomics data of mice retina...
-Finally, in the last case study we explore a subset of the QuickDraw dataset,
-we use `liminal` to identify instances of images from different labels that
-appear similiar.
+The next section steps through three use cases of our approach using simulated
+and real data. In the first case study, we step through data that lies on tree
+structure (similar to trajectories in single cell transcriptomics) to that
+highlights an iterative approach to modifying t-SNE views based on what we
+learn in the high-dimensional space. We show how through simple linked
+brushing, we can see when the assumptions of continuity of the embedding map
+are broken. In the second case study, we explore a well-clustered single cell
+transcriptomics data of mice retina... Finally, in the last case study we
+explore a subset of the QuickDraw dataset, we use `liminal` to identify
+instances of images from different labels that appear similar.
 
@@
 
-In this case study we explore some simulated noisy tree structured data (figure \@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break topology of the data, and then seeing if we can resolve this by tweaking the default parameters
-with reference to the global view of the data set.
+In this case study we explore some simulated noisy tree structured data (figure
+\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
+topology of the data, and then seeing if we can resolve this by tweaking the
+default parameters with reference to the global view of the data set.
 
@@
 
-First, we apply principal components and restrict the results down to the first twelve principal components (which makes up approximately 70% of the variance explained in the data) to use with the grand tour.
+First, we apply principal components and restrict the results down to the first
+twelve principal components (which makes up approximately 70% of the variance
+explained in the data) to use with the grand tour.
 
@@
 
-Moreover,
-we run t-SNE using the default arguments on the complete data (this keeps the first 50 PCs, sets the perpelxity to equal 30 and performs random initalisation) using the `Rtsne` package [@Rtsne].
-We then create a linked tour with t-SNE layout with `liminal` as shown in the following video :
+Moreover, we run t-SNE using the default arguments on the complete data (this
+keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
+initialisation) using the `Rtsne` package [@Rtsne]. We then create a linked
+tour with t-SNE layout with `liminal` as shown in the following video :
 
-<embededded video 1>
+<embedded video 1>
 
-From the video, we see that the t-SNE view has been unable to recapitulate the topology of the tree - the backbone (blue) branch has been split into three fragments.
-We can see this immediately via the linked highlighting over both plots.
-If we click on the legend for the zeroth branch, the blue colored points on each view are higlighted and the remaining points are made transparent.
-From, here it becomes apparent from the tour view that the blue branch forms the backbone of the tree and is connected to all other branches.
-We can also again see that cluster sizes formed via t-SNE can be misleading; from the tour view there is a lot of noise along the branches, while this does not appear to
+From the video, we see that the t-SNE view has been unable to recapitulate the
+topology of the tree - the backbone (blue) branch has been split into three
+fragments. We can see this immediately via the linked highlighting over both
+plots. If we click on the legend for the zero branch, the blue coloured points
+on each view are highlighted and the remaining points are made transparent.
+From, here it becomes apparent from the tour view that the blue branch forms
+the backbone of the tree and is connected to all other branches. We can also
+again see that cluster sizes formed via t-SNE can be misleading; from the tour
+view there is a lot of noise along the branches, while this does not appear to
 be the case for the t-SNE result.
 
-From the first view, we modify the inputs to the t-SNE view, to try and produce a better trade-off between local structure and retain the topology of the data.
-We keep every parameter the same except that we initialise $Y$ with the first two PCs (scaled to have standard deviation 1e-4) instead of the default random initialisation and increase the perplexity from 30 to 100.
-We then combine these results
-with our tour view as displayed in the video:
+From the first view, we modify the inputs to the t-SNE view, to try and produce
+a better trade-off between local structure and retain the topology of the data.
+We keep every parameter the same except that we initialise $Y$ with the first
+two PCs (scaled to have standard deviation 1e-4) instead of the default random
+initialisation and increase the perplexity from 30 to 100. We then combine
+these results with our tour view as displayed in the video:
 
@@
 
-The video shows that this selection of parameters results
-in the the tips of the branch
+The video shows that this selection of parameters results in the the tips of
+the branches (the three black dots in figure \@ref(fig:fake-trees)) being split
+into three clusters representing the final branches of the tree. However, there
+are perceptual issues following the placement of the three groupings on the
+t-SNE view that become apparent via simple linked brushing. If we brush the
+tips of the yellow and brown branches (which appear to be close to each other
+on the t-SNE view), we immediately see the placement is distorted in the t-SNE
+view, and in the tour view these tips are at opposite ends of the tree.
+Although, this is a known issue of the t-SNE algorithm, we can easily identify
+it via simple interactivity.
 

### 2020-04-23 scripts/liminal-quickdraw.R
add quickdraw images

@@ -1,0 +1,51 @@
+## ----init--------------------------------------------------------------------
+library(liminal)
+# remotes::install_github("huizezhang-sherry/quickdraw")
+library(quickdraw)
+
+# downsample the number of images to have 500 images in each category
+slice_image_meta <- function(a, size = 500L) {
+  res <- qd_read(a)
+  nr <- nrow(res)
+  slice <- sample.int(nr, size)
+  slab <- res[slice, , drop = FALSE]
+  slab[["row_number"]] <- slice
+  slab
+}
+
+slice_bitmap <- function(a, row_number) {
+  res <- qd_read_bitmap(a)
+  colnames(res) <- paste0("px", seq_len(ncol(res)))
+  tibble::as_tibble(res[row_number, , drop = FALSE])
+} 
+
+## ----check-cache--------------------------------------------------------------
+has_run <- file.exists(here::here("data/qd.rds"))
+
+## ----download-categories-----------------------------------------------------
+if (!has_run) {
+  animals <- c("cat", "dog", "octopus", "spider", "whale")
+  # download and cache
+  for (a in animals) {
+    qd_download(a)
+    qd_download_bitmap(a)
+  }
+  # read and slice sample meta
+  set.seed(4099)
+  slabs <- lapply(animals, slice_image_meta)
+  row_nums <- lapply(slabs, function(slab) slab[["row_number"]])
+  # extract bitmaps as matrices
+  bits <- mapply(slice_bitmap, 
+                 a = animals, 
+                 row_number = row_nums, 
+                 SIMPLIFY = FALSE)
+  # bind as a single data frame
+  slabs <- dplyr::bind_rows(slabs)
+  bits <- dplyr::bind_rows(bits)
+  drawings <- dplyr::bind_cols(slabs, bits)
+  # save results
+  saveRDS(drawings, file = here::here("data/qd.rds"))
+}
+
+## ----load-cache--------------------------------------------------------------
+drawings <- readRDS(here::here("data/qd.rds"))

### 2020-04-23 paper.Rmd
add quickdraw data

@@
 
-### Case Study 3: QuickDraw images
+This will be Macosko data
+
+### Case Study 3: How do people draw animals?
+
+In the final case study, we explore a subset of the Google QuickDraw dataset [@quickdraw]. The data are sketches drawn by people of common concepts
+like 'airplane' or 'shoe', and corresponding metadata about the humans who
+created the sketches, such as country, the word category, whether the sketch
+was recognised by Google's neural network, and the number of strokes taken.
+We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the `quickdraw` package in R.
+We sampled 500 images from each of the following  animal categories: "dog", "cat", "octopus", "spider", and "whale" and unravelled the bitmaps into a 784 dimension image vector.
 

### 2020-04-23 scripts/liminal-quickdraw.R
add tsne run

@@ -49,3 +49,9 @@ if (!has_run) {
 
 ## ----load-cache--------------------------------------------------------------
 drawings <- readRDS(here::here("data/qd.rds"))
+
+set.seed(5099)
+tsne <- Rtsne::Rtsne(dplyr::select(drawings, dplyr::starts_with("px")))
+tsne_df <- data.frame(x = tsne$Y[,1], y = tsne$Y[,2], animal = drawings$word)
+
+limn_xy(tsne_df, x = x , y = y, color = animal)

### 2020-04-30 paper.Rmd
update details here
add some tour details

@@
 
-?? include details of tour here
+synthesis of some other techniques like LLE, Isomap and PHATE
 
@@
 
+Recall our $n \times p$ data matrix $X$
+
+A tour is sequence of $p \times d$ orthonormal projection matrices (bases) $A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$.
+
+A tour allows us to explore the set of all $d$-dimensional subspaces via the projection of $X$ onto $A$. 
+]
+
+---
+
+# How does it work?
+
+.large[
+We visualise the sequence of low dimensional projections:
+
+$$Y_{t} = XA_{t}$$
+
+as animated scatter or density plots. 
+]
+
+---
+
+.large[**Grand Tour:** generate a set of random orthnormal bases, interpolate between them.]
+
+```{r basis-view, out.width="50%", echo = FALSE}
+include_graphics("img/bases-generation.png")
+```
+
+(From _Buja, Cook, Asimov and Hurley, 2004_)
+---
+
+class: middle
+
+.large[The tour is enhanced by **interactivity**:
+
+- play/pause the current projection
+- zooming to find 'interesting' points
+- highlighting sets of points
+- linking multiple views
+
+?? include details of tour here
+
 - don't need to realise the entire sequence

### 2020-05-01 paper.Rmd
update words

@@
 A tour allows us to explore the set of all $d$-dimensional subspaces via the projection of $X$ onto $A$.
-]
 
----
-
-# How does it work?
-
-.large[
 We visualise the sequence of low dimensional projections:
@@
 as animated scatter or density plots.
-]
 
----
+**Grand Tour:** generate a set of random orthnormal bases, interpolate between them.
 
-.large[**Grand Tour:** generate a set of random orthnormal bases, interpolate between them.]
-
-```{r basis-view, out.width="50%", echo = FALSE}
-include_graphics("img/bases-generation.png")
-```
-
-(From _Buja, Cook, Asimov and Hurley, 2004_)
----
-
-class: middle
-
-.large[The tour is enhanced by **interactivity**:
+The tour is enhanced by **interactivity**:
 

### 2020-05-04 paper.Rmd
clarifying tour section

--- 2020-04-23 paper.Rmd
@@
 the tour. A tour is a sequence of projections of a high dimensional dataset
-onto a low-dimensional orthonormal basis matrix, that is represented as a
-dynamic visualization. The sequence of generated bases are interpolated to form
-the tour path, allowing a user to explore the subspace of projections. A grand
-tour corresponds to choosing new bases at random, and can give an overview of
-the structure in the data. Instead of picking projections at random, a guided
-tour can be used to generate a sequence 'interesting' projections as quantified
-by an index function. Given the dynamic nature of the tour, user interaction is
-important for controlling and exploring the visualisation: the tour has been
-used previously by @Wickham2015-cx as tool for exploring statistical model fits
-and by @Buja1996-fk for exploring factorial experimental designs.
+onto a low-dimensional basis matrix, and is represented as a
+animated visualization.Given the dynamic nature of the tour, user interaction is important for controlling and exploring the visualisation: the tour has been
+used previously by @Wickham2015-cx as tool for exploring statistical model fits and by @Buja1996-fk for exploring factorial experimental designs.
 
@@
 
+### Tours explore the subspace $d$-dimension projections
+
+Here we provide an overview of the mathematical aspects of the tour for
+readers who are unfamiliar with the technique before discussing the computational aspects of how `liminal` interfaces with them. The tour
+is emphatically not a dimension reduction technique but rather a visualisation technique that is grounded in mathematical theory, that is able to ascertain the shape and global structure of a dataset via inspection
+of the subspace generated by the set of low-dimensional projections.
+In fact, when we use the tour in the case studies, we will first sphere our data via principal components to reduce dimensionality of our input.
+
+Like, when using DR techniques, the tour assumes we have a real data matrix
+$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
+a  sequence of $p \times d$ orthonormal projection matrices (bases) $A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$.
+For each pair of orthormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
+path between them is interpolated to form intermediate frames, and giving the sense of continuous movement from one basis to another.
+The tour is then the continous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t]$ as the tour path is interpolated between successive bases.
+A *grand tour* corresponds to choosing new orthoromal bases at random; allowing a user to ascertain structure via exploring the subspace of $d$-dimensional projections.
+Instead of picking projections at random, a *guided tour* can be used to generate a sequence 'interesting' projections as quantified by an index function.
+While `liminal` is able to visualise guided tours, our focus is on the grand tour.
+
 ### Tours as a streaming data problem
+
+The process of generating successive bases and interpolating between them before frames, results in the tour being a dynamic visualisation technique.
+Generally, if $d$ is one or two dimensions, the results are visualised as
+animated density or scatter plots. 
+The tour is enhanced by **interactivity**:
+
+- play/pause the current projection
+- zooming to find 'interesting' points
+- highlighting sets of points
 
@@
 ### Linking views via brushes
+
+- linking multiple views
 

### 2020-05-05 paper.Rmd
add tour section

@@
 path between them is interpolated to form intermediate frames, and giving the sense of continuous movement from one basis to another.
-The tour is then the continous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t]$ as the tour path is interpolated between successive bases.
+The tour is then the continous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t}$ as the tour path is interpolated between successive bases.
 A *grand tour* corresponds to choosing new orthoromal bases at random; allowing a user to ascertain structure via exploring the subspace of $d$-dimensional projections.
 Instead of picking projections at random, a *guided tour* can be used to generate a sequence 'interesting' projections as quantified by an index function.
-While `liminal` is able to visualise guided tours, our focus is on the grand tour.
+While `liminal` is able to visualise guided tours, our focus in the case studies uses the grand tour to provide overviews of global
+structure in the data.
 
@@
 
-The process of generating successive bases and interpolating between them before frames, results in the tour being a dynamic visualisation technique.
-Generally, if $d$ is one or two dimensions, the results are visualised as
-animated density or scatter plots. 
-The tour is enhanced by **interactivity**:
+The process of generating successive bases and interpolating between them to  construct intermediary frames, means the tour is a dynamic visualisation technique.
+Generally, the user would set $d=2$ the tour is visualised as
+a animated scatterplot.
+This process of constructing bases and intermediate frames and visualising the resulting projection is akin to making a "flipbook" animation.
+Like with a flipbook, an interface to the tour requires the ability
+to interact and modify it in real time. The user interface generated in `liminal`
+allows a user to play and pause the tour animation (via brushing),
+panning and zooming to modify the scales of the plot to provide context and
+click events to highlight groups of points (like clusters).
 
-- play/pause the current projection
-- zooming to find 'interesting' points
-- highlighting sets of points
-
-- don't need to realise the entire sequence
-- instead generate new bases according to a fixed frame rate
-- allows user interactions to play/pause on the current view
+These interactions are enabled by treating the basis generation as a reactive
+stream. Instead of realising the entire sequence, which limits the animation
+to have a discrete number of frames, new bases and their intermediate frames
+are generated dynamically via pushing the current projection to the visualisation interface. The interface listens to events like pressing a button
+or mouse-dragging and reacts by pausing the stream. This process allows the
+user to manipulate the tour in real time rather than having to fix the number
+of bases ahead of time.
 

### 2020-05-07 scripts/liminal-quickdraw.R
updating quickdraw example, looks like linked tour example is broken :(

@@ -19,6 +19,13 @@ slice_bitmap <- function(a, row_number) {
   tibble::as_tibble(res[row_number, , drop = FALSE])
 } 
 
+tidy_tsne <- function(model, data) {
+  enframe <- as.data.frame(model[["Y"]])
+  colnames(enframe) <- c("x", "y", "z")[seq_len(ncol(enframe))]
+  dplyr::bind_cols(enframe, data)
+} 
+
+
 ## ----check-cache--------------------------------------------------------------
 has_run <- file.exists(here::here("data/qd.rds"))
 
@@ -50,8 +57,52 @@ if (!has_run) {
 ## ----load-cache--------------------------------------------------------------
 drawings <- readRDS(here::here("data/qd.rds"))
 
+# default t-SNE does not split the categories
 set.seed(5099)
 tsne <- Rtsne::Rtsne(dplyr::select(drawings, dplyr::starts_with("px")))
-tsne_df <- data.frame(x = tsne$Y[,1], y = tsne$Y[,2], animal = drawings$word)
+tsne_df <- tidy_tsne(tsne, list(animal = drawings$word))
 
+# cat-dog gradient is spread through the other animal categories
 limn_xy(tsne_df, x = x , y = y, color = animal)
+
+# try the PCA initialisation trick
+qd_pca <- prcomp(dplyr::select(drawings, dplyr::starts_with("px")),
+                 center = TRUE)
+
+# bind the PCs to the drawings
+drawings_pcs <- dplyr::bind_cols(drawings,dplyr::as_tibble(qd_pca$x))
+
+# PCA plot, blobby like the tSNE view
+limn_xy(drawings_pcs, x = PC1, y = PC2, color = word)
+
+qd_var_explained <- data.frame(
+  component = seq_len(length(qd_pca$sdev)),
+  sd = qd_pca$sdev,
+  var_explained = qd_pca$sdev^2 / sum(qd_pca$sdev^2),
+  cum_var_explained = cumsum(qd_pca$sdev^2) / sum(qd_pca$sdev^2)
+)
+
+# a lot of variability unexplained by PCA, need quite a few PCs
+limn_xy(qd_var_explained, x = component, y = cum_var_explained)
+
+
+# try tSNE again this time scaling initialising at first two PCs
+y_init <- clamp_sd(as.matrix(dplyr::select(drawings_pcs, PC1, PC2)), 
+                   sd = 1e-4)
+
[…]

### 2020-05-07 scripts/liminal-quickdraw.R
more playing around with the quickdraw example

@@ -1,4 +1,5 @@
 ## ----init--------------------------------------------------------------------
+library(dplyr)
 library(liminal)
 # remotes::install_github("huizezhang-sherry/quickdraw")
 library(quickdraw)
@@ -16,7 +17,7 @@ slice_image_meta <- function(a, size = 500L) {
 slice_bitmap <- function(a, row_number) {
   res <- qd_read_bitmap(a)
   colnames(res) <- paste0("px", seq_len(ncol(res)))
-  tibble::as_tibble(res[row_number, , drop = FALSE])
+  dplyr::as_tibble(res[row_number, , drop = FALSE])
 } 
 
 tidy_tsne <- function(model, data) {
@@ -57,20 +58,31 @@ if (!has_run) {
 ## ----load-cache--------------------------------------------------------------
 drawings <- readRDS(here::here("data/qd.rds"))
 
+
+# normalise the pixel intensities for each image by centering each
+# image by their average pixel intensity 
+px_mat <- as.matrix(select(drawings, starts_with("px")))
+px_means <- rowMeans(px_mat)
+px_mat2 <- sweep(px_mat, 1, px_means)
+colnames(px_mat2) <- paste0("norm_", colnames(px_mat2))
+drawings <- bind_cols(drawings, as_tibble(px_mat2))
+
 # default t-SNE does not split the categories
 set.seed(5099)
-tsne <- Rtsne::Rtsne(dplyr::select(drawings, dplyr::starts_with("px")))
+tsne <- Rtsne::Rtsne(select(drawings, starts_with("norm_px")),
+                     pca_center = FALSE, 
+                     normalize = FALSE)
 tsne_df <- tidy_tsne(tsne, list(animal = drawings$word))
 
 # cat-dog gradient is spread through the other animal categories
 limn_xy(tsne_df, x = x , y = y, color = animal)
 
 # try the PCA initialisation trick
-qd_pca <- prcomp(dplyr::select(drawings, dplyr::starts_with("px")),
-                 center = TRUE)
+qd_pca <- prcomp(select(drawings, starts_with("norm_px")),
+                 center = FALSE)
 
 # bind the PCs to the drawings
-drawings_pcs <- dplyr::bind_cols(drawings,dplyr::as_tibble(qd_pca$x))
+drawings_pcs <- bind_cols(drawings,as_tibble(qd_pca$x))
 
[…]

### 2020-05-07 scripts/liminal-quickdraw.R
trying stuff out

@@ -63,9 +63,20 @@ drawings <- readRDS(here::here("data/qd.rds"))
 # image by their average pixel intensity 
 px_mat <- as.matrix(select(drawings, starts_with("px")))
 px_means <- rowMeans(px_mat)
-px_mat2 <- sweep(px_mat, 1, px_means)
-colnames(px_mat2) <- paste0("norm_", colnames(px_mat2))
-drawings <- bind_cols(drawings, as_tibble(px_mat2))
+px_norm <- sweep(px_mat, 1, px_means)
+
+# Sphere the images by taking the SVD of covariance matrix
+# this removes covariance between pixels
+# px_cov <- tcrossprod(px_norm) / ncol(px_norm)
+# svals <- svd(px_cov)
+# px_rot <-
+#   diag(1 / (sqrt(svals$d) + 1e-5)) %*%
+#   crossprod(svals$u, px_norm)
+
+
+colnames(px_norm) <- paste0("norm_", colnames(px_norm))
+
+drawings <- bind_cols(drawings, as_tibble(px_norm))
 
 # default t-SNE does not split the categories
 set.seed(5099)

### 2020-05-07 scripts/liminal-quickdraw.R
playing around with stuff

@@ -88,6 +88,7 @@ tsne_df <- tidy_tsne(tsne, list(animal = drawings$word))
 # cat-dog gradient is spread through the other animal categories
 limn_xy(tsne_df, x = x , y = y, color = animal)
 
+
 # try the PCA initialisation trick
 qd_pca <- prcomp(select(drawings, starts_with("norm_px")),
                  center = FALSE)
@@ -117,8 +118,7 @@ y_init <- clamp_sd(as.matrix(select(drawings_pcs, PC1, PC2)),
                    sd = 1e-4)
 
 qd_tsne <- Rtsne::Rtsne(
-  select(drawings, starts_with("norm_px")),
-  normalize = FALSE,
+  select(drawings_pcs, starts_with("norm_px")),
   pca_center = FALSE,
   Y_init = y_init,
   perplexity = 5,
@@ -129,14 +129,10 @@ qd_tsne_df <- tidy_tsne(qd_tsne, list(animal = drawings$word))
 
 limn_xy(qd_tsne_df, x = x , y = y, color = animal)
 
-
-
-limn_tour_xylink(x = select(drawings_pcs, PC1:PC20, animal = word),
-                 y = qd_tsne_df,
-                 x_color = animal, 
-                 y_color = animal)
-
-
+# limn_tour_xylink(x = select(drawings_pcs, PC1:PC20, animal = word),
+#                  y = qd_tsne_df,
+#                  x_color = animal, 
+#                  y_color = animal)
 
 
 # # try a projection pursuit guided tour

### 2020-05-07 scripts/liminal-quickdraw.R
try a DR with a randomforest

@@ -113,6 +113,11 @@ limn_xy(qd_var_explained, x = component, y = cum_var_explained)
 # look at tour, looks like each category forms a face of cube?
 limn_tour(drawings_pcs, PC1:PC20, color = word)
 
+rf_data <- select(drawings, word, starts_with("norm_px")) %>% 
+  mutate(word = factor(word)) %>% 
+  as.data.frame()
+rf <- randomForest::randomForest(word ~ . , data = rf_data)
+
 # try tSNE again this time scaling initialising at first two PCs
 y_init <- clamp_sd(as.matrix(select(drawings_pcs, PC1, PC2)), 
                    sd = 1e-4)

### 2020-05-08 scripts/liminal-mouse.R
add macosko example

@@ -1,0 +1,46 @@
+## ----init-----------------------------------------------------------
+library(scRNAseq)
+library(scater)
+library(scran)
+sce <- MacoskoRetinaData()
+
+
+## ----qc-------------------------------------------------------------
+is.mito <- grepl("^MT-", rownames(sce))
+qcstats <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
+filtered <- quickPerCellQC(qcstats, percent_subsets="subsets_Mito_percent")
+sce <- sce[, !filtered$discard]
+
+## ----normalise-counts-----------------------------------------------
+sce <- logNormCounts(sce)
+
+## ----gene-selection-------------------------------------------------
+dec <- modelGeneVar(sce) # fits mean-variance trend using `nls`.
+hvg <- getTopHVGs(dec, prop=0.2) # selects ~20% of nrow(sce) that are HVGs
+
+## ----pca------------------------------------------------------------
+sce <- runPCA(sce, ncomponents = 50, subset_row = hvg)
+# looking at elbow-plot looks like we get the first 10PCs explain most of the 
+# variation
+library(ggplot2)
+ggplot(data.frame(x = 1:50,  y = attr(reducedDim(sce), "percentVar")), 
+       aes(x,y)) + 
+  geom_point() +
+  labs(x = "Component", y =  "% Var Explained") +
+  scale_x_continuous(breaks = seq(0, 50, by = 5))
+
+## ----cluster--------------------------------------------------------
+# the recommended workflow is to use graph based clustering on the principal 
+# components
+g <- buildSNNGraph(sce, use.dimred = 'PCA')
+sce$cluster_membership <- factor(igraph::cluster_louvain(g)$membership)
+
+# extract PCs to tour and perform tSNE on, keep cluster labels too
+pc_df <- reducedDim(sce, "PCA")
+pc_df <- dplyr::as_tibble(pc_df)
+pc_df$cluster_membership <- sce$cluster_membership
+
+dplyr::count(pc_df, cluster_membership)
+
+## ---- tsne
+

### 2020-05-08 scripts/helpers.R
refactor code

@@ -1,0 +1,42 @@
+# -- utility functions, mostly used for quickdraw example
+# downsample the number of images to have 500 images in each category
+slice_image_meta <- function(a, size = 500L) {
+  res <- qd_read(a)
+  nr <- nrow(res)
+  slice <- sample.int(nr, size)
+  slab <- res[slice, , drop = FALSE]
+  slab[["row_number"]] <- slice
+  slab
+}
+
+slice_bitmap <- function(a, row_number) {
+  res <- qd_read_bitmap(a)
+  colnames(res) <- paste0("px", seq_len(ncol(res)))
+  dplyr::as_tibble(res[row_number, , drop = FALSE])
+} 
+
+tidy_tsne <- function(model, data) {
+  enframe <- as.data.frame(model[["Y"]])
+  colnames(enframe) <- c("x", "y", "z")[seq_len(ncol(enframe))]
+  dplyr::bind_cols(enframe, data)
+}
+
+tidy_var_explained <- function(object) {
+  if (methods::is(object, "SingleCellExperiment")) {
+    pve <- attr(SingleCellExperiment::reducedDim(object), "percentVar")
+    component <- seq_len(length(pve))
+    c_pve <- cumsum(pve)
+  }
+  
+  if (methods::is(object, "prcomp")) {
+    component <- seq_len(length(object$sdev))
+    var <- object$sdev^2
+    pve <- (var / sum(var)) * 100
+    c_pve <- cumsum(var) / sum(var)            
+  }
+  data.frame(
+    component = component,
+    var_explained = pve,
+    cum_var_explained = c_pve
+  )
+}

### 2020-05-08 scripts/liminal-mouse.R

@@ -1,12 +1,20 @@
 ## ----init-----------------------------------------------------------
-library(scRNAseq)
 library(scater)
 library(scran)
-sce <- MacoskoRetinaData()
+library(liminal)
+source(here::here("scripts", "helpers.R"))
+
+sce <- scRNAseq::MacoskoRetinaData(ensembl = TRUE)
+# annotation ensembl database
+anno_db <- AnnotationHub::AnnotationHub()[["AH73905"]] 
+anno <- AnnotationDbi::select(anno_db, 
+               keys=rownames(sce), 
+               keytype="GENEID", columns=c("SYMBOL", "SEQNAME", "GENEBIOTYPE"))
+rowData(sce) <- anno[match(rownames(sce), anno$GENEID), ]
 
 
 ## ----qc-------------------------------------------------------------
-is.mito <- grepl("^MT-", rownames(sce))
+is.mito <- grepl("^mt-", rowData(sce)$SYMBOL)
 qcstats <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
 filtered <- quickPerCellQC(qcstats, percent_subsets="subsets_Mito_percent")
 sce <- sce[, !filtered$discard]
@@ -16,18 +24,14 @@ sce <- logNormCounts(sce)
 
 ## ----gene-selection-------------------------------------------------
 dec <- modelGeneVar(sce) # fits mean-variance trend using `nls`.
-hvg <- getTopHVGs(dec, prop=0.2) # selects ~20% of nrow(sce) that are HVGs
+hvg <- getTopHVGs(dec, prop=0.1) # selects ~10% of nrow(sce) that are HVGs
 
 ## ----pca------------------------------------------------------------
 sce <- runPCA(sce, ncomponents = 50, subset_row = hvg)
-# looking at elbow-plot looks like we get the first 10PCs explain most of the 
-# variation
-library(ggplot2)
-ggplot(data.frame(x = 1:50,  y = attr(reducedDim(sce), "percentVar")), 
-       aes(x,y)) + 
-  geom_point() +
-  labs(x = "Component", y =  "% Var Explained") +
-  scale_x_continuous(breaks = seq(0, 50, by = 5))
+# looking at elbow-plot looks like we get the first ten PCs explain most of the 
+# variation, with the elbow at five
+var_explained <- tidy_var_explained(sce)
+limn_xy(var_explained, x = component, y = var_explained)
 
[…]

### 2020-05-08 scripts/liminal-quickdraw.R

@@ -3,32 +3,10 @@ library(dplyr)
 library(liminal)
 # remotes::install_github("huizezhang-sherry/quickdraw")
 library(quickdraw)
-
-# downsample the number of images to have 500 images in each category
-slice_image_meta <- function(a, size = 500L) {
-  res <- qd_read(a)
-  nr <- nrow(res)
-  slice <- sample.int(nr, size)
-  slab <- res[slice, , drop = FALSE]
-  slab[["row_number"]] <- slice
-  slab
-}
-
-slice_bitmap <- function(a, row_number) {
-  res <- qd_read_bitmap(a)
-  colnames(res) <- paste0("px", seq_len(ncol(res)))
-  dplyr::as_tibble(res[row_number, , drop = FALSE])
-} 
-
-tidy_tsne <- function(model, data) {
-  enframe <- as.data.frame(model[["Y"]])
-  colnames(enframe) <- c("x", "y", "z")[seq_len(ncol(enframe))]
-  dplyr::bind_cols(enframe, data)
-} 
-
+source(here::here("scripts", "helpers.R"))
 
 ## ----check-cache--------------------------------------------------------------
-has_run <- file.exists(here::here("data/qd.rds"))
+has_run <- file.exists(here::here("data", "qd.rds"))
 
 ## ----download-categories-----------------------------------------------------
 if (!has_run) {
@@ -99,16 +77,10 @@ drawings_pcs <- bind_cols(drawings,as_tibble(qd_pca$x))
 # PCA plot, blobby like the tSNE view
 limn_xy(drawings_pcs, x = PC1, y = PC2, color = word)
 
-qd_var_explained <- data.frame(
-  component = seq_len(length(qd_pca$sdev)),
-  sd = qd_pca$sdev,
-  var_explained = qd_pca$sdev^2 / sum(qd_pca$sdev^2),
-  cum_var_explained = cumsum(qd_pca$sdev^2) / sum(qd_pca$sdev^2)
-)
+qd_var_explained <- tidy_var_explained(qd_pca)
 
-# a lot of variability unexplained by PCA, 
-# need quite a few PCs to get to fifty percent
-limn_xy(qd_var_explained, x = component, y = cum_var_explained)
+# a lot of variability left unexplained by PCA, 
+limn_xy(qd_var_explained, x = component, y = var_explained)
 
 # look at tour, looks like each category forms a face of cube?
 limn_tour(drawings_pcs, PC1:PC20, color = word)

### 2020-05-08 scripts/liminal-mouse.R
update macosko

@@ -49,7 +49,25 @@ dplyr::count(pc_df, cluster_membership)
 ## ---- tsne
 tsne <- Rtsne::Rtsne(dplyr::select(pc_df, -cluster_membership),
                      pca = FALSE,
-                     Y_init = clamp_sd(reducedDim(sce)[ , 1:2], sd = 1e-4)
+                     Y_init = clamp_sd(reducedDim(sce)[ , 1:2], sd = 1e-4),
                      perplexity = 30)
+
+tsne_df <- tidy_tsne(tsne, list(cluster_membership = pc_df$cluster_membership))
+
+
+library(dplyr)
+# down sample so each group is down 10%
+set.seed(2099)
+pc_df_sub <- pc_df %>% 
+  mutate(row_number = row_number()) %>% 
+  group_by(cluster_membership) %>% 
+  sample_frac(size = 0.1) %>%
+  ungroup() 
+
+tsne_df %>% 
+  slice(pc_df_sub$row_number) %>% 
+  limn_xy(x = x, y = y , color = forcats::fct_relevel(cluster_membership)
+
+limn_tour(pc_df_sub, PC1:PC5, color = cluster_membership)
 
 

### 2020-06-22 paper.Rmd
update background section; correct case study 3

@@
 
-## Dimensionality Reduction Overview
+## Dimensionality Reduction Background
 
@@
 methods with respect to the data analyst, in addition to their high level the
-mathematical details. To begin we suppose the data is in the form rectangular
+mathematical details. Here we restrict our attention to two recent methods that are commonly used in the literature: t-distributed stochastic neighbour embedding 
+(t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
+mention several other techniques.
+
+To begin we suppose the data is in the form rectangular
 matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
 the number of observations in $p$ dimensions.  The purpose of any DR algorithm
-is to find low-dimensional representation of the data, $Y = [\mathbf{y_1},
-\dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$ matrix where $d \ll p$.
-The hope of the analyst is that the DR procedure to produces $Y$ will remove
-noise in the original dataset while retaining latent structure.
+is to find a low-dimensional representation of the data, 
+$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
+matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
+produces $Y$ will remove noise in the original dataset while retaining any 
+latent structure.
 
-DR methods can be classified into two broad groups: linear and non-linear
+DR methods can be classified into two broad classes: linear and non-linear
 methods. Linear methods perform a linear transformation of the data, that is,
@@
 covariance matrix. The eigenvalues are sorted in decreasing order and represent
-the variance explained by each component (eigenvector). A common approach to
-deciding on the number of principal components to retain is to plot the
-proportion of variance explained by each component and choose a cut-off.
+the variance explained by each component (eigenvector). 
+A common approach to deciding on the number of principal components to retain is
+to plot the proportion of variance explained by each component and choose a 
+cut-off.
 
@@
 $X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
-Here we restrict our attention to two recent methods that are commonly used in
-the literature: t-distributed stochastic neighbour embedding (t-SNE) and
-uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
-the pair-wise similarity of (Euclidean) distances of points in a high
+Multidimensional scaling (MDS) is a class of DR methods that aims to construct an embedding 
+$Y$ such that the pariwise distances (inner products) in $Y$ approximate the pairwise distances (inner products) in  $X$. There are many variants of MDS, such as 
+nonmetric scaling which uses ranks instead of distances. A related extension
+of MDS is Isomap which uses the $k$-nearest neighbor graphs to estimate the pairwise geodesic distance of points in $X$ then uses classical MDS to construct $Y$. Yet
+another approach are non-linear DR methods based on diffusion 
+processes. A recent example in the single cell genomics literature is PHATE.
+Here an affinity matrix is estimated via the pairwise distance matrix and k-nearest neighbors graph of $X$. The algorithm denoises estimated distances in 
+high-dimensional space via transforming the affinity matrix into a Markov transition probaility matrix and diffusing this matrix over a fixed number of time steps.
+Then the diffused probabilities are transformed once more to construct a distance matrix, and classical MDS is used to generate $Y$.
+A difficulty with using non-linear
+DR methods for exploratory data analysis are the choices required in selecting
+appropriate parameters. To see the magnitude of these choices 
+we now closely examine the precise details of t-SNE and UMAP.
+
+The t-SNE algorithm estimates
+the pairwise similarity of (Euclidean) distances of points in a high
 dimensional space using a Gaussian distribution and then estimates a
@@
 
-To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
+To begin, t-SNE transforms pairwise distances between $\mathbf{x_i}$ and
 $\mathbf{x_j}$ to similarities using a Gaussian kernel:
@@
 
-Next, the target low-dimensional space, $Y$, pair-wise distances between
+Next, the target low-dimensional space, $Y$, pairwise distances between
 $\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
@@
 
-The resulting configuration $Y$ is the one that minimizes the Kullback-Leibler
+The resulting embedding $Y$ is the one that minimizes the Kullback-Leibler
 divergence between the probability distributions formed via similarities of
@@
 So when the the loss function is minimised this corresponds to large attractive
-forces, that is, the pair-wise distances in $Y$ are small when there are
+forces, that is, the pairwise distances in $Y$ are small when there are
 non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
 should also be small when the loss function is minimised, that is, when
-pair-wise distances in $Y$ are large regardless of the magnitude of the
+pairwise distances in $Y$ are large regardless of the magnitude of the
 corresponding distances in $X$.
@@
 supposedly preservers global structure better than t-SNE [@Coenen2019-to].
-
-synthesis of some other techniques like LLE, Isomap and PHATE
 
@@
 a  sequence of $p \times d$ orthonormal projection matrices (bases) $A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$.
-For each pair of orthormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
+For each pair of orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
 path between them is interpolated to form intermediate frames, and giving the sense of continuous movement from one basis to another.
-The tour is then the continous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t}$ as the tour path is interpolated between successive bases.
-A *grand tour* corresponds to choosing new orthoromal bases at random; allowing a user to ascertain structure via exploring the subspace of $d$-dimensional projections.
+The tour is then the continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t}$ as the tour path is interpolated between successive bases.
+A *grand tour* corresponds to choosing new orthonormal bases at random; allowing a user to ascertain structure via exploring the subspace of $d$-dimensional projections.
 Instead of picking projections at random, a *guided tour* can be used to generate a sequence 'interesting' projections as quantified by an index function.
@@
 was recognised by Google's neural network, and the number of strokes taken.
-We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the `quickdraw` package in R.
-We sampled 500 images from each of the following  animal categories: "dog", "cat", "octopus", "spider", and "whale" and unravelled the bitmaps into a 784 dimension image vector.
+We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the `quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into a 784 dimension image vector.
+We sampled 500 images from each of the following  categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple", and "whale".
 

### 2020-06-23 paper.Rmd
start drafting case study

@@
 The data lies 2-d tree like structure consisting of ten branches.
-This data is available in the `phateR` package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
+This data is available in the `phateR` package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added.
+[@Moon2019-ce].
 
@@
 
-This will be Macosko data
+A common analysis task in single cell studies is performing clustering to
+identify groupings  of cells with similar expression profiles. Analysts
+in this area generally use non linear DR methods for verification and identifaction
+of clusters and developmental trajectories (i.e., case study 1). For clustering,
+workflows the primary task is verify the existence of clusters and then begin to 
+identify the clusters as cell types using the expression of "known" 
+marker genes. The task becomes harder if the goal is discovery of new cell
+types
+
+Here we describe a liminal view of mouse single cell data from @Macosko2015-ot
+following the workflow described in @Amezquita2020-at. To begin we obtained
+the count matrix from the `scRNAseq
 

### 2020-06-23 paper.Rmd
rejig
tidy up case study no. 2 section

@@
 onto a low-dimensional basis matrix, and is represented as a
-animated visualization.Given the dynamic nature of the tour, user interaction is important for controlling and exploring the visualisation: the tour has been
-used previously by @Wickham2015-cx as tool for exploring statistical model fits and by @Buja1996-fk for exploring factorial experimental designs.
+animated visualization.Given the dynamic nature of the tour, user interaction is 
+important for controlling and exploring the visualisation: the tour has been
+used previously by @Wickham2015-cx as tool for exploring statistical model fits 
+and by @Buja1996-fk for exploring factorial experimental designs.
 
@@
 visualisation for colouring points in an embedding according to their distance
-in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
+in the original high-dimensional space [@Ovchinnikova2020-sy]. The work by
 @Pezzotti2017-cz provides a user guided and modified form of the t-SNE
@@
 methods with respect to the data analyst, in addition to their high level the
-mathematical details. Here we restrict our attention to two recent methods that are commonly used in the literature: t-distributed stochastic neighbour embedding 
+mathematical details. Here we restrict our attention to two recent methods that
+are commonly used in the literature: t-distributed stochastic neighbour embedding 
 (t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
@@
 to have a discrete number of frames, new bases and their intermediate frames
-are generated dynamically via pushing the current projection to the visualisation interface. The interface listens to events like pressing a button
+are generated dynamically via pushing the current projection to the 
+visualisation interface. The interface listens to events like pressing a button
 or mouse-dragging and reacts by pausing the stream. This process allows the
@@
 identify the clusters as cell types using the expression of "known" 
-marker genes. The task becomes harder if the goal is discovery of new cell
-types
+marker genes. Here a 'faithful' a embedding should ideally preserve the topology
+of the data; cells that correspond to a cell type should lie in the same
+neighbourhood in high-dimensional space. In this case study we use our 
+spatial and linked brushing approaches to look at neighbourhood preservation
+(which can be seen via the tour) and to look at the affect of applying different
+feature selection techniques and distance metrics.
 
-Here we describe a liminal view of mouse single cell data from @Macosko2015-ot
-following the workflow described in @Amezquita2020-at. To begin we obtained
-the count matrix from the `scRNAseq
+First, we downloaded the raw mouse retinal single cell RNA-seq data from
+@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
+have followed a standard workflow for pre-processing and normalizing this data 
+(described by @Amezquita2020-at): we performed QC using the the `scater`
+by removing cells with high proportion of mitocondrial gene expression and low
+numbers of genes detected, we log-transformed and normalised the expression 
+values and  finally selected highly variable genes (HVGs) using `scran` 
+[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
+the normalised expression matrix and compute PCA using the first 25 components.
+Using the PCs we built a shared nearest neighours graph (with k = 10) and 
+used Louvain clustering to generate clusters. To check and verify the 
+clustering we construct a liminal view. Because most off the variablity in
+expression is explained by the first 5 PCs, we tour those. The t-SNE view
+is computed from all PCs
+
+- compare to just using the HVGs them selves 
+- neighbourhood based brushing to check how diffuse or distorted clusters are
+- different distance metrics?
 
@@
 
-In the final case study, we explore a subset of the Google QuickDraw dataset [@quickdraw]. The data are sketches drawn by people of common concepts
+In the final case study, we explore a subset of the Google QuickDraw dataset 
+[@quickdraw]. The data are sketches drawn by people of common concepts
 like 'airplane' or 'shoe', and corresponding metadata about the humans who
@@
 was recognised by Google's neural network, and the number of strokes taken.
-We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the `quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into a 784 dimension image vector.
-We sampled 500 images from each of the following  categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple", and "whale".
+We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the 
+`quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into 
+a 784 dimension image vector. We sampled 500 images from each of the following
+categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple",
+and "whale".
 

### 2020-06-23 paper.Rmd
notes

@@
 the normalised expression matrix and compute PCA using the first 25 components.
-Using the PCs we built a shared nearest neighours graph (with k = 10) and 
+Using the PCs we built a shared nearest neighours graph (with $k = 10$) and 
 used Louvain clustering to generate clusters. To check and verify the 
@@
 
-- compare to just using the HVGs them selves 
+- compare to just using the HVGs themselves with the tour
 - neighbourhood based brushing to check how diffuse or distorted clusters are
@@
 
+- limitations: large $n$, latency issues
+- future work: regularisation tour, slicing and sectioning
+
 ## Acknowledgements

### 2020-06-24 paper.Rmd
minor edits

@@
 sequencing experiment, or identifying a gradient or trajectory along the
-embedded form @Nguyen2019-yh. The second relates to using embeddings as part of
+embedded form [@Nguyen2019-yh]. The second relates to using embeddings as part of
 a clustering workflow. Here analysts are interested in identifying and naming
@@
 variables that are a-priori known to distinguish clusters. Both of these
-workflow steps rely on the embedding being 'faithful' or the original high
-dimensional dataset, and become much more difficult when there is no underlying
+workflow steps rely on the embedding being representative the original high
+dimensional dataset, and becomes much more difficult when there is no underlying
 ground truth.
@@
 2D scatter plots were mostly sufficient for detecting class separation, however
-they noted that often multiple embeddings were required. For the task of
+they also noted that often multiple embeddings were required. For the task of
 cluster identification, @Lewis2012-ai showed experimentally that novice users
@@
 important for controlling and exploring the visualisation: the tour has been
-used previously by @Wickham2015-cx as tool for exploring statistical model fits 
-and by @Buja1996-fk for exploring factorial experimental designs.
+used previously by @Wickham2015-cx and @Li2020-rh as a tool for 
+exploring statistical model fits and by @Buja1996-fk for exploring the space of 
+factorial experimental designs.
 
@@
 techniques: the Sleepwalk interface provides a click and highlight
-visualisation for colouring points in an embedding according to their distance
-in the original high-dimensional space [@Ovchinnikova2020-sy]. The work by
-@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
-algorithm, that allows users to modified optimisation parameters in real-time.
-Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
-PCA live in the browser and provides interactions to colour points, and
-highlights nearest neighbours [@Smilkov2016-hp].
+visualisation for colouring points in an embedding according to an estimated
+pairwise distance in the original high-dimensional space [@Ovchinnikova2020-sy].
+The idea behind this is that non-linear embedding methods should preserve 
+locality and that points that are 'close' to each other on the embedded view
+should be close according to the estimated pairwise distance. 
+Similarly, the TensorFlow embedding projector is a web interface to 
+running some non-liner embedding methods live in the browser and 
+provides interactions  to colour points, and select nearest neighbours
+[@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
+guided and modified form of the  t-SNE algorithm, that allows users to modified 
+optimisation parameters in real-time.
 
 There is no one-size fits all: finding an appropriate embedding for a given
-dataset is a difficult and somewhat poorly defined problem. For non-linear
+dataset is a difficult and a somewhat poorly defined problem. For non-linear
 methods, there are a lot of parameters to explore that can have an effect on
@@
 $X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
-Multidimensional scaling (MDS) is a class of DR methods that aims to construct an embedding 
-$Y$ such that the pariwise distances (inner products) in $Y$ approximate the pairwise distances (inner products) in  $X$. There are many variants of MDS, such as 
-nonmetric scaling which uses ranks instead of distances. A related extension
-of MDS is Isomap which uses the $k$-nearest neighbor graphs to estimate the pairwise geodesic distance of points in $X$ then uses classical MDS to construct $Y$. Yet
-another approach are non-linear DR methods based on diffusion 
-processes. A recent example in the single cell genomics literature is PHATE.
-Here an affinity matrix is estimated via the pairwise distance matrix and k-nearest neighbors graph of $X$. The algorithm denoises estimated distances in 
-high-dimensional space via transforming the affinity matrix into a Markov transition probaility matrix and diffusing this matrix over a fixed number of time steps.
-Then the diffused probabilities are transformed once more to construct a distance matrix, and classical MDS is used to generate $Y$.
-A difficulty with using non-linear
-DR methods for exploratory data analysis are the choices required in selecting
-appropriate parameters. To see the magnitude of these choices 
-we now closely examine the precise details of t-SNE and UMAP.
+Multidimensional scaling (MDS) is a class of DR methods that aims to construct
+an embedding  $Y$ such that the pariwise distances (inner products) in $Y$ 
+approximate the pairwise distances (inner products) in $X$ 
+[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
+variants of MDS, such as nonmetric scaling which amounts to replacing
+distances with ranks instead [@Kruskal1964-cw]. 
+A related technique is Isomap which uses a $k$-nearest neighbor graph
+to estimate the pairwise geodesic distance of points in $X$ then uses classical 
+MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
+processes such asdiffusion maps [@Coifman2005-ak]. A recent example of this 
+approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
+estimated via the pairwise distance matrix and k-nearest neighbors graph of $X$. 
+The algorithm denoises estimated distances in high-dimensional space via 
+transforming the affinity matrix into a Markov transition probability matrix 
+and diffusing this matrix over a fixed number of time steps. Then the diffused 
+probabilities are  transformed once more to construct a distance matrix, 
+and classical MDS is  used to generate $Y$. A general difficulty with using 
+non-linear DR methods for exploratory data analysis is selecting and tuning 
+appropriate parameters. To see the extent of these choices 
+we now examine the underpinnings of t-SNE and UMAP.
 
@@
 Another recent method UMAP has seen a large rise in popularity (at least in
-single cell transcriptomics) [@McInnes2018-co]. is a method that is related to
+single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
 LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
@@
 It appears to suffer from the same perceptual issues as t-SNE, however it
-supposedly preservers global structure better than t-SNE [@Coenen2019-to].
+supposedly preserves global structure better than t-SNE [@Coenen2019-to].
 
@@
 methods like t-SNE. We have made extensive use of ensemble graphics, that is
-aligning related plots alongside each other to provide context. As we will see
-in the case studies, this allows analysts to quickly compare views from
-embedding methods and allows them to see how the embedding method alters the
-global structure of their data. Using ensembles allows the use of interaction
-techniques, that allow analysts to perform cluster orientation tasks via
-linking multiple views. This approach allows our interface, to achieve the
-three principles for interactive high-dimensional data visualisation outlined
+aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
+As we will see in the case studies, this allows analysts to quickly compare 
+views fromembedding methods and allows them to see how the embedding method 
+alters theglobal structure of their data. Using ensembles allows the use of 
+interaction techniques, that allow analysts to perform cluster orientation 
+tasks vialinking multiple views. This approach allows our interface, to achieve 
+the three principles for interactive high-dimensional data visualisation outlined
 by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.
@@
 Here we provide an overview of the mathematical aspects of the tour for
-readers who are unfamiliar with the technique before discussing the computational aspects of how `liminal` interfaces with them. The tour
-is emphatically not a dimension reduction technique but rather a visualisation technique that is grounded in mathematical theory, that is able to ascertain the shape and global structure of a dataset via inspection
-of the subspace generated by the set of low-dimensional projections.
-In fact, when we use the tour in the case studies, we will first sphere our data via principal components to reduce dimensionality of our input.
+readers who are unfamiliar with the technique before discussing the 
+computational aspects of how `liminal` interfaces with them. The tour
+is emphatically not a dimension reduction technique but rather a visualisation 
+technique that is grounded in mathematical theory, that is able to ascertain 
+the shape and global structure of a dataset via inspection of the subspace 
+generated by the set of low-dimensional projections. In fact, in some
+of the case studies below, we  first sphere our data via principal 
+components to reduce dimensionality of our input.
 
@@
 $X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
-a  sequence of $p \times d$ orthonormal projection matrices (bases) $A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$.
-For each pair of orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
-path between them is interpolated to form intermediate frames, and giving the sense of continuous movement from one basis to another.
-The tour is then the continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t}$ as the tour path is interpolated between successive bases.
-A *grand tour* corresponds to choosing new orthonormal bases at random; allowing a user to ascertain structure via exploring the subspace of $d$-dimensional projections.
-Instead of picking projections at random, a *guided tour* can be used to generate a sequence 'interesting' projections as quantified by an index function.
-While `liminal` is able to visualise guided tours, our focus in the case studies uses the grand tour to provide overviews of global
-structure in the data.
+a  sequence of $p \times d$ orthonormal projection matrices (bases) 
+$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
+orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
+path between them is interpolated to form intermediate frames, and giving the 
+sense of continuous movement from one basis to another. The tour is then the 
+continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the 
+projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
+successive bases. A *grand tour* corresponds to choosing new orthonormal
+bases at random; allowing a user to ascertain structure via exploring the
+subspace of $d$-dimensional projections. Instead of picking projections at 
+random, a *guided tour* can be used to generate a sequence 'interesting' 
+projections as quantified by an index function. While `liminal` is able to 
+visualise guided tours, our focus in the case studies uses the grand tour to
+see global structure in the data.
 
@@
 
-The process of generating successive bases and interpolating between them to  construct intermediary frames, means the tour is a dynamic visualisation technique.
-Generally, the user would set $d=2$ the tour is visualised as
-a animated scatterplot.
-This process of constructing bases and intermediate frames and visualising the resulting projection is akin to making a "flipbook" animation.
-Like with a flipbook, an interface to the tour requires the ability
+The process of generating successive bases and interpolating between them to  
+construct intermediary frames, means the tour is a dynamic visualisation 
+technique. Generally, the user would set $d=2$ the tour is visualised as
+a animated scatterplot. This process of constructing bases and intermediate 
+frames and visualising the resulting projection is akin to making a "flipbook" 
+animation. Like with a flipbook, an interface to the tour requires the ability
 to interact and modify it in real time. The user interface generated in `liminal`

### 2020-07-09 paper.Rmd
reknit

@@
 Multidimensional scaling (MDS) is a class of DR methods that aims to construct
-an embedding  $Y$ such that the pariwise distances (inner products) in $Y$ 
+an embedding  $Y$ such that the pairwise distances (inner products) in $Y$ 
 approximate the pairwise distances (inner products) in $X$ 
 [@Torgerson1952-am; @Kruskal1964-cz]. There are many 
-variants of MDS, such as nonmetric scaling which amounts to replacing
+variants of MDS, such as non-metric scaling which amounts to replacing
 distances with ranks instead [@Kruskal1964-cw]. 
-A related technique is Isomap which uses a $k$-nearest neighbor graph
+A related technique is Isomap which uses a $k$-nearest neighbour graph
 to estimate the pairwise geodesic distance of points in $X$ then uses classical 
@@
 approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
-estimated via the pairwise distance matrix and k-nearest neighbors graph of $X$. 
-The algorithm denoises estimated distances in high-dimensional space via 
+estimated via the pairwise distance matrix and k-nearest neighbours graph of $X$. 
+The algorithm de-noises estimated distances in high-dimensional space via 
 transforming the affinity matrix into a Markov transition probability matrix 
@@
 As we will see in the case studies, this allows analysts to quickly compare 
-views fromembedding methods and allows them to see how the embedding method 
-alters theglobal structure of their data. Using ensembles allows the use of 
+views from embedding methods and allows them to see how the embedding method 
+alters the global structure of their data. Using ensembles allows the use of 
 interaction techniques, that allow analysts to perform cluster orientation 
-tasks vialinking multiple views. This approach allows our interface, to achieve 
+tasks via linking multiple views. This approach allows our interface, to achieve 
 the three principles for interactive high-dimensional data visualisation outlined
@@
 identify groupings  of cells with similar expression profiles. Analysts
-in this area generally use non linear DR methods for verification and identifaction
+in this area generally use non linear DR methods for verification and identification
 of clusters and developmental trajectories (i.e., case study 1). For clustering,
@@
 (described by @Amezquita2020-at): we performed QC using the the `scater`
-by removing cells with high proportion of mitocondrial gene expression and low
+by removing cells with high proportion of mitochondrial gene expression and low
 numbers of genes detected, we log-transformed and normalised the expression 
@@
 the normalised expression matrix and compute PCA using the first 25 components.
-Using the PCs we built a shared nearest neighours graph (with $k = 10$) and 
+Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
 used Louvain clustering to generate clusters. To check and verify the 
-clustering we construct a liminal view. Because most off the variablity in
+clustering we construct a liminal view. Because most off the variability in
 expression is explained by the first 5 PCs, we tour those. The t-SNE view

### 2020-07-09 paper.Rmd
update bibliography and add info on neighbors

@@
 
-### Tours explore the subspace $d$-dimension projections
+### Tours explore the subspace of $d$-dimensional projections
 
@@
 to interact and modify it in real time. The user interface generated in `liminal`
-allows a user to play and pause the tour animation (via brushing),
+allows a user to play, pause, and reset the tour animation,
 panning and zooming to modify the scales of the plot to provide context and
-click events to highlight groups of points (like clusters).
+click events to highlight groups of points if a labelling variable has
+been placed on the legend.
 
@@
 user to manipulate the tour in real time rather than having to fix the number
-of bases ahead of time.
+of bases ahead of time. Additionally, once the user has identified an 
+interesting projection or is done with the tour, the interface will return the 
+current basis for use downstream.
 
-### Linking views via brushes
+### Linking and highlighting views via interactions
 
-- linking multiple views
+The embedding and tour views are linked together via rectangular brushes;
+when a brush is active, points will be highlighted in the adjacent view. 
+Because the tour is dynamic, brush events that become active will pause the
+animation, so that a user can interrogate the current view. By default,
+brushing on the embedding view will produce a one-to-one linking with the tour
+view (the special case of $k=1$ nearest neighbors brush). It is possible
+to shift to one-to-many linking by increasing $k$ via the 'controls'
+panel of the interface.  If desired, the distance
+metric used to estimate nearest neighbours can by chosen by the user. 
+The user may provide a reference data set for computing
+nearest neighbours (for example a distance matrix, or the complete data matrix)
+independently of the tour or embedding views. The estimation of neighbours
+is done with the `FNN` package for datasets with less than 4096 observations
+and `RcppAnnoy` for larger datasets or if a user selects a non-euclidean
+distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
+with the interface, the brush regions are returned for downstream analysis.
 
@@
 The next section steps through three use cases of our approach using simulated
-and real data. In the first case study, we step through data that lies on tree
-structure (similar to trajectories in single cell transcriptomics) to that
+and real data. In the first case study, we step through three simulated datasets
+to show the `liminal` interface. The first simulation contains 3 clusters
+sampled from a 5-d gaussian distribution that are well embedded by t-SNE.
+The second is another example of spherical clusters but this time, they are
+tightly clustered and are distorted by t-SNE, which is immediately apparent in 
+the tour view. In the final simulation we
+look at a tree structure (similar to trajectories in single cell 
+transcriptomics) that
 highlights an iterative approach to modifying t-SNE views based on what we
@@
 brushing, we can see when the assumptions of continuity of the embedding map
-are broken. In the second case study, we explore a well-clustered single cell
-transcriptomics data of mice retina... Finally, in the last case study we
+are broken.
+
+In the second case study, we explore a well-clustered single cell
+transcriptomics data of mice retina...
+
+Finally, in the last case study we
 explore a subset of the QuickDraw dataset, we use `liminal` to identify
@@
 In the final case study, we explore a subset of the Google QuickDraw dataset 
-[@quickdraw]. The data are sketches drawn by people of common concepts
+[@quickdraw-paper]. The data are sketches drawn by people of common concepts
 like 'airplane' or 'shoe', and corresponding metadata about the humans who

### 2020-07-17 scripts/helpers.R
record examples

@@ -17,7 +17,7 @@ slice_bitmap <- function(a, row_number) {
 
 tidy_tsne <- function(model, data) {
   enframe <- as.data.frame(model[["Y"]])
-  colnames(enframe) <- c("x", "y", "z")[seq_len(ncol(enframe))]
+  colnames(enframe) <- c("tsneX", "tsneY", "tsneZ")[seq_len(ncol(enframe))]
   dplyr::bind_cols(enframe, data)
 }
 

### 2020-07-17 scripts/liminal-gaussian.R

@@ -1,0 +1,65 @@
+# adapted from Holmes and Ngyuen, 2019
+# Simple Gaussian 10-d
+source(here::here("scripts", "helpers.R"))
+library(liminal)
+library(ggplot2)
+
+set.seed(63968)
+m <- 500
+d <- 5
+p <- 10
+k <- 5
+sigma <- 1
+muMax <- 25
+mu <- matrix(runif(k*d, max = muMax), ncol = d)
+X <- lapply(seq_len(k), function(i) {
+  ix <- rep(1, m) %*% t(mu[i, ]) + matrix(sigma*rnorm(m*d), ncol = d)
+  as.data.frame(ix)
+})
+X_tbl <- dplyr::bind_rows(X, .id = "cluster")
+X <- as.matrix(X_tbl[,-1])
+Y <- X  %*% matrix(rnorm(d*p), ncol = p)
+colnames(Y) <- sprintf("dim%02d", 1:p)
+tour_tbl <- dplyr::bind_cols(as.data.frame(Y), 
+                             cluster = factor(X_tbl[["cluster"]]))
+
+
+tsne_y <- Rtsne::Rtsne(Y)
+embed_tbl <- tidy_tsne(tsne_y, 
+                       list(cluster = factor(X_tbl[["cluster"]])))
+colnames(embed_tbl) <- c("tsneX", "tsneY", "cluster")
+
+# run basic interface once
+limn_tour_link(
+  tour_data = tour_tbl,
+  embed_data = embed_tbl,
+  tour_color = cluster,
+  embed_color = cluster
+)
+
+
+# multichallenge data set
+library(dplyr)
+library(readr)
+url <- "http://ifs.tuwien.ac.at/dm/download/multiChallenge-matrix.txt"
+multi <- read_tsv(url, col_names = FALSE) %>% 
+  mutate(
+    index = rep(seq_len(200), 5),
+    key = rep(1:5, each = 200L)
+  ) %>%
+  filter(key %in% c(1, 4)) %>% 
+  mutate(group = rep(c("hierachical", "flat"), each = 200L)) %>% 
+  select(group, X1:X10)
+
+
+set.seed(143280)
+tsne_multi <- Rtsne::Rtsne(multi[, -1])
+embed_multi <- tidy_tsne(tsne_multi, list(group = multi[["group"]]))
+colnames(embed_multi) <- c("tsneX", "tsneY", "group")
+
+limn_tour_link(
+  tour_data = multi,
[…]

### 2020-07-17 scripts/liminal-pbmc3k.R

@@ -1,96 +1,0 @@
-## ---- include = FALSE------------------------------
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>"
-)
-
-
-## ----setup-----------------------------------------
-library(TENxPBMCData)
-library(scater)
-library(scran)
-
-pbmc3k <- TENxPBMCData("pbmc3k")
-pbmc3k
-
-
-## ----qc--------------------------------------------
-is.mito <- grep("MT", rowData(pbmc3k)$Symbol_TENx)
-stats <- perCellQCMetrics(pbmc3k, subsets=list(Mito=is.mito))
-high.mito <- isOutlier(stats$subsets_Mito_percent, nmads=3, type="higher")
-pbmc3k <- pbmc3k[,!high.mito]
-
-
-## --------------------------------------------------
-pbmc3k <- logNormCounts(pbmc3k)
-
-
-## --------------------------------------------------
-dec3k <- modelGeneVar(pbmc3k)
-chosen.hvgs <- which(dec3k$bio > 0)
-
-
-## --------------------------------------------------
-set.seed(2099)
-pbmc3k <- runPCA(pbmc3k, 
-                 ncomponents = 15,
-                 subset_row = chosen.hvgs,
-                 BSPARAM = BiocSingular::RandomParam(),
-                 name = "pca")
-
-set.seed(1999)
-pbmc3k <- runTSNE(pbmc3k,
-                  subset_row = chosen.hvgs,
-                  perplexity = 30,
-                  name = "tsne")
-
-
-## --------------------------------------------------
-g <- buildSNNGraph(pbmc3k, k=15, use.dimred = 'pca')
-clust <- igraph::cluster_louvain(g)
-pbmc3k$cluster <- factor(igraph::membership(clust))
-
-
-
-## ----flatten---------------------------------------
-tsne_tbl <- data.frame(
-  tsne_x = reducedDim(pbmc3k, "tsne")[,1],
-  tsne_y = reducedDim(pbmc3k, "tsne")[,2],
-  cluster = pbmc3k$cluster
-)
-
-express <- t(logcounts(pbmc3k)[chosen.hvgs,])
-colnames(express) <- rowData(pbmc3k)[chosen.hvgs, "Symbol_TENx"]
-express <- as.data.frame(express)
-tsne_tbl <- dplyr::bind_cols(
-  tsne_tbl,
-  express
-)
-
-
-## --------------------------------------------------
[…]

### 2020-07-17 paper.Rmd
update examples

@@
 The next section steps through three use cases of our approach using simulated
-and real data. In the first case study, we step through three simulated datasets
-to show the `liminal` interface. The first simulation contains 3 clusters
-sampled from a 5-d gaussian distribution that are well embedded by t-SNE.
-The second is another example of spherical clusters but this time, they are
-tightly clustered and are distorted by t-SNE, which is immediately apparent in 
-the tour view. In the final simulation we
-look at a tree structure (similar to trajectories in single cell 
-transcriptomics) that
-highlights an iterative approach to modifying t-SNE views based on what we
-learn in the high-dimensional space. We show how through simple linked
-brushing, we can see when the assumptions of continuity of the embedding map
-are broken.
+and real data.  The videos presented show the visual 
+appearance of the `liminal` interface, and how we can interact with the tour
+via the controls previously described.
+In the first case study, we look at three different examples
+where t-SNE is applied to three simulations where cluster structure and geometry
+is known with varying success.
 
@@
 
-### Case Study 1: Exploring tree structured data with tours and t-SNE
+### Case Study 1: Exploring simulation data with tours and t-SNE
 
-In this case study we explore some simulated noisy tree structured data (figure
+To begin we look at simulated datasets that reproduce known facts
+aboutthe  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
+Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
+covariance matrix. We then computed a t-SNE layout using default settings,
+and set up the `liminal` linked interface with grand tour on the 10-$d$
+observations (video - ).
+
+<embedded video liminal gaussian>
+
+We see in the video that t-SNE has correctly split out each cluster and
+laid them out in a star like formation. This is agrees with the tour view,
+where once we start the animation, the five clusters begin to appear but
+generally points are more concentrated in the projection view compared to 
+the tSNE layout. This can be seen from our brushing action on the tSNE side.
+
+<embedded video liminal hierachcial >
+
+We repeat the process of simulating Gaussian clusters in video <>, but this time
+we have a cluster that has more granular structure. This time we have two
+Gaussian clusters with equal covariance embedded in $10-d$, and a third cluster
+with hierachcial structure. This cluster has two 3-$d$ clusters embedded in 
+10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
+Again tSNE has correctly identified the sub-clusters but their relative 
+locations to each other is distorted, with the orange and blue groups being
+far from each other in the tour view. We see in this case that is difficult
+to see the sub-clusters in the tour view, however, once we zoom and highlight
+they become more apparent. When we brush the sub-clusters in the tSNE, their
+relative placement is again exaggerated, with the tour showing that they
+are indeed much closer than the impression the tSNE view gives.
+
+Next explore some simulated noisy tree structured data (figure
 \@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
@@
 explained in the data) to use with the grand tour.
-
-<!-- can we show R code? -->
 
@@
 expression is explained by the first 5 PCs, we tour those. The t-SNE view
-is computed from all PCs
+is computed from all PCs.
 

### 2020-07-17 paper.Rmd
update text

@@
 Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
-covariance matrix. We then computed a t-SNE layout using default settings,
+covariance matrix. We then computed a t-SNE layout with default settings 
+using the `Rtsne` package [@Rtsne],
 and set up the `liminal` linked interface with grand tour on the 10-$d$
@@
 keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
-initialisation) using the `Rtsne` package [@Rtsne]. We then create a linked
+initialisation). We then create a linked
 tour with t-SNE layout with `liminal` as shown in the following video :

### 2020-07-18 paper.Rmd
adjust figures and section

@@
 ```{r fake-trees, echo = FALSE, out.width = "50%", fig.align = "center", fig.cap="(ref:fake-trees)"}
-knitr::include_graphics(here::here("img", "tree.png"))
+knitr::include_graphics(here::here("img", "tree-view.png"))
 ```
@@
 The video shows that this selection of parameters results in the the tips of
-the branches (the three black dots in figure \@ref(fig:fake-trees)) being split
-into three clusters representing the final branches of the tree. However, there
+the branches (the three black dots in figure \@ref(fig:fake-trees) a) being split
+into three clusters representing the terminal branches of the tree. However, there
 are perceptual issues following the placement of the three groupings on the
@@
 Although, this is a known issue of the t-SNE algorithm, we can easily identify
-it via simple interactivity.
+it via simple interactivity without knowing the inner workings of the method.
 
@@
 neighbourhood in high-dimensional space. In this case study we use our 
-spatial and linked brushing approaches to look at neighbourhood preservation
-(which can be seen via the tour) and to look at the affect of applying different
-feature selection techniques and distance metrics.
+linked brushing approaches to look at neighbourhood preservation
+(which can be seen via the tour) and to look at marker genes through
+the lens of the tour.
 

### 2020-07-18 scripts/liminal-trees.R

@@ -1,6 +1,13 @@
 ## ----init-----------------------------------------------------------
 library(liminal)
 dim(fake_trees)
+library(cowplot)
+library(ggplot2)
+
+tree_img <- ggdraw() + 
+  draw_image(here::here("img", "tree.png"), scale = 0.8) +
+  draw_figure_label("Tree Structure", position = "bottom.right") +
+  theme_void()
 
 
 ## ----pc-view--------------------------------------------------------
@@ -10,8 +17,13 @@ summary(pcs)$importance[3,1:12]
 
 ## ----pca-xy---------------------------------------------------------
 fake_trees <- dplyr::bind_cols(fake_trees, as.data.frame(pcs$x))
-limn_xy(fake_trees, x = PC1, y = PC2, color = branches)
 
+pca_xy <- ggplot(fake_trees, aes(x = PC1, y = PC2, color = branches)) +
+  geom_point() +
+  scale_color_manual(values = limn_pal_tableau10()) +
+  labs(caption = "PCA") +
+  theme(aspect.ratio = 1) +
+  theme_void() 
 
 ## ----tsne-xy--------------------------------------------------------
 set.seed(2099)
@@ -21,14 +33,22 @@ tsne_df <- data.frame(tsneX = tsne$Y[,1],
                       tsneY = tsne$Y[,2],
                       branches = fake_trees$branches)
 
-limn_xy(tsne_df, x = tsneX, y = tsneY, color = branches)
+tsne_xy  <- ggplot(tsne_df, aes(x = tsneX, y = tsneY, color = branches)) +
+  geom_point() +
+  scale_color_manual(values = limn_pal_tableau10()) + 
+  guides(color = FALSE) +
+  labs(caption = "tSNE") +
+  theme(aspect.ratio = 1) +
+  theme_void() 
+
+## ----tour----------------------------------------------------------
 
 # now together
-limn_tour_xylink(
-  dplyr::select(fake_trees, PC1:PC12, branches),
-  tsne_df,
-  x_color = branches,
-  y_color = branches
+limn_tour_link(
+  tour_data = dplyr::select(fake_trees, PC1:PC12, branches),
+  embed_data = tsne_df,
+  embed_color = branches,
+  tour_color = branches
 )
 
 
[…]

### 2020-07-18 paper.Rmd
add citations; rejig section to note things aren't implemented yet

@@
 high-dimensional space and the embedding space. As distances can be noisy in
-high-dimensions, ranks can be used instead as has been proposed by ...
+high-dimensions, ranks can be used instead as has been proposed by @Lee2009-zb.
 Identifying distortions and diffusions allows an analyst to investigate the
-quality of their embedding and revise them iteratively. We propose this can be
-done visually using our side-by-side tour and embedding views. Some quality
-checks, such as finding distortions can be identified using simple linked
-brushing, however we can also interrogate them via spatial brushes and brush
-composition.
+quality of their embedding and revise them iteratively.
 
- Look up nearest neighbours graph from points that lie in a brushing region.
-
-Highlight the corresponding neighbours using colour or transparency in the
-linked view.
-
-- The $k$ nearest neighbours graph can be pre-computed quickly, for either $X$, $Y$ or both.
-- Instead of using the neighbour indices, we could use the neighbour distances instead.
-- Composition of multiple brushes could be used to show where there are matches/mismatches between nearest neighbour graphs.
-
-Using a linked neighbourhood brush, we can visually investigate the nearest
-neighbour relationships in the high-dimensional space via brushing in the
-embedding view. The user can select the number of nearest neighbours directly,
-and modify the distance metric used for determining the neighbours. This allows
-users to interrogate the stability of clusters generated in the embedding view.
-Multiple brushes can be used to pose queries in either the tour view or the
-embedding view; interface controls allow these brushes to combine using logical
-operators such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
-simple colour highlights, allowing analysts to get a more holistic view of the
-effect of an embedding algorithm.
+We propose this can be
+done visually using our side-by-side tour and embedding views.
+In the simplest case, a local continuity check can be assessed via one to
+one linked brushing from the embedding to the tour view. This can be generalised further with one to many brushing, via spatial brushing. This form
+of brushing acts directly on $k$-nearest neighbors ($l$-nn) graph computed from a
+reference dataset, when a user brushes over a region in the embedding,
+all the points that match the graphs edges are selected on the 
+corresponding tour view. In place of highlighting, one could use opacity
+or binned colour scales to encode distances or ranks instead of the 
+neighboring points, or by highlighting.
+Brush compostion, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
+This is left for
+future implementations of `liminal`.
 
@@
 brushing on the embedding view will produce a one-to-one linking with the tour
-view (the special case of $k=1$ nearest neighbors brush). It is possible
-to shift to one-to-many linking by increasing $k$ via the 'controls'
-panel of the interface.  If desired, the distance
-metric used to estimate nearest neighbours can by chosen by the user. 
+view (the special case of $k=1$ nearest neighbors brush). In future
+implementations it will be possible to shift from one-to-many linking by increasing $k$ via the 'controls' panel of the interface.
+If desired, the distance metric used to estimate nearest neighbours can by chosen by the user.
 The user may provide a reference data set for computing
@@
 (ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
-The data lies 2-d tree like structure consisting of ten branches.
-This data is available in the `phateR` package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added.
-[@Moon2019-ce].
+**(a)** The true data lies 2-d tree like structure consisting of ten branches.
+This data is available in the `phateR` package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
+**(b)** The first two principal components, which form the initial projection for the tour, note that the backbone of the three is obscured by this view.
+**(c)** The default tSNE view breaks the global structure of the tree.
+**(d)** Altering the initialision of tSNE using the first two principal components, clusters the tree at its branching points.
 
-```{r fake-trees, echo = FALSE, out.width = "50%", fig.align = "center", fig.cap="(ref:fake-trees)"}
+```{r fake-trees, echo = FALSE, fig.align = "center", fig.cap="(ref:fake-trees)"}
 knitr::include_graphics(here::here("img", "tree-view.png"))
@@
 linked brushing approaches to look at neighbourhood preservation
-(which can be seen via the tour) and to look at marker genes through
-the lens of the tour.
+and look at marker genes through the lens of the tour.
 
@@
 Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
-used Louvain clustering to generate clusters. To check and verify the 
-clustering we construct a liminal view. Because most off the variability in
+used Louvain clustering to generate clusters [@Blondel2008-bx]. 
+To check and verify the clustering we construct a liminal view. Because most off the variability in
 expression is explained by the first 5 PCs, we tour those. The t-SNE view

### 2020-07-18 paper.Rmd
update code and text

@@
 where t-SNE is applied to three simulations where cluster structure and geometry
-is known with varying success.
+of the underlying data is known.
 
-In the second case study, we explore a well-clustered single cell
+In the second case study, we look at well-clustered single cell
 transcriptomics data of mice retina...
@@
 used Louvain clustering to generate clusters [@Blondel2008-bx]. 
-To check and verify the clustering we construct a liminal view. Because most off the variability in
-expression is explained by the first 5 PCs, we tour those. The t-SNE view
-is computed from all PCs.
+To check and verify the clustering we construct a liminal view. 
+We tour the first five PCs (approximately 20% of the variance in expression),
+along side the tSNE view which was computed from all 25 PCs. Due to latency
+of the interface we do a ten percent weighted sample of rows based on
+cluster membership.
 

### 2020-07-18 scripts/liminal-mouse.R

@@ -2,6 +2,7 @@
 suppressPackageStartupMessages(library(scater))
 suppressPackageStartupMessages(library(scran))
 library(liminal)
+library(ggplot2)
 source(here::here("scripts", "helpers.R"))
 
 sce <- scRNAseq::MacoskoRetinaData(ensembl = TRUE)
@@ -26,34 +27,29 @@ sce <- logNormCounts(sce)
 dec <- modelGeneVar(sce) # fits mean-variance trend using `nls`.
 hvg <- getTopHVGs(dec, prop=0.1) # selects ~10% of nrow(sce) that are HVGs
 
-# looking at top ten most variable genes as a tour
-hvg_df <- tibble::as_tibble(as.matrix(t(logcounts(sce[hvg[1:10],]))))
-colnames(hvg_df) <- rowData(sce)[hvg[1:10],]$SYMBOL
 
-limn_tour(hvg_df[sample.int(nrow(hvg_df), 3000),], cols = 1:10)
 
 ## ----pca------------------------------------------------------------
 sce <- runPCA(sce, ncomponents = 25, subset_row = hvg)
-# looking at elbow-plot looks like we get the first ten PCs explain most of the 
-# variation, with the elbow at five
+# looking at elbow-plot looks like we get the first five PCs explain most of the 
+# variation
 var_explained <- tidy_var_explained(sce)
-limn_xy(var_explained, x = component, y = var_explained)
+ggplot(var_explained, aes(x = component, y = var_explained)) +
+  geom_point()
 
 ## ----cluster--------------------------------------------------------
 # the recommended workflow is to use graph based clustering on the principal 
 # components
 g <- buildSNNGraph(sce, use.dimred = 'PCA')
-sce$cluster_membership <- factor(igraph::cluster_louvain(g)$membership)
+colData(sce)$cluster_membership <- factor(igraph::cluster_louvain(g)$membership)
 
 # extract PCs to tour and perform tSNE on, keep cluster labels too
 pc_df <- reducedDim(sce, "PCA")
 pc_df <- dplyr::as_tibble(pc_df)
-pc_df$cluster_membership <- sce$cluster_membership
+pc_df$cluster_membership <- colData(sce)$cluster_membership
+pc_df$inx <- seq_len(nrow(pc_df))
 
-dplyr::count(pc_df, cluster_membership)
-
[…]

### 2020-07-18 paper.Rmd
add marker gene example

@@
 used Louvain clustering to generate clusters [@Blondel2008-bx].
-To check and verify the clustering we construct a liminal view. 
+
+To check and verify the clustering we construct a `liminal` view. 
 We tour the first five PCs (approximately 20% of the variance in expression),
-along side the tSNE view which was computed from all 25 PCs. Due to latency
-of the interface we do a ten percent weighted sample of rows based on
-cluster membership.
+along side the tSNE view which was computed from all 25 PCs. 
+ Due to latency of the interface we do weighted sample of rows based on
+cluster membership, leaving us with approximately 10 per cent of the original
+data size - 4,590 cells. Although this is not ideal, it still allows us to
+get a sense of the shape of the clusters as seen in the video below:
 
-- compare to just using the HVGs themselves with the tour
-- neighbourhood based brushing to check how diffuse or distorted clusters are
-- different distance metrics?
+< embedded video mouse 1 >
+
+In the following video we can observe that the embedding has mostly captured the
+clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
+linked brushing. As expected avoids tSNE mitigates the crowding problem that is
+an issue for tour in this case, where points in clusters 2,4,6, and 11 are
+clumped together in tour view, but are blown up in the embedding view. 
+The tour appears to form a tetrahedron like shape, with points lying on
+surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
+a phenomena that has also been observed in @Korem2015-af. Brushing on
+the tour view, reveals points in cluster 9 that are diffuse in the embedding 
+view, points in cluster 9 are relatively far away and spread apart from other clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding.
+
+Next, we identify marker genes for clusters using one sided Welch t-tests with
+a minimum log fold change of one as recommended by @Amezquita2020-at, which
+uses the testing framework from @McCarthy2009-qx. We select the top 10 marker genes that are upregulated in cluster 2, which was one of the clumped clusters
+when we toured on principal components. Here the tour becomes an alternative
+to a standard heatmap view for assessing shared markers; the basis
+generation (shown as the biplot on the left view) reflects the relative 
+weighting of each gene. We run the tour directly on the 
+log normalised expression values using the same subset as before:
+
+< embedded video mouse 2 >
+
+We see that the expression of the marker genes using t, appear to separate
+the previously clumped clusters 2,4,6, and 11 from the other clusters, 
+indicating that these genes are expressed in all four clusters, and the 
+magnitude of expression in these genes is similar accross these groups, 
+consistent with the results of marker gene analysis.
 

### 2020-07-18 scripts/liminal-mouse.R

@@ -52,9 +52,8 @@ pc_df$inx <- seq_len(nrow(pc_df))
 table(colData(sce)$cluster_membership)
 
 ## ---- tsne
-tsne <- Rtsne::Rtsne(dplyr::select(pc_df, -cluster_membership),
+tsne <- Rtsne::Rtsne(reducedDim(sce, "PCA"),
                      pca = FALSE,
-                     Y_init = clamp_sd(reducedDim(sce)[ , 1:2], sd = 1e-4),
                      perplexity = 30)
 
 tsne_df <- tidy_tsne(tsne, list(cluster_membership = pc_df$cluster_membership))
@@ -76,4 +75,26 @@ limn_tour_link(embed_data,
                tour_color = cluster_membership
 )
 
+# ---
 
+markers <- findMarkers(sce, 
+                       groups = colData(sce)$cluster_membership,
+                       direction = "up",
+                       lfc = 1,
+                       pval.type = "some")
+
+marker_genes <- markers[[2]]
+
+sub_sce <- sce[rownames(sce) %in% rownames(marker_genes)[1:10]]
+
+exprs <- t(logcounts(sub_sce))
+colnames(exprs) <- rowData(sub_sce)$SYMBOL
+
+tour_marker <- dplyr::bind_cols(
+  dplyr::as_tibble(as.matrix(exprs)),
+  cluster_membership = colData(sub_sce)$cluster_membership
+) %>% 
+  janitor::clean_names() %>% 
+  slice(tour_data$inx)
+
+limn_tour(tour_marker, 1:10, color = cluster_membership)

### 2020-07-18 paper.Rmd
add videos; remove last example for now

@@
 useful for cluster orientation tasks.  The implementation of our framework is
-available as an R package called `liminal` available at
+available as an R package called **liminal** available at
 https://github.com/sa-lee/liminal.
@@
 cluster orientation tasks. This workflow is incorporated into an R package
-called `liminal` (Available: https://github.com/sa-lee/liminal).
+called **liminal** (Available: https://github.com/sa-lee/liminal).
 
@@
 one linked brushing from the embedding to the tour view. This can be generalised further with one to many brushing, via spatial brushing. This form
-of brushing acts directly on $k$-nearest neighbors ($l$-nn) graph computed from a
+of brushing acts directly on $k$-nearest neighbours ($l$-nn) graph computed from a
 reference dataset, when a user brushes over a region in the embedding,
@@
 or binned colour scales to encode distances or ranks instead of the 
-neighboring points, or by highlighting.
-Brush compostion, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
+neighbouring points, or by highlighting.
+Brush composition, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
 This is left for
-future implementations of `liminal`.
+future implementations of **liminal**.
 
@@
 We have implemented the above design as an open source R package called
-`liminal` [@baseR]. The package allows analysts to construct concatenated
+**liminal** [@baseR]. The package allows analysts to construct concatenated
 visualisations, via the Vega-Lite grammar of interactive graphics (using
 `vegawidget` package) and provides an interface for constructing brushes and
-manipulating tour paths using the `shiny` and `tourr` packages. `liminal` also
+manipulating tour paths using the `shiny` and `tourr` packages. **liminal** also
 provides a stand-alone interface to the tour, in addition to the linked
-scatterplot interfaces discussed above. Below we display the `liminal` API, and
+scatterplot interfaces discussed above. Below we display the **liminal** API, and
 our approach to generating tour paths and user interactions.
@@
 readers who are unfamiliar with the technique before discussing the 
-computational aspects of how `liminal` interfaces with them. The tour
+computational aspects of how **liminal** interfaces with them. The tour
 is emphatically not a dimension reduction technique but rather a visualisation 
@@
 random, a *guided tour* can be used to generate a sequence 'interesting' 
-projections as quantified by an index function. While `liminal` is able to 
+projections as quantified by an index function. While **liminal** is able to 
 visualise guided tours, our focus in the case studies uses the grand tour to
@@
 animation. Like with a flipbook, an interface to the tour requires the ability
-to interact and modify it in real time. The user interface generated in `liminal`
+to interact and modify it in real time. The user interface generated in **liminal**
 allows a user to play, pause, and reset the tour animation,
@@
 brushing on the embedding view will produce a one-to-one linking with the tour
-view (the special case of $k=1$ nearest neighbors brush). In future
+view (the special case of $k=1$ nearest neighbours brush). In future
 implementations it will be possible to shift from one-to-many linking by increasing $k$ via the 'controls' panel of the interface.
@@
 is done with the `FNN` package for datasets with less than 4096 observations
-and `RcppAnnoy` for larger datasets or if a user selects a non-euclidean
+and `RcppAnnoy` for larger datasets or if a user selects a non-Euclidean
 distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
@@
 
-The next section steps through three use cases of our approach using simulated
+The next section steps through case studies cases of our approach using simulated
 and real data.  The videos presented show the visual 
-appearance of the `liminal` interface, and how we can interact with the tour
+appearance of the **liminal** interface, and how we can interact with the tour
 via the controls previously described.
@@
 
-In the second case study, we look at well-clustered single cell
-transcriptomics data of mice retina...
+In the second case study, we apply our approach to clustering 
+the mouse retina data from @Macosko2015-ot, and use apply the tour
+to the process of verifying marker genes.
 
+<!--
 Finally, in the last case study we
-explore a subset of the QuickDraw dataset, we use `liminal` to identify
+explore a subset of the QuickDraw dataset, we use **liminal** to identify
 instances of images from different labels that appear similar.
+-->
 
@@
 To begin we look at simulated datasets that reproduce known facts
-aboutthe  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
+about the  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
 Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
 covariance matrix. We then computed a t-SNE layout with default settings 
-using the `Rtsne` package [@Rtsne],
-and set up the `liminal` linked interface with grand tour on the 10-$d$
+using the **Rtsne** package [@Rtsne],
+and set up the **liminal** linked interface with grand tour on the 10-$d$
 observations (video - ).
@@
 generally points are more concentrated in the projection view compared to 
-the tSNE layout. This can be seen from our brushing action on the tSNE side.
+the t-SNE layout. This can be seen from our brushing action on the t-SNE side.
 
@@
 10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
-Again tSNE has correctly identified the sub-clusters but their relative 
+Again t-SNE has correctly identified the sub-clusters but their relative 
 locations to each other is distorted, with the orange and blue groups being
@@
 to see the sub-clusters in the tour view, however, once we zoom and highlight
-they become more apparent. When we brush the sub-clusters in the tSNE, their
+they become more apparent. When we brush the sub-clusters in the t-SNE, their
 relative placement is again exaggerated, with the tour showing that they
-are indeed much closer than the impression the tSNE view gives.
+are indeed much closer than the impression the t-SNE view gives.
 
@@
 (ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
-**(a)** The true data lies 2-d tree like structure consisting of ten branches.
-This data is available in the `phateR` package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
-**(b)** The first two principal components, which form the initial projection for the tour, note that the backbone of the three is obscured by this view.
-**(c)** The default tSNE view breaks the global structure of the tree.
-**(d)** Altering the initialision of tSNE using the first two principal components, clusters the tree at its branching points.
+*(a)* The true data lies 2-d tree like structure consisting of ten branches.
+This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
+*(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the three is obscured by this view.
+*(c)* The default t-SNE view breaks the global structure of the tree.
+*(d)* Altering the initialision of t-SNE using the first two principal components, clusters the tree at its branching points.
 
@@
 initialisation). We then create a linked
-tour with t-SNE layout with `liminal` as shown in the following video :
+tour with t-SNE layout with **liminal** as shown in the following video :
 
@@
 First, we downloaded the raw mouse retinal single cell RNA-seq data from
-@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
+@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
 have followed a standard workflow for pre-processing and normalizing this data 
@@
 
-To check and verify the clustering we construct a `liminal` view. 
+To check and verify the clustering we construct a **liminal** view. 
 We tour the first five PCs (approximately 20% of the variance in expression),
-along side the tSNE view which was computed from all 25 PCs. 
+along side the t-SNE view which was computed from all 25 PCs. 
  Due to latency of the interface we do weighted sample of rows based on
@@
 clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
-linked brushing. As expected avoids tSNE mitigates the crowding problem that is
+linked brushing. As expected avoids t-SNE mitigates the crowding problem that is
 an issue for tour in this case, where points in clusters 2,4,6, and 11 are
@@
 
+<!--
 ### Case Study 3: How do people draw animals?
@@
 categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple",
-and "whale".
+and "whale". We are interested in discovering images that are similiar but
+belong to different classifications 
+-->
 
@@
 
-- limitations: large $n$, latency issues
-- future work: regularisation tour, slicing and sectioning
+We have shown that the use of tours as a tool for interacting with
+high dimensional data provides an addiontal insight for interrogating views
+generated from embeddings. The interface we have designed in the **liminal**
+package, allows a user to gain a deeper understanding of an embedding 
+algorithm, and rectifies perceptual issues associated with NLDR methods via 
+linked interactions with the tour. As we have shown in the simulation case
+studies, the t-SNE method can produce misleading embeddings but can be
+detected through the use linked brushing and highlighting. In the case when
+the data has a piecewise linear geometry, like the tree simulation, the tour
+preserves the shape of the data which can be obscured by the embedding method.
 
-## Acknowledgements
+Our holistic framework is also useful in real data situations, as displayed
+in our second case study. The tour when combined with t-SNE allowed us
+to identify clusters, while giving us an idea of their orientation to 
+each other. Moreover, we could visually inspect the separation clusters
+using a tour on marker gene sets. This case study revealed some limitations
+in using the **liminal** interface for larger datasets. First,
+t-SNE avoids the crowding problem, points are seperated into distinct regions
+on the graph canvas, while for the tour points are concentrated in the
+center of the projection and become difficult to see. 
+Second, as $n$ increases the animation becomes slower as there is more data passing from the server to the client.
 

### 2020-07-19 paper.Rmd
update paper

@@
 to identify clusters, while giving us an idea of their orientation to 
-each other. Moreover, we could visually inspect the separation clusters
-using a tour on marker gene sets. This case study revealed some limitations
+each other. Moreover, we could visually inspect the separation of clusters
+using a tour on marker gene sets. We see our approach as being valuable
+to the single cell analyst who wants to make their embeddings more 
+interpretable.
+
+There are some limitations
 in using the **liminal** interface for larger datasets. First,
 t-SNE avoids the crowding problem, points are seperated into distinct regions
-on the graph canvas, while for the tour points are concentrated in the
+on the display canvas, while for the tour, points are concentrated in the
 center of the projection and become difficult to see. 
-Second, as $n$ increases the animation becomes slower as there is more data passing from the server to the client.
+Second, as $n$ increases the animation becomes slower as there is more data passing from the server to the client. Both of these issues could be mitigated
+through alternate tour paths and displays. The recent work
+by @Laa2020-wr is a promising area for further investigation, as well as
+work from topological statistics [@Rieck2017-kk; @Genovese2017-iq].
 

### 2020-07-28 scripts/figures.R
add figures

@@ -1,0 +1,41 @@
+# pull screenshots together using cowplot
+library(ggplot2)
+library(cowplot)
+
+draw_screenshot <- function(file) {
+  ggdraw() + 
+    draw_image(here::here("img", file), scale = 0.8) +
+    theme_void()
+}
+
+enframe_screenshot <- function(lst) {
+  plot_grid(plotlist = lst, 
+            ncol = 2, 
+            labels = "auto",
+            align = "hv",
+            axis = "t" )
+}
+
+case_fig <- function(case) {
+  files <- list.files(here::here("img"), 
+                      pattern = ".png$")
+  
+  files <- files[grep(paste0("^", case), files)]
+
+  enframe_screenshot(lapply(files, draw_screenshot))
+}
+
+cases <- c("gaussian", 
+           "hierarchical", 
+           "trees-01", 
+           "trees-02", 
+           "mouse-01",
+           "mouse-02")
+
+for (case in cases) {
+  fout <- paste0("liminal-screenshot-", case, ".png")
+  pout <- case_fig(case)
+  
+  ggsave2(here::here("img", fout), plot = pout)
+  
+}

### 2020-07-29 paper.Rmd
update paper and include videos

@@
 
+```{r setup-videos, cache = FALSE}
+vid_lookup <- tibble::tribble(
+  ~key, ~url, ~path,
+  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("img", "liminal-screenshot-gaussian.png"),
+  "hierarchical", "https://player.vimeo.com/video/439635905", here::here("img", "liminal-screenshot-hierarchical.png"),
+  "trees-01", "https://player.vimeo.com/video/439635892", here::here("img", "liminal-screenshot-trees-01.png"),
+  "trees-02", "https://player.vimeo.com/video/439635863", here::here("img", "liminal-screenshot-trees-02.png"),
+  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("img", "liminal-screenshot-mouse-01.png"),
+  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("img", "liminal-screenshot-mouse-02.png")
+)
+
+display_vid_html <- function(key, lookup) {
+  video <- lookup[lookup$key == key, ]
+  knitr::include_url(video[["url"]])
+}
+
+display_screenshot <- function(key, lookup) {
+  img <- lookup[lookup$key == key, ]
+  knitr::include_graphics(img[["path"]])
+}
+
+```
+
+# Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings {#ch:tsne}
+
 There has been a rapid uptake in the use of non-linear dimensionality reduction
@@
 the preservation of local and global structure in the resulting visualisation.
-We present a visual analytics framework for the pragmatic usage of NLDR methods
+We present visual diagnostics for the pragmatic usage of NLDR methods
 by combining them with a technique called the tour. A tour is a sequence of
@@
 linking the tour to a view obtained from an NLDR method, we can preserve global
-structure and through user interactions like spatial linked brushing observe
+structure and through user interactions like linked brushing observe
 where the NLDR view may be misleading. We display several case studies from
-single cell transciptomics and image processing, that shows our approach is
-useful for cluster orientation tasks.  The implementation of our framework is
-available as an R package called **liminal** available at
+both simulated and real data from single cell transcriptomics, that shows our
+approach is useful for cluster orientation tasks.  The implementation of our 
+framework is available as an R package called **liminal** available at
 https://github.com/sa-lee/liminal.
@@
 High dimensional data is increasingly prevalent in the natural sciences and
-beyond but presents a challenge to the analyst in terms of both data cleaning /
+beyond but presents a challenge to the analyst in terms of data cleaning,
 pre-processing and visualisation. Methods to embed data from a high-dimensional
@@
 workflow where they are used to ascertain hidden structure and de-noise data
-for downstream analysis (thereby nullifying the 'curse of dimensionality').
+for downstream analysis .
 
@@
 variables that are a-priori known to distinguish clusters. Both of these
-workflow steps rely on the embedding being representative the original high
+workflow steps rely on the embedding being representative of the original high
 dimensional dataset, and becomes much more difficult when there is no underlying
@@
 of non-linear embedding techniques were more likely to consider clusters of
-points on a 2d scatter plot to be the result of a spurious embedding compared
+points on a 2D scatter plot to be the result of a spurious embedding compared
 to advanced users who were aware of the inner workings of the embedding
@@
 onto a low-dimensional basis matrix, and is represented as a
-animated visualization.Given the dynamic nature of the tour, user interaction is 
-important for controlling and exploring the visualisation: the tour has been
-used previously by @Wickham2015-cx and @Li2020-rh as a tool for 
+animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic nature
+of the tour, user interaction is  important for controlling and exploring the 
+visualisation: the tour has been used previously by @Wickham2015-cx and  
 exploring statistical model fits and by @Buja1996-fk for exploring the space of 
@@
 
-While there has been much work on the algorithmic details of the aforementioned
-embedding methods, there has been relatively few tools designed to assist users
-to interact with these techniques and assist them in making comparisons between
-embeddings and performing the aforementioned cluster orientation tasks. Several
+While there has been much work on the algorithmic details of
+embedding methods, there are relatively few tools designed to assist users
+to interact with these techniques: when is an embedding sufficient for the
+task at hand? Several
 interactive interfaces have been proposed for evaluating or using embedding
-techniques: the Sleepwalk interface provides a click and highlight
-visualisation for colouring points in an embedding according to an estimated
-pairwise distance in the original high-dimensional space [@Ovchinnikova2020-sy].
-The idea behind this is that non-linear embedding methods should preserve 
-locality and that points that are 'close' to each other on the embedded view
-should be close according to the estimated pairwise distance. 
-Similarly, the TensorFlow embedding projector is a web interface to 
+techniques. @Buja2008-fn used tours to guide analysts during the optimisation
+of multidimensional scaling methods by extending their interactive visualisation
+software called **XGobi** and **GGobi** into a new tool called **GGvis** 
+[@Swayne1998-uq; @Swayne2003-qd; @Swayne2004-ua].
+Their interface allows the analyst to dynamically modify and check whether
+an MDS configuration has preserved the locality and closeness
+of points between the configuration and the original data.
+@Ovchinnikova2020-sy created the **Sleepwalk**  interface for checking
+non-linear embeddings in single cell RNA-seq data. It provides a click and 
+highlight visualisation for colouring points in an embedding according to an 
+estimated pairwise distance in the original high-dimensional space.
+Similarly, the **TensorFlow**  embedding projector is a web interface to 
 running some non-liner embedding methods live in the browser and 
@@
 [@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
-guided and modified form of the  t-SNE algorithm, that allows users to modified 
+guided and modified form of the  t-SNE algorithm, that allows users to modify 
 optimisation parameters in real-time.
@@
 embeddings require interaction but should also be able to be incorporated into
-an analysts workflow. We propose a more pragmatic workflow inspired by
+an analysts workflow. Instead, we implement a more pragmatic workflow by
 incorporating interactive graphics and tours with embeddings that allows users
 to see a global overview of their high dimensional data and assists them with
-cluster orientation tasks. This workflow is incorporated into an R package
-called **liminal** (Available: https://github.com/sa-lee/liminal).
+cluster orientation tasks.
 
-## Dimensionality Reduction Background
+The rest of the paper is organised as follows. The next section provides
+background on dimension reduction methods, including an overview of the tour.
+Then we describe the visual design of **liminal**, followed by implementation
+details. Next we provide case studies that show how our interface assists
+in using embedding methods. Finally, we describe the insights gained by using
+**liminal** and plans for extensions to the software.
+
+## Overview of Dimension Reduction
 
@@
 (t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
-mention several other techniques.
+mention several other techniques [@Maaten2008-sk; @McInnes2018-co].
 
@@
 Multidimensional scaling (MDS) is a class of DR methods that aims to construct
-an embedding  $Y$ such that the pairwise distances (inner products) in $Y$ 
-approximate the pairwise distances (inner products) in $X$ 
+an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
+approximate the pair-wise distances (inner products) in $X$ 
 [@Torgerson1952-am; @Kruskal1964-cz]. There are many 
@@
 A related technique is Isomap which uses a $k$-nearest neighbour graph
-to estimate the pairwise geodesic distance of points in $X$ then uses classical 
+to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
 MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
-processes such asdiffusion maps [@Coifman2005-ak]. A recent example of this 
+processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
 approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
-estimated via the pairwise distance matrix and k-nearest neighbours graph of $X$. 
+estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$. 
 The algorithm de-noises estimated distances in high-dimensional space via 
@@
 The t-SNE algorithm estimates
-the pairwise similarity of (Euclidean) distances of points in a high
+the pair-wise similarity of (Euclidean) distances of points in a high
 dimensional space using a Gaussian distribution and then estimates a
 configuration in the low dimensional embedding space by modelling similarities
-using a t-distribution with 1 degree of freedom. There are several subtleties
+using a t-distribution with 1 degree of freedom [@Maaten2008-sk]. There are several subtleties
 to the to use of the algorithm that are revealed by stepping through its
@@
 
-To begin, t-SNE transforms pairwise distances between $\mathbf{x_i}$ and
+To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
 $\mathbf{x_j}$ to similarities using a Gaussian kernel:
@@
 
-Next, the target low-dimensional space, $Y$, pairwise distances between
+Next, the target low-dimensional space, $Y$, pair-wise distances between
 $\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
@@
 So when the the loss function is minimised this corresponds to large attractive
-forces, that is, the pairwise distances in $Y$ are small when there are
+forces, that is, the pair-wise distances in $Y$ are small when there are
 non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
 should also be small when the loss function is minimised, that is, when
-pairwise distances in $Y$ are large regardless of the magnitude of the
+pair-wise distances in $Y$ are large regardless of the magnitude of the
 corresponding distances in $X$.
@@
 
-Another recent method UMAP has seen a large rise in popularity (at least in
+Another recent method, UMAP, has seen a large rise in popularity (at least in
 single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
@@
 which is optimized using stochastic gradient descent and defines a different
-kernel for similarities in the low dimensional space. Due to it's computational
-speed it's possible to generate UMAP embeddings in more than three dimensions.
-It appears to suffer from the same perceptual issues as t-SNE, however it
-supposedly preserves global structure better than t-SNE [@Coenen2019-to].
+kernel for similarities in the low dimensional space. Due to its computational
+speed it is possible to generate UMAP embeddings in more than three dimensions.
+It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE [@Coenen2019-to].
+
+### Tours explore the subspace of $d$-dimensional projections
+
+The tour is a visualisation technique that is grounded in 
+mathematical theory, and is able to ascertain 
+the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections [@Asimov1985-vp; @Buja1986-zr].
+
+Like, when using other DR techniques, the tour assumes we have a real data matrix
+$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
+a  sequence of $p \times d$ orthonormal projection matrices (bases) 
+$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
+orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
+path between them is interpolated to form intermediate frames, and giving the 
+sense of continuous movement from one basis to another. The tour is then the 
+continuous visualisation of the projections $Y_{t} = XA_{t}$, that is the 
+projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
+successive bases. A *grand tour* corresponds to choosing new orthonormal
+bases at random; allowing a user to ascertain structure via exploring the
+subspace of $d$-dimensional projections.  In practice, we first sphere 
+our data via principal components to reduce dimensionality of $X$ prior to 
+running the tour. Instead of picking projections at 
+random, a *guided tour* can be used to generate a sequence of 'interesting' 
+projections as quantified by an index function [@Cook1995-bi]. While our software, 
+**liminal** is able to  visualise guided tours, our focus in the case studies 
+uses the grand tour to see global structure in the data.
 
@@
 
-We propose using tours as part of an analyst's workflow when using embedding
-methods like t-SNE. We have made extensive use of ensemble graphics, that is
-aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
-As we will see in the case studies, this allows analysts to quickly compare 
-views from embedding methods and allows them to see how the embedding method 
-alters the global structure of their data. Using ensembles allows the use of 
-interaction techniques, that allow analysts to perform cluster orientation 
-tasks via linking multiple views. This approach allows our interface, to achieve 
+Tours provide a supportive visualisation to NLDR graphics, and 
+can be easily incorporated into an analysts workflow with our software package, 
+**liminal**.  Our interface allows analysts to quickly compare 
+views from embedding methods and see how an embedding method 
+preserves or alters the geometry of their data. Using multiple concatenated and
+linked views with the tour enhances interaction techniques, and allows 
+analysts to perform  cluster orientation tasks via linked highlighting and 
+brushing [@McDonald1982-ew; @Becker1987-gd]. 
+This approach allows our interface to achieve 
 the three principles for interactive high-dimensional data visualisation outlined
@@
 tour is dynamic, it is often useful to be able to pause and highlight a
-particular view. In our interface, brushing will pause the tour path, allowing
-users to identify 'interesting' projections. The domain of the axis scales from
-running a tour is called the half range, and is computed by rescaling the input
-data onto hyper-dimensional unit cube. We bind the half range to a mouse wheel
-event, allowing a user to pan and zoom on the tour view dynamically. This is
-useful for peeling back dense clumps of points to reveal structure.
+particular view. In addition to pause, play and reset buttons, brushing will 
+pause the tour path, allowing users to identify 'interesting' projections. The
+domain of  the axis scales from running a tour is called the half range, and is 
+computed by rescaling the input data onto hyper-dimensional unit cube. We bind 
+the half range to a mouse wheel event, allowing a user to pan and zoom on the 
+tour view dynamically. This is useful for peeling back dense clumps of points 
+to reveal structure.
 
@@
 We have combined the tour view in a side by side layout with a scatter plot
-view that represents the output of an embedding algorithm.  These views are
+view as has been done in previous tour interfaces **XGobi** and **DataViewer** 
+[@Buja1986-ku; @Swayne1998-uq]. These views are
 linked; analysts can brush regions or highlight collections of points in either
@@
 
-We propose this can be
-done visually using our side-by-side tour and embedding views.
+These checks are done visually using our side-by-side tour and embedding views.
 In the simplest case, a local continuity check can be assessed via one to
-one linked brushing from the embedding to the tour view. This can be generalised further with one to many brushing, via spatial brushing. This form
-of brushing acts directly on $k$-nearest neighbours ($l$-nn) graph computed from a
-reference dataset, when a user brushes over a region in the embedding,
-all the points that match the graphs edges are selected on the 
-corresponding tour view. In place of highlighting, one could use opacity
-or binned colour scales to encode distances or ranks instead of the 
-neighbouring points, or by highlighting.
-Brush composition, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
-This is left for
-future implementations of **liminal**.
+one linked brushing from the embedding to the tour view. Similarly, diffusions
+are identified from linked brushing on the tour view, highlighting points
+in the embedding view.
 
@@
 We have implemented the above design as an open source R package called
-**liminal** [@baseR]. The package allows analysts to construct concatenated
-visualisations, via the Vega-Lite grammar of interactive graphics (using
-`vegawidget` package) and provides an interface for constructing brushes and
-manipulating tour paths using the `shiny` and `tourr` packages. **liminal** also
-provides a stand-alone interface to the tour, in addition to the linked
-scatterplot interfaces discussed above. Below we display the **liminal** API, and
-our approach to generating tour paths and user interactions.
-
-### Tours explore the subspace of $d$-dimensional projections
-
-Here we provide an overview of the mathematical aspects of the tour for
-readers who are unfamiliar with the technique before discussing the 
-computational aspects of how **liminal** interfaces with them. The tour
-is emphatically not a dimension reduction technique but rather a visualisation 
-technique that is grounded in mathematical theory, that is able to ascertain 
-the shape and global structure of a dataset via inspection of the subspace 
-generated by the set of low-dimensional projections. In fact, in some
-of the case studies below, we  first sphere our data via principal 
-components to reduce dimensionality of our input.
-
-Like, when using DR techniques, the tour assumes we have a real data matrix
-$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
-a  sequence of $p \times d$ orthonormal projection matrices (bases) 
-$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
-orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
-path between them is interpolated to form intermediate frames, and giving the 
-sense of continuous movement from one basis to another. The tour is then the 
-continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the 
-projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
-successive bases. A *grand tour* corresponds to choosing new orthonormal
-bases at random; allowing a user to ascertain structure via exploring the
-subspace of $d$-dimensional projections. Instead of picking projections at 
-random, a *guided tour* can be used to generate a sequence 'interesting' 
-projections as quantified by an index function. While **liminal** is able to 
-visualise guided tours, our focus in the case studies uses the grand tour to
-see global structure in the data.
+**liminal** [@r-liminal]. The package allows analysts to construct concatenated
+visualisations, drawn with the **Vega-Lite** grammar of interactive graphics via
+the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides 
+an interface for constructing linked and stand alone interfaces for 
+manipulating tour paths via  the **shiny** and **tourr** packages 
+[@Chang2020-bq; @Wickham2011-st].
 
@@
 technique. Generally, the user would set $d=2$ the tour is visualised as
-a animated scatterplot. This process of constructing bases and intermediate 
-frames and visualising the resulting projection is akin to making a "flipbook" 
-animation. Like with a flipbook, an interface to the tour requires the ability
+a animated scatter plot. This process of constructing bases and intermediate 
+frames and visualising the resulting projection is akin to making a "flip book" 
+animation. Like with a flip book, an interface to the tour requires the ability
 to interact and modify it in real time. The user interface generated in **liminal**
@@
 brushing on the embedding view will produce a one-to-one linking with the tour
-view (the special case of $k=1$ nearest neighbours brush). In future
-implementations it will be possible to shift from one-to-many linking by increasing $k$ via the 'controls' panel of the interface.
-If desired, the distance metric used to estimate nearest neighbours can by chosen by the user.
-The user may provide a reference data set for computing
-nearest neighbours (for example a distance matrix, or the complete data matrix)
-independently of the tour or embedding views. The estimation of neighbours
-is done with the `FNN` package for datasets with less than 4096 observations
-and `RcppAnnoy` for larger datasets or if a user selects a non-Euclidean
-distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
-with the interface, the brush regions are returned for downstream analysis.
+view. For interpreting specific combinations of clusters, the multiple
+guides on the legend can be selected in order to see their relative orientations.
+The interface is constructed as a **shiny** gadget specifically designed for
+interactive data analysis. Selections such as brushing regions and the current
+tour path are returned after the user clicks done on the interface and become
+available for further investigation.
 
@@
 The next section steps through case studies cases of our approach using simulated
-and real data.  The videos presented show the visual 
+and real data.
+
+The first three case studies use simulations where the cluster structure and
+geometry of the underlying data is known. We start with a simple example
+where we generated spherical clusters that are embedded well by t-SNE. Then
+we move onto more complex examples where the tour provides insight, such
+as clusters that have substructure and where there is more complex geometry
+in the data.
+
+In the final case study, we apply our approach to clustering 
+the mouse retina data from @Macosko2015-ot, and apply the tour
+to the process of verifying marker genes that separate clusters.
+
+The videos presented show the visual 
 appearance of the **liminal** interface, and how we can interact with the tour
-via the controls previously described.
-In the first case study, we look at three different examples
-where t-SNE is applied to three simulations where cluster structure and geometry
-of the underlying data is known.
+via the controls previously described. Both code and videos available for 
+reproducing this paper are available at https://github.com/sa-lee/paper-liminal.
 
-In the second case study, we apply our approach to clustering 
-the mouse retina data from @Macosko2015-ot, and use apply the tour
-to the process of verifying marker genes.
-
-<!--
-Finally, in the last case study we
-explore a subset of the QuickDraw dataset, we use **liminal** to identify
-instances of images from different labels that appear similar.
--->
-
-### Case Study 1: Exploring simulation data with tours and t-SNE
+### Case Study 1: Exploring spherical Gaussian clusters
 
@@
 and set up the **liminal** linked interface with grand tour on the 10-$d$
-observations (video - ).
+observations.
 
-<embedded video liminal gaussian>
+(ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921.
+*(a)*: t-SNE (left) and tour (right) in side by side layout coloured by cluster.
+*(b)*: brushing the red cluster on the t-SNE view, shows the corresponding points on the tour.
 
-We see in the video that t-SNE has correctly split out each cluster and
-laid them out in a star like formation. This is agrees with the tour view,
+```{r gaussian, fig.cap = "(ref:liminal-gaussian)"}
+display_screenshot("gaussian", vid_lookup)
+```
+
+```{r gaussian-h, eval = knitr::is_html_output()}
+display_vid_html("gaussian", vid_lookup)
+```
+
+We see in figure \@ref(fig:gaussian) and linked video that t-SNE has 
+correctly split out each cluster and laid them out in a star like formation. 
+This is agrees with the tour view,
 where once we start the animation, the five clusters begin to appear but
 generally points are more concentrated in the projection view compared to 
-the t-SNE layout. This can be seen from our brushing action on the t-SNE side.
+the t-SNE layout (figure \@ref(fig:gaussian)a). 
+This can be confirmed via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
 
-<embedded video liminal hierachcial >
+### Case Study 2: Exploring spherical Gaussian clusters with hierarchical structure
 
-We repeat the process of simulating Gaussian clusters in video <>, but this time
-we have a cluster that has more granular structure. This time we have two
-Gaussian clusters with equal covariance embedded in $10-d$, and a third cluster
-with hierachcial structure. This cluster has two 3-$d$ clusters embedded in 
-10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
+We repeat the process of viewing Gaussian clusters from the 
+_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
+[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
+embedded in $10-d$, and a third cluster
+with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
+10-$d$, where the second cluster is subdivided into three smaller clusters, 
+that are each equidistant from each other and have the same covariance structure.
 Again t-SNE has correctly identified the sub-clusters but their relative 
 locations to each other is distorted, with the orange and blue groups being
-far from each other in the tour view. We see in this case that is difficult
+far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
+in this case that is difficult
 to see the sub-clusters in the tour view, however, once we zoom and highlight
-they become more apparent. When we brush the sub-clusters in the t-SNE, their
+they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
+the sub-clusters in the t-SNE, their
 relative placement is again exaggerated, with the tour showing that they
 are indeed much closer than the impression the t-SNE view gives.
+
+(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
+*(a)* t-SNE has identified and spread out the clusters with substructure.
+*(b)* Using linked highlighting and brushing, the tour view shows that sub-clusters are much closer together in high dimensional space than the t-SNE view implies.
+
+```{r hierarchical, fig.cap="(ref:liminal-hierarchical)"}
+display_screenshot("hierarchical", vid_lookup)
+```
+
+```{r hierarchicall-h, eval = knitr::is_html_output()}
+display_vid_html("hierarchical", vid_lookup)
+```
+
+### Case Study 3: Exploring data with piecewise linear structure
 
@@
 This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
-*(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the three is obscured by this view.
+*(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view.
 *(c)* The default t-SNE view breaks the global structure of the tree.
-*(d)* Altering the initialision of t-SNE using the first two principal components, clusters the tree at its branching points.
+*(d)* Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.
 
@@
 initialisation). We then create a linked
-tour with t-SNE layout with **liminal** as shown in the following video :
+tour with t-SNE layout with **liminal** as shown in figure \@ref(fig:trees-01).
 
-<embedded video 1>
+(ref:trees-01) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635892.
+*(a)*: Default t-SNE has difficulties embedding the tree.
+*(b)* linked highlighting shows that the backbone of the tree has been broken by the embedding.
 
-From the video, we see that the t-SNE view has been unable to recapitulate the
+```{r trees-01, fig.cap = "(ref:trees-01)"}
+display_screenshot("trees-01", vid_lookup)
+```
+
+```{r trees01-h, eval = knitr::is_html_output()}
+display_vid_html("trees-01", vid_lookup)
+```
+
+We see that the t-SNE view has been unable to recapitulate the
 topology of the tree - the backbone (blue) branch has been split into three
-fragments. We can see this immediately via the linked highlighting over both
+fragments (figure \@ref(fig:trees-01)a). We can see this immediately via the linked highlighting over both
 plots. If we click on the legend for the zero branch, the blue coloured points
@@
 view there is a lot of noise along the branches, while this does not appear to
-be the case for the t-SNE result.
+be the case for the t-SNE result (figure \@ref(fig:trees-01)b).
 
@@
 initialisation and increase the perplexity from 30 to 100. We then combine
-these results with our tour view as displayed in the video:
+these results with our tour view as displayed in figure \@ref(fig:trees-02).
 
-<embedded video 2>
+(ref:trees-02) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635863.
 
-The video shows that this selection of parameters results in the the tips of
-the branches (the three black dots in figure \@ref(fig:fake-trees) a) being split
-into three clusters representing the terminal branches of the tree. However, there
-are perceptual issues following the placement of the three groupings on the
-t-SNE view that become apparent via simple linked brushing. If we brush the
+```{r trees-02, fig.cap = "(ref:trees-02)"}
+display_screenshot("trees-02", vid_lookup)
+```
+
+```{r trees-02-h, eval = knitr::is_html_output()}
+display_vid_html("trees-02", vid_lookup)
+```
+
+Figure \@(ref:trees-02) shows that this selection of parameters results in the 
+tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
+being split into three clusters representing the terminal branches of the tree. 
+However, there are perceptual issues following the placement of the three 
+groupings on the t-SNE view that become apparent via simple linked brushing. 
+If we brush the
 tips of the yellow and brown branches (which appear to be close to each other
 on the t-SNE view), we immediately see the placement is distorted in the t-SNE
-view, and in the tour view these tips are at opposite ends of the tree.
+view, and in the tour view these tips are at opposite ends of the tree 
+(figure \@ref(fig:trees-02)b).
 Although, this is a known issue of the t-SNE algorithm, we can easily identify
@@
 
-### Case Study 2: Clustering single cell RNA-seq data
+### Case Study 4: Clustering single cell RNA-seq data
 
@@
 data size - 4,590 cells. Although this is not ideal, it still allows us to
-get a sense of the shape of the clusters as seen in the video below:
+get a sense of the shape of the clusters as seen in figure \@ref(fig:mouse-01).
 
-< embedded video mouse 1 >
+(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data.
+A video of the tour animation is available at https://player.vimeo.com/video/439635812.
+*(a)*: linked brushing identifies distortions in cluster placement by the embedding.
+*(b)*: t-SNE spreads out larger clusters like cluster 2, 4, 6, and 11 which are concentrated in the tour view.
+*(c)*: the tour view gives a sense of the geometry of the data; the PCs appear to form a tetrahedron like shape.
+*(d)*: brushing on the tour view highlights diffusions, points in cluster 9 are close together in the tour view are far away in the t-SNE view.
 
-In the following video we can observe that the embedding has mostly captured the
+```{r mouse-01, fig.cap = "(ref:mouse-01)" }
+display_screenshot("mouse-01", vid_lookup)
+```
+
+```{r mouse-01-h, eval = knitr::is_html_output()}
+display_vid_html("mouse-01", vid_lookup)
+```
+
+We observe that the embedding has mostly captured the
 clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
-linked brushing. As expected avoids t-SNE mitigates the crowding problem that is
+linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
 an issue for tour in this case, where points in clusters 2,4,6, and 11 are
-clumped together in tour view, but are blown up in the embedding view. 
+clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
 The tour appears to form a tetrahedron like shape, with points lying on
 surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
-a phenomena that has also been observed in @Korem2015-af. Brushing on
+a phenomena that has also been observed in @Korem2015-af 
+(figure \@ref(fig:mouse-01)c). Brushing on
 the tour view, reveals points in cluster 9 that are diffuse in the embedding 
-view, points in cluster 9 are relatively far away and spread apart from other clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding.
+view, points in cluster 9 are relatively far away and spread apart from other 
+clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
+(figure \@ref(fig:mouse-01)d).
 
@@
 a minimum log fold change of one as recommended by @Amezquita2020-at, which
-uses the testing framework from @McCarthy2009-qx. We select the top 10 marker genes that are upregulated in cluster 2, which was one of the clumped clusters
+uses the testing framework from @McCarthy2009-qx. We select the top 10 marker 
+genes that are upregulated in cluster 2, which was one of the clumped clusters
 when we toured on principal components. Here the tour becomes an alternative
@@
 weighting of each gene. We run the tour directly on the 
-log normalised expression values using the same subset as before:
+log normalised expression values using the same subset as before figure \@ref(fig:mouse-02).
 
-< embedded video mouse 2 >
+(ref:mouse-02) Screenshots of the **liminal** tour applied to a marker gene set.
+A video of the tour animation is available at https://player.vimeo.com/video/439635843.
+*(a)*: the axis (left) and tour (right) view of 10 marker genes, using linked highlighting we see that the marker genes separate clusters 2, 4, 6, and 11 from the other groups.
+*(b)*: after zooming, it appears that the marker genes give rise to a gradient along the clusters.
 
-We see that the expression of the marker genes using t, appear to separate
+```{r mouse-02, fig.cap = "(ref:mouse-02)"}
+display_screenshot("mouse-02", vid_lookup)
+```
+
+```{r mouse-02-h, eval = knitr::is_html_output()}
+display_vid_html("mouse-02", vid_lookup)
+```
+
+We see that the expression of the marker genes, appear to separate
 the previously clumped clusters 2,4,6, and 11 from the other clusters, 
-indicating that these genes are expressed in all four clusters, and the 
-magnitude of expression in these genes is similar accross these groups, 
-consistent with the results of marker gene analysis.
-
-<!--
-### Case Study 3: How do people draw animals?
-
-In the final case study, we explore a subset of the Google QuickDraw dataset 
-[@quickdraw-paper]. The data are sketches drawn by people of common concepts
-like 'airplane' or 'shoe', and corresponding metadata about the humans who
-created the sketches, such as country, the word category, whether the sketch
-was recognised by Google's neural network, and the number of strokes taken.
-We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the 
-`quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into 
-a 784 dimension image vector. We sampled 500 images from each of the following
-categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple",
-and "whale". We are interested in discovering images that are similiar but
-belong to different classifications 
--->
+indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). Using zooming, we can see a trajectory forming along
+the clusters, while the axis view shows that magnitude of expression in 
+the marker genes is similar across these separated clusters which is 
+consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).
 
@@
 We have shown that the use of tours as a tool for interacting with
-high dimensional data provides an addiontal insight for interrogating views
+high dimensional data provides an additional insight for interrogating views
 generated from embeddings. The interface we have designed in the **liminal**
@@
 linked interactions with the tour. As we have shown in the simulation case
-studies, the t-SNE method can produce misleading embeddings but can be
+studies, the t-SNE method can produce misleading embeddings which can be
 detected through the use linked brushing and highlighting. In the case when
@@
 
+In future implementations it will be possible to shift from one to one linked
+brushing to one to many linked brushing. This form of brushing acts directly on
+a $k$-nearest neighbours ($k$-nn) graph computed from a
+reference dataset: when a user brushes over a region in the embedding,
+all the points that match the graphs edges are selected on the 
+corresponding tour view.  The reference data set for computing
+nearest neighbours (for example a distance matrix, or the complete data matrix)
+can be independent of the tour or embedding views. The estimation of neighbours
+can be computed efficiently with the `FNN` or `RcppAnnoy` packages
+on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 
+could use opacity or binned colour scales to encode distances or ranks instead 
+of the neighbouring points. Brush composition, such as 'and', 'or',
+or 'not' brushes, could be used to interrogate mismatches between the 
+$k$-nn graphs estimated from both the embedding and tour views.
+
 There are some limitations
 in using the **liminal** interface for larger datasets. First,
-t-SNE avoids the crowding problem, points are seperated into distinct regions
+t-SNE avoids the crowding problem, points are separated into distinct regions
 on the display canvas, while for the tour, points are concentrated in the
-center of the projection and become difficult to see. 
-Second, as $n$ increases the animation becomes slower as there is more data passing from the server to the client. Both of these issues could be mitigated
+centre of the projection and become difficult to see. 
+Second, as $n$ increases the animation becomes slower as there is more data
+passing from the server to the client. Both of these issues could be mitigated
 through alternate tour paths and displays. The recent work
@@
 
+## Acknowledgements {-}
+
+We would like to thank Dr David Frazier, Dr Ursula Laa and Dr Paul Harrison for 
+their feedback on this work. The screenshots and images were compiled using the 
+**cowplot** and **ggplot2** packages [@r-cowplot; @Wickham2016-gz].
+
 ## References

### 2020-08-17 paper.Rmd
inline with thesis version

@@
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-```
 
@@
 and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
-2D scatter plots were mostly sufficient for detecting class separation, however
+scatter plots were mostly sufficient for detecting class separation, however
 they also noted that often multiple embeddings were required. For the task of
@@
 of non-linear embedding techniques were more likely to consider clusters of
-points on a 2D scatter plot to be the result of a spurious embedding compared
+points on a scatter plot to be the result of a spurious embedding compared
 to advanced users who were aware of the inner workings of the embedding
@@
 the tour. A tour is a sequence of projections of a high dimensional dataset
-onto a low-dimensional basis matrix, and is represented as a
-animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic nature
-of the tour, user interaction is  important for controlling and exploring the 
-visualisation: the tour has been used previously by @Wickham2015-cx and  
-exploring statistical model fits and by @Buja1996-fk for exploring the space of 
-factorial experimental designs.
+onto a low-dimensional basis matrix, and is represented as an
+animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic 
+nature of the tour, user interaction is  important for controlling and 
+exploring the visualisation: the tour has been used previously by 
+@Wickham2015-cx for exploring statistical model fits and by @Buja1996-fk for 
+exploring the space of factorial experimental designs.
 
@@
 $X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
-Multidimensional scaling (MDS) is a class of DR methods that aims to construct
+Multidimensional scaling (MDS) is a class of DR method that aims to construct
 an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
@@
 $\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
-distribution using a Cauchy kernel:
+distribution using a t-distribution with one degree of freedom (Cauchy kernel):
 
@@
 
-Like, when using other DR techniques, the tour assumes we have a real data matrix
+Like, when, using other DR techniques, the tour assumes we have a real data matrix
 $X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
@@
 Diffusions occur when points are far from each other in the embedding view are
-near in the original data.  Whether, points are near or far is reliant on the
+near in the original data.  Whether, points are near, or far is reliant on the
 distance metric used; distortions and diffusions can be thought of as the
@@
 construct intermediary frames, means the tour is a dynamic visualisation 
-technique. Generally, the user would set $d=2$ the tour is visualised as
-a animated scatter plot. This process of constructing bases and intermediate 
+technique. Generally, the user would set $d=2$ and the tour is visualised as
+an animated scatter plot. This process of constructing bases and intermediate 
 frames and visualising the resulting projection is akin to making a "flip book" 
@@
 
-The next section steps through case studies cases of our approach using simulated
-and real data.
+The next section steps through case studies of our approach using simulations
+and an application to single cell RNA-seq data.
 
@@
 
-The videos presented show the visual 
+We *strongly* recommend viewing the linked videos for each case study
+while reading. Links to the videos are available in table \@ref(tab:vimeo) 
+and in the figures for each case study. The videos presented show the visual 
 appearance of the **liminal** interface, and how we can interact with the tour
-via the controls previously described. Both code and videos available for 
-reproducing this paper are available at https://github.com/sa-lee/paper-liminal.
+via the controls previously described. If you are unable to view the videos,
+the figures in each case study consist of screenshots that summarise what is
+learned from combining the tour and an embedding view.
 
@@
 (ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921.
-*(a)*: t-SNE (left) and tour (right) in side by side layout coloured by cluster.
-*(b)*: brushing the red cluster on the t-SNE view, shows the corresponding points on the tour.
 
-```{r gaussian, fig.cap = "(ref:liminal-gaussian)"}
+```{r gaussian, fig.cap = "(ref:liminal-gaussian)", out.height = "75%"}
 display_screenshot("gaussian", vid_lookup)
@@
 
-We see in figure \@ref(fig:gaussian) and linked video that t-SNE has 
+From the video linked inn figure \@ref(fig:gaussian), we learn that t-SNE has 
 correctly split out each cluster and laid them out in a star like formation. 
@@
 the t-SNE layout (figure \@ref(fig:gaussian)a). 
-This can be confirmed via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
+This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
 
@@
 
-We repeat the process of viewing Gaussian clusters from the 
+Next we view Gaussian clusters from the 
 _Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
 [@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
-embedded in $10-d$, and a third cluster
+embedded in 10-$d$, and a third cluster
 with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
@@
 that are each equidistant from each other and have the same covariance structure.
-Again t-SNE has correctly identified the sub-clusters but their relative 
+From the video linked in figure \@ref(fig:hierarchical), we see 
+that t-SNE has correctly identified the sub-clusters. However, their relative 
 locations to each other is distorted, with the orange and blue groups being
@@
 (ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
-*(a)* t-SNE has identified and spread out the clusters with substructure.
-*(b)* Using linked highlighting and brushing, the tour view shows that sub-clusters are much closer together in high dimensional space than the t-SNE view implies.
 
-```{r hierarchical, fig.cap="(ref:liminal-hierarchical)"}
+```{r hierarchical, fig.cap="(ref:liminal-hierarchical)", out.height = "75%"}
 display_screenshot("hierarchical", vid_lookup)
@@
 
-Next explore some simulated noisy tree structured data (figure
+Next we explore some simulated noisy tree structured data (figure
 \@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
@@
 default parameters with reference to the global view of the data set. 
+This simulation aims to mimic branching trajectories of cell differentiation:
+if there were only mature cells, we would just see the tips of the branches 
+which have a hierarchical pattern of clustering.
 
 (ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
-*(a)* The true data lies 2-d tree like structure consisting of ten branches.
+*(a)* The true data lies on 2-$d$ tree consisting of ten branches.
 This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
@@
 (ref:trees-01) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635892.
-*(a)*: Default t-SNE has difficulties embedding the tree.
-*(b)* linked highlighting shows that the backbone of the tree has been broken by the embedding.
 
-```{r trees-01, fig.cap = "(ref:trees-01)"}
+```{r trees-01, fig.cap = "(ref:trees-01)", out.height = "75%" }
 display_screenshot("trees-01", vid_lookup)
@@
 
-We see that the t-SNE view has been unable to recapitulate the
-topology of the tree - the backbone (blue) branch has been split into three
-fragments (figure \@ref(fig:trees-01)a). We can see this immediately via the linked highlighting over both
+From the linked video, we see that the t-SNE view has been unable to 
+recapitulate the topology of the tree - the backbone (blue) branch has been 
+split into three fragments (figure \@ref(fig:trees-01)a). We can see 
+this immediately via the linked highlighting over both
 plots. If we click on the legend for the zero branch, the blue coloured points
@@
 From, here it becomes apparent from the tour view that the blue branch forms
-the backbone of the tree and is connected to all other branches. We can also
-again see that cluster sizes formed via t-SNE can be misleading; from the tour
+the backbone of the tree and is connected to all other branches. From the video 
+it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
 view there is a lot of noise along the branches, while this does not appear to
@@
 initialisation and increase the perplexity from 30 to 100. We then combine
-these results with our tour view as displayed in figure \@ref(fig:trees-02).
+these results with our tour view as displayed in the linked video in the caption
+of figure \@ref(fig:trees-02).
 
@@
 
-Figure \@(ref:trees-02) shows that this selection of parameters results in the 
+The video linked in figure \@ref(fig:trees-02) shows that this selection of 
+parameters results in the 
 tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
@@
 in this area generally use non linear DR methods for verification and identification
-of clusters and developmental trajectories (i.e., case study 1). For clustering,
+of clusters and developmental trajectories (i.e., case study 1). For clustering
 workflows the primary task is verify the existence of clusters and then begin to 
@@
 linked brushing approaches to look at neighbourhood preservation
-and look at marker genes through the lens of the tour.
+and look at marker genes through the lens of the tour. The data we have
+selected for this case study has features similar to those found in case studies
+2 and 3.
 
@@
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the the `scater`
+(described by @Amezquita2020-at): we performed QC using the the **scater**
 by removing cells with high proportion of mitochondrial gene expression and low
 numbers of genes detected, we log-transformed and normalised the expression 
-values and  finally selected highly variable genes (HVGs) using `scran` 
+values and  finally selected highly variable genes (HVGs) using **scran** 
 [@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
@@
 We tour the first five PCs (approximately 20% of the variance in expression),
-along side the t-SNE view which was computed from all 25 PCs. 
-Due to latency of the interface we do weighted sample of rows based on
+alongside the t-SNE view which was computed from all 25 PCs. 
+Due to latency of the interface we do a weighted sample of  the rows based on
 cluster membership, leaving us with approximately 10 per cent of the original
 data size - 4,590 cells. Although this is not ideal, it still allows us to
-get a sense of the shape of the clusters as seen in figure \@ref(fig:mouse-01).
+get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01).
 
-(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data.
-A video of the tour animation is available at https://player.vimeo.com/video/439635812.
-*(a)*: linked brushing identifies distortions in cluster placement by the embedding.
-*(b)*: t-SNE spreads out larger clusters like cluster 2, 4, 6, and 11 which are concentrated in the tour view.
-*(c)*: the tour view gives a sense of the geometry of the data; the PCs appear to form a tetrahedron like shape.
-*(d)*: brushing on the tour view highlights diffusions, points in cluster 9 are close together in the tour view are far away in the t-SNE view.
+(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data, a video of the tour animation is available at https://player.vimeo.com/video/439635812.
 
-```{r mouse-01, fig.cap = "(ref:mouse-01)" }
+```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "85%"}
 display_screenshot("mouse-01", vid_lookup)
@@
 
-We observe that the embedding has mostly captured the
+From the video linked in figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
 clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
@@
 weighting of each gene. We run the tour directly on the 
-log normalised expression values using the same subset as before figure \@ref(fig:mouse-02).
+log normalised expression values using the same subset as before.
 
-(ref:mouse-02) Screenshots of the **liminal** tour applied to a marker gene set.
-A video of the tour animation is available at https://player.vimeo.com/video/439635843.
-*(a)*: the axis (left) and tour (right) view of 10 marker genes, using linked highlighting we see that the marker genes separate clusters 2, 4, 6, and 11 from the other groups.
-*(b)*: after zooming, it appears that the marker genes give rise to a gradient along the clusters.
+(ref:mouse-02) Screenshots of the **liminal** tour applied to a marker gene set, a video of the tour animation is available at https://player.vimeo.com/video/439635843.
 
-```{r mouse-02, fig.cap = "(ref:mouse-02)"}
+```{r mouse-02, fig.cap = "(ref:mouse-02)", out.height = "75%"}
 display_screenshot("mouse-02", vid_lookup)
@@
 
-We see that the expression of the marker genes, appear to separate
+From the video linked in figure \@ref(fig:mouse-02), we see that the 
+expression of the marker genes, appear to separate
 the previously clumped clusters 2,4,6, and 11 from the other clusters, 
-indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). Using zooming, we can see a trajectory forming along
+indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
 the clusters, while the axis view shows that magnitude of expression in 
@@
 
-Our holistic framework is also useful in real data situations, as displayed
-in our second case study. The tour when combined with t-SNE allowed us
+Our framework can also be useful in practice, as displayed
+in the fourth case study. The tour when combined with t-SNE allowed us
 to identify clusters, while giving us an idea of their orientation to 
@@
 
-In future implementations it will be possible to shift from one to one linked
-brushing to one to many linked brushing. This form of brushing acts directly on
-a $k$-nearest neighbours ($k$-nn) graph computed from a
+We have shown in the case studies, that one to one linked brushing can
+be used to identify distortions in the embedding, however we would
+like extend this to one to many linked brushing, which would allow us to directly interrogate neighbourhood preservation.
+This form of brushing acts directly on a $k$-nearest neighbours ($k$-nn) graph computed from a
 reference dataset: when a user brushes over a region in the embedding,
@@
 nearest neighbours (for example a distance matrix, or the complete data matrix) 
-can be independent of the tour or embedding views. The estimation of neighbours
-can be computed efficiently with the `FNN` or `RcppAnnoy` packages
-on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 
-could use opacity or binned colour scales to encode distances or ranks instead 
-of the neighbouring points. Brush composition, such as 'and', 'or',
-or 'not' brushes, could be used to interrogate mismatches between the 
-$k$-nn graphs estimated from both the embedding and tour views.
+can be independent of the tour or embedding views. In place of 
+highlighting, one  could use opacity or binned colour scales to encode 
+distances or ranks instead of the neighbouring points. We have begun 
+implementing this brush in **liminal**,  using the  **FNN** or **RcppAnnoy**
+packages for fast neighbourhood estimation on the server side, however there
+are still technicalities that need be resolved [@fnn-pkg; @rcpp-annoy-pkg]. 
+Brush composition, such as 'and', 'or', or 'not' brushes, could be used to 
+further investigate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
 
@@
 
-## References
+## Supplementary Materials {-}
+
+Code, data, and video for reproducing this paper are available at https://github.com/sa-lee/paper-liminal. Direct links to videos for viewing
+online are available in table \@ref(tab:vimeo).
+
+```{r vimeo}
+vid_lookup$`Case Study` <- c(1, 2, 3,3,4,4)
+knitr::kable(dplyr::select(vid_lookup, `Case Study`, 
+                           `Example` = key, `URL` = url), 
+             caption = "Case Study Videos",
+             booktabs = TRUE)
+```

### 2020-09-18 paper.Rmd
most recent version of paper

@@
 ---
-
 ```{r setup-videos, cache = FALSE}

### 2020-09-18 paper.Rmd
fix typos

@@
 
-So when the the loss function is minimised this corresponds to large attractive
+So when the loss function is minimised this corresponds to large attractive
 forces, that is, the pair-wise distances in $Y$ are small when there are
@@
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the the **scater**
+(described by @Amezquita2020-at): we performed QC using the **scater** package
 by removing cells with high proportion of mitochondrial gene expression and low
@@
 The tour appears to form a tetrahedron like shape, with points lying on
-surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
+surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
 a phenomena that has also been observed in @Korem2015-af 

### 2020-09-18 paper.Rmd
rejig output

@@
 ---
-```{r setup-videos, cache = FALSE}
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE
+)
+```
+
+```{r setup-videos, cache = FALSE, echo = FALSE}
 vid_lookup <- tibble::tribble(

### 2020-09-24 paper.Rmd
add template things -> seems to work

@@
 ---
-title: " Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings"
+documentclass: jdssv
 author:
   - name: Stuart Lee
-  - name: Di Cook
-date: "`r Sys.Date()`"
+    affiliation: Monash University
+    # use this syntax to add text on several lines
+    address: |
+      | Department of Econometrics and Business Statistics,
+      | Monash University
+    email: \email{stuart.lee1@monash.edu}
+  - name: Ursula Laa 
+    affiliation: Monash University
+    address: |
+      | Department of Econometrics and Business Statistics,
+      | Monash University
+    # To add another line, use \AND at the end of the previous one as above
+  - name: Dianne Cook
+    address: |
+      | Department of Econometrics and Business Statistics,
+      | Monash University
+    affiliation: Monash University
+title:
+  formatted: "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
+  # If you use tex in the formatted title, also supply version without
+  plain:     "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
+  # For running headers, if needed
+  short:     "\\pkg{liminal}: High-Dimensional Interactive Data Visualisation"
+abstract: >
+  There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows.
+The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of local and global structure in the resulting visualisation.
+We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
+A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
+The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view.
+By linking the tour to a view obtained from an NLDR method, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
+We display several case studies from both simulated and real data from single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
+The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.
+keywords:
+  # at least one keyword must be supplied
+  formatted: [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, "\\proglang{R}"]
+  plain:     [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, R]
 output: 
-  bookdown::html_document2:
+  bookdown::pdf_document2:
+    template: template/jdssv_template.tex
+    keep_tex: true
     toc: false
-    number_sections: no
 bibliography: liminal.bib
+biblio-style: apalike
 ---
@@
   message = FALSE,
-  warning = FALSE
+  warning = FALSE,
+  fig.align = "center"
 )
@@
 ```
-
-# Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings {#ch:tsne}
-
-There has been a rapid uptake in the use of non-linear dimensionality reduction
-(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
-the natural sciences as part of cluster orientation and dimension reduction
-workflows. The appropriate use of these methods is made difficult by their
-complex parameterisations and the multitude of decisions required to balance
-the preservation of local and global structure in the resulting visualisation.
-We present visual diagnostics for the pragmatic usage of NLDR methods
-by combining them with a technique called the tour. A tour is a sequence of
-interpolated linear projections of multivariate data onto a lower dimensional
-space. The sequence is displayed as a dynamic visualisation, allowing a user to
-see the shadows the high-dimensional data casts in a lower dimensional view. By
-linking the tour to a view obtained from an NLDR method, we can preserve global
-structure and through user interactions like linked brushing observe
-where the NLDR view may be misleading. We display several case studies from
-both simulated and real data from single cell transcriptomics, that shows our
-approach is useful for cluster orientation tasks.  The implementation of our 
-framework is available as an R package called **liminal** available at
-https://github.com/sa-lee/liminal.
-
-## Introduction
+# Introduction
 
@@
 of multidimensional scaling methods by extending their interactive visualisation
-software called **XGobi** and **GGobi** into a new tool called **GGvis** 
+software called \pkg{XGobi} and \pkg{GGobi} into a new tool called \pkg{GGvis} 
 [@Swayne1998-uq; @Swayne2003-qd; @Swayne2004-ua].
@@
 of points between the configuration and the original data.
-@Ovchinnikova2020-sy created the **Sleepwalk**  interface for checking
+@Ovchinnikova2020-sy created the \pkg{Sleepwalk}  interface for checking
 non-linear embeddings in single cell RNA-seq data. It provides a click and 
@@
 estimated pairwise distance in the original high-dimensional space.
-Similarly, the **TensorFlow**  embedding projector is a web interface to 
+Similarly, the \pkg{TensorFlow}  embedding projector is a web interface to 
 running some non-liner embedding methods live in the browser and 
@@
 background on dimension reduction methods, including an overview of the tour.
-Then we describe the visual design of **liminal**, followed by implementation
+Then we describe the visual design of \pkg{liminal}, followed by implementation
 details. Next we provide case studies that show how our interface assists
 in using embedding methods. Finally, we describe the insights gained by using
-**liminal** and plans for extensions to the software.
+\pkg{liminal} and plans for extensions to the software.
 
-## Overview of Dimension Reduction
+# Overview of Dimension Reduction
 
@@
 
-$$ p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 /
+$$ 
+p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 /
 2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 /
-2\sigma_i^2)} $$
+2\sigma_i^2)} 
+$$
 
@@
 
-### Tours explore the subspace of $d$-dimensional projections
+## Tours explore the subspace of low dimensional projections
 
@@
 a  sequence of $p \times d$ orthonormal projection matrices (bases) 
-$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
+$A_t$, where $d$ is typically 1 or 2. For each pair of 
 orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
@@
 projections as quantified by an index function [@Cook1995-bi]. While our software, 
-**liminal** is able to  visualise guided tours, our focus in the case studies 
+\pkg{liminal} is able to  visualise guided tours, our focus in the case studies 
 uses the grand tour to see global structure in the data.
 
-## Visual Design
+# Visual Design
 
@@
 can be easily incorporated into an analysts workflow with our software package, 
-**liminal**.  Our interface allows analysts to quickly compare 
+\pkg{liminal}.  Our interface allows analysts to quickly compare 
 views from embedding methods and see how an embedding method 
@@
 
-### Finding Gestalt: focus and context
+## Finding Gestalt: focus and context
 
@@
 
-### Posing Queries: multiple views, many contexts
+## Posing Queries: multiple views, many contexts
 
 We have combined the tour view in a side by side layout with a scatter plot
-view as has been done in previous tour interfaces **XGobi** and **DataViewer** 
+view as has been done in previous tour interfaces \pkg{XGobi} and \pkg{DataViewer} 
 [@Buja1986-ku; @Swayne1998-uq]. These views are
@@
 
-### Making comparisons: revising embeddings
+## Making comparisons: revising embeddings
 
@@
 
-## Software Infrastructure
+# Software Infrastructure
 
-We have implemented the above design as an open source R package called
-**liminal** [@r-liminal]. The package allows analysts to construct concatenated
-visualisations, drawn with the **Vega-Lite** grammar of interactive graphics via
-the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides 
+We have implemented the above design as an open source \proglang{R} package called
+\pkg{liminal} [@r-liminal]. The package allows analysts to construct concatenated
+visualisations, drawn with the \pkg{Vega-Lite} grammar of interactive graphics via
+the \pkg{vegawidget} package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides 
 an interface for constructing linked and stand alone interfaces for 
-manipulating tour paths via  the **shiny** and **tourr** packages 
+manipulating tour paths via  the \pkg{shiny} and \pkg{tourr} packages 
 [@Chang2020-bq; @Wickham2011-st].
 
-### Tours as a streaming data problem
+## Tours as a streaming data problem
 
@@
 animation. Like with a flip book, an interface to the tour requires the ability
-to interact and modify it in real time. The user interface generated in **liminal**
+to interact and modify it in real time. The user interface generated in \pkg{liminal}
 allows a user to play, pause, and reset the tour animation,
@@
 
-### Linking and highlighting views via interactions
+## Linking and highlighting views via interactions
 
@@
 guides on the legend can be selected in order to see their relative orientations.
-The interface is constructed as a **shiny** gadget specifically designed for
+The interface is constructed as a \pkg{shiny} gadget specifically designed for
 interactive data analysis. Selections such as brushing regions and the current
@@
 
-## Case Studies
+# Case Studies
 
@@
 and in the figures for each case study. The videos presented show the visual 
-appearance of the **liminal** interface, and how we can interact with the tour
+appearance of the \pkg{liminal} interface, and how we can interact with the tour
 via the controls previously described. If you are unable to view the videos,
@@
 
-### Case Study 1: Exploring spherical Gaussian clusters
+## Case Study 1: Exploring spherical Gaussian clusters
 
@@
 covariance matrix. We then computed a t-SNE layout with default settings 
-using the **Rtsne** package [@Rtsne],
-and set up the **liminal** linked interface with grand tour on the 10-$d$
+using the \pkg{Rtsne} package [@Rtsne],
+and set up the \pkg{liminal} linked interface with grand tour on the 10-$d$
 observations.
 
-(ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921.
+(ref:liminal-gaussian) Screenshots of the \pkg{liminal} interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921.
 
@@
 
-### Case Study 2: Exploring spherical Gaussian clusters with hierarchical structure
+## Case Study 2: Exploring spherical Gaussian clusters with hierarchical structure
 
@@
 
-(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
+(ref:liminal-hierarchical) Screenshots of the \pkg{liminal} interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
 
@@
 
-### Case Study 3: Exploring data with piecewise linear structure
+## Case Study 3: Exploring data with piecewise linear structure
 
@@
 *(a)* The true data lies on 2-$d$ tree consisting of ten branches.
-This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
+This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
 *(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view.
@@
 initialisation). We then create a linked
-tour with t-SNE layout with **liminal** as shown in figure \@ref(fig:trees-01).
+tour with t-SNE layout with \pkg{liminal} as shown in figure \@ref(fig:trees-01).
 
-(ref:trees-01) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635892.
+(ref:trees-01) Screenshots of the \pkg{liminal} interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635892.
 
@@
 
-(ref:trees-02) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635863.
+(ref:trees-02) Screenshots of the \pkg{liminal} interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635863.
 
@@
 
-### Case Study 4: Clustering single cell RNA-seq data
+## Case Study 4: Clustering single cell RNA-seq data
 
@@
 First, we downloaded the raw mouse retinal single cell RNA-seq data from
-@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
+@Macosko2015-ot using the \pkg{scRNAseq} Bioconductor package [@scRNAseq-d]. We
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the **scater** package
+(described by @Amezquita2020-at): we performed QC using the \pkg{scater} package
 by removing cells with high proportion of mitochondrial gene expression and low
 numbers of genes detected, we log-transformed and normalised the expression 
-values and  finally selected highly variable genes (HVGs) using **scran** 
+values and  finally selected highly variable genes (HVGs) using \pkg{scran} 
 [@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
@@
 
-To check and verify the clustering we construct a **liminal** view. 
+To check and verify the clustering we construct a \pkg{liminal} view. 
 We tour the first five PCs (approximately 20% of the variance in expression),
@@
 
-(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data, a video of the tour animation is available at https://player.vimeo.com/video/439635812.
+(ref:mouse-01) Screenshots of the \pkg{liminal} interface applied to single cell data, a video of the tour animation is available at https://player.vimeo.com/video/439635812.
 
@@
 
-(ref:mouse-02) Screenshots of the **liminal** tour applied to a marker gene set, a video of the tour animation is available at https://player.vimeo.com/video/439635843.
+(ref:mouse-02) Screenshots of the \pkg{liminal} tour applied to a marker gene set, a video of the tour animation is available at https://player.vimeo.com/video/439635843.
 
@@
 high dimensional data provides an additional insight for interrogating views
-generated from embeddings. The interface we have designed in the **liminal**
+generated from embeddings. The interface we have designed in the \pkg{liminal}
 package, allows a user to gain a deeper understanding of an embedding 
@@
 distances or ranks instead of the neighbouring points. We have begun 
-implementing this brush in **liminal**,  using the  **FNN** or **RcppAnnoy**
+implementing this brush in \pkg{liminal},  using the  \pkg{FNN} or \pkg{RcppAnnoy}
 packages for fast neighbourhood estimation on the server side, however there
@@
 There are some limitations
-in using the **liminal** interface for larger datasets. First,
+in using the \pkg{liminal} interface for larger datasets. First,
 t-SNE avoids the crowding problem, points are separated into distinct regions
@@
 
-## Acknowledgements {-}
+# Acknowledgements {-}
 
-We would like to thank Dr David Frazier, Dr Ursula Laa and Dr Paul Harrison for 
+We would like to thank Dr David Frazier and Dr Paul Harrison for 
 their feedback on this work. The screenshots and images were compiled using the 
-**cowplot** and **ggplot2** packages [@r-cowplot; @Wickham2016-gz].
+\pkg{cowplot} and \pkg{ggplot2} packages [@r-cowplot; @Wickham2016-gz].
 
-## Supplementary Materials {-}
+# Supplementary Materials {-}
 

### 2020-09-24 template/JDSSV_editorialprocess.tex
add template things -> seems to work

\documentclass[article]{jdssv}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{thumbpdf,lmodern}
\usepackage{bm}
\usepackage{todonotes} %% Remove this package when finishing the manuscript

%% another package (only for this demo article)
\usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\newcommand{\ma}[1]{\ensuremath{\mathbf{#1}}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Patrick J.F. Groenen\\Erasmus University Rotterdam
   \And Stefan Van Aelst\\KU Leuven}
\Plainauthor{Patrick J.F. Groenen, Stefan Van Aelst}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Guidelines for the Journal of Data Science, Statistics, and Visualisation}
\Plaintitle{Guidelines for the Journal of Data Science, Statistics, and Visualisation Version 0.2}
\Shorttitle{JDSSV Guidelines, Version 0.2}

%% - \Abstract{} almost as usual
\Abstract{
  This short article illustrates how to write a manuscript for the   \emph{Journal of Data Science, Statistics and Visualisation} (JDSSV) using its {\LaTeX} style files.   Please follow JDSSV's style guidelines precisely. Also, it is recommended to keep the {\LaTeX} code as simple as possible, that is, avoid inclusion of packages/commands that are not necessary. 
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{JDSSV, style guidelines, comma-separated, not capitalized, \proglang{R}}
\Plainkeywords{JDSSV, style guidelines, comma-separated, not capitalized, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Patrick J.F. Groenen\\
  Econometric Institute\\
  Erasmus School of Economics\\
  Erasmus University Rotterdam\\
  P.O. Box  1738 \\
  3000 DR Rotterdam, The Netherlands\\
  E-mail: \email{groenen@ese.eur.nl}\\
  URL: \url{https://personal.eur.nl/groenen/}\\
\ \\
%}
%
%\Address{
  Stefan Van Aelst\\
  Department of Mathematics\\
  KU Leuven\\
  Celestijnenlaan 200B\\
  3001 Leuven, Belgium\\
  E-mail: \email{Stefan.VanAelst@kuleuven.be}\\
  URL: \url{https://wis.kuleuven.be/stat/robust}
}

\begin{document}

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section{Mission}

JDSSV\footnote{This document is an adaptation of the style guide of the Journal of Statistical Software \citep{jss_style_guide}.} is an international refereed journal that creates a forum to present recent progress and ideas in the different disciplines of data science, statistics, and visualisation. It welcomes contributions to data science, statistics, and visualisation, and in particular those aspects which link and integrate these subject areas. Articles can cover topics such as machine learning and statistical learning, the visualisation and verbalisation of data, visual analytics, big data infrastructures and analytics, interactive learning, and advanced computing. Papers that discuss two or more research areas of the journal are favoured. Scientific contributions should be of a high standard. Articles should be oriented towards a wide scientific audience of statisticians, data scientists, computer scientists, data analysts, etc.

The journal welcomes original contributions that are not being considered for publication elsewhere and contain a high level of novelty. Papers with a thorough but concise review of a certain topic with the potential to provide new insights are also welcome. Manuscripts submitted to the journal generally are accompanied by supplementary material that containing software code, technical derivations or detailed explanations, additional examples, etc. All submitted material will be reviewed by the assigned associate editor and reviewers of the manuscript.

Manuscripts may have a substantial theoretical component, but it is expected that all manuscripts contain at least one application on empirical or simulated data. The journal emphasizes the reproducibility of the results presented its papers. Therefore, all data and software code that is necessary to reproduce the empirical results in the manuscript should be made available in a user friendly manner. If the empirical data cannot be released for reasons of confidentiality or otherwise, then a generated dataset with comparable properties should be provided.

\section[]{Preparing your Manuscript for Submission}

All submissions to JDSSV should be written in {\LaTeX} using the JDSSV style files provided on the journal website \url{https://jdssv.org}. For initial submission it suffices providing the pdf version of the manuscript together with all the necessary supplementary material including software code to reproduce all results in the manuscript. The final version of accepted manuscripts should adhere to all the style guidelines and should incorporate all changes requested by the production editor. All {\LaTeX}  source files of the final manuscript should be submitted and accepted manuscripts are only published if these files comply with all guidelines and instructions provided by the journal.

\section[]{Software}

The journal expects that submissions contain accompanying
software with the aim of reproducibility of the results and application of the proposed methodology to other data by the reader.  All existing software used in the paper should be properly referenced. Code should be delivered in an easily readable manner with clear instructions on its use, and preferably is accompanied by instructive examples of its use. We highly recommend to provide code that can be used in open-source software such as \proglang{R} \citep{R}, \proglang{Python} \citep{python}, \proglang{Julia} \citep{Julia}, \proglang{Octave} \citep{octave}, etc.

To make code widely accessible, we advise making it available in a repository such as \url{https://zenodo.org} where it will receive a permanent Digital Object Identifier (DOI) which can be included in the manuscript. 
%\todo{\tiny Alternatives: (1) We store the data/code on our JDSSV website with the article. (2) We upload it on Zenodo ourselves (not sure if that can be done).}

The provided code should at least consist of a file or set of files for the functions that execute the core of the method. For reproducibility, another necessary file is the script that creates the results (tables and figures) of the paper. For methods that run very long, provide a toy example that runs sufficiently fast and highlights the properties of the method. Nonproprietary data should be provided or a permanent link to these data should be given. The corresponding script to analyze the data should read these data.

\begin{sloppypar}
For increased readability, please apply the following naming conventions for code: Start function names with a verb, e.g., \code{set_initialisation()}, \code{compute_loss()}, \code{update_X()}, etc. when appropriate. Give  objects descriptive names or closely follow the notation in the paper.
For programming conventions (particularly in \proglang{R}), see Hadley Wickham's guidelines (\url{http://r-pkgs.had.co.nz/style.html}) and the use of the \pkg{styler} package is recommended. For \proglang{python}, consider Google's recommendations (\url{https://github.com/google/styleguide/blob/gh-pages/pyguide.md}) Sufficient comments should be added to the code to make it understandable. For \proglang{Octave} and \proglang{MatLab} \citep{matlab}, consider Richard Johnson's MatLab Style Guidelines 2.0 (\url{https://www.mathworks.com/matlabcentral/fileexchange/46056-matlab-style-guidelines-2-0}).
\end{sloppypar}

For writing about software JDSSV requires authors to use the markup
\verb|\proglang{}| (programming languages and large programmable systems), \verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands, arguments, etc.). If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the {\LaTeX} command as well. Below we also illustrate how abbrevations should be introduced and citation commands can be employed. See the {\LaTeX} code for more details.

\section{Review process}

All manuscripts should be submitted online at  \url{https://jdssv.org}. JDSSV uses a single blind review process. Upon submission of their manuscript, authors have the opportunity to provide a short list of suggested reviewers as well as a few names of researchers that preferably should not be contacted for reviewing the manuscript.  All submitted manuscript will undergo automatic checking for plagiarism and will not be considered for further review in case of plagiarism. The manuscript will be assigned to one of the editors who will make an initial screening to check the quality of the submitted work, possibly with the help of an associate editor. After positive screening, the manuscript will be assigned to an associate editor who will seek the opinion of at least two reviewers. Reviewers are asked to send their report within one month. Associate editors should typically make their recommendation one week after reception of the review reports. The initial review process should normally not take more than three months. All communication between the authors and the journal will be administered by the journal editors.

%% -- Manuscript ---------------------------------------------------------------

%% - In principle "as usual" again.
%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
%%   avoid empty lines before and after the equation (which would signal a new
%%   paragraph.
%% - When describing longer chunks of code that are _not_ meant for execution
%%   (e.g., a function synopsis or list of arguments), the environment {Code}
%%   is recommended. Alternatively, a plain {verbatim} can also be used.
%%   (For executed code see the next section.)

\section{Correspondence Analysis} \label{sec:models}

As an example, we provide a simple implementation of correspondence analysis \citep[see, e.g.,][]{greenacre2010biplots}. For the analysis of bivariate categorical data or other tables that contain counts, correspondence analysis can be a useful technique to visualize important relations between the categories. Consider word count data on explanations of data science by several websources provided by \citet{Lubbe2018}, see Table~\ref{tab:word_counts}.

\begin{table}
\caption{Part of the table with word counts describing data science by different web sources as collected by \citet{Lubbe2018}.}
\label{tab:word_counts}
\begin{center}
\begin{tabular}{lccccccc} \hline
            & dsc.test & dsctechniques & datadiversity & dsc.tips & $\ldots$ & wikipedia \\ \hline
statistics  &       6  &           2   &          2    &    0     & $\ldots$ &  2 \\
big data    &       2  &           0   &          3    &    5     & $\ldots$ &  0 \\
analytics   &       1  &           0   &          2    &    0     & $\ldots$ &  2 \\
database    &       2  &           0   &          0    &    2     & $\ldots$ &  1 \\
insight     &       0  &           0   &          4    &    0     & $\ldots$ &  2 \\
$\vdots$    & $\vdots$ &    $\vdots$   &   $\vdots$    & $\vdots$ &$ \ddots$ & $\vdots$ \\
programming &       7  &           0   &          0    &    0     & $\ldots$ &  0 \\
\hline
\end{tabular}
\end{center}
\end{table}

Let $\ma{F}$ be the matrix containing the values in Table~\ref{tab:word_counts}. The most simple model to fit such a table is the independence model in matrix $\ma{E}$ defined as 
\begin{eqnarray*}
  \ma{E} = n^{-1}\ma{D}_r \ma{11}^\top \ma{D}_c
\end{eqnarray*}
with $\ma{1}$ a vector of ones of appropriate length, $\ma{D}_r = \textrm{Diag}(\ma{F1})$ the diagonal matrix with row sums of $\ma{F}$, $\ma{D}_c = \textrm{Diag}(\ma{F^\top1})$ the diagonal matrix with column sums of $\ma{F}$, and $n = \ma{1^\top F1}$ be the sum of all elements in $\ma{F}$. The goal of correspondence analysis is to find a matrices of row and column coordinates $\ma{R}$ and $\ma{C}$ such that
\begin{eqnarray*}
  L(\ma{R}, \ma{C}) = \|\ma{D}_r^{-1/2} (\ma{F} - \ma{E} - \ma{D}_r\ma{RC}\ma{D}_c^\top) \ma{D}_c^{-1/2}\|^2
\end{eqnarray*}
is minimized \citep[see, for example,][]{vandeveldenetal2009seriation}. It may be verified that the least-squares optimal solution is obtained for by computing the singular value decomposition (SVD) 
\begin{eqnarray*}
  \ma{D}_r^{-1/2} (\ma{F} - \ma{E})\ma{D}_c^{-1/2} = \ma{UDV}^\top
\end{eqnarray*}
and computing
\begin{eqnarray*}
  \ma{R} &=& n^{1/2}\ma{D}_r^{-1/2} \ma{UD}^{\alpha}\\
  \ma{C} &=& n^{1/2}\ma{D}_c^{-1/2} \ma{VD}^{1-\alpha}
\end{eqnarray*}
for some $\alpha$. Note that the least-squares optimal rank $p$ solution is obtained by taking the first $p$ columns of $\ma{R}$ and $\ma{C}$. If $\alpha = 1$ the so-called row principal solution is obtained where the row points are the weighted centroids of the column points, $\alpha = 0$ gives the column principal solution with column points being the weighted average of the row points, and $\alpha = 1/2$ results in the symmetric solution. The importance measure of the dimensions are given by the inertia, that is, the diagonal elements of $\ma{D}^2$. For more details, see, for example, \cite{greenacre2010biplots}.

A simple implementation of correspondence analysis in \proglang{R} is given by the function \code{corana()} given by
\begin{Code}
corana <- function(dat, alpha = 0.5){
  # Perform correspondence analysis
  # Input: 
  ##  dat   numeric matrix dat with nonnegative entries
  ##  alpha = 1 is row principal standardisation
  ##        = 0 is column principle standardisation
  ##        = 0.5 is symmetric standardisation
  dat <- as.matrix(dat)
  Dr <- rowSums(dat)
  Dc <- colSums(dat)
  n  <- sum(Dr)
  # Compute expected values under the independence model
  E  <- outer(Dr, Dc)/n
  tt <- svd(diag(Dr^-0.5) %*% (dat - E) %*% diag(Dc^-0.5))
  ## Remove dimensions with singular value zero
  ind <- tt$d > 1e-10
  tt$d <- tt$d[ind]
  tt$u <- tt$u[, ind]
  tt$v <- tt$v[, ind]
  ## Compute row scores R and column scores C 
  R  <- n^0.5 * diag(Dr^-0.5) %*% tt$u %*% diag(tt$d^alpha)
  C  <- n^0.5 * diag(Dc^-0.5) %*% tt$v %*% diag(tt$d^(1 - alpha))
  rownames(R) <- rownames(dat)
  rownames(C) <- colnames(dat)
  ## Compute relative contribution to inertia per dimension 
  row.inert   <- outer(Dr/n, tt$d^(2*alpha),       "/") * R^2
  col.inert   <- outer(Dc/n, tt$d^(2*(1 - alpha)), "/") * C^2
  ## Compute reconstructed Chi-square distance for row and column points
  row.dist    <- R^2 %*% diag(tt$d^(2 - 2*alpha))
  row.dist    <- row.dist / outer(rowSums(row.dist), rep(1, ncol(R))) 
  col.dist    <- C^2 %*% diag(tt$d^(2 - 2*(1 - alpha)))
  col.dist    <- col.dist / outer(rowSums(col.dist), rep(1, ncol(C))) 
  ## Prepare list of results
  out <- list(R = R, C = C, sing.val = tt$d, 
              row.inert = row.inert, col.inert = col.inert,
              row.dist  = row.dist,  col.dist  = col.dist)
  class(out) <- "corana"
  return(out)
}\end{Code}
A plot is given by the \code{plot.corana()} method
\begin{Code}
plot.corana <- function(out, dims = 1:2, ...){
  R <- out$R[, dims]
  C <- out$C[, dims]
  ## Set up coordinate system
  coord <- rbind(R, C)
  plot(coord[, 1], coord[, 2], type = "n", asp = 1, las = 1,
       xlab = paste0("Dim ", dims[1]), ylab = paste0("Dim ", dims[2]), ...)
  abline(h = 0, v = 0, col = "gray")
  ## Use reconstructed distance as importance measure for transparency and size
  row.alpha <- rowSums(out$row.dist[, dims])
  row.alpha <- (row.alpha/max(row.alpha))^0.7
  col.alpha <- rowSums(out$col.dist[, dims])
  col.alpha <- (col.alpha/max(col.alpha))^0.7
  ## Plot row and column points 
  points(R[, 1], R[, 2], pch = 20, col = rgb(1, 0, 0, row.alpha))
  points(C[, 1], C[, 2], pch = 20, col = rgb(0, 0, 1, col.alpha))
  ## Write text labels
  text(R[, 1], R[, 2], rownames(R), pos = compute.pos(R), cex = 2*row.alpha,
       col = rgb(1, 0, 0, ifelse(row.alpha > 0.1, row.alpha, 0)))
  text(C[, 1], C[, 2], rownames(C), pos = compute.pos(R), cex = 2*col.alpha,
       col = rgb(0, 0, 1, ifelse(col.alpha > 0.1, col.alpha, 0)))
}
\end{Code}
Assuming that \code{ds.word.cnt} contains the full matrix of which a part is shown in Table~\ref*{tab:word_counts}, then the following code does a correspondence analysis on these data:
\begin{Code}
out.corana <- corana(ds.word.cnt)
plot(out.corana, dims = 2:3)
\end{Code}
that yields the plot of Dimensions 2 and 3 in Figure~\ref{fig:ca_biplot_dims23}. In this plot, the points and labels are made more transparent and the labels decrease in size as they are represented worse in these two dimensions.

\begin{figure}
\begin{center}
\includegraphics[width=10cm]{ca_example.pdf}
\end{center}
\caption{Biplot of a correspondence analysis on word counts used in descriptions of data science by different web sources in dimensions 2 and 3.} \label{fig:ca_biplot_dims23}
\end{figure}

%% -- Summary/conclusions/discussion -------------------------------------------

%\section{Summary and discussion} \label{sec:summary}
%
%\begin{leftbar}
%As usual \dots
%\end{leftbar}

%% -- Optional special unnumbered sections -------------------------------------
\newpage
\section*{Computational Details}

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, \dots) that are not cited in the main text can be credited here.

The results in this paper were obtained using
\proglang{R}~3.5.1. \proglang{R} itself
and all packages used are available from the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/}.

\section*{Acknowledgments}

All acknowledgments should be collected in this
unnumbered section before the references. It may contain the usual information
about funding and feedback from colleagues/reviewers/etc. Furthermore,
information such as relative contributions of the authors may be added here
(if any).

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\bibliography{refs}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\newpage

\begin{appendix}

\section{More Technical Details} \label{app:technical}

Appendices can be included after the bibliography (with a page break). Each section within the appendix should have a proper section title (rather than just \emph{Appendix}).

%For more technical style details, please check out JSS's style FAQ at
%\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
%which includes the following topics:
%\begin{itemize}
%  \item All main words in titles and headers start with a capital.
%  \item Use vectorized graphics formats such as eps and pdf when possible. Graphics should be of high quality while trying to minimize the file size.
%\end{itemize}

\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}

References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|, \verb|\citealp| etc.\ (and never hard-coded). This commands yield different formats of author-year citations and allow to include additional details (e.g., pages, chapters, \dots) in brackets. 
%In case you are not familiar with these commands see the JSS style FAQ for details.

Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially when acquiring the entries automatically from mixed online sources. However, it is important that informations are complete and presented in a consistent style to avoid confusions. JDSSV requires the following format.
\begin{itemize}
  \item Specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
    be used in the references.
  \item Titles should be inserted in title case.
  \item Journal titles should not be abbreviated and in title case.
  \item DOIs should be included where available.
  \item Software should be properly cited as well. For \proglang{R} packages
    \code{citation("pkgname")} typically provides a good starting point.
\end{itemize}
\end{appendix}
\newpage
%% -----------------------------------------------------------------------------

\end{document}

### 2020-09-24 README.Rmd
update repo and dependencies

@@
   collapse = TRUE,
-  comment = "#>"
+  comment = "R>"
 )
@@
 
-`liminal` aims to increase the effective use of embeddings by combining
-them with interactive and dynamic graphics (via the vegawidget and shiny
-packages) and with a technique from multivariate stastics called the tour (via
-the tourr package).  Briefly, a tour is a sequence of interploated projections
-of multivariate data onto lower dimensional space. The sequence is displayed as
-a dynamic visualisation, and enables us to see the shadows the high dimensional
-data makes in a lower dimensional view. By combining the tour with embedding
-algorithms, we can see the following:
+> There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows.
+The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of local and global structure in the resulting visualisation.
+We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
+A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
+The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view.
+By linking the tour to a view obtained from an NLDR method, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
+We display several case studies from both simulated and real data from single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
+The implementation of our framework is available as an R package called **liminal** available at https://github.com/sa-lee/liminal.
 
-1. whether distances in the embedding view are meaningful 
-2. the local and global structure of the data 
-3. identify 'interesting' shapes or points in the data
+## Reproducibility
 
-## Outline
+Package dependencies for the case studies can be installed with the **BiocManager** package
 
-- review of embedding algorithms, interactive graphics, tours
-- technology
-- design
-- examples/benchmarks
-  - Kevin moon
-  - pdfsense
-  - single cell
-  - quick draw
+```{r, eval = FALSE}
+# install.packages("BiocManager")
+BiocManager::install("sa-lee/paper-liminal")
+```
+
+## Repo Structure
+
+The case studies that produce the videos and still images found in the paper
+can be reproduced in R via running the files found in the `scripts` directory.
+
+```{r, echo = FALSE}
+fs::dir_tree(recurse = TRUE,  
+             type = c("file", "directory"),
+             glob = c("*.png|*.pdf|*.cls|*.bib|*.html"),
+             invert = TRUE)
+```

### 2020-10-01 paper.Rmd
Comments added to introduction.

@@
 ground truth.
+\textcolor{red}{Ursula: maybe work out the "dimension reduction" examples in a
+bit more detail? They are difficult to understand for someone who does not know
+about them already (or at least for me), and it wasn't very clear why clustering
+is not part of that (I guess because we want to actually tag clusters for further
+analysis down the line?).}
 
@@
 algorithm.
+
+\textcolor{red}{Ursula: suggestion for the structure of the following paragraphs:
+start with "no one-size fits all", then talk about interactive interfaces to assess
+embeddings, and introduce tours when talking about existing interfaces, and finish by
+talking about liminal approach.}
 

### 2020-10-02 paper.Rmd
Some restructuring of Section 2.

@@
 
+<!--
+Ursula: I don't think this paragraph is needed.
+
 In the following section, we provide a brief overview of the goals of DR
@@
 mention several other techniques [@Maaten2008-sk; @McInnes2018-co]. 
+-->
 
-To begin we suppose the data is in the form rectangular
+To begin we suppose the data is in the form of a rectangular
 matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
@@
 methods. Linear methods perform a linear transformation of the data, that is,
-$Y$ is a linear transformation of $X$; one example is principal components
+$Y$ is a linear transformation of $X$. One example is principal components
 analysis (PCA) which performs an eigendecomposition of the estimated sample
@@
 cut-off.
+When working with linear transformations we often need more than two dimensions
+to capture the latent structure. In this case we can use tour methods
+[@Asimov1985-vp; @Buja1986-zr] to show interpolated sequences of projections,
+providing the viewer with intuition about structure in more than two
+dimensions, as described below.
 
@@
 estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$.
+
+<!--
+Ursula: too much detail?
+
 The algorithm de-noises estimated distances in high-dimensional space via 
@@
 probabilities are  transformed once more to construct a distance matrix, 
-and classical MDS is  used to generate $Y$. A general difficulty with using 
+and classical MDS is  used to generate $Y$.
+-->
+
+A general difficulty with using 
 non-linear DR methods for exploratory data analysis is selecting and tuning 
-appropriate parameters. To see the extent of these choices 
-we now examine the underpinnings of t-SNE and UMAP.
+appropriate parameters. For concretenes here we focus on t-distributed
+stochastic neighbour embedding (t-SNE), and we will examine
+its underpinning in some detail below [@Maaten2008-sk]. Similar considerations hold for related
+methods, for example UMAP [@McInnes2018-co].
 
-The t-SNE algorithm estimates
-the pair-wise similarity of (Euclidean) distances of points in a high
-dimensional space using a Gaussian distribution and then estimates a
-configuration in the low dimensional embedding space by modelling similarities
-using a t-distribution with 1 degree of freedom [@Maaten2008-sk]. There are several subtleties
+## Introduction to t-SNE
+
+The t-SNE algorithm estimates the pair-wise similarity of points in a high
+dimensional space based on their (Euclidean) distances using a Gaussian
+distribution. A configuration in the low dimensional embedding space is then
+estimated by modelling similarities using a t-distribution with 1 degree of
+freedom [@Maaten2008-sk]. There are several subtleties
 to the to use of the algorithm that are revealed by stepping through its
@@
 $$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
+\textcolor{red}{Ursula: is there an intuitive explanation why we want to tune
+perplexity rather than working with sigma directly?}
 As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded above
 by the number of observations , $n-1$, in the data, corresponding to
-$\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a
+$\sigma^2_{i} \rightarrow \infty$. This essentially turns t-SNE into a
 nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all
@@
 
-Next, the target low-dimensional space, $Y$, pair-wise distances between
+Next, in the target low-dimensional space, $Y$, pair-wise distances between
 $\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
@@
 
-So when the loss function is minimised this corresponds to large attractive
+Minimising  the loss function corresponds to large attractive
 forces, that is, the pair-wise distances in $Y$ are small when there are
 non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
-should also be small when the loss function is minimised, that is, when
-pair-wise distances in $Y$ are large regardless of the magnitude of the
-corresponding distances in $X$.
+should also be small, that is, overall the
+pair-wise distances in $Y$ should be large regardless of the magnitude of the
+corresponding distances in $X$. As a result, clusters that are separate in $X$
+will be placed far from each other in $Y$.
+This minimisation is done via stochastic gradient decent,
+and introduces a number of hyperparameters, for example the number of iterations,
+the learning rate, and early exaggeration, a multiplier of the attractive force
+used at the beginning of the optimisation.
 
 Taken together, these details reveal the sheer number of decisions that an
-analyst must make. How does one choose the perplexity? How should the
-parameters that control the optimisation of the loss function (done with
-stochastic gradient descent), like the number of iterations, or early
-exaggeration ( a multiplier of the attractive force at the beginning of the
-optimisation), or the learning rate be selected?  It is a known problem that
+analyst must make. How does one choose the perplexity? And the
+parameters that control the optimisation of the loss function?
+It is a known problem that
 t-SNE can have trouble recovering topology and that configurations can be
@@
 approaches for tuning the optimisation parameters. However, the cluster sizes
-and their relative orientation from a $t-SNE$ view can be misleading
+and their relative orientation from a t-SNE view can be misleading
 perceptually, due to the algorithms emphasis on locality.
@@
 The tour is a visualisation technique that is grounded in 
-mathematical theory, and is able to ascertain 
+mathematical theory, and allows the viewer to ascertain 
 the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections [@Asimov1985-vp; @Buja1986-zr].
 
-Like, when, using other DR techniques, the tour assumes we have a real data matrix
+As with other DR techniques, the tour assumes we have a real data matrix
 $X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
@@
 orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
-path between them is interpolated to form intermediate frames, and giving the 
+path between them is interpolated to form intermediate frames, giving the 
 sense of continuous movement from one basis to another. The tour is then the 
-continuous visualisation of the projections $Y_{t} = XA_{t}$, that is the 
+continuous visualisation of the projected data $Y_{t} = XA_{t}$, that is the 
 projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
-successive bases. A *grand tour* corresponds to choosing new orthonormal
+successive bases.
+
+A *grand tour* corresponds to choosing new orthonormal
 bases at random; allowing a user to ascertain structure via exploring the

### 2020-10-02 paper.Rmd
Suggestions for Section 3.

@@
 
+\textcolor{red}{Ursula: here you are following the ordering from the cited
+paper, but maybe it makes sense to discuss making comparisons before
+talking about posing queries?}
+
+<!--
+Ursula: I think for these subsections we need to have a first paragraph
+that specifies what the concept means and why it is important here,
+before describing how it is incorporated in liminal
+-->
+
 ## Finding Gestalt: focus and context
 
-To investigate latent structure and the shape of a high dimensional dataset, a
-tour can be run without the use of an external embedding. It is often useful to
+To understand the data structure, we look for Gestalt features such as
+(non-)linearities, clusters or outliers. A tour display is
+preferred for this task, since it accurately captures the geometry, while NLDR
+methods typically introduce distortions.
+
+To investigate latent structure and the shape of a high dimensional dataset in
+\pkg{liminal}, a tour can be run without the use of an external embedding.
+It is often useful to
 first run principal components on the input as an initial dimension reduction
@@
 layout displaying the magnitude and direction of each basis vector. Since the
-tour is dynamic, it is often useful to be able to pause and highlight a
+tour is dynamic, it is useful to be able to pause and highlight a
 particular view. In addition to pause, play and reset buttons, brushing will 
-pause the tour path, allowing users to identify 'interesting' projections. The
+pause the tour path, allowing users to identify 'interesting' projections. 
+\textcolor{red}{Interesting projections or interesting points?}
+The
 domain of the axis scales from running a tour is called the half range, and is 
-computed by rescaling the input data onto hyper-dimensional unit cube. We bind 
+computed by rescaling the input data onto hyper-dimensional
+\textcolor{red}{d-dimensional?}
+unit cube. We bind 
 the half range to a mouse wheel event, allowing a user to pan and zoom on the 
@@
 ## Posing Queries: multiple views, many contexts
+
+The initial visualisation gives an overview of the data structure,
+and naturally leads to queries that investigate observed features with
+the aim to further characterise them. This is an essential aspect of
+our framework, where we use a tour to better characterise
+features observed in a NLDR display.
 
@@
 
+Combining multiple views of a single data set allows the analyst
+to make meaniningful comparisons. In \pkg{liminal} the embedding
+and tour views are arranged side-by-side for direct cross-checks.
+
 As mentioned previously, when using any DR method, we are assuming the
 embedding is representative of the high-dimensional dataset it was computed
-from. Defining what it means for embedding to be 'representative` or 'faithful'
+from. Defining what it means for embeddings to be 'representative` or 'faithful'
 to high-dimensional data is ill-posed and depends on the underlying task an

### 2020-11-26 paper.Rmd
Comment added in Sec 4
resectioning and spelling

@@
 non-linear DR methods for exploratory data analysis is selecting and tuning 
-appropriate parameters. For concretenes here we focus on t-distributed
+appropriate parameters. For concreteness here we focus on t-distributed
 stochastic neighbour embedding (t-SNE), and we will examine
@@
 the three principles for interactive high-dimensional data visualisation outlined
-by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.
+by @Buja1996-fk:  finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.
 
@@
 Combining multiple views of a single data set allows the analyst
-to make meaniningful comparisons. In \pkg{liminal} the embedding
+to make meaningful comparisons. In \pkg{liminal} the embedding
 and tour views are arranged side-by-side for direct cross-checks.
@@
 
+\textcolor{red}{Ursula: can I use restart to replay the same tour path (i.e. find a
+previously seen projection), or will it start a new random selection of bases?}
+
 ## Linking and highlighting views via interactions
@@
 
-From the video linked inn figure \@ref(fig:gaussian), we learn that t-SNE has 
+From the video linked in Figure \@ref(fig:gaussian), we learn that t-SNE has 
 correctly split out each cluster and laid them out in a star like formation. 
@@
 generally points are more concentrated in the projection view compared to 
-the t-SNE layout (figure \@ref(fig:gaussian)a). 
-This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
+the t-SNE layout (Figure \@ref(fig:gaussian)a). 
+This can be seen via brushing the t-SNE view (Figure \@ref(fig:gaussian)b).
 
@@
 that are each equidistant from each other and have the same covariance structure.
-From the video linked in figure \@ref(fig:hierarchical), we see 
+From the video linked in Figure \@ref(fig:hierarchical), we see 
 that t-SNE has correctly identified the sub-clusters. However, their relative 
 locations to each other is distorted, with the orange and blue groups being
-far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
+far from each other in the tour view (Figure \@ref(fig:hierarchical)a). We see 
 in this case that is difficult
 to see the sub-clusters in the tour view, however, once we zoom and highlight
-they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
+they become more apparent (Figure \@ref(fig:hierarchical)b). When we brush 
 the sub-clusters in the t-SNE, their
@@
 
-(ref:liminal-hierarchical) Screenshots of the \pkg{liminal} interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
+(ref:liminal-hierarchical) Screenshots of the \pkg{liminal} interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635905.
 
@@
 
-Next we explore some simulated noisy tree structured data (figure
-\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
+Next we explore some simulated noisy tree structured data (Figure
+\@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
 topology of the data, and then seeing if we can resolve this by tweaking the
@@
 *(a)* The true data lies on 2-$d$ tree consisting of ten branches.
-This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
+This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
 *(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view.
@@
 initialisation). We then create a linked
-tour with t-SNE layout with \pkg{liminal} as shown in figure \@ref(fig:trees-01).
+tour with t-SNE layout with \pkg{liminal} as shown in Figure \@ref(fig:trees-01).
 
@@
 recapitulate the topology of the tree - the backbone (blue) branch has been 
-split into three fragments (figure \@ref(fig:trees-01)a). We can see 
+split into three fragments (Figure \@ref(fig:trees-01)a). We can see 
 this immediately via the linked highlighting over both
@@
 view there is a lot of noise along the branches, while this does not appear to
-be the case for the t-SNE result (figure \@ref(fig:trees-01)b).
+be the case for the t-SNE result (Figure \@ref(fig:trees-01)b).
 
@@
 these results with our tour view as displayed in the linked video in the caption
-of figure \@ref(fig:trees-02).
+of Figure \@ref(fig:trees-02).
 
@@
 
-The video linked in figure \@ref(fig:trees-02) shows that this selection of 
+The video linked in Figure \@ref(fig:trees-02) shows that this selection of 
 parameters results in the 
-tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
+tips of the branches (the three black dots in Figure \@ref(fig:fake-trees)a) 
 being split into three clusters representing the terminal branches of the tree. 
@@
 view, and in the tour view these tips are at opposite ends of the tree 
-(figure \@ref(fig:trees-02)b).
+(Figure \@ref(fig:trees-02)b).
 Although, this is a known issue of the t-SNE algorithm, we can easily identify
@@
 First, we downloaded the raw mouse retinal single cell RNA-seq data from
-@Macosko2015-ot using the \pkg{scRNAseq} Bioconductor package [@scRNAseq-d]. We
+@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the \pkg{scater} package
+(described by @Amezquita2020-at): we performed QC using the **scater** package
 by removing cells with high proportion of mitochondrial gene expression and low
 numbers of genes detected, we log-transformed and normalised the expression 
-values and  finally selected highly variable genes (HVGs) using \pkg{scran} 
+values and  finally selected highly variable genes (HVGs) using **scran** 
 [@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
@@
 alongside the t-SNE view which was computed from all 25 PCs.
-Due to latency of the interface we do a weighted sample of  the rows based on
+We have selected only the first five PCs because there is a large drop in the percentage of variance explained after the fifth component, with each component after contributing less than one percent of variance.
+Consequently, increasing the number of PCs to tour would increase the dimensionality and volume of the subspace we are touring but without adding any additional signal to the view.
+
+Due to latency of the \pkg{liminal} interface we do a weighted sample of  the rows based on
 cluster membership, leaving us with approximately 10 per cent of the original
 data size - 4,590 cells. Although this is not ideal, it still allows us to
-get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01).
+get a sense of the shape of the clusters as seen from the linked video in Figure \@ref(fig:mouse-01).
+If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.
 
@@
 
-```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "85%"}
+```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "90%"}
 display_screenshot("mouse-01", vid_lookup)
@@
 
-From the video linked in figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
+From the video linked in Figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
 clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
-linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
+linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
 an issue for tour in this case, where points in clusters 2,4,6, and 11 are
-clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
+clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
 The tour appears to form a tetrahedron like shape, with points lying on
@@
 a phenomena that has also been observed in @Korem2015-af 
-(figure \@ref(fig:mouse-01)c). Brushing on
+(Figure \@ref(fig:mouse-01)c). Brushing on
 the tour view, reveals points in cluster 9 that are diffuse in the embedding 
@@
 clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
-(figure \@ref(fig:mouse-01)d).
+(Figure \@ref(fig:mouse-01)d).
 
@@
 
-From the video linked in figure \@ref(fig:mouse-02), we see that the 
+From the video linked in Figure \@ref(fig:mouse-02), we see that the 
 expression of the marker genes, appear to separate
 the previously clumped clusters 2,4,6, and 11 from the other clusters, 
-indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
+indicating that these genes are expressed in all four clusters (Figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
 the clusters, while the axis view shows that magnitude of expression in 
 the marker genes is similar across these separated clusters which is 
-consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).
+consistent with the results of marker gene analysis (Figure \@ref(fig:mouse-02)b).
 
-## Discussion
+# Discussion
 
@@
 distances or ranks instead of the neighbouring points. We have begun 
-implementing this brush in \pkg{liminal},  using the  \pkg{FNN} or \pkg{RcppAnnoy}
+implementing this brush in \pkg{liminal},  using the  **FNN** or **RcppAnnoy**
 packages for fast neighbourhood estimation on the server side, however there
@@
 t-SNE avoids the crowding problem, points are separated into distinct regions
-on the display canvas, while for the tour, points are concentrated in the
-centre of the projection and become difficult to see. 
-Second, as $n$ increases the animation becomes slower as there is more data
-passing from the server to the client. Both of these issues could be mitigated
-through alternate tour paths and displays. The recent work
+on the display canvas. For the tour, points are concentrated in the
+centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem [@Laa2020-vk]. 
+Second, as $n$ increases both the embedding view and tour view become harder to read due to over-plotting, while the interactivity and animation become slower as there is more data passing from the server to the client.
+For the tasks we have looked at in this paper, where shape and density are important to the analyst, we think that better displays and sub-sampling strategies are more useful than being able to look at every single point on the canvas.
+We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however there are alternative sampling approaches that could be applied, like selecting points close to the cluster centres.
+Alternative displays via statistical transformations could also mitigate the need to show all of the data.
+Recent work
 by @Laa2020-wr is a promising area for further investigation, as well as

### 2020-11-26 paper.Rmd
compress abstract

@@
 abstract: >
-  There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows.
-The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of local and global structure in the resulting visualisation.
+  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences as part of cluster orientation and dimension reduction workflows.
+The appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
 We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
@@
 The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view.
-By linking the tour to a view obtained from an NLDR method, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
-We display several case studies from both simulated and real data from single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
+By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
+We display several case studies from both simulateions and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
 The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.

### 2020-11-27 paper.Rmd
resolve comments

@@
 sequencing experiment, or identifying a gradient or trajectory along the
-embedded form [@Nguyen2019-yh]. The second relates to using embeddings as part of
+embedded form like changes in cell development or species abundance [@Nguyen2019-yh]. The second relates to using embeddings as part of
 a clustering workflow. Here analysts are interested in identifying and naming
 clusters and verifying them by either applying known labels or colouring by
-variables that are a-priori known to distinguish clusters. Both of these
-workflow steps rely on the embedding being representative of the original high
-dimensional dataset, and becomes much more difficult when there is no underlying
-ground truth.
-\textcolor{red}{Ursula: maybe work out the "dimension reduction" examples in a
-bit more detail? They are difficult to understand for someone who does not know
-about them already (or at least for me), and it wasn't very clear why clustering
-is not part of that (I guess because we want to actually tag clusters for further
-analysis down the line?).}
+variables that are a-priori known to distinguish clusters like the expression of a marker gene to identify a cell type.
+Once clusters are identified they are used for further analysis to identify what features in the data make them distinguishable.
+Both of these
+workflow steps rely on the embedding being representative of the original high dimensional dataset, and becomes much more difficult when there is no underlying ground truth.
 
-As part of a visualization workflow, it's important to consider the perception
+As part of a visualization workflow, it is important to consider the perception
 and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
@@
 
-\textcolor{red}{Ursula: suggestion for the structure of the following paragraphs:
-start with "no one-size fits all", then talk about interactive interfaces to assess
-embeddings, and introduce tours when talking about existing interfaces, and finish by
-talking about liminal approach.}
-
-A complementary approach for visualizing structure in high dimensional data is 
-the tour. A tour is a sequence of projections of a high dimensional dataset
-onto a low-dimensional basis matrix, and is represented as an
-animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic 
-nature of the tour, user interaction is  important for controlling and 
-exploring the visualisation: the tour has been used previously by 
-@Wickham2015-cx for exploring statistical model fits and by @Buja1996-fk for 
-exploring the space of factorial experimental designs.
-
+There is no one-size fits all: finding an appropriate embedding for a given
+dataset is a difficult and a somewhat poorly defined problem. For non-linear
+methods, there are a lot of parameters to explore that can have an effect on
+the resulting visualisation and interpretation. 
 While there has been much work on the algorithmic details of
@@
 
-There is no one-size fits all: finding an appropriate embedding for a given
-dataset is a difficult and a somewhat poorly defined problem. For non-linear
-methods, there are a lot of parameters to explore that can have an effect on
-the resulting visualisation and interpretation. Interfaces for evaluating
+A complementary approach for visualizing structure in high dimensional data is 
+the tour. A tour is a sequence of projections of a high dimensional dataset
+onto a low-dimensional basis matrix, and is represented as an
+animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic 
+nature of the tour, user interaction is  important for controlling and 
+exploring the visualisation: the tour has been used previously by 
+@Wickham2015-cx for exploring statistical model fits and by @Buja1996-fk for 
+exploring the space of factorial experimental designs.
+
+The approach used in this paper is to augment the results of an NLDR method with the tour with our  \proglang{R} package called \pkg{liminal}. Interfaces for evaluating
 embeddings require interaction but should also be able to be incorporated into
-an analysts workflow. Instead, we implement a more pragmatic workflow by
+an analysts workflow. Here we implement a more pragmatic workflow by
 incorporating interactive graphics and tours with embeddings that allows users
-to see a global overview of their high dimensional data and assists them with
-cluster orientation tasks.
+to see a global overview of their high dimensional data and assists them with cluster orientation tasks.
 
@@
 # Overview of Dimension Reduction
-
-<!--
-Ursula: I don't think this paragraph is needed.
-
-In the following section, we provide a brief overview of the goals of DR
-methods with respect to the data analyst, in addition to their high level the
-mathematical details. Here we restrict our attention to two recent methods that
-are commonly used in the literature: t-distributed stochastic neighbour embedding 
-(t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
-mention several other techniques [@Maaten2008-sk; @McInnes2018-co]. 
--->
 
@@
 processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
-approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
-estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$.
-
-<!--
-Ursula: too much detail?
-
-The algorithm de-noises estimated distances in high-dimensional space via 
-transforming the affinity matrix into a Markov transition probability matrix 
-and diffusing this matrix over a fixed number of time steps. Then the diffused 
-probabilities are  transformed once more to construct a distance matrix, 
-and classical MDS is  used to generate $Y$.
--->
+approach is the PHATE algorithm [@Moon2019-ce].
 
@@
 $$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
-\textcolor{red}{Ursula: is there an intuitive explanation why we want to tune
-perplexity rather than working with sigma directly?}
+
 As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded above
@@
 
-\textcolor{red}{Ursula: here you are following the ordering from the cited
-paper, but maybe it makes sense to discuss making comparisons before
-talking about posing queries?}
-
-<!--
-Ursula: I think for these subsections we need to have a first paragraph
-that specifies what the concept means and why it is important here,
-before describing how it is incorporated in liminal
--->
-
 ## Finding Gestalt: focus and context
@@
 pause the tour path, allowing users to identify 'interesting' projections. 
-\textcolor{red}{Interesting projections or interesting points?}
 The
 domain of the axis scales from running a tour is called the half range, and is 
-computed by rescaling the input data onto hyper-dimensional
-\textcolor{red}{d-dimensional?}
-unit cube. We bind 
+computed by rescaling the input data onto $d$-dimensional unit cube. We bind 
 the half range to a mouse wheel event, allowing a user to pan and zoom on the 
@@
 current basis for use downstream.
-
-\textcolor{red}{Ursula: can I use restart to replay the same tour path (i.e. find a
-previously seen projection), or will it start a new random selection of bases?}
 
@@
 Code, data, and video for reproducing this paper are available at https://github.com/sa-lee/paper-liminal. Direct links to videos for viewing
-online are available in table \@ref(tab:vimeo).
+online are available in Table \@ref(tab:vimeo).
 

### 2020-11-27 paper.Rmd
reknit

@@
 abstract: >
-  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences as part of cluster orientation and dimension reduction workflows.
-The appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
+  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
 We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
@@
   warning = FALSE,
-  fig.align = "center"
+  fig.align = "center",
+  fig.show = 'hold',
+  dpi = 300,
+  out.width = "\\textwidth",
+  fig.path = here::here('img/')
 )
@@
 ```
+
+# References {-}

### 2020-11-27 paper.tex
reknit

\documentclass[article,notitle]{jdssv}
\AtBeginDocument{\maketitle}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Add any LaTeX packages and other preamble here if required
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{thumbpdf,lmodern}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{booktabs}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\newcommand{\ma}[1]{\ensuremath{\mathbf{#1}}}

%% begin template
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Stuart Lee\\Monash University \And Ursula Laa\\Monash University \And Dianne Cook\\Monash University
}
\title{Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings}

\Plainauthor{Stuart Lee, Ursula Laa, Dianne Cook}
\Plaintitle{Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings}
\Shorttitle{\pkg{liminal}: High-Dimensional Interactive Data Visualisation}

\Abstract{
Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique. We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour. A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space. The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view. By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading. We display several case studies from both simulateions and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks. The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.
}

\Keywords{dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, \proglang{R}}
\Plainkeywords{dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, R}

\Address{
    Stuart Lee\\
    Monash University\\
    Department of Econometrics and Business Statistics,\\
Monash University\\
  E-mail: \email{stuart.lee1@monash.edu}\\

      Ursula Laa\\
    Monash University\\
    Department of Econometrics and Business Statistics,\\
Monash University\\

      Dianne Cook\\
    Monash University\\
    Department of Econometrics and Business Statistics,\\
Monash University\\

  }

% Pandoc header

\begin{document}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of data cleaning,
pre-processing and visualisation. Methods to embed data from a high-dimensional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and de-noise data
for downstream analysis .

Choosing an appropriate embedding presents a challenge to the analyst. How does
an analyst know whether the embedding has captured the underlying topology and
geometry of the high dimensional space? The answer depends on the analyst's
workflow. Brehmer et al. (2014) characterised two main workflow steps that an
analyst performs when using embedding techniques: dimension reduction and
cluster orientation. The first relates to dimension reduction achieved by using
an embedding method, here an analyst wants to characterise and map meaning onto
the embedded form, for example identifying batch effects from a high throughput
sequencing experiment, or identifying a gradient or trajectory along the
embedded form like changes in cell development or species abundance (Nguyen and Holmes 2019). The second relates to using embeddings as part of
a clustering workflow. Here analysts are interested in identifying and naming
clusters and verifying them by either applying known labels or colouring by
variables that are a-priori known to distinguish clusters like the expression of a marker gene to identify a cell type. Once clusters are identified they are used for further analysis to identify what features in the data make them distinguishable. Both of these
workflow steps rely on the embedding being representative of the original high dimensional dataset, and becomes much more difficult when there is no underlying ground truth.

As part of a visualization workflow, it is important to consider the perception
and interpretation of embedding methods as well. Sedlmair, Munzner, and Tory (2013) showed that
scatter plots were mostly sufficient for detecting class separation, however
they also noted that often multiple embeddings were required. For the task of
cluster identification, Lewis, Van der Maaten, and Sa (2012) showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and a somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation.
While there has been much work on the algorithmic details of
embedding methods, there are relatively few tools designed to assist users
to interact with these techniques: when is an embedding sufficient for the
task at hand? Several
interactive interfaces have been proposed for evaluating or using embedding
techniques. Buja et al. (2008) used tours to guide analysts during the optimisation
of multidimensional scaling methods by extending their interactive visualisation
software called \pkg{XGobi} and \pkg{GGobi} into a new tool called \pkg{GGvis}
(Swayne, Cook, and Buja 1998; Swayne et al. 2003; Swayne and Buja 2004).
Their interface allows the analyst to dynamically modify and check whether
an MDS configuration has preserved the locality and closeness
of points between the configuration and the original data.
Ovchinnikova and Anders (2020) created the \pkg{Sleepwalk} interface for checking
non-linear embeddings in single cell RNA-seq data. It provides a click and
highlight visualisation for colouring points in an embedding according to an
estimated pairwise distance in the original high-dimensional space.
Similarly, the \pkg{TensorFlow} embedding projector is a web interface to
running some non-liner embedding methods live in the browser and
provides interactions to colour points, and select nearest neighbours
(Smilkov et al. 2016). Finally, the work by Pezzotti et al. (2017) provides a user
guided and modified form of the t-SNE algorithm, that allows users to modify
optimisation parameters in real-time.

A complementary approach for visualizing structure in high dimensional data is
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional basis matrix, and is represented as an
animated visualization (Asimov 1985; Buja and Asimov 1986). Given the dynamic
nature of the tour, user interaction is important for controlling and
exploring the visualisation: the tour has been used previously by
Wickham, Cook, and Hofmann (2015) for exploring statistical model fits and by Buja, Cook, and Swayne (1996) for
exploring the space of factorial experimental designs.

The approach used in this paper is to augment the results of an NLDR method with the tour with our \proglang{R} package called \pkg{liminal}. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. Here we implement a more pragmatic workflow by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with cluster orientation tasks.

The rest of the paper is organised as follows. The next section provides
background on dimension reduction methods, including an overview of the tour.
Then we describe the visual design of \pkg{liminal}, followed by implementation
details. Next we provide case studies that show how our interface assists
in using embedding methods. Finally, we describe the insights gained by using
\pkg{liminal} and plans for extensions to the software.

\hypertarget{overview-of-dimension-reduction}{%
\section{Overview of Dimension Reduction}\label{overview-of-dimension-reduction}}

To begin we suppose the data is in the form of a rectangular
matrix of real numbers, \(X = [\mathbf{x_1}, \dots, \mathbf{x_n}]\), where \(n\) is
the number of observations in \(p\) dimensions. The purpose of any DR algorithm
is to find a low-dimensional representation of the data,
\(Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]\), such that \(Y\) is an \(n \times d\)
matrix where \(d \ll p\). The hope of the analyst is that the DR procedure to
produces \(Y\) will remove noise in the original dataset while retaining any
latent structure.

DR methods can be classified into two broad classes: linear and non-linear
methods. Linear methods perform a linear transformation of the data, that is,
\(Y\) is a linear transformation of \(X\). One example is principal components
analysis (PCA) which performs an eigendecomposition of the estimated sample
covariance matrix. The eigenvalues are sorted in decreasing order and represent
the variance explained by each component (eigenvector).
A common approach to deciding on the number of principal components to retain is
to plot the proportion of variance explained by each component and choose a
cut-off.
When working with linear transformations we often need more than two dimensions
to capture the latent structure. In this case we can use tour methods
(Asimov 1985; Buja and Asimov 1986) to show interpolated sequences of projections,
providing the viewer with intuition about structure in more than two
dimensions, as described below.

For non-linear methods \(Y\) is generated via a pre-processed form of the input
\(X\) such as the \(k\)-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR method that aims to construct
an embedding \(Y\) such that the pair-wise distances (inner products) in \(Y\)
approximate the pair-wise distances (inner products) in \(X\)
(Torgerson 1952; Kruskal 1964a). There are many
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead (Kruskal 1964b).
A related technique is Isomap which uses a \(k\)-nearest neighbour graph
to estimate the pair-wise geodesic distance of points in \(X\) then uses classical
MDS to construct \(Y\) (Silva and Tenenbaum 2003). Other approaches are based on diffusion
processes such as diffusion maps (Coifman et al. 2005). A recent example of this
approach is the PHATE algorithm (Moon et al. 2019).

A general difficulty with using
non-linear DR methods for exploratory data analysis is selecting and tuning
appropriate parameters. For concreteness here we focus on t-distributed
stochastic neighbour embedding (t-SNE), and we will examine
its underpinning in some detail below (Maaten and Hinton 2008). Similar considerations hold for related
methods, for example UMAP (McInnes, Healy, and Melville 2018).

\hypertarget{introduction-to-t-sne}{%
\subsection{Introduction to t-SNE}\label{introduction-to-t-sne}}

The t-SNE algorithm estimates the pair-wise similarity of points in a high
dimensional space based on their (Euclidean) distances using a Gaussian
distribution. A configuration in the low dimensional embedding space is then
estimated by modelling similarities using a t-distribution with 1 degree of
freedom (Maaten and Hinton 2008). There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pair-wise distances between \(\mathbf{x_i}\) and
\(\mathbf{x_j}\) to similarities using a Gaussian kernel:

\[ 
p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 /
2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 /
2\sigma_i^2)} 
\]

The conditional probabilities are then normalised and symmetrised to form a
joint probability distribution via averaging:

\[ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n} \]

The variance parameter of the Gaussian kernel is controlled by the analyst
using a fixed value of perplexity for all observations:

\[ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) \]

As the perplexity increases, \(\sigma^2_{i}\) increases, until its bounded above
by the number of observations , \(n-1\), in the data, corresponding to
\(\sigma^2_{i} \rightarrow \infty\). This essentially turns t-SNE into a
nearest neighbours algorithm, \(p_{i|j}\) will be close to zero for all
observations that are not in the \(\mathcal{O}(\text{perplexity}_i)\)
neighbourhood graph of the \(i\)th observation (Maaten 2014).

Next, in the target low-dimensional space, \(Y\), pair-wise distances between
\(\mathbf{y_i}\) and \(\mathbf{y_j}\) are modelled as a symmetric probability
distribution using a t-distribution with one degree of freedom (Cauchy kernel):

\[ q_{ij} = \frac{w_{ij}}{Z} \\ \text{where } w_{ij} = \frac{1}{ 1 + \lVert
\mathbf{y_i - y_j} \rVert ^ 2} \text{ and } Z = \sum_{k \ne l} w_{kl}. \]

The resulting embedding \(Y\) is the one that minimizes the Kullback-Leibler
divergence between the probability distributions formed via similarities of
observations in \(X\), \(\mathcal{P}\) and similarities of observations in \(Y\),
\(\mathcal{Q}\):

\[ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log
\frac{p_{ij}}{q_{ij}}\]

Re-writing the loss function in terms of attractive (right) and repulsive
(left) forces we obtain:

\[ \mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} +
\log\sum_{i \ne j} w_{ij} \]

Minimising the loss function corresponds to large attractive
forces, that is, the pair-wise distances in \(Y\) are small when there are
non-zero \(p_{ij}\), i.e.~\(x_i\) and \(x_j\) are close together. The repulsive force
should also be small, that is, overall the
pair-wise distances in \(Y\) should be large regardless of the magnitude of the
corresponding distances in \(X\). As a result, clusters that are separate in \(X\)
will be placed far from each other in \(Y\).
This minimisation is done via stochastic gradient decent,
and introduces a number of hyperparameters, for example the number of iterations,
the learning rate, and early exaggeration, a multiplier of the attractive force
used at the beginning of the optimisation.

Taken together, these details reveal the sheer number of decisions that an
analyst must make. How does one choose the perplexity? And the
parameters that control the optimisation of the loss function?
It is a known problem that
t-SNE can have trouble recovering topology and that configurations can be
highly dependent on how the algorithm is initialised and parameterized
(Wattenberg, Viégas, and Johnson 2016; Kobak and Berens 2019; Melville 2020). If the goal is cluster
orientation a recent theoretical contribution by Linderman and Steinerberger (2019) proved that
t-SNE can recover spherical and well separated cluster shapes, and proposed new
approaches for tuning the optimisation parameters. However, the cluster sizes
and their relative orientation from a t-SNE view can be misleading
perceptually, due to the algorithms emphasis on locality.

Another recent method, UMAP, has seen a large rise in popularity (at least in
single cell transcriptomics) (McInnes, Healy, and Melville 2018). It is a method that is related to
LargeVis (Tang et al. 2016), and like t-SNE acts on the k-nearest neighbour graph.
Its main differences are that it uses a different cost function (cross entropy)
which is optimized using stochastic gradient descent and defines a different
kernel for similarities in the low dimensional space. Due to its computational
speed it is possible to generate UMAP embeddings in more than three dimensions.
It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE (Coenen and Pearce 2019).

\hypertarget{tours-explore-the-subspace-of-low-dimensional-projections}{%
\subsection{Tours explore the subspace of low dimensional projections}\label{tours-explore-the-subspace-of-low-dimensional-projections}}

The tour is a visualisation technique that is grounded in
mathematical theory, and allows the viewer to ascertain
the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections (Asimov 1985; Buja and Asimov 1986).

As with other DR techniques, the tour assumes we have a real data matrix
\(X\) consisting of \(n\) observations in \(p\) dimensions. First, the tour generates
a sequence of \(p \times d\) orthonormal projection matrices (bases)
\(A_t\), where \(d\) is typically 1 or 2. For each pair of
orthonormal bases \(A_{t}\) and \(A_{t+1}\) that are generated, the geodesic
path between them is interpolated to form intermediate frames, giving the
sense of continuous movement from one basis to another. The tour is then the
continuous visualisation of the projected data \(Y_{t} = XA_{t}\), that is the
projection of \(X\) onto \(A_{t}\) as the tour path is interpolated between
successive bases.

A \emph{grand tour} corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of \(d\)-dimensional projections. In practice, we first sphere
our data via principal components to reduce dimensionality of \(X\) prior to
running the tour. Instead of picking projections at
random, a \emph{guided tour} can be used to generate a sequence of `interesting'
projections as quantified by an index function (Cook et al. 1995). While our software,
\pkg{liminal} is able to visualise guided tours, our focus in the case studies
uses the grand tour to see global structure in the data.

\hypertarget{visual-design}{%
\section{Visual Design}\label{visual-design}}

Tours provide a supportive visualisation to NLDR graphics, and
can be easily incorporated into an analysts workflow with our software package,
\pkg{liminal}. Our interface allows analysts to quickly compare
views from embedding methods and see how an embedding method
preserves or alters the geometry of their data. Using multiple concatenated and
linked views with the tour enhances interaction techniques, and allows
analysts to perform cluster orientation tasks via linked highlighting and
brushing (McDonald 1982; Becker and Cleveland 1987).
This approach allows our interface to achieve
the three principles for interactive high-dimensional data visualisation outlined
by Buja, Cook, and Swayne (1996): finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.

\hypertarget{finding-gestalt-focus-and-context}{%
\subsection{Finding Gestalt: focus and context}\label{finding-gestalt-focus-and-context}}

To understand the data structure, we look for Gestalt features such as
(non-)linearities, clusters or outliers. A tour display is
preferred for this task, since it accurately captures the geometry, while NLDR
methods typically introduce distortions.

To investigate latent structure and the shape of a high dimensional dataset in
\pkg{liminal}, a tour can be run without the use of an external embedding.
It is often useful to
first run principal components on the input as an initial dimension reduction
step, and then tour a subset of those components instead, i.e.~by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is useful to be able to pause and highlight a
particular view. In addition to pause, play and reset buttons, brushing will
pause the tour path, allowing users to identify `interesting' projections.
The
domain of the axis scales from running a tour is called the half range, and is
computed by rescaling the input data onto \(d\)-dimensional unit cube. We bind
the half range to a mouse wheel event, allowing a user to pan and zoom on the
tour view dynamically. This is useful for peeling back dense clumps of points
to reveal structure.

\hypertarget{posing-queries-multiple-views-many-contexts}{%
\subsection{Posing Queries: multiple views, many contexts}\label{posing-queries-multiple-views-many-contexts}}

The initial visualisation gives an overview of the data structure,
and naturally leads to queries that investigate observed features with
the aim to further characterise them. This is an essential aspect of
our framework, where we use a tour to better characterise
features observed in a NLDR display.

We have combined the tour view in a side by side layout with a scatter plot
view as has been done in previous tour interfaces \pkg{XGobi} and \pkg{DataViewer}
(Buja, Hurley, and McDonald 1986; Swayne, Cook, and Buja 1998). These views are
linked; analysts can brush regions or highlight collections of points in either
view. Linked highlighting can be performed when points have been previously
labelled according to some discrete structure, i.e.~cluster labels are
available. This is achieved via the analyst clicking on groups in the legend,
which causes unselected groupings to have their points become less opaque.
Consequently, simple linked highlighting can alleviate a known downfall of
methods such as UMAP or t-SNE: that is distances between clusters are
misleading. By highlighting corresponding clusters in the tour view, the
analyst can see the relationship between clusters, and therefore obtain a more
accurate representation of the topology of their data.

Simple linked brushing is achieved via mouse-click and drag movements. By
default, when brushing occurs in the tour view, the current projection is
paused and corresponding points in the embedding view are highlighted.
Likewise, when brushing occurs in the embedding view, corresponding points in
the tour view are highlighted. In this case, an analyst can use brushing for
manually identifying clusters and verifying cluster locations and shapes:
brushing in the embedding view gives analysts a sense of the shape and
proximity of cluster in high-dimensional space.

\hypertarget{making-comparisons-revising-embeddings}{%
\subsection{Making comparisons: revising embeddings}\label{making-comparisons-revising-embeddings}}

Combining multiple views of a single data set allows the analyst
to make meaningful comparisons. In \pkg{liminal} the embedding
and tour views are arranged side-by-side for direct cross-checks.

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embeddings to be `representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data. Whether, points are near, or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbours graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by Lee and Verleysen (2009).
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively.

These checks are done visually using our side-by-side tour and embedding views.
In the simplest case, a local continuity check can be assessed via one to
one linked brushing from the embedding to the tour view. Similarly, diffusions
are identified from linked brushing on the tour view, highlighting points
in the embedding view.

\hypertarget{software-infrastructure}{%
\section{Software Infrastructure}\label{software-infrastructure}}

We have implemented the above design as an open source \proglang{R} package called
\pkg{liminal} (Lee and Cook 2020). The package allows analysts to construct concatenated
visualisations, drawn with the \pkg{Vega-Lite} grammar of interactive graphics via
the \pkg{vegawidget} package (Satyanarayan et al. 2017; Lyttle and Vega/Vega-Lite Developers 2020). It provides
an interface for constructing linked and stand alone interfaces for
manipulating tour paths via the \pkg{shiny} and \pkg{tourr} packages
(Chang et al. 2020; Wickham et al. 2011).

\hypertarget{tours-as-a-streaming-data-problem}{%
\subsection{Tours as a streaming data problem}\label{tours-as-a-streaming-data-problem}}

The process of generating successive bases and interpolating between them to\\
construct intermediary frames, means the tour is a dynamic visualisation
technique. Generally, the user would set \(d=2\) and the tour is visualised as
an animated scatter plot. This process of constructing bases and intermediate
frames and visualising the resulting projection is akin to making a ``flip book''
animation. Like with a flip book, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in \pkg{liminal}
allows a user to play, pause, and reset the tour animation,
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points if a labelling variable has
been placed on the legend.

These interactions are enabled by treating the basis generation as a reactive
stream. Instead of realising the entire sequence, which limits the animation
to have a discrete number of frames, new bases and their intermediate frames
are generated dynamically via pushing the current projection to the
visualisation interface. The interface listens to events like pressing a button
or mouse-dragging and reacts by pausing the stream. This process allows the
user to manipulate the tour in real time rather than having to fix the number
of bases ahead of time. Additionally, once the user has identified an
interesting projection or is done with the tour, the interface will return the
current basis for use downstream.

\hypertarget{linking-and-highlighting-views-via-interactions}{%
\subsection{Linking and highlighting views via interactions}\label{linking-and-highlighting-views-via-interactions}}

The embedding and tour views are linked together via rectangular brushes;
when a brush is active, points will be highlighted in the adjacent view.
Because the tour is dynamic, brush events that become active will pause the
animation, so that a user can interrogate the current view. By default,
brushing on the embedding view will produce a one-to-one linking with the tour
view. For interpreting specific combinations of clusters, the multiple
guides on the legend can be selected in order to see their relative orientations.
The interface is constructed as a \pkg{shiny} gadget specifically designed for
interactive data analysis. Selections such as brushing regions and the current
tour path are returned after the user clicks done on the interface and become
available for further investigation.

\hypertarget{case-studies}{%
\section{Case Studies}\label{case-studies}}

The next section steps through case studies of our approach using simulations
and an application to single cell RNA-seq data.

The first three case studies use simulations where the cluster structure and
geometry of the underlying data is known. We start with a simple example
where we generated spherical clusters that are embedded well by t-SNE. Then
we move onto more complex examples where the tour provides insight, such
as clusters that have substructure and where there is more complex geometry
in the data.

In the final case study, we apply our approach to clustering
the mouse retina data from Macosko et al. (2015), and apply the tour
to the process of verifying marker genes that separate clusters.

We \emph{strongly} recommend viewing the linked videos for each case study
while reading. Links to the videos are available in table \ref{tab:vimeo}
and in the figures for each case study. The videos presented show the visual
appearance of the \pkg{liminal} interface, and how we can interact with the tour
via the controls previously described. If you are unable to view the videos,
the figures in each case study consist of screenshots that summarise what is
learned from combining the tour and an embedding view.

\hypertarget{case-study-1-exploring-spherical-gaussian-clusters}{%
\subsection{Case Study 1: Exploring spherical Gaussian clusters}\label{case-study-1-exploring-spherical-gaussian-clusters}}

To begin we look at simulated datasets that reproduce known facts
about the t-SNE algorithm. Our first data set consists of five spherical 5-\(d\)
Gaussian clusters embedded in 10-\(d\) space, each cluster has the same
covariance matrix. We then computed a t-SNE layout with default settings
using the \pkg{Rtsne} package (Krijthe 2015),
and set up the \pkg{liminal} linked interface with grand tour on the 10-\(d\)
observations.

\begin{figure}

{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-gaussian}

}

\caption{Screenshots of the \pkg{liminal} interface applied to well clustered data, a video of the tour animation is available at \url{https://player.vimeo.com/video/439635921}.}\label{fig:gaussian}
\end{figure}

From the video linked in Figure \ref{fig:gaussian}, we learn that t-SNE has
correctly split out each cluster and laid them out in a star like formation.
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to
the t-SNE layout (Figure \ref{fig:gaussian}a).
This can be seen via brushing the t-SNE view (Figure \ref{fig:gaussian}b).

\hypertarget{case-study-2-exploring-spherical-gaussian-clusters-with-hierarchical-structure}{%
\subsection{Case Study 2: Exploring spherical Gaussian clusters with hierarchical structure}\label{case-study-2-exploring-spherical-gaussian-clusters-with-hierarchical-structure}}

Next we view Gaussian clusters from the
\emph{Multi Challenge Dataset}, a benchmark simulation data set for clustering tasks
(Rauber 2009). This dataset has two Gaussian clusters with equal covariance
embedded in 10-\(d\), and a third cluster
with hierarchical structure. This cluster has two 3-\(d\) clusters embedded in
10-\(d\), where the second cluster is subdivided into three smaller clusters,
that are each equidistant from each other and have the same covariance structure.
From the video linked in Figure \ref{fig:hierarchical}, we see
that t-SNE has correctly identified the sub-clusters. However, their relative
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (Figure \ref{fig:hierarchical}a). We see
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (Figure \ref{fig:hierarchical}b). When we brush
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

\begin{figure}

{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-hierarchical}

}

\caption{Screenshots of the \pkg{liminal} interface applied to sub-clustered data, a video of the tour animation is available at \url{https://player.vimeo.com/video/439635905}.}\label{fig:hierarchical}
\end{figure}

\hypertarget{case-study-3-exploring-data-with-piecewise-linear-structure}{%
\subsection{Case Study 3: Exploring data with piecewise linear structure}\label{case-study-3-exploring-data-with-piecewise-linear-structure}}

Next we explore some simulated noisy tree structured data (Figure
\ref{fig:fake-trees}). Our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set.
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches
which have a hierarchical pattern of clustering.

\begin{figure}

{\centering \includegraphics[width=\textwidth]{/Users/slee0046/paper-liminal/img/tree-view}

}

\caption{Example high-dimensional tree shaped data, \(n = 3000\) and \(p = 100\). \emph{(a)} The true data lies on 2-\(d\) tree consisting of ten branches. This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added (Moon et al. 2019). \emph{(b)} The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view. \emph{(c)} The default t-SNE view breaks the global structure of the tree. \emph{(d)} Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.}\label{fig:fake-trees}
\end{figure}

First, we apply principal components and restrict the results down to the first
twelve principal components (which makes up approximately 70\% of the variance
explained in the data) to use with the grand tour.

Moreover, we run t-SNE using the default arguments on the complete data (this
keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
initialisation). We then create a linked
tour with t-SNE layout with \pkg{liminal} as shown in Figure \ref{fig:trees-01}.

\begin{figure}

{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-trees-01}

}

\caption{Screenshots of the \pkg{liminal} interface applied to tree structured data, a video of the tour animation is available at \url{https://player.vimeo.com/video/439635892}.}\label{fig:trees-01}
\end{figure}

From the linked video, we see that the t-SNE view has been unable to
recapitulate the topology of the tree - the backbone (blue) branch has been
split into three fragments (Figure \ref{fig:trees-01}a). We can see
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (Figure \ref{fig:trees-01}b).

From the first view, we modify the inputs to the t-SNE view, to try and produce
a better trade-off between local structure and retain the topology of the data.
We keep every parameter the same except that we initialise \(Y\) with the first
two PCs (scaled to have standard deviation 1e-4) instead of the default random
initialisation and increase the perplexity from 30 to 100. We then combine
these results with our tour view as displayed in the linked video in the caption
of Figure \ref{fig:trees-02}.

\begin{figure}

{\centering \includegraphics[width=\textwidth]{/Users/slee0046/paper-liminal/img/liminal-screenshot-trees-02}

}

\caption{Screenshots of the \pkg{liminal} interface applied to tree structured data, a video of the tour animation is available at \url{https://player.vimeo.com/video/439635863}.}\label{fig:trees-02}
\end{figure}

The video linked in Figure \ref{fig:trees-02} shows that this selection of
parameters results in the
tips of the branches (the three black dots in Figure \ref{fig:fake-trees}a)
being split into three clusters representing the terminal branches of the tree.
However, there are perceptual issues following the placement of the three
groupings on the t-SNE view that become apparent via simple linked brushing.
If we brush the
tips of the yellow and brown branches (which appear to be close to each other
on the t-SNE view), we immediately see the placement is distorted in the t-SNE
view, and in the tour view these tips are at opposite ends of the tree
(Figure \ref{fig:trees-02}b).
Although, this is a known issue of the t-SNE algorithm, we can easily identify
it via simple interactivity without knowing the inner workings of the method.

\hypertarget{case-study-4-clustering-single-cell-rna-seq-data}{%
\subsection{Case Study 4: Clustering single cell RNA-seq data}\label{case-study-4-clustering-single-cell-rna-seq-data}}

A common analysis task in single cell studies is performing clustering to
identify groupings of cells with similar expression profiles. Analysts
in this area generally use non linear DR methods for verification and identification
of clusters and developmental trajectories (i.e., case study 1). For clustering
workflows the primary task is verify the existence of clusters and then begin to
identify the clusters as cell types using the expression of ``known''
marker genes. Here a `faithful' a embedding should ideally preserve the topology
of the data; cells that correspond to a cell type should lie in the same
neighbourhood in high-dimensional space. In this case study we use our
linked brushing approaches to look at neighbourhood preservation
and look at marker genes through the lens of the tour. The data we have
selected for this case study has features similar to those found in case studies
2 and 3.

First, we downloaded the raw mouse retinal single cell RNA-seq data from
Macosko et al. (2015) using the \textbf{scRNAseq} Bioconductor package (Risso and Cole 2019). We
have followed a standard workflow for pre-processing and normalizing this data
(described by Amezquita et al. (2020)): we performed QC using the \textbf{scater} package
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression
values and finally selected highly variable genes (HVGs) using \textbf{scran}
(McCarthy et al. 2017; Lun, McCarthy, and Marioni 2016). The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
used Louvain clustering to generate clusters (Blondel et al. 2008).

To check and verify the clustering we construct a \pkg{liminal} view.
We tour the first five PCs (approximately 20\% of the variance in expression),
alongside the t-SNE view which was computed from all 25 PCs. We have selected only the first five PCs because there is a large drop in the percentage of variance explained after the fifth component, with each component after contributing less than one percent of variance.
Consequently, increasing the number of PCs to tour would increase the dimensionality and volume of the subspace we are touring but without adding any additional signal to the view.

Due to latency of the \pkg{liminal} interface we do a weighted sample of the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in Figure \ref{fig:mouse-01}. If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.

\begin{figure}

{\centering \includegraphics[width=\textwidth,height=0.9\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-mouse-01}

}

\caption{Screenshots of the \pkg{liminal} interface applied to single cell data, a video of the tour animation is available at \url{https://player.vimeo.com/video/439635812}.}\label{fig:mouse-01}
\end{figure}

From the video linked in Figure \ref{fig:mouse-01}, we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (Figure \ref{fig:mouse-01}a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (Figure \ref{fig:mouse-01}b).
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the tetrahedron in 5-\(d\) PCA space -
a phenomena that has also been observed in Korem et al. (2015)
(Figure \ref{fig:mouse-01}c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding
view, points in cluster 9 are relatively far away and spread apart from other
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(Figure \ref{fig:mouse-01}d).

Next, we identify marker genes for clusters using one sided Welch t-tests with
a minimum log fold change of one as recommended by Amezquita et al. (2020), which
uses the testing framework from McCarthy and Smyth (2009). We select the top 10 marker
genes that are upregulated in cluster 2, which was one of the clumped clusters
when we toured on principal components. Here the tour becomes an alternative
to a standard heatmap view for assessing shared markers; the basis
generation (shown as the biplot on the left view) reflects the relative
weighting of each gene. We run the tour directly on the
log normalised expression values using the same subset as before.

\begin{figure}

{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-mouse-02}

}

\caption{Screenshots of the \pkg{liminal} tour applied to a marker gene set, a video of the tour animation is available at \url{https://player.vimeo.com/video/439635843}.}\label{fig:mouse-02}
\end{figure}

From the video linked in Figure \ref{fig:mouse-02}, we see that the
expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters,
indicating that these genes are expressed in all four clusters (Figure \ref{fig:mouse-02}a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in
the marker genes is similar across these separated clusters which is
consistent with the results of marker gene analysis (Figure \ref{fig:mouse-02}b).

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

We have shown that the use of tours as a tool for interacting with
high dimensional data provides an additional insight for interrogating views
generated from embeddings. The interface we have designed in the \pkg{liminal}
package, allows a user to gain a deeper understanding of an embedding
algorithm, and rectifies perceptual issues associated with NLDR methods via
linked interactions with the tour. As we have shown in the simulation case
studies, the t-SNE method can produce misleading embeddings which can be
detected through the use linked brushing and highlighting. In the case when
the data has a piecewise linear geometry, like the tree simulation, the tour
preserves the shape of the data which can be obscured by the embedding method.

Our framework can also be useful in practice, as displayed
in the fourth case study. The tour when combined with t-SNE allowed us
to identify clusters, while giving us an idea of their orientation to
each other. Moreover, we could visually inspect the separation of clusters
using a tour on marker gene sets. We see our approach as being valuable
to the single cell analyst who wants to make their embeddings more
interpretable.

We have shown in the case studies, that one to one linked brushing can
be used to identify distortions in the embedding, however we would
like extend this to one to many linked brushing, which would allow us to directly interrogate neighbourhood preservation. This form of brushing acts directly on a \(k\)-nearest neighbours (\(k\)-nn) graph computed from a
reference dataset: when a user brushes over a region in the embedding,
all the points that match the graphs edges are selected on the
corresponding tour view. The reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix)
can be independent of the tour or embedding views. In place of
highlighting, one could use opacity or binned colour scales to encode
distances or ranks instead of the neighbouring points. We have begun
implementing this brush in \pkg{liminal}, using the \textbf{FNN} or \textbf{RcppAnnoy}
packages for fast neighbourhood estimation on the server side, however there
are still technicalities that need be resolved (Beygelzimer et al. 2019; Eddelbuettel 2020).
Brush composition, such as `and', `or', or `not' brushes, could be used to
further investigate mismatches between the \(k\)-nn graphs estimated from both the embedding and tour views.

There are some limitations
in using the \pkg{liminal} interface for larger datasets. First,
t-SNE avoids the crowding problem, points are separated into distinct regions
on the display canvas. For the tour, points are concentrated in the
centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem (Laa, Cook, and Lee 2020).
Second, as \(n\) increases both the embedding view and tour view become harder to read due to over-plotting, while the interactivity and animation become slower as there is more data passing from the server to the client.
For the tasks we have looked at in this paper, where shape and density are important to the analyst, we think that better displays and sub-sampling strategies are more useful than being able to look at every single point on the canvas. We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however there are alternative sampling approaches that could be applied, like selecting points close to the cluster centres. Alternative displays via statistical transformations could also mitigate the need to show all of the data. Recent work
by Laa et al. (2020) is a promising area for further investigation, as well as
work from topological statistics (Rieck 2017; Genovese et al. 2017).

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

We would like to thank Dr David Frazier and Dr Paul Harrison for
their feedback on this work. The screenshots and images were compiled using the
\pkg{cowplot} and \pkg{ggplot2} packages (Wilke 2019; Wickham 2016).

\hypertarget{supplementary-materials}{%
\section*{Supplementary Materials}\label{supplementary-materials}}
\addcontentsline{toc}{section}{Supplementary Materials}

Code, data, and video for reproducing this paper are available at \url{https://github.com/sa-lee/paper-liminal}. Direct links to videos for viewing
online are available in Table \ref{tab:vimeo}.

\begin{table}

\caption{\label{tab:vimeo}Case Study Videos}
\centering
\begin{tabular}[t]{rll}
\toprule
Case Study & Example & URL\\
\midrule
1 & gaussian & https://player.vimeo.com/video/439635921\\
2 & hierarchical & https://player.vimeo.com/video/439635905\\
3 & trees-01 & https://player.vimeo.com/video/439635892\\
3 & trees-02 & https://player.vimeo.com/video/439635863\\
4 & mouse-01 & https://player.vimeo.com/video/439635812\\
\addlinespace
4 & mouse-02 & https://player.vimeo.com/video/439635843\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Amezquita2020-at}{}%
Amezquita, Robert A, Aaron T L Lun, Etienne Becht, Vince J Carey, Lindsay N Carpp, Ludwig Geistlinger, Federico Marini, et al. 2020. ``Orchestrating Single-Cell Analysis with Bioconductor.'' \emph{Nat. Methods} 17 (2): 137--45. \url{https://doi.org/10.1038/s41592-019-0654-x}.

\leavevmode\hypertarget{ref-Asimov1985-vp}{}%
Asimov, Daniel. 1985. ``The Grand Tour: A Tool for Viewing Multidimensional Data.'' \emph{SIAM J. Sci. And Stat. Comput.} 6 (1): 128--43. \url{https://doi.org/10.1137/0906011}.

\leavevmode\hypertarget{ref-Becker1987-gd}{}%
Becker, Richard A, and William S Cleveland. 1987. ``Brushing Scatterplots.'' \emph{Technometrics} 29 (2): 127--42. \url{https://doi.org/10.2307/1269768}.

\leavevmode\hypertarget{ref-fnn-pkg}{}%
Beygelzimer, Alina, Sham Kakadet, John Langford, Sunil Arya, David Mount, and Shengqiao Li. 2019. \emph{FNN: Fast Nearest Neighbor Search Algorithms and Applications}. \url{https://CRAN.R-project.org/package=FNN}.

\leavevmode\hypertarget{ref-Blondel2008-bx}{}%
Blondel, Vincent D, Jean-Loup Guillaume, Renaud Lambiotte, and Etienne Lefebvre. 2008. ``Fast Unfolding of Communities in Large Networks.'' \emph{J. Stat. Mech.} 2008 (10): P10008. \url{https://doi.org/10.1088/1742-5468/2008/10/P10008}.

\leavevmode\hypertarget{ref-Brehmer2014-hk}{}%
Brehmer, Matthew, Michael Sedlmair, Stephen Ingram, and Tamara Munzner. 2014. ``Visualizing Dimensionally-Reduced Data: Interviews with Analysts and a Characterization of Task Sequences.'' In \emph{Proceedings of the Fifth Workshop on Beyond Time and Errors: Novel Evaluation Methods for Visualization}, 1--8. dl.acm.org. \url{https://dl.acm.org/doi/abs/10.1145/2669557.2669559}.

\leavevmode\hypertarget{ref-Buja1986-zr}{}%
Buja, Andreas, and Daniel Asimov. 1986. ``Grand Tour Methods: An Outline.'' In \emph{Proceedings of the Seventeenth Symposium on the Interface of Computer Sciences and Statistics on Computer Science and Statistics}, 63--67. USA: Elsevier North-Holland, Inc. \url{https://dl.acm.org/doi/10.5555/26036.26046}.

\leavevmode\hypertarget{ref-Buja1996-fk}{}%
Buja, Andreas, Dianne Cook, and Deborah F Swayne. 1996. ``Interactive High-Dimensional Data Visualization.'' \emph{J. Comput. Graph. Stat.} 5 (1): 78--99. \url{https://doi.org/10.2307/1390754}.

\leavevmode\hypertarget{ref-Buja1986-ku}{}%
Buja, Andreas, Catherine Hurley, and John McDonald. 1986. ``A Data Viewer for Multivariate Data.'' In \emph{Computing Science and Statistics: Proceedings of the 18th Symposium on the Interface}, 171--74. Washington: American Statistical Association.

\leavevmode\hypertarget{ref-Buja2008-fn}{}%
Buja, Andreas, Deborah F Swayne, Michael L Littman, Nathaniel Dean, Heike Hofmann, and Lisha Chen. 2008. ``Data Visualization with Multidimensional Scaling.'' \emph{J. Comput. Graph. Stat.} 17 (2): 444--72. \url{https://doi.org/10.1198/106186008X318440}.

\leavevmode\hypertarget{ref-Chang2020-bq}{}%
Chang, Winston, Joe Cheng, J J Allaire, Yihui Xie, and Jonathan McPherson. 2020. ``Shiny: Web Application Framework for R.'' \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-Coenen2019-to}{}%
Coenen, Andy, and Adam Pearce. 2019. ``Understanding UMAP.'' \url{https://pair-code.github.io/understanding-umap/}. \url{https://pair-code.github.io/understanding-umap/}.

\leavevmode\hypertarget{ref-Coifman2005-ak}{}%
Coifman, R R, S Lafon, A B Lee, M Maggioni, B Nadler, F Warner, and S W Zucker. 2005. ``Geometric Diffusions as a Tool for Harmonic Analysis and Structure Definition of Data: Diffusion Maps.'' \emph{Proc. Natl. Acad. Sci. U. S. A.} 102 (21): 7426--31. \url{https://doi.org/10.1073/pnas.0500334102}.

\leavevmode\hypertarget{ref-Cook1995-bi}{}%
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995. ``Grand Tour and Projection Pursuit.'' \emph{J. Comput. Graph. Stat.} 4 (3): 155--72. \url{https://doi.org/10.1080/10618600.1995.10474674}.

\leavevmode\hypertarget{ref-rcpp-annoy-pkg}{}%
Eddelbuettel, Dirk. 2020. \emph{RcppAnnoy: 'Rcpp' Bindings for 'Annoy', a Library for Approximate Nearest Neighbors}. \url{https://CRAN.R-project.org/package=RcppAnnoy}.

\leavevmode\hypertarget{ref-Genovese2017-iq}{}%
Genovese, Christopher, Marco Perone-Pacifico, Isabella Verdinelli, and Larry Wasserman. 2017. ``Finding Singular Features.'' \emph{J. Comput. Graph. Stat.} 26 (3): 598--609. \url{https://doi.org/10.1080/10618600.2016.1260472}.

\leavevmode\hypertarget{ref-Kobak2019-lm}{}%
Kobak, Dmitry, and Philipp Berens. 2019. ``The Art of Using t-SNE for Single-Cell Transcriptomics.'' \emph{Nat. Commun.} 10 (1): 5416. \url{https://doi.org/10.1038/s41467-019-13056-x}.

\leavevmode\hypertarget{ref-Korem2015-af}{}%
Korem, Yael, Pablo Szekely, Yuval Hart, Hila Sheftel, Jean Hausser, Avi Mayo, Michael E Rothenberg, Tomer Kalisky, and Uri Alon. 2015. ``Geometry of the Gene Expression Space of Individual Cells.'' \emph{PLoS Comput. Biol.} 11 (7): e1004224. \url{https://doi.org/10.1371/journal.pcbi.1004224}.

\leavevmode\hypertarget{ref-Rtsne}{}%
Krijthe, Jesse H. 2015. \emph{Rtsne: T-Distributed Stochastic Neighbor Embedding Using Barnes-Hut Implementation}. \url{https://github.com/jkrijthe/Rtsne}.

\leavevmode\hypertarget{ref-Kruskal1964-cz}{}%
Kruskal, J B. 1964a. ``Multidimensional Scaling by Optimizing Goodness of Fit to a Nonmetric Hypothesis.'' \emph{Psychometrika} 29 (1): 1--27. \url{https://doi.org/10.1007/BF02289565}.

\leavevmode\hypertarget{ref-Kruskal1964-cw}{}%
---------. 1964b. ``Nonmetric Multidimensional Scaling: A Numerical Method.'' \emph{Psychometrika} 29 (2): 115--29. \url{https://doi.org/10.1007/BF02289694}.

\leavevmode\hypertarget{ref-Laa2020-wr}{}%
Laa, Ursula, Dianne Cook, Andreas Buja, and German Valencia. 2020. ``Hole or Grain? A Section Pursuit Index for Finding Hidden Structure in Multiple Dimensions,'' April. \url{http://arxiv.org/abs/2004.13327}.

\leavevmode\hypertarget{ref-Laa2020-vk}{}%
Laa, Ursula, Dianne Cook, and Stuart Lee. 2020. ``Burning Sage: Reversing the Curse of Dimensionality in the Visualization of High-Dimensional Data,'' September. \url{http://arxiv.org/abs/2009.10979}.

\leavevmode\hypertarget{ref-Lee2009-zb}{}%
Lee, John A, and Michel Verleysen. 2009. ``Quality Assessment of Dimensionality Reduction: Rank-Based Criteria.'' \emph{Neurocomputing} 72 (7): 1431--43. \url{https://doi.org/10.1016/j.neucom.2008.12.017}.

\leavevmode\hypertarget{ref-r-liminal}{}%
Lee, Stuart, and Dianne Cook. 2020. \emph{Liminal: Multivariate Data Visualization with Tours and Embeddings}. \url{https://github.com/sa-lee/liminal}.

\leavevmode\hypertarget{ref-Lewis2012-ai}{}%
Lewis, Joshua, Laurens Van der Maaten, and Virginia de Sa. 2012. ``A Behavioral Investigation of Dimensionality Reduction.'' In \emph{Proceedings of the Annual Meeting of the Cognitive Science Society}. Vol. 34.

\leavevmode\hypertarget{ref-Linderman2019-dq}{}%
Linderman, George C, and Stefan Steinerberger. 2019. ``Clustering with t-SNE, Provably.'' \emph{SIAM Journal on Mathematics of Data Science} 1 (2): 313--32. \url{https://doi.org/10.1137/18M1216134}.

\leavevmode\hypertarget{ref-Lun2016}{}%
Lun, Aaron T. L., Davis J. McCarthy, and John C. Marioni. 2016. ``A Step-by-Step Workflow for Low-Level Analysis of Single-Cell Rna-Seq Data with Bioconductor.'' \emph{F1000Res.} 5: 2122. \url{https://doi.org/10.12688/f1000research.9501.2}.

\leavevmode\hypertarget{ref-Lyttle2020-hp}{}%
Lyttle, Ian, and Vega/Vega-Lite Developers. 2020. ``Vegawidget: 'Htmlwidget' for 'Vega' and 'Vega-Lite'.'' \url{https://CRAN.R-project.org/package=vegawidget}.

\leavevmode\hypertarget{ref-Van_Der_Maaten2014-zn}{}%
Maaten, Laurens van der. 2014. ``Accelerating t-SNE Using Tree-Based Algorithms.'' \emph{J. Mach. Learn. Res.} \url{http://www.jmlr.org/papers/volume15/vandermaaten14a/vandermaaten14a.pdf}.

\leavevmode\hypertarget{ref-Maaten2008-sk}{}%
Maaten, Laurens van der, and Geoffrey Hinton. 2008. ``Visualizing Data Using t-SNE.'' \emph{J. Mach. Learn. Res.} 9 (Nov): 2579--2605. \url{http://www.jmlr.org/papers/v9/vandermaaten08a.html}.

\leavevmode\hypertarget{ref-Macosko2015-ot}{}%
Macosko, Evan Z, Anindita Basu, Rahul Satija, James Nemesh, Karthik Shekhar, Melissa Goldman, Itay Tirosh, et al. 2015. ``Highly Parallel Genome-Wide Expression Profiling of Individual Cells Using Nanoliter Droplets.'' \emph{Cell} 161 (5): 1202--14. \url{https://doi.org/10.1016/j.cell.2015.05.002}.

\leavevmode\hypertarget{ref-McCarthy2017}{}%
McCarthy, Davis J., Kieran R. Campbell, Aaron T. L. Lun, and Quin F. Willis. 2017. ``Scater: Pre-Processing, Quality Control, Normalisation and Visualisation of Single-Cell RNA-Seq Data in R.'' \emph{Bioinformatics} 33 (8): 1179--86. \url{https://doi.org/10.1093/bioinformatics/btw777}.

\leavevmode\hypertarget{ref-McCarthy2009-qx}{}%
McCarthy, Davis J, and Gordon K Smyth. 2009. ``Testing Significance Relative to a Fold-Change Threshold Is a TREAT.'' \emph{Bioinformatics} 25 (6): 765--71. \url{https://doi.org/10.1093/bioinformatics/btp053}.

\leavevmode\hypertarget{ref-McDonald1982-ew}{}%
McDonald, J A. 1982. ``Interactive Graphics for Data Analysis.'' PhD thesis, Stanford University. \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.331.6673\&rep=rep1\&type=pdf}.

\leavevmode\hypertarget{ref-McInnes2018-co}{}%
McInnes, Leland, John Healy, and James Melville. 2018. ``UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction,'' February. \url{http://arxiv.org/abs/1802.03426}.

\leavevmode\hypertarget{ref-Melville2020}{}%
Melville, James. 2020. ``t-SNE Initialization Options.'' \url{https://jlmelville.github.io/smallvis/init.html}. \url{https://jlmelville.github.io/smallvis/init.html}.

\leavevmode\hypertarget{ref-Moon2019-ce}{}%
Moon, Kevin R, David van Dijk, Zheng Wang, Scott Gigante, Daniel B Burkhardt, William S Chen, Kristina Yim, et al. 2019. ``Visualizing Structure and Transitions for Biological Data Exploration.'' \emph{bioRxiv}. \url{https://doi.org/10.1101/120378}.

\leavevmode\hypertarget{ref-Nguyen2019-yh}{}%
Nguyen, Lan Huong, and Susan Holmes. 2019. ``Ten Quick Tips for Effective Dimensionality Reduction.'' \emph{PLoS Comput. Biol.} 15 (6): e1006907. \url{https://doi.org/10.1371/journal.pcbi.1006907}.

\leavevmode\hypertarget{ref-Ovchinnikova2020-sy}{}%
Ovchinnikova, Svetlana, and Simon Anders. 2020. ``Exploring Dimension-Reduced Embeddings with Sleepwalk.'' \emph{Genome Res.} 30 (5): 749--56. \url{https://doi.org/10.1101/gr.251447.119}.

\leavevmode\hypertarget{ref-Pezzotti2017-cz}{}%
Pezzotti, Nicola, Boudewijn P F Lelieveldt, Laurens Van Der Maaten, Thomas Hollt, Elmar Eisemann, and Anna Vilanova. 2017. ``Approximated and User Steerable tSNE for Progressive Visual Analytics.'' \emph{IEEE Trans. Vis. Comput. Graph.} 23 (7): 1739--52. \url{https://doi.org/10.1109/TVCG.2016.2570755}.

\leavevmode\hypertarget{ref-Rauber2009-vh}{}%
Rauber, Andreas. 2009. ``Multi-Challenge Data Set.'' \url{http://ifs.tuwien.ac.at/dm/dataSets.html}. \url{http://ifs.tuwien.ac.at/dm/dataSets.html}.

\leavevmode\hypertarget{ref-Rieck2017-kk}{}%
Rieck, Bastian. 2017. ``Persistent Homology in Multivariate Data Visualization.'' Edited by Heike Little and Michael Gertz. PhD thesis, Ruprecht-Karls-Universität Heidelberg.

\leavevmode\hypertarget{ref-scRNAseq-d}{}%
Risso, Davide, and Michael Cole. 2019. \emph{ScRNAseq: Collection of Public Single-Cell Rna-Seq Datasets}.

\leavevmode\hypertarget{ref-Satyanarayan2017-gs}{}%
Satyanarayan, Arvind, Dominik Moritz, Kanit Wongsuphasawat, and Jeffrey Heer. 2017. ``Vega-Lite: A Grammar of Interactive Graphics.'' \emph{IEEE Trans. Vis. Comput. Graph.} 23 (1): 341--50. \url{https://doi.org/10.1109/TVCG.2016.2599030}.

\leavevmode\hypertarget{ref-Sedlmair2013-pn}{}%
Sedlmair, Michael, Tamara Munzner, and Melanie Tory. 2013. ``Empirical Guidance on Scatterplot and Dimension Reduction Technique Choices.'' \emph{IEEE Trans. Vis. Comput. Graph.} 19 (12): 2634--43. \url{https://doi.org/10.1109/TVCG.2013.153}.

\leavevmode\hypertarget{ref-Silva2003-xw}{}%
Silva, V D, and J B Tenenbaum. 2003. ``Global Versus Local Methods in Nonlinear Dimensionality Reduction.'' \emph{Adv. Neural Inf. Process. Syst.} \url{http://papers.nips.cc/paper/2141-global-versus-local-methods-in-nonlinear-dimensionality-reduction.pdf}.

\leavevmode\hypertarget{ref-Smilkov2016-hp}{}%
Smilkov, Daniel, Nikhil Thorat, Charles Nicholson, Emily Reif, Fernanda B Viégas, and Martin Wattenberg. 2016. ``Embedding Projector: Interactive Visualization and Interpretation of Embeddings,'' November. \url{http://arxiv.org/abs/1611.05469}.

\leavevmode\hypertarget{ref-Swayne2004-ua}{}%
Swayne, Deborah F, and Andreas Buja. 2004. ``Exploratory Visual Analysis of Graphs in GGOBI.'' In \emph{COMPSTAT 2004 --- Proceedings in Computational Statistics}, 477--88. Physica-Verlag HD. \url{https://doi.org/10.1007/978-3-7908-2656-2/_39}.

\leavevmode\hypertarget{ref-Swayne1998-uq}{}%
Swayne, Deborah F, Dianne Cook, and Andreas Buja. 1998. ``XGobi: Interactive Dynamic Data Visualization in the X Window System.'' \emph{J. Comput. Graph. Stat.} 7 (1): 113--30. \url{https://doi.org/10.1080/10618600.1998.10474764}.

\leavevmode\hypertarget{ref-Swayne2003-qd}{}%
Swayne, Deborah F, Duncan Temple Lang, Andreas Buja, and Dianne Cook. 2003. ``GGobi: Evolving from XGobi into an Extensible Framework for Interactive Data Visualization.'' \emph{Comput. Stat. Data Anal.} 43 (4): 423--44. \url{https://doi.org/10.1016/S0167-9473(02)00286-4}.

\leavevmode\hypertarget{ref-Tang2016-oz}{}%
Tang, Jian, Jingzhou Liu, Ming Zhang, and Qiaozhu Mei. 2016. ``Visualizing Large-Scale and High-Dimensional Data,'' February. \url{http://arxiv.org/abs/1602.00370}.

\leavevmode\hypertarget{ref-Torgerson1952-am}{}%
Torgerson, Warren S. 1952. ``Multidimensional Scaling: I. Theory and Method.'' \emph{Psychometrika} 17 (4): 401--19. \url{https://doi.org/10.1007/BF02288916}.

\leavevmode\hypertarget{ref-Wattenberg2016-ji}{}%
Wattenberg, Martin, Fernanda Viégas, and Ian Johnson. 2016. ``How to Use t-SNE Effectively.'' \emph{Distill} 1 (10). \url{https://doi.org/10.23915/distill.00002}.

\leavevmode\hypertarget{ref-Wickham2016-gz}{}%
Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data Analysis}. Use R! Springer International Publishing. \url{https://doi.org/10.1007/978-3-319-24277-4}.

\leavevmode\hypertarget{ref-Wickham2015-cx}{}%
Wickham, Hadley, Dianne Cook, and Heike Hofmann. 2015. ``Visualizing Statistical Models: Removing the Blindfold.'' \emph{Statistical Analysis and Data Mining: The ASA Data Science Journal} 8 (4): 203--25. \url{https://doi.org/10.1002/sam.11271}.

\leavevmode\hypertarget{ref-Wickham2011-st}{}%
Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011. ``Tourr: An R Package for Exploring Multivariate Data with Projections.'' \emph{Journal of Statistical Software, Articles} 40 (2): 1--18. \url{https://doi.org/10.18637/jss.v040.i02}.

\leavevmode\hypertarget{ref-r-cowplot}{}%
Wilke, Clause O. 2019. \emph{Cowplot: Streamlined Plot Theme and Plot Annotations for 'Ggplot2'}. \url{https://CRAN.R-project.org/package=cowplot}.

\end{document}

### 2020-11-28 paper.Rmd
Typos fixed.

@@
 By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
-We display several case studies from both simulateions and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
+We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
 The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.
@@
 $X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
-Multidimensional scaling (MDS) is a class of DR method that aims to construct
+Multidimensional scaling (MDS) is a class of DR methods that aims to construct
 an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
@@
 correctly split out each cluster and laid them out in a star like formation. 
-This is agrees with the tour view,
+This agrees with the tour view,
 where once we start the animation, the five clusters begin to appear but
@@
 (ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
-*(a)* The true data lies on 2-$d$ tree consisting of ten branches.
+*(a)* The true data lies on a 2-$d$ tree consisting of ten branches.
 This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
@@
 on each view are highlighted and the remaining points are made transparent.
-From, here it becomes apparent from the tour view that the blue branch forms
+From here it becomes apparent from the tour view that the blue branch forms
 the backbone of the tree and is connected to all other branches. From the video 
@@
 linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
-an issue for tour in this case, where points in clusters 2,4,6, and 11 are
+an issue for tour in this case, where points in clusters 2, 4, 6, and 11 are
 clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
 The tour appears to form a tetrahedron like shape, with points lying on
-surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
+the surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
 a phenomena that has also been observed in @Korem2015-af 
@@
 expression of the marker genes, appear to separate
-the previously clumped clusters 2,4,6, and 11 from the other clusters, 
+the previously clumped clusters 2, 4, 6, and 11 from the other clusters, 
 indicating that these genes are expressed in all four clusters (Figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along

### 2020-12-10 paper.tex
add acknowledgment

@@
 \(X\) such as the \(k\)-nearest neighbours graph or via a kernel transformation.
-Multidimensional scaling (MDS) is a class of DR method that aims to construct
+Multidimensional scaling (MDS) is a class of DR methods that aims to construct
 an embedding \(Y\) such that the pair-wise distances (inner products) in \(Y\)
@@
 correctly split out each cluster and laid them out in a star like formation.
-This is agrees with the tour view,
+This agrees with the tour view,
 where once we start the animation, the five clusters begin to appear but
@@
 \caption{Example high-dimensional tree shaped data, \(n = 3000\) and \(p = 100\).
-\emph{(a)} The true data lies on 2-\(d\) tree consisting of ten branches.
+\emph{(a)} The true data lies on a 2-\(d\) tree consisting of ten branches.
 This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added (Moon et al.
@@
 on each view are highlighted and the remaining points are made transparent.
-From, here it becomes apparent from the tour view that the blue branch forms
+From here it becomes apparent from the tour view that the blue branch forms
 the backbone of the tree and is connected to all other branches. From the video
@@
 linked brushing (Figure \ref{fig:mouse-01}a). As expected, t-SNE mitigates the crowding problem that is
-an issue for tour in this case, where points in clusters 2,4,6, and 11 are
+an issue for tour in this case, where points in clusters 2, 4, 6, and 11 are
 clumped together in tour view, but are blown up in the embedding view (Figure \ref{fig:mouse-01}b).
 The tour appears to form a tetrahedron like shape, with points lying on
-surface and along the vertices of the tetrahedron in 5-\(d\) PCA space -
+the surface and along the vertices of the tetrahedron in 5-\(d\) PCA space -
 a phenomena that has also been observed in Korem et al. (2015)
@@
 expression of the marker genes, appear to separate
-the previously clumped clusters 2,4,6, and 11 from the other clusters,
+the previously clumped clusters 2, 4, 6, and 11 from the other clusters,
 indicating that these genes are expressed in all four clusters (Figure \ref{fig:mouse-02}a). After zooming, we can see a trajectory forming along
@@
 
+The authors gratefully acknowledge the support of the Australian Research Council.
 We would like to thank Dr David Frazier and Dr Paul Harrison for

### 2020-12-11 scripts/liminal-quickdraw.R
remove old files

@@ -1,115 +1,0 @@
-## ----init--------------------------------------------------------------------
-library(dplyr)
-library(liminal)
-# remotes::install_github("huizezhang-sherry/quickdraw")
-library(quickdraw)
-source(here::here("scripts", "helpers.R"))
-
-## ----check-cache--------------------------------------------------------------
-has_run <- file.exists(here::here("data", "qd.rds"))
-
-## ----download-categories-----------------------------------------------------
-if (!has_run) {
-  animals <- c("broccoli", "flying saucer", "ice cream", "octopus", "pineapple", "whale")
-  # download and cache
-  for (a in animals) {
-    qd_download(a)
-    qd_download_bitmap(a)
-  }
-  # read and slice sample meta
-  set.seed(4099)
-  slabs <- lapply(animals, slice_image_meta)
-  row_nums <- lapply(slabs, function(slab) slab[["row_number"]])
-  # extract bitmaps as matrices
-  bits <- mapply(slice_bitmap, 
-                 a = animals, 
-                 row_number = row_nums, 
-                 SIMPLIFY = FALSE)
-  # bind as a single data frame
-  slabs <- dplyr::bind_rows(slabs)
-  bits <- dplyr::bind_rows(bits)
-  drawings <- dplyr::bind_cols(slabs, bits)
-  # save results
-  saveRDS(drawings, file = here::here("data/qd.rds"))
-}
-
-## ----load-cache--------------------------------------------------------------
-drawings <- readRDS(here::here("data/qd.rds"))
-
-
-# normalise the pixel intensities for each image by centering each
-# image by their average pixel intensity 
-# px_mat <- as.matrix(select(drawings, starts_with("px")))
-# px_means <- rowMeans(px_mat)
-# px_norm <- sweep(px_mat, 1, px_means)
-# colnames(px_norm) <- paste0("norm_", colnames(px_norm))
-# drawings <- bind_cols(drawings, as_tibble(px_norm))
-
-# Sphere the images by taking the SVD of covariance matrix
-# this removes covariance between pixels
-# px_cov <- tcrossprod(px_norm) / ncol(px_norm)
-# svals <- svd(px_cov)
-# px_rot <-
-#   diag(1 / (sqrt(svals$d) + 1e-5)) %*%
-#   crossprod(svals$u, px_norm)
-
-
-# default t-SNE does pretty well at splitting the categories
[…]

### 2020-12-11 liminal.Rmd
rename for arxiv
reformat for arxiv

--- 2020-11-28 paper.Rmd
@@
     toc: false
+    citation_package: natbib
 bibliography: liminal.bib
@@
 ```{r setup, include = FALSE}
+options(tinytex.clean = FALSE)
 knitr::opts_chunk$set(
@@
 
+The authors gratefully acknowledge the support of the Australian Research Council.
 We would like to thank Dr David Frazier and Dr Paul Harrison for 

### 2020-12-11 liminal.tex
reformat for arxiv

--- 2020-12-10 paper.tex
@@
 geometry of the high dimensional space? The answer depends on the analyst's
-workflow. Brehmer et al. (2014) characterised two main workflow steps that an
+workflow. \citet{Brehmer2014-hk} characterised two main workflow steps that an
 analyst performs when using embedding techniques: dimension reduction and
@@
 sequencing experiment, or identifying a gradient or trajectory along the
-embedded form like changes in cell development or species abundance (Nguyen and Holmes 2019). The second relates to using embeddings as part of
+embedded form like changes in cell development or species abundance \citep{Nguyen2019-yh}. The second relates to using embeddings as part of
 a clustering workflow. Here analysts are interested in identifying and naming
@@
 As part of a visualization workflow, it is important to consider the perception
-and interpretation of embedding methods as well. Sedlmair, Munzner, and Tory (2013) showed that
+and interpretation of embedding methods as well. \citet{Sedlmair2013-pn} showed that
 scatter plots were mostly sufficient for detecting class separation, however
 they also noted that often multiple embeddings were required. For the task of
-cluster identification, Lewis, Van der Maaten, and Sa (2012) showed experimentally that novice users
+cluster identification, \citet{Lewis2012-ai} showed experimentally that novice users
 of non-linear embedding techniques were more likely to consider clusters of
@@
 interactive interfaces have been proposed for evaluating or using embedding
-techniques. Buja et al. (2008) used tours to guide analysts during the optimisation
+techniques. \citet{Buja2008-fn} used tours to guide analysts during the optimisation
 of multidimensional scaling methods by extending their interactive visualisation
 software called \pkg{XGobi} and \pkg{GGobi} into a new tool called \pkg{GGvis}
-(Swayne, Cook, and Buja 1998; Swayne et al. 2003; Swayne and Buja 2004).
+\citep{Swayne1998-uq, Swayne2003-qd, Swayne2004-ua}.
 Their interface allows the analyst to dynamically modify and check whether
@@
 of points between the configuration and the original data.
-Ovchinnikova and Anders (2020) created the \pkg{Sleepwalk} interface for checking
+\citet{Ovchinnikova2020-sy} created the \pkg{Sleepwalk} interface for checking
 non-linear embeddings in single cell RNA-seq data. It provides a click and
@@
 provides interactions to colour points, and select nearest neighbours
-(Smilkov et al. 2016). Finally, the work by Pezzotti et al. (2017) provides a user
+\citep{Smilkov2016-hp}. Finally, the work by \citet{Pezzotti2017-cz} provides a user
 guided and modified form of the t-SNE algorithm, that allows users to modify
@@
 onto a low-dimensional basis matrix, and is represented as an
-animated visualization (Asimov 1985; Buja and Asimov 1986). Given the dynamic
+animated visualization \citep{Asimov1985-vp, Buja1986-zr}. Given the dynamic
 nature of the tour, user interaction is important for controlling and
 exploring the visualisation: the tour has been used previously by
-Wickham, Cook, and Hofmann (2015) for exploring statistical model fits and by Buja, Cook, and Swayne (1996) for
+\citet{Wickham2015-cx} for exploring statistical model fits and by \citet{Buja1996-fk} for
 exploring the space of factorial experimental designs.
@@
 to capture the latent structure. In this case we can use tour methods
-(Asimov 1985; Buja and Asimov 1986) to show interpolated sequences of projections,
+\citep{Asimov1985-vp, Buja1986-zr} to show interpolated sequences of projections,
 providing the viewer with intuition about structure in more than two
@@
 approximate the pair-wise distances (inner products) in \(X\)
-(Torgerson 1952; Kruskal 1964a). There are many
+\citep{Torgerson1952-am, Kruskal1964-cz}. There are many
 variants of MDS, such as non-metric scaling which amounts to replacing
-distances with ranks instead (Kruskal 1964b).
+distances with ranks instead \citep{Kruskal1964-cw}.
 A related technique is Isomap which uses a \(k\)-nearest neighbour graph
 to estimate the pair-wise geodesic distance of points in \(X\) then uses classical
-MDS to construct \(Y\) (Silva and Tenenbaum 2003). Other approaches are based on diffusion
-processes such as diffusion maps (Coifman et al. 2005). A recent example of this
-approach is the PHATE algorithm (Moon et al. 2019).
+MDS to construct \(Y\) \citep{Silva2003-xw}. Other approaches are based on diffusion
+processes such as diffusion maps \citep{Coifman2005-ak}. A recent example of this
+approach is the PHATE algorithm \citep{Moon2019-ce}.
 
@@
 stochastic neighbour embedding (t-SNE), and we will examine
-its underpinning in some detail below (Maaten and Hinton 2008). Similar considerations hold for related
-methods, for example UMAP (McInnes, Healy, and Melville 2018).
+its underpinning in some detail below \citep{Maaten2008-sk}. Similar considerations hold for related
+methods, for example UMAP \citep{McInnes2018-co}.
 
@@
 estimated by modelling similarities using a t-distribution with 1 degree of
-freedom (Maaten and Hinton 2008). There are several subtleties
+freedom \citep{Maaten2008-sk}. There are several subtleties
 to the to use of the algorithm that are revealed by stepping through its
@@
 observations that are not in the \(\mathcal{O}(\text{perplexity}_i)\)
-neighbourhood graph of the \(i\)th observation (Maaten 2014).
+neighbourhood graph of the \(i\)th observation \citep{Van_Der_Maaten2014-zn}.
 
@@
 highly dependent on how the algorithm is initialised and parameterized
-(Wattenberg, Viégas, and Johnson 2016; Kobak and Berens 2019; Melville 2020). If the goal is cluster
-orientation a recent theoretical contribution by Linderman and Steinerberger (2019) proved that
+\citep{Wattenberg2016-ji, Kobak2019-lm, Melville2020}. If the goal is cluster
+orientation a recent theoretical contribution by \citet{Linderman2019-dq} proved that
 t-SNE can recover spherical and well separated cluster shapes, and proposed new
@@
 Another recent method, UMAP, has seen a large rise in popularity (at least in
-single cell transcriptomics) (McInnes, Healy, and Melville 2018). It is a method that is related to
-LargeVis (Tang et al. 2016), and like t-SNE acts on the k-nearest neighbour graph.
+single cell transcriptomics) \citep{McInnes2018-co}. It is a method that is related to
+LargeVis \citep{Tang2016-oz}, and like t-SNE acts on the k-nearest neighbour graph.
 Its main differences are that it uses a different cost function (cross entropy)
@@
 speed it is possible to generate UMAP embeddings in more than three dimensions.
-It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE (Coenen and Pearce 2019).
+It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE \citep{Coenen2019-to}.
 
@@
 mathematical theory, and allows the viewer to ascertain
-the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections (Asimov 1985; Buja and Asimov 1986).
+the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections \citep{Asimov1985-vp, Buja1986-zr}.
 
@@
 random, a \emph{guided tour} can be used to generate a sequence of `interesting'
-projections as quantified by an index function (Cook et al. 1995). While our software,
+projections as quantified by an index function \citep{Cook1995-bi}. While our software,
 \pkg{liminal} is able to visualise guided tours, our focus in the case studies
@@
 analysts to perform cluster orientation tasks via linked highlighting and
-brushing (McDonald 1982; Becker and Cleveland 1987).
+brushing \citep{McDonald1982-ew, Becker1987-gd}.
 This approach allows our interface to achieve
 the three principles for interactive high-dimensional data visualisation outlined
-by Buja, Cook, and Swayne (1996): finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.
+by \citet{Buja1996-fk}: finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.
 
@@
 view as has been done in previous tour interfaces \pkg{XGobi} and \pkg{DataViewer}
-(Buja, Hurley, and McDonald 1986; Swayne, Cook, and Buja 1998). These views are
+\citep{Buja1986-ku, Swayne1998-uq}. These views are
 linked; analysts can brush regions or highlight collections of points in either
@@
 high-dimensional space and the embedding space. As distances can be noisy in
-high-dimensions, ranks can be used instead as has been proposed by Lee and Verleysen (2009).
+high-dimensions, ranks can be used instead as has been proposed by \citet{Lee2009-zb}.
 Identifying distortions and diffusions allows an analyst to investigate the
@@
 We have implemented the above design as an open source \proglang{R} package called
-\pkg{liminal} (Lee and Cook 2020). The package allows analysts to construct concatenated
+\pkg{liminal} \citep{r-liminal}. The package allows analysts to construct concatenated
 visualisations, drawn with the \pkg{Vega-Lite} grammar of interactive graphics via
-the \pkg{vegawidget} package (Satyanarayan et al. 2017; Lyttle and Vega/Vega-Lite Developers 2020). It provides
+the \pkg{vegawidget} package \citep{Satyanarayan2017-gs, Lyttle2020-hp}. It provides
 an interface for constructing linked and stand alone interfaces for
 manipulating tour paths via the \pkg{shiny} and \pkg{tourr} packages
-(Chang et al. 2020; Wickham et al. 2011).
+\citep{Chang2020-bq, Wickham2011-st}.
 
@@
 In the final case study, we apply our approach to clustering
-the mouse retina data from Macosko et al. (2015), and apply the tour
+the mouse retina data from \citet{Macosko2015-ot}, and apply the tour
 to the process of verifying marker genes that separate clusters.
@@
 covariance matrix. We then computed a t-SNE layout with default settings
-using the \pkg{Rtsne} package (Krijthe 2015),
+using the \pkg{Rtsne} package \citep{Rtsne},
 and set up the \pkg{liminal} linked interface with grand tour on the 10-\(d\)
@@
 \emph{Multi Challenge Dataset}, a benchmark simulation data set for clustering tasks
-(Rauber 2009). This dataset has two Gaussian clusters with equal covariance
+\citep{Rauber2009-vh}. This dataset has two Gaussian clusters with equal covariance
 embedded in 10-\(d\), and a third cluster
@@
 \emph{(a)} The true data lies on a 2-\(d\) tree consisting of ten branches.
-This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added (Moon et al.
-2019).
+This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added \citep{Moon2019-ce}.
 \emph{(b)} The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view.
@@
 First, we downloaded the raw mouse retinal single cell RNA-seq data from
-Macosko et al. (2015) using the \textbf{scRNAseq} Bioconductor package (Risso and Cole 2019). We
+\citet{Macosko2015-ot} using the \textbf{scRNAseq} Bioconductor package \citep{scRNAseq-d}. We
 have followed a standard workflow for pre-processing and normalizing this data
-(described by Amezquita et al. (2020)): we performed QC using the \textbf{scater} package
+(described by \citet{Amezquita2020-at}): we performed QC using the \textbf{scater} package
 by removing cells with high proportion of mitochondrial gene expression and low
@@
 values and finally selected highly variable genes (HVGs) using \textbf{scran}
-(McCarthy et al. 2017; Lun, McCarthy, and Marioni 2016). The top ten percent of HVGs were used to subset
+\citep{McCarthy2017, Lun2016}. The top ten percent of HVGs were used to subset
 the normalised expression matrix and compute PCA using the first 25 components.
 Using the PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
-used Louvain clustering to generate clusters (Blondel et al. 2008).
+used Louvain clustering to generate clusters \citep{Blondel2008-bx}.
 
@@
 the surface and along the vertices of the tetrahedron in 5-\(d\) PCA space -
-a phenomena that has also been observed in Korem et al. (2015)
+a phenomena that has also been observed in \citet{Korem2015-af}
 (Figure \ref{fig:mouse-01}c). Brushing on
@@
 Next, we identify marker genes for clusters using one sided Welch t-tests with
-a minimum log fold change of one as recommended by Amezquita et al. (2020), which
-uses the testing framework from McCarthy and Smyth (2009). We select the top 10 marker
+a minimum log fold change of one as recommended by \citet{Amezquita2020-at}, which
+uses the testing framework from \citet{McCarthy2009-qx}. We select the top 10 marker
 genes that are upregulated in cluster 2, which was one of the clumped clusters
@@
 packages for fast neighbourhood estimation on the server side, however there
-are still technicalities that need be resolved (Beygelzimer et al. 2019; Eddelbuettel 2020).
+are still technicalities that need be resolved \citep{fnn-pkg, rcpp-annoy-pkg}.
 Brush composition, such as `and', `or', or `not' brushes, could be used to
@@
 on the display canvas. For the tour, points are concentrated in the
-centre of the projection and become difficult to see.
-We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem (Laa, Cook, and Lee 2020).
+centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem \citep{Laa2020-vk}.
 Second, as \(n\) increases both the embedding view and tour view become harder to read due to over-plotting, while the interactivity and animation become slower as there is more data passing from the server to the client.
@@
 Recent work
-by Laa et al. (2020) is a promising area for further investigation, as well as
-work from topological statistics (Rieck 2017; Genovese et al. 2017).
+by \citet{Laa2020-wr} is a promising area for further investigation, as well as
+work from topological statistics \citep{Rieck2017-kk, Genovese2017-iq}.
 
@@
 their feedback on this work. The screenshots and images were compiled using the
-\pkg{cowplot} and \pkg{ggplot2} packages (Wilke 2019; Wickham 2016).
+\pkg{cowplot} and \pkg{ggplot2} packages \citep{r-cowplot, Wickham2016-gz}.
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Amezquita2020-at}{}%
-Amezquita, Robert A, Aaron T L Lun, Etienne Becht, Vince J Carey, Lindsay N Carpp, Ludwig Geistlinger, Federico Marini, et al.
-2020.
-``Orchestrating Single-Cell Analysis with Bioconductor.'' \emph{Nat.
-Methods} 17 (2): 137--45.
-\url{https://doi.org/10.1038/s41592-019-0654-x}.
-
-\leavevmode\hypertarget{ref-Asimov1985-vp}{}%
-Asimov, Daniel. 1985. ``The Grand Tour: A Tool for Viewing Multidimensional Data.'' \emph{SIAM J. Sci. And Stat. Comput.} 6 (1): 128--43. \url{https://doi.org/10.1137/0906011}.
-
-\leavevmode\hypertarget{ref-Becker1987-gd}{}%
-Becker, Richard A, and William S Cleveland. 1987. ``Brushing Scatterplots.'' \emph{Technometrics} 29 (2): 127--42. \url{https://doi.org/10.2307/1269768}.
-
-\leavevmode\hypertarget{ref-fnn-pkg}{}%
-Beygelzimer, Alina, Sham Kakadet, John Langford, Sunil Arya, David Mount, and Shengqiao Li.
-2019.
-\emph{FNN: Fast Nearest Neighbor Search Algorithms and Applications}.
-\url{https://CRAN.R-project.org/package=FNN}.
-
-\leavevmode\hypertarget{ref-Blondel2008-bx}{}%
-Blondel, Vincent D, Jean-Loup Guillaume, Renaud Lambiotte, and Etienne Lefebvre.
-2008.
-``Fast Unfolding of Communities in Large Networks.'' \emph{J.
-Stat.
-Mech.} 2008 (10): P10008.
-\url{https://doi.org/10.1088/1742-5468/2008/10/P10008}.
-
-\leavevmode\hypertarget{ref-Brehmer2014-hk}{}%
-Brehmer, Matthew, Michael Sedlmair, Stephen Ingram, and Tamara Munzner.
-2014.
-``Visualizing Dimensionally-Reduced Data: Interviews with Analysts and a Characterization of Task Sequences.'' In \emph{Proceedings of the Fifth Workshop on Beyond Time and Errors: Novel Evaluation Methods for Visualization}, 1--8.
-dl.acm.org.
-\url{https://dl.acm.org/doi/abs/10.1145/2669557.2669559}.
-
-\leavevmode\hypertarget{ref-Buja1986-zr}{}%
-Buja, Andreas, and Daniel Asimov.
-1986.
-``Grand Tour Methods: An Outline.'' In \emph{Proceedings of the Seventeenth Symposium on the Interface of Computer Sciences and Statistics on Computer Science and Statistics}, 63--67.
-USA: Elsevier North-Holland, Inc.
-\url{https://dl.acm.org/doi/10.5555/26036.26046}.
-
-\leavevmode\hypertarget{ref-Buja1996-fk}{}%
-Buja, Andreas, Dianne Cook, and Deborah F Swayne. 1996. ``Interactive High-Dimensional Data Visualization.'' \emph{J. Comput. Graph. Stat.} 5 (1): 78--99. \url{https://doi.org/10.2307/1390754}.
-
-\leavevmode\hypertarget{ref-Buja1986-ku}{}%
-Buja, Andreas, Catherine Hurley, and John McDonald.
-1986.
-``A Data Viewer for Multivariate Data.'' In \emph{Computing Science and Statistics: Proceedings of the 18th Symposium on the Interface}, 171--74.
-Washington: American Statistical Association.
-
-\leavevmode\hypertarget{ref-Buja2008-fn}{}%
-Buja, Andreas, Deborah F Swayne, Michael L Littman, Nathaniel Dean, Heike Hofmann, and Lisha Chen.
-2008.
-``Data Visualization with Multidimensional Scaling.'' \emph{J.
-Comput.
-Graph.
-Stat.} 17 (2): 444--72.
-\url{https://doi.org/10.1198/106186008X318440}.
-
-\leavevmode\hypertarget{ref-Chang2020-bq}{}%
-Chang, Winston, Joe Cheng, J J Allaire, Yihui Xie, and Jonathan McPherson. 2020. ``Shiny: Web Application Framework for R.'' \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\hypertarget{ref-Coenen2019-to}{}%
-Coenen, Andy, and Adam Pearce. 2019. ``Understanding UMAP.'' \url{https://pair-code.github.io/understanding-umap/}. \url{https://pair-code.github.io/understanding-umap/}.
-
-\leavevmode\hypertarget{ref-Coifman2005-ak}{}%
-Coifman, R R, S Lafon, A B Lee, M Maggioni, B Nadler, F Warner, and S W Zucker.
-2005.
-``Geometric Diffusions as a Tool for Harmonic Analysis and Structure Definition of Data: Diffusion Maps.'' \emph{Proc.
-Natl.
-Acad.
-Sci.
-U.
-S.
-A.} 102 (21): 7426--31.
-\url{https://doi.org/10.1073/pnas.0500334102}.
-
-\leavevmode\hypertarget{ref-Cook1995-bi}{}%
-Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley.
-1995.
-``Grand Tour and Projection Pursuit.'' \emph{J.
-Comput.
-Graph.
-Stat.} 4 (3): 155--72.
-\url{https://doi.org/10.1080/10618600.1995.10474674}.
-
-\leavevmode\hypertarget{ref-rcpp-annoy-pkg}{}%
-Eddelbuettel, Dirk. 2020. \emph{RcppAnnoy: 'Rcpp' Bindings for 'Annoy', a Library for Approximate Nearest Neighbors}. \url{https://CRAN.R-project.org/package=RcppAnnoy}.
-
-\leavevmode\hypertarget{ref-Genovese2017-iq}{}%
-Genovese, Christopher, Marco Perone-Pacifico, Isabella Verdinelli, and Larry Wasserman.
-2017.
-``Finding Singular Features.'' \emph{J.
-Comput.
-Graph.
-Stat.} 26 (3): 598--609.
-\url{https://doi.org/10.1080/10618600.2016.1260472}.
-
-\leavevmode\hypertarget{ref-Kobak2019-lm}{}%
-Kobak, Dmitry, and Philipp Berens. 2019. ``The Art of Using t-SNE for Single-Cell Transcriptomics.'' \emph{Nat. Commun.} 10 (1): 5416. \url{https://doi.org/10.1038/s41467-019-13056-x}.
-
-\leavevmode\hypertarget{ref-Korem2015-af}{}%
-Korem, Yael, Pablo Szekely, Yuval Hart, Hila Sheftel, Jean Hausser, Avi Mayo, Michael E Rothenberg, Tomer Kalisky, and Uri Alon.
-2015.
-``Geometry of the Gene Expression Space of Individual Cells.'' \emph{PLoS Comput.
-Biol.} 11 (7): e1004224.
-\url{https://doi.org/10.1371/journal.pcbi.1004224}.
-
-\leavevmode\hypertarget{ref-Rtsne}{}%
-Krijthe, Jesse H. 2015. \emph{Rtsne: T-Distributed Stochastic Neighbor Embedding Using Barnes-Hut Implementation}. \url{https://github.com/jkrijthe/Rtsne}.
-
-\leavevmode\hypertarget{ref-Kruskal1964-cz}{}%
-Kruskal, J B. 1964a. ``Multidimensional Scaling by Optimizing Goodness of Fit to a Nonmetric Hypothesis.'' \emph{Psychometrika} 29 (1): 1--27. \url{https://doi.org/10.1007/BF02289565}.
-
-\leavevmode\hypertarget{ref-Kruskal1964-cw}{}%
----------. 1964b. ``Nonmetric Multidimensional Scaling: A Numerical Method.'' \emph{Psychometrika} 29 (2): 115--29. \url{https://doi.org/10.1007/BF02289694}.
-
-\leavevmode\hypertarget{ref-Laa2020-wr}{}%
-Laa, Ursula, Dianne Cook, Andreas Buja, and German Valencia.
-2020.
-``Hole or Grain?
-A Section Pursuit Index for Finding Hidden Structure in Multiple Dimensions,'' April.
-\url{http://arxiv.org/abs/2004.13327}.
-
-\leavevmode\hypertarget{ref-Laa2020-vk}{}%
-Laa, Ursula, Dianne Cook, and Stuart Lee. 2020. ``Burning Sage: Reversing the Curse of Dimensionality in the Visualization of High-Dimensional Data,'' September. \url{http://arxiv.org/abs/2009.10979}.
-
-\leavevmode\hypertarget{ref-Lee2009-zb}{}%
-Lee, John A, and Michel Verleysen.
-2009.
-``Quality Assessment of Dimensionality Reduction: Rank-Based Criteria.'' \emph{Neurocomputing} 72 (7): 1431--43.
-\url{https://doi.org/10.1016/j.neucom.2008.12.017}.
-
-\leavevmode\hypertarget{ref-r-liminal}{}%
-Lee, Stuart, and Dianne Cook. 2020. \emph{Liminal: Multivariate Data Visualization with Tours and Embeddings}. \url{https://github.com/sa-lee/liminal}.
-
-\leavevmode\hypertarget{ref-Lewis2012-ai}{}%
-Lewis, Joshua, Laurens Van der Maaten, and Virginia de Sa.
-2012.
-``A Behavioral Investigation of Dimensionality Reduction.'' In \emph{Proceedings of the Annual Meeting of the Cognitive Science Society}.
-Vol.
-34.
-
-\leavevmode\hypertarget{ref-Linderman2019-dq}{}%
-Linderman, George C, and Stefan Steinerberger. 2019. ``Clustering with t-SNE, Provably.'' \emph{SIAM Journal on Mathematics of Data Science} 1 (2): 313--32. \url{https://doi.org/10.1137/18M1216134}.
-
-\leavevmode\hypertarget{ref-Lun2016}{}%
-Lun, Aaron T.
-L., Davis J.
-McCarthy, and John C.
-Marioni.
-2016.
-``A Step-by-Step Workflow for Low-Level Analysis of Single-Cell Rna-Seq Data with Bioconductor.'' \emph{F1000Res.} 5: 2122.
-\url{https://doi.org/10.12688/f1000research.9501.2}.
-
-\leavevmode\hypertarget{ref-Lyttle2020-hp}{}%
-Lyttle, Ian, and Vega/Vega-Lite Developers. 2020. ``Vegawidget: 'Htmlwidget' for 'Vega' and 'Vega-Lite'.'' \url{https://CRAN.R-project.org/package=vegawidget}.
-
-\leavevmode\hypertarget{ref-Van_Der_Maaten2014-zn}{}%
-Maaten, Laurens van der. 2014. ``Accelerating t-SNE Using Tree-Based Algorithms.'' \emph{J. Mach. Learn. Res.} \url{http://www.jmlr.org/papers/volume15/vandermaaten14a/vandermaaten14a.pdf}.
-
-\leavevmode\hypertarget{ref-Maaten2008-sk}{}%
-Maaten, Laurens van der, and Geoffrey Hinton. 2008. ``Visualizing Data Using t-SNE.'' \emph{J. Mach. Learn. Res.} 9 (Nov): 2579--2605. \url{http://www.jmlr.org/papers/v9/vandermaaten08a.html}.
-
-\leavevmode\hypertarget{ref-Macosko2015-ot}{}%
-Macosko, Evan Z, Anindita Basu, Rahul Satija, James Nemesh, Karthik Shekhar, Melissa Goldman, Itay Tirosh, et al.
-2015.
-``Highly Parallel Genome-Wide Expression Profiling of Individual Cells Using Nanoliter Droplets.'' \emph{Cell} 161 (5): 1202--14.
-\url{https://doi.org/10.1016/j.cell.2015.05.002}.
-
-\leavevmode\hypertarget{ref-McCarthy2017}{}%
-McCarthy, Davis J., Kieran R.
-Campbell, Aaron T.
-L.
-Lun, and Quin F.
-Willis.
-2017.
-``Scater: Pre-Processing, Quality Control, Normalisation and Visualisation of Single-Cell RNA-Seq Data in R.'' \emph{Bioinformatics} 33 (8): 1179--86.
-\url{https://doi.org/10.1093/bioinformatics/btw777}.
-
-\leavevmode\hypertarget{ref-McCarthy2009-qx}{}%
-McCarthy, Davis J, and Gordon K Smyth.
-2009.
-``Testing Significance Relative to a Fold-Change Threshold Is a TREAT.'' \emph{Bioinformatics} 25 (6): 765--71.
-\url{https://doi.org/10.1093/bioinformatics/btp053}.
-
-\leavevmode\hypertarget{ref-McDonald1982-ew}{}%
-McDonald, J A. 1982. ``Interactive Graphics for Data Analysis.'' PhD thesis, Stanford University. \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.331.6673\&rep=rep1\&type=pdf}.
-
-\leavevmode\hypertarget{ref-McInnes2018-co}{}%
-McInnes, Leland, John Healy, and James Melville. 2018. ``UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction,'' February. \url{http://arxiv.org/abs/1802.03426}.
-
-\leavevmode\hypertarget{ref-Melville2020}{}%
-Melville, James. 2020. ``t-SNE Initialization Options.'' \url{https://jlmelville.github.io/smallvis/init.html}. \url{https://jlmelville.github.io/smallvis/init.html}.
-
-\leavevmode\hypertarget{ref-Moon2019-ce}{}%
-Moon, Kevin R, David van Dijk, Zheng Wang, Scott Gigante, Daniel B Burkhardt, William S Chen, Kristina Yim, et al.
-2019.
-``Visualizing Structure and Transitions for Biological Data Exploration.'' \emph{bioRxiv}.
-\url{https://doi.org/10.1101/120378}.
-
-\leavevmode\hypertarget{ref-Nguyen2019-yh}{}%
-Nguyen, Lan Huong, and Susan Holmes. 2019. ``Ten Quick Tips for Effective Dimensionality Reduction.'' \emph{PLoS Comput. Biol.} 15 (6): e1006907. \url{https://doi.org/10.1371/journal.pcbi.1006907}.
-
-\leavevmode\hypertarget{ref-Ovchinnikova2020-sy}{}%
-Ovchinnikova, Svetlana, and Simon Anders. 2020. ``Exploring Dimension-Reduced Embeddings with Sleepwalk.'' \emph{Genome Res.} 30 (5): 749--56. \url{https://doi.org/10.1101/gr.251447.119}.
-
-\leavevmode\hypertarget{ref-Pezzotti2017-cz}{}%
-Pezzotti, Nicola, Boudewijn P F Lelieveldt, Laurens Van Der Maaten, Thomas Hollt, Elmar Eisemann, and Anna Vilanova.
-2017.
-``Approximated and User Steerable tSNE for Progressive Visual Analytics.'' \emph{IEEE Trans.
-Vis.
-Comput.
-Graph.} 23 (7): 1739--52.
-\url{https://doi.org/10.1109/TVCG.2016.2570755}.
-
-\leavevmode\hypertarget{ref-Rauber2009-vh}{}%
-Rauber, Andreas. 2009. ``Multi-Challenge Data Set.'' \url{http://ifs.tuwien.ac.at/dm/dataSets.html}. \url{http://ifs.tuwien.ac.at/dm/dataSets.html}.
-
-\leavevmode\hypertarget{ref-Rieck2017-kk}{}%
-Rieck, Bastian. 2017. ``Persistent Homology in Multivariate Data Visualization.'' Edited by Heike Little and Michael Gertz. PhD thesis, Ruprecht-Karls-Universität Heidelberg.
-
-\leavevmode\hypertarget{ref-scRNAseq-d}{}%
-Risso, Davide, and Michael Cole. 2019. \emph{ScRNAseq: Collection of Public Single-Cell Rna-Seq Datasets}.
-
-\leavevmode\hypertarget{ref-Satyanarayan2017-gs}{}%
-Satyanarayan, Arvind, Dominik Moritz, Kanit Wongsuphasawat, and Jeffrey Heer.
-2017.
-``Vega-Lite: A Grammar of Interactive Graphics.'' \emph{IEEE Trans.
-Vis.
-Comput.
-Graph.} 23 (1): 341--50.
-\url{https://doi.org/10.1109/TVCG.2016.2599030}.
-
-\leavevmode\hypertarget{ref-Sedlmair2013-pn}{}%
-Sedlmair, Michael, Tamara Munzner, and Melanie Tory.
-2013.
-``Empirical Guidance on Scatterplot and Dimension Reduction Technique Choices.'' \emph{IEEE Trans.
-Vis.
-Comput.
-Graph.} 19 (12): 2634--43.
-\url{https://doi.org/10.1109/TVCG.2013.153}.
-
-\leavevmode\hypertarget{ref-Silva2003-xw}{}%
-Silva, V D, and J B Tenenbaum.
-2003.
-``Global Versus Local Methods in Nonlinear Dimensionality Reduction.'' \emph{Adv.
-Neural Inf.
-Process.
-Syst.} \url{http://papers.nips.cc/paper/2141-global-versus-local-methods-in-nonlinear-dimensionality-reduction.pdf}.
-
-\leavevmode\hypertarget{ref-Smilkov2016-hp}{}%
-Smilkov, Daniel, Nikhil Thorat, Charles Nicholson, Emily Reif, Fernanda B Viégas, and Martin Wattenberg.
-2016.
-``Embedding Projector: Interactive Visualization and Interpretation of Embeddings,'' November.
-\url{http://arxiv.org/abs/1611.05469}.
-
-\leavevmode\hypertarget{ref-Swayne2004-ua}{}%
-Swayne, Deborah F, and Andreas Buja.
-2004.
-``Exploratory Visual Analysis of Graphs in GGOBI.'' In \emph{COMPSTAT 2004 --- Proceedings in Computational Statistics}, 477--88.
-Physica-Verlag HD.
-\url{https://doi.org/10.1007/978-3-7908-2656-2/_39}.
-
-\leavevmode\hypertarget{ref-Swayne1998-uq}{}%
-Swayne, Deborah F, Dianne Cook, and Andreas Buja.
-1998.
-``XGobi: Interactive Dynamic Data Visualization in the X Window System.'' \emph{J.
-Comput.
-Graph.
-Stat.} 7 (1): 113--30.
-\url{https://doi.org/10.1080/10618600.1998.10474764}.
-
-\leavevmode\hypertarget{ref-Swayne2003-qd}{}%
-Swayne, Deborah F, Duncan Temple Lang, Andreas Buja, and Dianne Cook.
-2003.
-``GGobi: Evolving from XGobi into an Extensible Framework for Interactive Data Visualization.'' \emph{Comput.
-Stat.
-Data Anal.} 43 (4): 423--44.
-\url{https://doi.org/10.1016/S0167-9473(02)00286-4}.
-
-\leavevmode\hypertarget{ref-Tang2016-oz}{}%
-Tang, Jian, Jingzhou Liu, Ming Zhang, and Qiaozhu Mei. 2016. ``Visualizing Large-Scale and High-Dimensional Data,'' February. \url{http://arxiv.org/abs/1602.00370}.
-
-\leavevmode\hypertarget{ref-Torgerson1952-am}{}%
-Torgerson, Warren S. 1952. ``Multidimensional Scaling: I. Theory and Method.'' \emph{Psychometrika} 17 (4): 401--19. \url{https://doi.org/10.1007/BF02288916}.
-
-\leavevmode\hypertarget{ref-Wattenberg2016-ji}{}%
-Wattenberg, Martin, Fernanda Viégas, and Ian Johnson. 2016. ``How to Use t-SNE Effectively.'' \emph{Distill} 1 (10). \url{https://doi.org/10.23915/distill.00002}.
-
-\leavevmode\hypertarget{ref-Wickham2016-gz}{}%
-Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data Analysis}. Use R! Springer International Publishing. \url{https://doi.org/10.1007/978-3-319-24277-4}.
-
-\leavevmode\hypertarget{ref-Wickham2015-cx}{}%
-Wickham, Hadley, Dianne Cook, and Heike Hofmann.
-2015.
-``Visualizing Statistical Models: Removing the Blindfold.'' \emph{Statistical Analysis and Data Mining: The ASA Data Science Journal} 8 (4): 203--25.
-\url{https://doi.org/10.1002/sam.11271}.
-
-\leavevmode\hypertarget{ref-Wickham2011-st}{}%
-Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja.
-2011.
-``Tourr: An R Package for Exploring Multivariate Data with Projections.'' \emph{Journal of Statistical Software, Articles} 40 (2): 1--18.
-\url{https://doi.org/10.18637/jss.v040.i02}.
-
-\leavevmode\hypertarget{ref-r-cowplot}{}%
-Wilke, Clause O. 2019. \emph{Cowplot: Streamlined Plot Theme and Plot Annotations for 'Ggplot2'}. \url{https://CRAN.R-project.org/package=cowplot}.
+\renewcommand\refname{References}
+\bibliography{liminal.bib}
 

### 2020-12-11 liminal.Rmd
add arxiv version

@@
   ~key, ~url, ~path,
-  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("img", "liminal-screenshot-gaussian.png"),
-  "hierarchical", "https://player.vimeo.com/video/439635905", here::here("img", "liminal-screenshot-hierarchical.png"),
-  "trees-01", "https://player.vimeo.com/video/439635892", here::here("img", "liminal-screenshot-trees-01.png"),
-  "trees-02", "https://player.vimeo.com/video/439635863", here::here("img", "liminal-screenshot-trees-02.png"),
-  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("img", "liminal-screenshot-mouse-01.png"),
-  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("img", "liminal-screenshot-mouse-02.png")
+  "gaussian",  "https://player.vimeo.com/video/439635921", "./img/liminal-screenshot-gaussian.png",
+  "hierarchical", "https://player.vimeo.com/video/439635905", "./img/liminal-screenshot-hierarchical.png",
+  "trees-01", "https://player.vimeo.com/video/439635892", "./img/liminal-screenshot-trees-01.png",
+  "trees-02", "https://player.vimeo.com/video/439635863","./img/liminal-screenshot-trees-02.png",
+  "mouse-01", "https://player.vimeo.com/video/439635812","img/liminal-screenshot-mouse-01.png",
+  "mouse-02", "https://player.vimeo.com/video/439635843", "./img/liminal-screenshot-mouse-02.png"
 )
@@
 ```{r fake-trees, echo = FALSE, fig.align = "center", fig.cap="(ref:fake-trees)"}
-knitr::include_graphics(here::here("img", "tree-view.png"))
+knitr::include_graphics("./img/tree-view.png")
 ```

### 2020-12-11 liminal.tex
add arxiv version

@@
 
-{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-gaussian}
+{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{./img/liminal-screenshot-gaussian}
 
@@
 
-{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-hierarchical}
+{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{./img/liminal-screenshot-hierarchical}
 
@@
 
-{\centering \includegraphics[width=\textwidth]{/Users/slee0046/paper-liminal/img/tree-view}
+{\centering \includegraphics[width=\textwidth]{./img/tree-view}
 
@@
 
-{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-trees-01}
+{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{./img/liminal-screenshot-trees-01}
 
@@
 
-{\centering \includegraphics[width=\textwidth]{/Users/slee0046/paper-liminal/img/liminal-screenshot-trees-02}
+{\centering \includegraphics[width=\textwidth]{./img/liminal-screenshot-trees-02}
 
@@
 
-{\centering \includegraphics[width=\textwidth,height=0.9\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-mouse-01}
+{\centering \includegraphics[width=\textwidth,height=0.9\textheight]{img/liminal-screenshot-mouse-01}
 
@@
 
-{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{/Users/slee0046/paper-liminal/img/liminal-screenshot-mouse-02}
+{\centering \includegraphics[width=\textwidth,height=0.75\textheight]{./img/liminal-screenshot-mouse-02}
 

### 2021-06-02 reviews/response.md
responses

Dear editor and reviewers,

We would like to thank for your thorough reviews of our manuscript and
appreciate your constructive criticisms.

In the following response, I would like to address each of your comments
and highlight the amendments we have made to the manuscript and the liminal software.

# Reviewer A

> Unfortunately, I was not able to run the provided code. My guess is that there is some incompatibility with the newest R-version 4-0-3. Please check!

This was a problem with a newer version of the tourr package causing liminal
to not animate the tour correctly. It has been fixed and the package is now 
available on CRAN and has been tested against several versions of R.

# Reviewer B

> For example, the various clusters came pre-colored.  I would like to see a little more about how the analyst would identify these clusters, especially in a case such as your tree example where it doesn't appear that you would have the coloring/clustering you have without a priori knowledge of the different branches.  Is it possible to create that clustering starting from the tour or the embedding?

> Would it be worthwhile to link a pairwise scatterplot matrix for all or a subset of the original variables or components so as to be able to relate clusters and other features via brushing back to those variables for interpretation?  I guess the biplot in Figure 7 may be intended to show something similar?  Is it possible to expand and clarify your results from this figure a bit for those of us who are not genomics specialists?

### 2021-06-03 liminal.Rmd
changes

@@
 matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
-the number of observations in $p$ dimensions.  The purpose of any DR algorithm
+the number of observations in $p$ dimensions.  The purpose of any dimension reduction (DR) algorithm
 is to find a low-dimensional representation of the data, 
@@
 estimated by modelling similarities using a t-distribution with 1 degree of
-freedom [@Maaten2008-sk]. There are several subtleties
-to the to use of the algorithm that are revealed by stepping through its
-machinery.
+freedom [@Maaten2008-sk]. There are several subtleties of the algorithm that 
+are revealed by stepping through its machinery.
 
@@
 interesting projection or is done with the tour, the interface will return the 
-current basis for use downstream.
+current basis (as a matrix) for use downstream.
 
@@
 far from each other in the tour view (Figure \@ref(fig:hierarchical)a). We see 
-in this case that is difficult
+in this case that it is difficult
 to see the sub-clusters in the tour view, however, once we zoom and highlight
@@
 of clusters and developmental trajectories (i.e., case study 1). For clustering
-workflows the primary task is verify the existence of clusters and then begin to 
+workflows the primary task is to verify the existence of clusters and then begin to 
 identify the clusters as cell types using the expression of "known" 
@@
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the **scater** package
+(described by @Amezquita2020-at): we performed quality control using the **scater** package
 by removing cells with high proportion of mitochondrial gene expression and low

### 2021-06-28 liminal.Rmd
changes

@@
 By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
-We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
-The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.
+We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks and for correcting an NLDR method.
+The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available on CRAN.
 keywords:
@@
 
-High dimensional data is increasingly prevalent in the natural sciences and
+High-dimensional data is increasingly prevalent in the natural sciences and
 beyond but presents a challenge to the analyst in terms of data cleaning,
@@
 workflow where they are used to ascertain hidden structure and de-noise data
-for downstream analysis .
+for downstream analysis.
 
@@
 an analyst know whether the embedding has captured the underlying topology and
-geometry of the high dimensional space? The answer depends on the analyst's
+geometry of the high-dimensional space? The answer depends on the analyst's
 workflow. @Brehmer2014-hk characterised two main workflow steps that an
@@
 Both of these
-workflow steps rely on the embedding being representative of the original high dimensional dataset, and becomes much more difficult when there is no underlying ground truth.
+workflow steps rely on the embedding being representative of the original high-dimensional dataset, and becomes much more difficult when there is no underlying ground truth.
 
@@
 
-A complementary approach for visualizing structure in high dimensional data is 
-the tour. A tour is a sequence of projections of a high dimensional dataset
+A complementary approach for visualizing structure in high-dimensional data is 
+the tour. A tour is a sequence of projections of a high-dimensional dataset
 onto a low-dimensional basis matrix, and is represented as an
@@
 
-The approach used in this paper is to augment the results of an NLDR method with the tour with our  \proglang{R} package called \pkg{liminal}. Interfaces for evaluating
+The approach used in this paper is to augment the results of an NLDR method with the tour with our\proglang{R} package called \pkg{liminal}. Interfaces for evaluating
 embeddings require interaction but should also be able to be incorporated into
@@
 incorporating interactive graphics and tours with embeddings that allows users
-to see a global overview of their high dimensional data and assists them with cluster orientation tasks.
+to see a global overview of their high-dimensional data, allowing them to 
+adjust an NLDR view and assist them with cluster orientation tasks.
 
@@
 in using embedding methods. Finally, we describe the insights gained by using
-\pkg{liminal} and plans for extensions to the software.
+\pkg{liminal} and plans for extensions to the software. Throughout the paper
+when we refer to high-dimensional data, we mean it in a broad sense not 
+specifically referring to the case of small $N$ large $p$ data.
 
@@
 
-To investigate latent structure and the shape of a high dimensional dataset in
+To investigate latent structure and the shape of a high-dimensional dataset in
 \pkg{liminal}, a tour can be run without the use of an external embedding.
@@
 From the video linked in Figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
-clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
+clusters relative location to each other to their location in high-dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
 linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
@@
 We have shown that the use of tours as a tool for interacting with
-high dimensional data provides an additional insight for interrogating views
+high-dimensional data provides an additional insight for interrogating views
 generated from embeddings. The interface we have designed in the \pkg{liminal}

### 2021-06-29 reviews/response.md
changes
finalise response

@@
 This was a problem with a newer version of the tourr package causing liminal
-to not animate the tour correctly. It has been fixed and the package is now 
-available on CRAN and has been tested against several versions of R.
+to not animate the tour correctly. It has been fixed and the liminal package is now 
+available on CRAN and has been tested against several versions of R on multiple
+operating systems.
+
+## Major remarks
+
+> The manuscript is well written and presents the approach in a balanced way. Despite
+this, the goal and primary focus of the paper gets not always clear as it it switches focus
+between presenting an enhancement of NLDR methods, presenting a better alternative to
+NLDR methods for certain tasks (clustering) and describing the general functionality of
+interactions with grand tour representations. A revision should clarify the goals of the
+manuscript.
+> Secondly, the relevance of having a cluster orientation task at hand seems relevant for
+some arguments but not for others. Here also, the authors should clearly provide guidance
+to the reader on which data and research contexts are essential for certain conclusions.
+
+I have added comments to the abstract and throughout the manuscript
+to clarify that the goal of the liminal interface is to both correct an embedding
+(as shown in case study 2) and for clustering (case study 3). The discussion
+also clarifies the goals of the method.
+
+> Thirdly, the term high-dimensional data has gained fairly specfic meaning in some
+contexts. Please clarify whether you use the term in its rather broad sense or whether you
+want to allude particularly to small-N-large-p situations and their resulting conditions.
+
+This has been clarified in the introduction.
+
+> Fourthly, the manuscript has been enhanced by links to video sequences of tour ani-
+mations for the case studies. While these videos add a lot to the static description in the
+papers, additional explanations to the videos (voice-over or captions) would further ease
+the understanding of the animations and hence improve further the manuscript.
+
+Thank you for the suggestion, I have now added descriptions and
+captions and descriptions to all videos.
+
+> Minor remarks
+
+These have all been amended in the text.
 
@@
 
+This is possible, but we considered it out of scope for the paper to go into
+specifics of how one would perform a cluster analysis. 
+I've added some sentences to the main text about how one might go about using
+the interface to generate clusters manually..  The GGobi book by the last author of the manuscript details how one might spin and brush using a tour, to 
+build up a cluster analysis manually.
+
 > Would it be worthwhile to link a pairwise scatterplot matrix for all or a subset of the original variables or components so as to be able to relate clusters and other features via brushing back to those variables for interpretation?
@@
 Is it possible to expand and clarify your results from this figure a bit for those of us who are not genomics specialists?
+
+I have expanded what specific terms mean in the main text, to clarify the purposes
+of this type of analysis and the graphic. Essentially marker genes are of interest
+to biologists for determining signatures that represent cells. The biplot
+on this view was intended to show how the relative expression of each gene could
+differentiate between possible cell types (clusters). We didn't use a
+scatterplot matrix here because it would be difficult to read with the large number
+of variables. As the liminal interface is a shiny app, it could be hooked up to
+other visualisations by the user.
+
+Yours Sincerely,
+
+Dr Stuart Lee

### 2021-06-29 liminal.Rmd
update main text and email

@@
       | Monash University
-    email: \email{stuart.lee1@monash.edu}
+    email: \email{stuart.andrew.lee@gmail.com}
   - name: Ursula Laa 
@@
 abstract: >
-  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
-We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
+  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complexity; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
+We present visual diagnostics for the usage of NLDR methods by combining them with a technique called the tour.
 A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
@@
 By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
-We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks and for correcting an NLDR method.
-The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available on CRAN.
+We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks and for correcting an NLDR embedding.
+The implementation of our framework is available as an \proglang{R} package called \pkg{liminal}.
 keywords:
@@
 (described by @Amezquita2020-at): we performed quality control using the **scater** package
-by removing cells with high proportion of mitochondrial gene expression and low
+by removing cells with high proportion of mitochondrial gene expression (as this indicates poor sample quality) and low
 numbers of genes detected, we log-transformed and normalised the expression 
@@
 If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.
+One could perform this approach manually via the \pkg{liminal} interface by returning the regions identified by brushing on the tour or embedding view.
 
@@
 uses the testing framework from @McCarthy2009-qx. We select the top 10 marker 
-genes that are upregulated in cluster 2, which was one of the clumped clusters
+genes that are expressed in relatively larger quantities (upregulated) in cluster 2, which was one of the clumped clusters
 when we toured on principal components. Here the tour becomes an alternative

### 2021-06-29 liminal.tex
update build

@@
 \Abstract{
-Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
-We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
+Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complexity; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
+We present visual diagnostics for the usage of NLDR methods by combining them with a technique called the tour.
 A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
@@
 By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading.
-We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks.
-The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.
+We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks and for correcting an NLDR embedding.
+The implementation of our framework is available as an \proglang{R} package called \pkg{liminal}.
 }
@@
 Monash University\\
-  E-mail: \email{stuart.lee1@monash.edu}\\
+  E-mail: \email{stuart.andrew.lee@gmail.com}\\
 
@@
 an analyst know whether the embedding has captured the underlying topology and
-geometry of the high dimensional space? The answer depends on the analyst's
+geometry of the high-dimensional space? The answer depends on the analyst's
 workflow. \citet{Brehmer2014-hk} characterised two main workflow steps that an
@@
 
-A complementary approach for visualizing structure in high dimensional data is
-the tour. A tour is a sequence of projections of a high dimensional dataset
+A complementary approach for visualizing structure in high-dimensional data is
+the tour. A tour is a sequence of projections of a high-dimensional dataset
 onto a low-dimensional basis matrix, and is represented as an
@@
 
-The approach used in this paper is to augment the results of an NLDR method with the tour with our \proglang{R} package called \pkg{liminal}. Interfaces for evaluating
+The approach used in this paper is to augment the results of an NLDR method with the tour with our\proglang{R} package called \pkg{liminal}. Interfaces for evaluating
 embeddings require interaction but should also be able to be incorporated into
@@
 incorporating interactive graphics and tours with embeddings that allows users
-to see a global overview of their high dimensional data and assists them with cluster orientation tasks.
+to see a global overview of their high-dimensional data, allowing them to
+adjust an NLDR view and assist them with cluster orientation tasks.
 
@@
 in using embedding methods. Finally, we describe the insights gained by using
-\pkg{liminal} and plans for extensions to the software.
+\pkg{liminal} and plans for extensions to the software. Throughout the paper
+when we refer to high-dimensional data, we mean it in a broad sense not
+specifically referring to the case of small \(N\) large \(p\) data.
 
@@
 matrix of real numbers, \(X = [\mathbf{x_1}, \dots, \mathbf{x_n}]\), where \(n\) is
-the number of observations in \(p\) dimensions. The purpose of any DR algorithm
+the number of observations in \(p\) dimensions. The purpose of any dimension reduction (DR) algorithm
 is to find a low-dimensional representation of the data,
@@
 estimated by modelling similarities using a t-distribution with 1 degree of
-freedom \citep{Maaten2008-sk}. There are several subtleties
-to the to use of the algorithm that are revealed by stepping through its
-machinery.
+freedom \citep{Maaten2008-sk}. There are several subtleties of the algorithm that
+are revealed by stepping through its machinery.
 
@@
 interesting projection or is done with the tour, the interface will return the
-current basis for use downstream.
+current basis (as a matrix) for use downstream.
 
@@
 far from each other in the tour view (Figure \ref{fig:hierarchical}a). We see
-in this case that is difficult
+in this case that it is difficult
 to see the sub-clusters in the tour view, however, once we zoom and highlight
@@
 of clusters and developmental trajectories (i.e., case study 1). For clustering
-workflows the primary task is verify the existence of clusters and then begin to
+workflows the primary task is to verify the existence of clusters and then begin to
 identify the clusters as cell types using the expression of ``known''
@@
 have followed a standard workflow for pre-processing and normalizing this data
-(described by \citet{Amezquita2020-at}): we performed QC using the \textbf{scater} package
-by removing cells with high proportion of mitochondrial gene expression and low
+(described by \citet{Amezquita2020-at}): we performed quality control using the \textbf{scater} package
+by removing cells with high proportion of mitochondrial gene expression (as this indicates poor sample quality) and low
 numbers of genes detected, we log-transformed and normalised the expression
@@
 From the video linked in Figure \ref{fig:mouse-01}, we learn that the embedding has mostly captured the
-clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
+clusters relative location to each other to their location in high-dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
 linked brushing (Figure \ref{fig:mouse-01}a). As expected, t-SNE mitigates the crowding problem that is
@@
 uses the testing framework from \citet{McCarthy2009-qx}. We select the top 10 marker
-genes that are upregulated in cluster 2, which was one of the clumped clusters
+genes that are expressed in relatively larger quantities (upregulated) in cluster 2, which was one of the clumped clusters
 when we toured on principal components. Here the tour becomes an alternative

### 2022-02-26 liminal.Rmd
copy edits

@@
 abstract: >
-  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complexity; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
+  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences.
+However, the appropriate use of these methods is difficult because of their complexity; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
 We present visual diagnostics for the usage of NLDR methods by combining them with a technique called the tour.
@@
 ```
+\newpage
+
 # Introduction
@@
 beyond but presents a challenge to the analyst in terms of data cleaning,
-pre-processing and visualisation. Methods to embed data from a high-dimensional
+pre-processing and visualization. Methods to embed data from a high-dimensional
 space into a low-dimensional one now form a core step of the data analysis
@@
 geometry of the high-dimensional space? The answer depends on the analyst's
-workflow. @Brehmer2014-hk characterised two main workflow steps that an
+workflow. @Brehmer2014-hk characterized two main workflow steps that an
 analyst performs when using embedding techniques: dimension reduction and
 cluster orientation. The first relates to dimension reduction achieved by using
-an embedding method, here an analyst wants to characterise and map meaning onto
+an embedding method, here an analyst wants to characterize and map meaning onto
 the embedded form, for example identifying batch effects from a high throughput
@@
 a clustering workflow. Here analysts are interested in identifying and naming
-clusters and verifying them by either applying known labels or colouring by
+clusters and verifying them by either applying known labels or coloring by
 variables that are a-priori known to distinguish clusters like the expression of a marker gene to identify a cell type.
@@
 and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
-scatter plots were mostly sufficient for detecting class separation, however
+scatter plots were mostly sufficient for detecting class separation, however,
 they also noted that often multiple embeddings were required. For the task of
@@
 
-There is no one-size fits all: finding an appropriate embedding for a given
+There is no one-size fits all; finding an appropriate embedding for a given
 dataset is a difficult and a somewhat poorly defined problem. For non-linear
 methods, there are a lot of parameters to explore that can have an effect on
-the resulting visualisation and interpretation. 
+the resulting visualization and interpretation. 
 While there has been much work on the algorithmic details of
-embedding methods, there are relatively few tools designed to assist users
+embedding methods; there are relatively few tools designed to assist users
 to interact with these techniques: when is an embedding sufficient for the
@@
 interactive interfaces have been proposed for evaluating or using embedding
-techniques. @Buja2008-fn used tours to guide analysts during the optimisation
-of multidimensional scaling methods by extending their interactive visualisation
+techniques. @Buja2008-fn used tours to guide analysts during the optimization
+of multidimensional scaling methods by extending their interactive visualization
 software called \pkg{XGobi} and \pkg{GGobi} into a new tool called \pkg{GGvis} 
@@
 non-linear embeddings in single cell RNA-seq data. It provides a click and 
-highlight visualisation for colouring points in an embedding according to an 
+highlight visualization for coloring points in an embedding according to an 
 estimated pairwise distance in the original high-dimensional space.
 Similarly, the \pkg{TensorFlow}  embedding projector is a web interface to 
-running some non-liner embedding methods live in the browser and 
-provides interactions to colour points, and select nearest neighbours
+running some non-linear embedding methods live in the browser and it
+provides interactions to color points, and select nearest neighbors
 [@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
 guided and modified form of the  t-SNE algorithm, that allows users to modify 
-optimisation parameters in real-time.
+optimization parameters in real-time.
 
@@
 nature of the tour, user interaction is  important for controlling and 
-exploring the visualisation: the tour has been used previously by 
+exploring the visualization; the tour has been used previously by 
 @Wickham2015-cx for exploring statistical model fits and by @Buja1996-fk for 
@@
 
-The rest of the paper is organised as follows. The next section provides
+The rest of the paper is organized as follows. The next section provides
 background on dimension reduction methods, including an overview of the tour.
@@
 
-To begin we suppose the data is in the form of a rectangular
+To begin, we suppose the data is in the form of a rectangular
 matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
@@
 matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
-produces $Y$ will remove noise in the original dataset while retaining any 
+produce $Y$ will remove noise in the original dataset while retaining any 
 latent structure.
@@
 $Y$ is a linear transformation of $X$. One example is principal components
-analysis (PCA) which performs an eigendecomposition of the estimated sample
+analysis (PCA) which performs an eigen-decomposition of the estimated sample
 covariance matrix. The eigenvalues are sorted in decreasing order and represent
@@
 cut-off.
-When working with linear transformations we often need more than two dimensions
+When working with linear transformations, we often need more than two dimensions
 to capture the latent structure. In this case we can use tour methods
@@
 
-For non-linear methods $Y$ is generated via a pre-processed form of the input
-$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
+For non-linear methods, $Y$ is generated via a pre-processed form of the input
+$X$ such as the $k$-nearest neighbors graph or via a kernel transformation.
 Multidimensional scaling (MDS) is a class of DR methods that aims to construct
@@
 distances with ranks instead [@Kruskal1964-cw]. 
-A related technique is Isomap which uses a $k$-nearest neighbour graph
+A related technique is Isomap which uses a $k$-nearest neighbor graph
 to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
@@
 non-linear DR methods for exploratory data analysis is selecting and tuning 
-appropriate parameters. For concreteness here we focus on t-distributed
-stochastic neighbour embedding (t-SNE), and we will examine
+appropriate parameters. For concreteness here, we focus on t-distributed
+stochastic neighbor embedding (t-SNE), and we will examine
 its underpinning in some detail below [@Maaten2008-sk]. Similar considerations hold for related
@@
 2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 /
-2\sigma_i^2)} 
+2\sigma_i^2)}. 
 $$
 
-The conditional probabilities are then normalised and symmetrised to form a
+The conditional probabilities are then normalized and symmetrized to form a
 joint probability distribution via averaging:
 
-$$ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n} $$
+$$ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n}. $$
 
@@
 
-$$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
+$$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})).$$
 
@@
 $\sigma^2_{i} \rightarrow \infty$. This essentially turns t-SNE into a
-nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all
+nearest neighbors algorithm, $p_{i|j}$ will be close to zero for all
 observations that are not in the $\mathcal{O}(\text{perplexity}_i)$
-neighbourhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].
+neighborhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].
 
 Next, in the target low-dimensional space, $Y$, pair-wise distances between
-$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
+$\mathbf{y_i}$ and $\mathbf{y_j}$ are modeled as a symmetric probability
 distribution using a t-distribution with one degree of freedom (Cauchy kernel):
@@
 $$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log
-\frac{p_{ij}}{q_{ij}}$$
+\frac{p_{ij}}{q_{ij}}.$$
 
@@
 $$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} +
-\log\sum_{i \ne j} w_{ij} $$
+\log\sum_{i \ne j} w_{ij}. $$
 
-Minimising  the loss function corresponds to large attractive
+Minimizing  the loss function corresponds to large attractive
 forces, that is, the pair-wise distances in $Y$ are small when there are
-non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
+non-zero $p_{ij}$, i.e., $x_i$ and $x_j$ are close together. The repulsive force
 should also be small, that is, overall the
@@
 will be placed far from each other in $Y$.
-This minimisation is done via stochastic gradient decent,
-and introduces a number of hyperparameters, for example the number of iterations,
+This minimization is done via stochastic gradient decent,
+and introduces a number of hyperparameters, for example, the number of iterations,
 the learning rate, and early exaggeration, a multiplier of the attractive force
-used at the beginning of the optimisation.
+used at the beginning of the optimization.
 
 Taken together, these details reveal the sheer number of decisions that an
-analyst must make. How does one choose the perplexity? And the
-parameters that control the optimisation of the loss function?
+analyst must make. How does one choose the perplexity and the
+parameters that control the optimization of the loss function?
 It is a known problem that
 t-SNE can have trouble recovering topology and that configurations can be
-highly dependent on how the algorithm is initialised and parameterized
+highly dependent on how the algorithm is initialized and parameterized
 [@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is cluster
@@
 t-SNE can recover spherical and well separated cluster shapes, and proposed new
-approaches for tuning the optimisation parameters. However, the cluster sizes
+approaches for tuning the optimization parameters. However, the cluster sizes
 and their relative orientation from a t-SNE view can be misleading
@@
 single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
-LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
+LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbor graph.
 Its main differences are that it uses a different cost function (cross entropy)
@@
 speed it is possible to generate UMAP embeddings in more than three dimensions.
-It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE [@Coenen2019-to].
+It appears to suffer from the same perceptual issues as t-SNE, however, it supposedly preserves global structure better than t-SNE [@Coenen2019-to].
 
@@
 
-The tour is a visualisation technique that is grounded in 
+The tour is a visualization technique that is grounded in 
 mathematical theory, and allows the viewer to ascertain 
@@
 sense of continuous movement from one basis to another. The tour is then the 
-continuous visualisation of the projected data $Y_{t} = XA_{t}$, that is the 
+continuous visualization of the projected data $Y_{t} = XA_{t}$, that is the 
 projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
@@
 bases at random; allowing a user to ascertain structure via exploring the
-subspace of $d$-dimensional projections.  In practice, we first sphere 
-our data via principal components to reduce dimensionality of $X$ prior to 
-running the tour. Instead of picking projections at 
-random, a *guided tour* can be used to generate a sequence of 'interesting' 
-projections as quantified by an index function [@Cook1995-bi]. While our software, 
-\pkg{liminal} is able to  visualise guided tours, our focus in the case studies 
-uses the grand tour to see global structure in the data.
+subspace of $d$-dimensional projections.
+In practice, we first form  our data into a sphere via principal components to reduce dimensionality of $X$ prior to running the tour.
+Instead of picking projections at  random, a *guided tour* can be used to generate a sequence of 'interesting'
+projections as quantified by an index function [@Cook1995-bi].
+While our software,\pkg{liminal} is able to visualize guided tours, our focus in the case studies uses the grand tour to see global structure in the data.
 
@@
 
-Tours provide a supportive visualisation to NLDR graphics, and 
+Tours provide a supportive visualization to NLDR graphics, and 
 can be easily incorporated into an analysts workflow with our software package, 
@@
 This approach allows our interface to achieve 
-the three principles for interactive high-dimensional data visualisation outlined
+the three principles for interactive high-dimensional data visualization outlined
 by @Buja1996-fk:  finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.
@@
 first run principal components on the input as an initial dimension reduction
-step, and then tour a subset of those components instead, i.e. by selecting
+step, and then tour a subset of those components instead, i.e., by selecting
 them from a scree plot. The default tour layout is a scatter plot with an axis
@@
 
-The initial visualisation gives an overview of the data structure,
+The initial visualization gives an overview of the data structure,
 and naturally leads to queries that investigate observed features with
-the aim to further characterise them. This is an essential aspect of
-our framework, where we use a tour to better characterise
+the aim to further characterize them. This is an essential aspect of
+our framework, where we use a tour to better characterize
 features observed in a NLDR display.
@@
 view. Linked highlighting can be performed when points have been previously
-labelled according to some discrete structure, i.e. cluster labels are
+labelled according to some discrete structure, i.e., cluster labels are
 available. This is achieved via the analyst clicking on groups in the legend,
@@
 Consequently, simple linked highlighting can alleviate a known downfall of
-methods such as UMAP or t-SNE: that is distances between clusters are
+methods such as UMAP or t-SNE; that is, distances between clusters are
 misleading. By highlighting corresponding clusters in the tour view, the
@@
 the tour view are highlighted. In this case, an analyst can use brushing for
-manually identifying clusters and verifying cluster locations and shapes:
+manually identifying clusters and verifying cluster locations and shapes;
 brushing in the embedding view gives analysts a sense of the shape and
@@
 distance metric used; distortions and diffusions can be thought of as the
-preservation of distances or the nearest neighbours graphs between the
+preservation of distances or the nearest neighbors graphs between the
 high-dimensional space and the embedding space. As distances can be noisy in
@@
 \pkg{liminal} [@r-liminal]. The package allows analysts to construct concatenated
-visualisations, drawn with the \pkg{Vega-Lite} grammar of interactive graphics via
+visualizations, drawn with the \pkg{Vega-Lite} grammar of interactive graphics via
 the \pkg{vegawidget} package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides 
@@
 The process of generating successive bases and interpolating between them to  
-construct intermediary frames, means the tour is a dynamic visualisation 
-technique. Generally, the user would set $d=2$ and the tour is visualised as
+construct intermediary frames, means the tour is a dynamic visualization 
+technique. Generally, the user would set $d=2$ and the tour is visualized as
 an animated scatter plot. This process of constructing bases and intermediate 
-frames and visualising the resulting projection is akin to making a "flip book" 
+frames and visualizing the resulting projection is akin to making a "flip book" 
 animation. Like with a flip book, an interface to the tour requires the ability
@@
 panning and zooming to modify the scales of the plot to provide context and
-click events to highlight groups of points if a labelling variable has
+click events to highlight groups of points if a labeling variable has
 been placed on the legend.
@@
 These interactions are enabled by treating the basis generation as a reactive
-stream. Instead of realising the entire sequence, which limits the animation
+stream. Instead of realizing the entire sequence, which limits the animation
 to have a discrete number of frames, new bases and their intermediate frames
 are generated dynamically via pushing the current projection to the 
-visualisation interface. The interface listens to events like pressing a button
+visualization interface. The interface listens to events like pressing a button
 or mouse-dragging and reacts by pausing the stream. This process allows the
@@
 via the controls previously described. If you are unable to view the videos,
-the figures in each case study consist of screenshots that summarise what is
+the figures in each case study consist of screenshots that summarize what is
 learned from combining the tour and an embedding view.
@@
 
-To begin we look at simulated datasets that reproduce known facts
+To begin, we look at simulated datasets that reproduce known facts
 about the  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
@@
 
-Next we view Gaussian clusters from the 
+Next, we view Gaussian clusters from the 
 _Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
@@
 
-Next we explore some simulated noisy tree structured data (Figure
-\@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
+Next, we explore some simulated noisy tree structured data (Figure
+\@ref(fig:fake-trees)). Our interest here is how t-SNE visualizations break the
 topology of the data, and then seeing if we can resolve this by tweaking the
 default parameters with reference to the global view of the data set. 
-This simulation aims to mimic branching trajectories of cell differentiation:
+This simulation aims to mimic branching trajectories of cell differentiation;
 if there were only mature cells, we would just see the tips of the branches 
@@
 twelve principal components (which makes up approximately 70% of the variance
-explained in the data) to use with the grand tour.
+explained in the data), to use with the grand tour.
 
@@
 keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
-initialisation). We then create a linked
+initialization). We then create a linked
 tour with t-SNE layout with \pkg{liminal} as shown in Figure \@ref(fig:trees-01).
@@
 this immediately via the linked highlighting over both
-plots. If we click on the legend for the zero branch, the blue coloured points
+plots. If we click on the legend for the zero branch, the blue colored points
 on each view are highlighted and the remaining points are made transparent.
 From here it becomes apparent from the tour view that the blue branch forms
-the backbone of the tree and is connected to all other branches. From the video 
+the backbone of the tree and is connected to all other branches. From the video, 
 it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
@@
 a better trade-off between local structure and retain the topology of the data.
-We keep every parameter the same except that we initialise $Y$ with the first
+We keep every parameter the same except that we initialize $Y$ with the first
 two PCs (scaled to have standard deviation 1e-4) instead of the default random
-initialisation and increase the perplexity from 30 to 100. We then combine
+initialization and increase the perplexity from 30 to 100. We then combine
 these results with our tour view as displayed in the linked video in the caption
@@
 identify groupings  of cells with similar expression profiles. Analysts
-in this area generally use non linear DR methods for verification and identification
+in this area generally use non-linear DR methods for verification and identification
 of clusters and developmental trajectories (i.e., case study 1). For clustering
-workflows the primary task is to verify the existence of clusters and then begin to 
+workflows, the primary task is to verify the existence of clusters and then begin to 
 identify the clusters as cell types using the expression of "known" 
-marker genes. Here a 'faithful' a embedding should ideally preserve the topology
+marker genes. Here a 'faithful'  embedding should ideally preserve the topology
 of the data; cells that correspond to a cell type should lie in the same
-neighbourhood in high-dimensional space. In this case study we use our 
-linked brushing approaches to look at neighbourhood preservation
+neighborhood in high-dimensional space. In this case study, we use our 
+linked brushing approaches to look at neighborhood preservation
 and look at marker genes through the lens of the tour. The data we have
@@
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed quality control using the **scater** package
-by removing cells with high proportion of mitochondrial gene expression (as this indicates poor sample quality) and low
-numbers of genes detected, we log-transformed and normalised the expression 
+(described by @Amezquita2020-at). We performed quality control using the **scater** package
+by removing cells with a high proportion of mitochondrial gene expression (as this indicates poor sample quality), and low
+numbers of genes detected. We then log-transformed and normalized the expression 
 values and  finally selected highly variable genes (HVGs) using **scran** 
 [@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
-the normalised expression matrix and compute PCA using the first 25 components.
-Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
+the normalized expression matrix and compute PCA using the first 25 components.
+Using the PCs we built a shared nearest neighbors graph (with $k = 10$) and 
 used Louvain clustering to generate clusters [@Blondel2008-bx].
@@
 alongside the t-SNE view which was computed from all 25 PCs.
-We have selected only the first five PCs because there is a large drop in the percentage of variance explained after the fifth component, with each component after contributing less than one percent of variance.
+We have selected only the first five PCs because there is a large drop in the percentage of variance explained after the fifth component, with each component after that contributing less than one percent of variance.
 Consequently, increasing the number of PCs to tour would increase the dimensionality and volume of the subspace we are touring but without adding any additional signal to the view.
@@
 get a sense of the shape of the clusters as seen from the linked video in Figure \@ref(fig:mouse-01).
-If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.
+If one was interested in performing more in-depth cluster analysis, we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.
 One could perform this approach manually via the \pkg{liminal} interface by returning the regions identified by brushing on the tour or embedding view.
@@
 clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
-The tour appears to form a tetrahedron like shape, with points lying on
+The tour appears to form a tetrahedron-like shape, with points lying on
 the surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
@@
 genes that are expressed in relatively larger quantities (upregulated) in cluster 2, which was one of the clumped clusters
-when we toured on principal components. Here the tour becomes an alternative
+when we toured on principal components. Here, the tour becomes an alternative
 to a standard heatmap view for assessing shared markers; the basis
@@
 weighting of each gene. We run the tour directly on the 
-log normalised expression values using the same subset as before.
+log normalized expression values using the same subset as before.
 
@@
 We have shown in the case studies, that one to one linked brushing can
-be used to identify distortions in the embedding, however we would
-like extend this to one to many linked brushing, which would allow us to directly interrogate neighbourhood preservation.
-This form of brushing acts directly on a $k$-nearest neighbours ($k$-nn) graph computed from a
+be used to identify distortions in the embedding, however, we would
+like extend this to one to many linked brushing, which would allow us to directly interrogate neighborhood preservation.
+This form of brushing acts directly on a $k$-nearest neighbors ($k$-nn) graph computed from a
 reference dataset: when a user brushes over a region in the embedding,
@@
 corresponding tour view. The reference data set for computing
-nearest neighbours (for example a distance matrix, or the complete data matrix) 
+nearest neighbors (for example, a distance matrix, or the complete data matrix) 
 can be independent of the tour or embedding views. In place of 
-highlighting, one  could use opacity or binned colour scales to encode 
-distances or ranks instead of the neighbouring points. We have begun 
+highlighting, one  could use opacity or binned color scales to encode 
+distances or ranks instead of the neighboring points. We have begun 
 implementing this brush in \pkg{liminal},  using the  **FNN** or **RcppAnnoy**
-packages for fast neighbourhood estimation on the server side, however there
+packages for fast neighborhood estimation on the server side, however, there
 are still technicalities that need be resolved [@fnn-pkg; @rcpp-annoy-pkg]. 
@@
 in using the \pkg{liminal} interface for larger datasets. First,
-t-SNE avoids the crowding problem, points are separated into distinct regions
+t-SNE avoids the crowding problem; points are separated into distinct regions
 on the display canvas. For the tour, points are concentrated in the
@@
 For the tasks we have looked at in this paper, where shape and density are important to the analyst, we think that better displays and sub-sampling strategies are more useful than being able to look at every single point on the canvas.
-We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however there are alternative sampling approaches that could be applied, like selecting points close to the cluster centres.
+We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however, there are alternative sampling approaches that could be applied, like selecting points close to the cluster centers.
 Alternative displays via statistical transformations could also mitigate the need to show all of the data.
@@
 
+\newpage
 # Acknowledgements {-}
@@
 vid_lookup$`Case Study` <- c(1, 2, 3,3,4,4)
+kableExtra::kable_styling(
   knitr::kable(dplyr::select(vid_lookup, `Case Study`, 
@@
                caption = "Case Study Videos",
-             booktabs = TRUE)
+               booktabs = TRUE),
+  latex_options = "HOLD_position"
+)
 ```

### 2022-02-26 liminal.tex
add tex files

@@
 \Abstract{
-Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complexity; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
+Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences.
+However, the appropriate use of these methods is difficult because of their complexity; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique.
 We present visual diagnostics for the usage of NLDR methods by combining them with a technique called the tour.
@@
 % Pandoc header
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
 \begin{document}
+
+\newpage
 
@@
 geometry of the high-dimensional space? The answer depends on the analyst's
-workflow. \citet{Brehmer2014-hk} characterised two main workflow steps that an
+workflow. \citet{Brehmer2014-hk} characterized two main workflow steps that an
 analyst performs when using embedding techniques: dimension reduction and
 cluster orientation. The first relates to dimension reduction achieved by using
-an embedding method, here an analyst wants to characterise and map meaning onto
+an embedding method, here an analyst wants to characterize and map meaning onto
 the embedded form, for example identifying batch effects from a high throughput
@@
 and interpretation of embedding methods as well. \citet{Sedlmair2013-pn} showed that
-scatter plots were mostly sufficient for detecting class separation, however
+scatter plots were mostly sufficient for detecting class separation, however,
 they also noted that often multiple embeddings were required. For the task of
@@
 
-There is no one-size fits all: finding an appropriate embedding for a given
+There is no one-size fits all; finding an appropriate embedding for a given
 dataset is a difficult and a somewhat poorly defined problem. For non-linear
 methods, there are a lot of parameters to explore that can have an effect on
-the resulting visualisation and interpretation.
+the resulting visualization and interpretation.
 While there has been much work on the algorithmic details of
-embedding methods, there are relatively few tools designed to assist users
+embedding methods; there are relatively few tools designed to assist users
 to interact with these techniques: when is an embedding sufficient for the
@@
 interactive interfaces have been proposed for evaluating or using embedding
-techniques. \citet{Buja2008-fn} used tours to guide analysts during the optimisation
-of multidimensional scaling methods by extending their interactive visualisation
+techniques. \citet{Buja2008-fn} used tours to guide analysts during the optimization
+of multidimensional scaling methods by extending their interactive visualization
 software called \pkg{XGobi} and \pkg{GGobi} into a new tool called \pkg{GGvis}
@@
 non-linear embeddings in single cell RNA-seq data. It provides a click and
-highlight visualisation for colouring points in an embedding according to an
+highlight visualization for coloring points in an embedding according to an
 estimated pairwise distance in the original high-dimensional space.
 Similarly, the \pkg{TensorFlow} embedding projector is a web interface to
-running some non-liner embedding methods live in the browser and
-provides interactions to colour points, and select nearest neighbours
+running some non-linear embedding methods live in the browser and it
+provides interactions to color points, and select nearest neighbors
 \citep{Smilkov2016-hp}. Finally, the work by \citet{Pezzotti2017-cz} provides a user
 guided and modified form of the t-SNE algorithm, that allows users to modify
-optimisation parameters in real-time.
+optimization parameters in real-time.
 
@@
 nature of the tour, user interaction is important for controlling and
-exploring the visualisation: the tour has been used previously by
+exploring the visualization; the tour has been used previously by
 \citet{Wickham2015-cx} for exploring statistical model fits and by \citet{Buja1996-fk} for
@@
 
-To begin we suppose the data is in the form of a rectangular
+To begin, we suppose the data is in the form of a rectangular
 matrix of real numbers, \(X = [\mathbf{x_1}, \dots, \mathbf{x_n}]\), where \(n\) is
@@
 matrix where \(d \ll p\). The hope of the analyst is that the DR procedure to
-produces \(Y\) will remove noise in the original dataset while retaining any
+produce \(Y\) will remove noise in the original dataset while retaining any
 latent structure.
@@
 \(Y\) is a linear transformation of \(X\). One example is principal components
-analysis (PCA) which performs an eigendecomposition of the estimated sample
+analysis (PCA) which performs an eigen-decomposition of the estimated sample
 covariance matrix. The eigenvalues are sorted in decreasing order and represent
@@
 
-For non-linear methods \(Y\) is generated via a pre-processed form of the input
-\(X\) such as the \(k\)-nearest neighbours graph or via a kernel transformation.
+For non-linear methods, \(Y\) is generated via a pre-processed form of the input
+\(X\) such as the \(k\)-nearest neighbors graph or via a kernel transformation.
 Multidimensional scaling (MDS) is a class of DR methods that aims to construct
@@
 distances with ranks instead \citep{Kruskal1964-cw}.
-A related technique is Isomap which uses a \(k\)-nearest neighbour graph
+A related technique is Isomap which uses a \(k\)-nearest neighbor graph
 to estimate the pair-wise geodesic distance of points in \(X\) then uses classical
@@
 non-linear DR methods for exploratory data analysis is selecting and tuning
-appropriate parameters. For concreteness here we focus on t-distributed
-stochastic neighbour embedding (t-SNE), and we will examine
+appropriate parameters. For concreteness here, we focus on t-distributed
+stochastic neighbor embedding (t-SNE), and we will examine
 its underpinning in some detail below \citep{Maaten2008-sk}. Similar considerations hold for related
@@
 2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 /
-2\sigma_i^2)} 
+2\sigma_i^2)}. 
 \]
 
-The conditional probabilities are then normalised and symmetrised to form a
+The conditional probabilities are then normalized and symmetrized to form a
 joint probability distribution via averaging:
 
-\[ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n} \]
+\[ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n}. \]
 
@@
 
-\[ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) \]
+\[ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})).\]
 
@@
 \(\sigma^2_{i} \rightarrow \infty\). This essentially turns t-SNE into a
-nearest neighbours algorithm, \(p_{i|j}\) will be close to zero for all
+nearest neighbors algorithm, \(p_{i|j}\) will be close to zero for all
 observations that are not in the \(\mathcal{O}(\text{perplexity}_i)\)
-neighbourhood graph of the \(i\)th observation \citep{Van_Der_Maaten2014-zn}.
+neighborhood graph of the \(i\)th observation \citep{Van_Der_Maaten2014-zn}.
 
 Next, in the target low-dimensional space, \(Y\), pair-wise distances between
-\(\mathbf{y_i}\) and \(\mathbf{y_j}\) are modelled as a symmetric probability
+\(\mathbf{y_i}\) and \(\mathbf{y_j}\) are modeled as a symmetric probability
 distribution using a t-distribution with one degree of freedom (Cauchy kernel):
@@
 \[ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log
-\frac{p_{ij}}{q_{ij}}\]
+\frac{p_{ij}}{q_{ij}}.\]
 
@@
 \[ \mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} +
-\log\sum_{i \ne j} w_{ij} \]
+\log\sum_{i \ne j} w_{ij}. \]
 
-Minimising the loss function corresponds to large attractive
+Minimizing the loss function corresponds to large attractive
 forces, that is, the pair-wise distances in \(Y\) are small when there are
-non-zero \(p_{ij}\), i.e.~\(x_i\) and \(x_j\) are close together. The repulsive force
+non-zero \(p_{ij}\), i.e., \(x_i\) and \(x_j\) are close together. The repulsive force
 should also be small, that is, overall the
@@
 will be placed far from each other in \(Y\).
-This minimisation is done via stochastic gradient decent,
-and introduces a number of hyperparameters, for example the number of iterations,
+This minimization is done via stochastic gradient decent,
+and introduces a number of hyperparameters, for example, the number of iterations,
 the learning rate, and early exaggeration, a multiplier of the attractive force
-used at the beginning of the optimisation.
+used at the beginning of the optimization.
 
 Taken together, these details reveal the sheer number of decisions that an
-analyst must make. How does one choose the perplexity? And the
-parameters that control the optimisation of the loss function?
+analyst must make. How does one choose the perplexity and the
+parameters that control the optimization of the loss function?
 It is a known problem that
 t-SNE can have trouble recovering topology and that configurations can be
-highly dependent on how the algorithm is initialised and parameterized
+highly dependent on how the algorithm is initialized and parameterized
 \citep{Wattenberg2016-ji, Kobak2019-lm, Melville2020}. If the goal is cluster
@@
 single cell transcriptomics) \citep{McInnes2018-co}. It is a method that is related to
-LargeVis \citep{Tang2016-oz}, and like t-SNE acts on the k-nearest neighbour graph.
+LargeVis \citep{Tang2016-oz}, and like t-SNE acts on the k-nearest neighbor graph.
 Its main differences are that it uses a different cost function (cross entropy)
@@
 speed it is possible to generate UMAP embeddings in more than three dimensions.
-It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE \citep{Coenen2019-to}.
+It appears to suffer from the same perceptual issues as t-SNE, however, it supposedly preserves global structure better than t-SNE \citep{Coenen2019-to}.
 
@@
 
-The tour is a visualisation technique that is grounded in
+The tour is a visualization technique that is grounded in
 mathematical theory, and allows the viewer to ascertain
@@
 sense of continuous movement from one basis to another. The tour is then the
-continuous visualisation of the projected data \(Y_{t} = XA_{t}\), that is the
+continuous visualization of the projected data \(Y_{t} = XA_{t}\), that is the
 projection of \(X\) onto \(A_{t}\) as the tour path is interpolated between
@@
 bases at random; allowing a user to ascertain structure via exploring the
-subspace of \(d\)-dimensional projections. In practice, we first sphere
-our data via principal components to reduce dimensionality of \(X\) prior to
-running the tour. Instead of picking projections at
-random, a \emph{guided tour} can be used to generate a sequence of `interesting'
-projections as quantified by an index function \citep{Cook1995-bi}. While our software,
-\pkg{liminal} is able to visualise guided tours, our focus in the case studies
-uses the grand tour to see global structure in the data.
+subspace of \(d\)-dimensional projections.
+In practice, we first form our data into a sphere via principal components to reduce dimensionality of \(X\) prior to running the tour.
+Instead of picking projections at random, a \emph{guided tour} can be used to generate a sequence of `interesting'
+projections as quantified by an index function \citep{Cook1995-bi}.
+While our software,\pkg{liminal} is able to visualize guided tours, our focus in the case studies uses the grand tour to see global structure in the data.
 
@@
 
-Tours provide a supportive visualisation to NLDR graphics, and
+Tours provide a supportive visualization to NLDR graphics, and
 can be easily incorporated into an analysts workflow with our software package,
@@
 This approach allows our interface to achieve
-the three principles for interactive high-dimensional data visualisation outlined
+the three principles for interactive high-dimensional data visualization outlined
 by \citet{Buja1996-fk}: finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.
@@
 first run principal components on the input as an initial dimension reduction
-step, and then tour a subset of those components instead, i.e.~by selecting
+step, and then tour a subset of those components instead, i.e., by selecting
 them from a scree plot. The default tour layout is a scatter plot with an axis
@@
 view. Linked highlighting can be performed when points have been previously
-labelled according to some discrete structure, i.e.~cluster labels are
+labelled according to some discrete structure, i.e., cluster labels are
 available. This is achieved via the analyst clicking on groups in the legend,
@@
 \pkg{liminal} \citep{r-liminal}. The package allows analysts to construct concatenated
-visualisations, drawn with the \pkg{Vega-Lite} grammar of interactive graphics via
+visualizations, drawn with the \pkg{Vega-Lite} grammar of interactive graphics via
 the \pkg{vegawidget} package \citep{Satyanarayan2017-gs, Lyttle2020-hp}. It provides
@@
 The process of generating successive bases and interpolating between them to\\
-construct intermediary frames, means the tour is a dynamic visualisation
-technique. Generally, the user would set \(d=2\) and the tour is visualised as
+construct intermediary frames, means the tour is a dynamic visualization
+technique. Generally, the user would set \(d=2\) and the tour is visualized as
 an animated scatter plot. This process of constructing bases and intermediate
-frames and visualising the resulting projection is akin to making a ``flip book''
+frames and visualizing the resulting projection is akin to making a ``flip book''
 animation. Like with a flip book, an interface to the tour requires the ability
@@
 These interactions are enabled by treating the basis generation as a reactive
-stream. Instead of realising the entire sequence, which limits the animation
+stream. Instead of realizing the entire sequence, which limits the animation
 to have a discrete number of frames, new bases and their intermediate frames
 are generated dynamically via pushing the current projection to the
-visualisation interface. The interface listens to events like pressing a button
+visualization interface. The interface listens to events like pressing a button
 or mouse-dragging and reacts by pausing the stream. This process allows the
@@
 
-To begin we look at simulated datasets that reproduce known facts
+To begin, we look at simulated datasets that reproduce known facts
 about the t-SNE algorithm. Our first data set consists of five spherical 5-\(d\)
@@
 
-Next we view Gaussian clusters from the
+Next, we view Gaussian clusters from the
 \emph{Multi Challenge Dataset}, a benchmark simulation data set for clustering tasks
@@
 
-Next we explore some simulated noisy tree structured data (Figure
-\ref{fig:fake-trees}). Our interest here is how t-SNE visualisations break
+Next, we explore some simulated noisy tree structured data (Figure
+\ref{fig:fake-trees}). Our interest here is how t-SNE visualizations break the
 topology of the data, and then seeing if we can resolve this by tweaking the
 default parameters with reference to the global view of the data set.
-This simulation aims to mimic branching trajectories of cell differentiation:
+This simulation aims to mimic branching trajectories of cell differentiation;
 if there were only mature cells, we would just see the tips of the branches
@@
 twelve principal components (which makes up approximately 70\% of the variance
-explained in the data) to use with the grand tour.
+explained in the data), to use with the grand tour.
 
@@
 keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
-initialisation). We then create a linked
+initialization). We then create a linked
 tour with t-SNE layout with \pkg{liminal} as shown in Figure \ref{fig:trees-01}.
@@
 this immediately via the linked highlighting over both
-plots. If we click on the legend for the zero branch, the blue coloured points
+plots. If we click on the legend for the zero branch, the blue colored points
 on each view are highlighted and the remaining points are made transparent.
 From here it becomes apparent from the tour view that the blue branch forms
-the backbone of the tree and is connected to all other branches. From the video
+the backbone of the tree and is connected to all other branches. From the video,
 it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
@@
 a better trade-off between local structure and retain the topology of the data.
-We keep every parameter the same except that we initialise \(Y\) with the first
+We keep every parameter the same except that we initialize \(Y\) with the first
 two PCs (scaled to have standard deviation 1e-4) instead of the default random
-initialisation and increase the perplexity from 30 to 100. We then combine
+initialization and increase the perplexity from 30 to 100. We then combine
 these results with our tour view as displayed in the linked video in the caption
@@
 identify groupings of cells with similar expression profiles. Analysts
-in this area generally use non linear DR methods for verification and identification
+in this area generally use non-linear DR methods for verification and identification
 of clusters and developmental trajectories (i.e., case study 1). For clustering
-workflows the primary task is to verify the existence of clusters and then begin to
+workflows, the primary task is to verify the existence of clusters and then begin to
 identify the clusters as cell types using the expression of ``known''
-marker genes. Here a `faithful' a embedding should ideally preserve the topology
+marker genes. Here a `faithful' embedding should ideally preserve the topology
 of the data; cells that correspond to a cell type should lie in the same
-neighbourhood in high-dimensional space. In this case study we use our
-linked brushing approaches to look at neighbourhood preservation
+neighborhood in high-dimensional space. In this case study, we use our
+linked brushing approaches to look at neighborhood preservation
 and look at marker genes through the lens of the tour. The data we have
@@
 have followed a standard workflow for pre-processing and normalizing this data
-(described by \citet{Amezquita2020-at}): we performed quality control using the \textbf{scater} package
-by removing cells with high proportion of mitochondrial gene expression (as this indicates poor sample quality) and low
-numbers of genes detected, we log-transformed and normalised the expression
+(described by \citet{Amezquita2020-at}). We performed quality control using the \textbf{scater} package
+by removing cells with a high proportion of mitochondrial gene expression (as this indicates poor sample quality), and low
+numbers of genes detected. We then log-transformed and normalized the expression
 values and finally selected highly variable genes (HVGs) using \textbf{scran}
 \citep{McCarthy2017, Lun2016}. The top ten percent of HVGs were used to subset
-the normalised expression matrix and compute PCA using the first 25 components.
-Using the PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
+the normalized expression matrix and compute PCA using the first 25 components.
+Using the PCs we built a shared nearest neighbors graph (with \(k = 10\)) and
 used Louvain clustering to generate clusters \citep{Blondel2008-bx}.
@@
 get a sense of the shape of the clusters as seen from the linked video in Figure \ref{fig:mouse-01}.
-If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.
+If one was interested in performing more in-depth cluster analysis, we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.
 One could perform this approach manually via the \pkg{liminal} interface by returning the regions identified by brushing on the tour or embedding view.
@@
 clumped together in tour view, but are blown up in the embedding view (Figure \ref{fig:mouse-01}b).
-The tour appears to form a tetrahedron like shape, with points lying on
+The tour appears to form a tetrahedron-like shape, with points lying on
 the surface and along the vertices of the tetrahedron in 5-\(d\) PCA space -
@@
 weighting of each gene. We run the tour directly on the
-log normalised expression values using the same subset as before.
+log normalized expression values using the same subset as before.
 
@@
 We have shown in the case studies, that one to one linked brushing can
-be used to identify distortions in the embedding, however we would
-like extend this to one to many linked brushing, which would allow us to directly interrogate neighbourhood preservation.
-This form of brushing acts directly on a \(k\)-nearest neighbours (\(k\)-nn) graph computed from a
+be used to identify distortions in the embedding, however, we would
+like extend this to one to many linked brushing, which would allow us to directly interrogate neighborhood preservation.
+This form of brushing acts directly on a \(k\)-nearest neighbors (\(k\)-nn) graph computed from a
 reference dataset: when a user brushes over a region in the embedding,
@@
 corresponding tour view. The reference data set for computing
-nearest neighbours (for example a distance matrix, or the complete data matrix)
+nearest neighbors (for example, a distance matrix, or the complete data matrix)
 can be independent of the tour or embedding views. In place of
-highlighting, one could use opacity or binned colour scales to encode
-distances or ranks instead of the neighbouring points. We have begun
+highlighting, one could use opacity or binned color scales to encode
+distances or ranks instead of the neighboring points. We have begun
 implementing this brush in \pkg{liminal}, using the \textbf{FNN} or \textbf{RcppAnnoy}
-packages for fast neighbourhood estimation on the server side, however there
+packages for fast neighborhood estimation on the server side, however, there
 are still technicalities that need be resolved \citep{fnn-pkg, rcpp-annoy-pkg}.
@@
 work from topological statistics \citep{Rieck2017-kk, Genovese2017-iq}.
+
+\newpage
 
@@
 
-\begin{table}
+\begin{table}[H]
 
